### 2015-07-28 sig-alternate.tex
First commit

% This is "sig-alternate.tex" V2.0 May 2012
% This file should be compiled with V2.5 of "sig-alternate.cls" May 2012
%
% This example file demonstrates the use of the 'sig-alternate.cls'
% V2.5 LaTeX2e document class file. It is for those submitting
% articles to ACM Conference Proceedings WHO DO NOT WISH TO
% STRICTLY ADHERE TO THE SIGS (PUBS-BOARD-ENDORSED) STYLE.
% The 'sig-alternate.cls' file will produce a similar-looking,
% albeit, 'tighter' paper resulting in, invariably, fewer pages.
%
% ----------------------------------------------------------------------------------------------------------------
% This .tex file (and associated .cls V2.5) produces:
%       1) The Permission Statement
%       2) The Conference (location) Info information
%       3) The Copyright Line with ACM data
%       4) NO page numbers
%
% as against the acm_proc_article-sp.cls file which
% DOES NOT produce 1) thru' 3) above.
%
% Using 'sig-alternate.cls' you have control, however, from within
% the source .tex file, over both the CopyrightYear
% (defaulted to 200X) and the ACM Copyright Data
% (defaulted to X-XXXXX-XX-X/XX/XX).
% e.g.
% \CopyrightYear{2007} will cause 2007 to appear in the copyright line.
% \crdata{0-12345-67-8/90/12} will cause 0-12345-67-8/90/12 to appear in the copyright line.
%
% ---------------------------------------------------------------------------------------------------------------
% This .tex source is an example which *does* use
% the .bib file (from which the .bbl file % is produced).
% REMEMBER HOWEVER: After having produced the .bbl file,
% and prior to final submission, you *NEED* to 'insert'
% your .bbl file into your source .tex file so as to provide
% ONE 'self-contained' source file.
%
% ================= IF YOU HAVE QUESTIONS =======================
% Questions regarding the SIGS styles, SIGS policies and
% procedures, Conferences etc. should be sent to
% Adrienne Griscti (griscti@acm.org)
%
% Technical questions _only_ to
% Gerald Murray (murray@hq.acm.org)
% ===============================================================
%
% For tracking purposes - this is V2.0 - May 2012

\documentclass{sig-alternate}

\begin{document}
%
% --- Author Metadata here ---
\conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}
%\CopyrightYear{2007} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
%\crdata{0-12345-67-8/90/01}  % Allows default copyright data (0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
% --- End of Author Metadata ---

\title{Alternate {\ttlit ACM} SIG Proceedings Paper in LaTeX
Format\titlenote{(Produces the permission block, and
copyright information). For use with
SIG-ALTERNATE.CLS. Supported by ACM.}}
\subtitle{[Extended Abstract]
\titlenote{A full version of this paper is available as
\textit{Author's Guide to Preparing ACM SIG Proceedings Using
\LaTeX$2_\epsilon$\ and BibTeX} at
\texttt{www.acm.org/eaddress.htm}}}
%
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.

\numberofauthors{8} %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors section.
%
\author{
% You can go ahead and credit any number of authors here,
% e.g. one 'row of three' or two rows (consisting of one row of three
% and a second row of one, two or three).
%
% The command \alignauthor (no curly braces needed) should
% precede each author name, affiliation/snail-mail address and
% e-mail address. Additionally, tag each line of
% affiliation/address with \affaddr, and tag the
% e-mail address with \email.
%
% 1st. author
\alignauthor
Ben Trovato\titlenote{Dr.~Trovato insisted his name be first.}\\
       \affaddr{Institute for Clarity in Documentation}\\
       \affaddr{1932 Wallamaloo Lane}\\
       \affaddr{Wallamaloo, New Zealand}\\
       \email{trovato@corporation.com}
% 2nd. author
\alignauthor
G.K.M. Tobin\titlenote{The secretary disavows
any knowledge of this author's actions.}\\
       \affaddr{Institute for Clarity in Documentation}\\
       \affaddr{P.O. Box 1212}\\
       \affaddr{Dublin, Ohio 43017-6221}\\
       \email{webmaster@marysville-ohio.com}
% 3rd. author
\alignauthor Lars Th{\o}rv{\"a}ld\titlenote{This author is the
one who did all the really hard work.}\\
       \affaddr{The Th{\o}rv{\"a}ld Group}\\
       \affaddr{1 Th{\o}rv{\"a}ld Circle}\\
       \affaddr{Hekla, Iceland}\\
       \email{larst@affiliation.org}
\and  % use '\and' if you need 'another row' of author names
% 4th. author
\alignauthor Lawrence P. Leipuner\\
       \affaddr{Brookhaven Laboratories}\\
       \affaddr{Brookhaven National Lab}\\
       \affaddr{P.O. Box 5000}\\
       \email{lleipuner@researchlabs.org}
% 5th. author
\alignauthor Sean Fogarty\\
       \affaddr{NASA Ames Research Center}\\
       \affaddr{Moffett Field}\\
       \affaddr{California 94035}\\
       \email{fogartys@amesres.org}
% 6th. author
\alignauthor Charles Palmer\\
       \affaddr{Palmer Research Laboratories}\\
       \affaddr{8600 Datapoint Drive}\\
       \affaddr{San Antonio, Texas 78229}\\
       \email{cpalmer@prl.com}
}
% There's nothing stopping you putting the seventh, eighth, etc.
% author on the opening page (as the 'third row') but we ask,
% for aesthetic reasons that you place these 'additional authors'
% in the \additional authors block, viz.
\additionalauthors{Additional authors: John Smith (The Th{\o}rv{\"a}ld Group,
email: {\texttt{jsmith@affiliation.org}}) and Julius P.~Kumquat
(The Kumquat Consortium, email: {\texttt{jpkumquat@consortium.net}}).}
\date{30 July 1999}
% Just remember to make sure that the TOTAL number of authors
% is the number that will appear on the first page PLUS the
% number that will appear in the \additionalauthors section.

\maketitle
\begin{abstract}
This paper provides a sample of a \LaTeX\ document which conforms,
somewhat loosely, to the formatting guidelines for
ACM SIG Proceedings. It is an {\em alternate} style which produces
a {\em tighter-looking} paper and was designed in response to
concerns expressed, by authors, over page-budgets.
It complements the document \textit{Author's (Alternate) Guide to
Preparing ACM SIG Proceedings Using \LaTeX$2_\epsilon$\ and Bib\TeX}.
This source file has been written with the intention of being
compiled under \LaTeX$2_\epsilon$\ and BibTeX.

The developers have tried to include every imaginable sort
of ``bells and whistles", such as a subtitle, footnotes on
title, subtitle and authors, as well as in the text, and
every optional component (e.g. Acknowledgments, Additional
Authors, Appendices), not to mention examples of
equations, theorems, tables and figures.

To make best use of this sample document, run it through \LaTeX\
and BibTeX, and compare this source code with the printed
output produced by the dvi file. A compiled PDF version
is available on the web page to help you with the
`look and feel'.
\end{abstract}

% A category with the (minimum) three required fields
\category{H.4}{Information Systems Applications}{Miscellaneous}
%A category including the fourth, optional field follows...
\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]

\terms{Theory}

\keywords{ACM proceedings, \LaTeX, text tagging}

\section{Introduction}
The \textit{proceedings} are the records of a conference.
ACM seeks to give these conference by-products a uniform,
high-quality appearance.  To do this, ACM has some rigid
requirements for the format of the proceedings documents: there
is a specified format (balanced  double columns), a specified
set of fonts (Arial or Helvetica and Times Roman) in
certain specified sizes (for instance, 9 point for body copy),
a specified live area (18 $\times$ 23.5 cm [7" $\times$ 9.25"]) centered on
the page, specified size of margins (1.9 cm [0.75"]) top, (2.54 cm [1"]) bottom
and (1.9 cm [.75"]) left and right; specified column width
(8.45 cm [3.33"]) and gutter size (.83 cm [.33"]).

The good news is, with only a handful of manual
settings\footnote{Two of these, the {\texttt{\char'134 numberofauthors}}
and {\texttt{\char'134 alignauthor}} commands, you have
already used; another, {\texttt{\char'134 balancecolumns}}, will
be used in your very last run of \LaTeX\ to ensure
balanced column heights on the last page.}, the \LaTeX\ document
class file handles all of this for you.

The remainder of this document is concerned with showing, in
the context of an ``actual'' document, the \LaTeX\ commands
specifically available for denoting the structure of a
proceedings paper, rather than with giving rigorous descriptions
or explanations of such commands.

\section{The {\secit Body} of The Paper}
Typically, the body of a paper is organized
into a hierarchical structure, with numbered or unnumbered
headings for sections, subsections, sub-subsections, and even
smaller sections.  The command \texttt{{\char'134}section} that
precedes this paragraph is part of such a
hierarchy.\footnote{This is the second footnote.  It
starts a series of three footnotes that add nothing
informational, but just give an idea of how footnotes work
and look. It is a wordy one, just so you see
how a longish one plays out.} \LaTeX\ handles the numbering
and placement of these headings for you, when you use
the appropriate heading commands around the titles
of the headings.  If you want a sub-subsection or
smaller part to be unnumbered in your output, simply append an
asterisk to the command name.  Examples of both
numbered and unnumbered headings will appear throughout the
balance of this sample document.

Because the entire article is contained in
the \textbf{document} environment, you can indicate the
start of a new paragraph with a blank line in your
input file; that is why this sentence forms a separate paragraph.

\subsection{Type Changes and {\subsecit Special} Characters}
We have already seen several typeface changes in this sample.  You
can indicate italicized words or phrases in your text with
the command \texttt{{\char'134}textit}; emboldening with the
command \texttt{{\char'134}textbf}
and typewriter-style (for instance, for computer code) with
\texttt{{\char'134}texttt}.  But remember, you do not
have to indicate typestyle changes when such changes are
part of the \textit{structural} elements of your
article; for instance, the heading of this subsection will
be in a sans serif\footnote{A third footnote, here.
Let's make this a rather short one to
see how it looks.} typeface, but that is handled by the
document class file. Take care with the use
of\footnote{A fourth, and last, footnote.}
the curly braces in typeface changes; they mark
the beginning and end of
the text that is to be in the different typeface.

You can use whatever symbols, accented characters, or
non-English characters you need anywhere in your document;
you can find a complete list of what is
available in the \textit{\LaTeX\
User's Guide}\cite{Lamport:LaTeX}.

\subsection{Math Equations}
You may want to display math equations in three distinct styles:
inline, numbered or non-numbered display.  Each of
the three are discussed in the next sections.

\subsubsection{Inline (In-text) Equations}
A formula that appears in the running text is called an
inline or in-text formula.  It is produced by the
\textbf{math} environment, which can be
invoked with the usual \texttt{{\char'134}begin. . .{\char'134}end}
construction or with the short form \texttt{\$. . .\$}. You
can use any of the symbols and structures,
from $\alpha$ to $\omega$, available in
\LaTeX\cite{Lamport:LaTeX}; this section will simply show a
few examples of in-text equations in context. Notice how
this equation: \begin{math}\lim_{n\rightarrow \infty}x=0\end{math},
set here in in-line math style, looks slightly different when
set in display style.  (See next section).

\subsubsection{Display Equations}
A numbered display equation -- one set off by vertical space
from the text and centered horizontally -- is produced
by the \textbf{equation} environment. An unnumbered display
equation is produced by the \textbf{displaymath} environment.

Again, in either environment, you can use any of the symbols
and structures available in \LaTeX; this section will just
give a couple of examples of display equations in context.
First, consider the equation, shown as an inline equation above:
\begin{equation}\lim_{n\rightarrow \infty}x=0\end{equation}
Notice how it is formatted somewhat differently in
the \textbf{displaymath}
environment.  Now, we'll enter an unnumbered equation:
\begin{displaymath}\sum_{i=0}^{\infty} x + 1\end{displaymath}
and follow it with another numbered equation:
\begin{equation}\sum_{i=0}^{\infty}x_i=\int_{0}^{\pi+2} f\end{equation}
just to demonstrate \LaTeX's able handling of numbering.

\subsection{Citations}
Citations to articles \cite{bowman:reasoning,
clark:pct, braams:babel, herlihy:methodology},
conference proceedings \cite{clark:pct} or
books \cite{salas:calculus, Lamport:LaTeX} listed
in the Bibliography section of your
article will occur throughout the text of your article.
You should use BibTeX to automatically produce this bibliography;
you simply need to insert one of several citation commands with
a key of the item cited in the proper location in
the \texttt{.tex} file \cite{Lamport:LaTeX}.
The key is a short reference you invent to uniquely
identify each work; in this sample document, the key is
the first author's surname and a
word from the title.  This identifying key is included
with each item in the \texttt{.bib} file for your article.

The details of the construction of the \texttt{.bib} file
are beyond the scope of this sample document, but more
information can be found in the \textit{Author's Guide},
and exhaustive details in the \textit{\LaTeX\ User's
Guide}\cite{Lamport:LaTeX}.

This article shows only the plainest form
of the citation command, using \texttt{{\char'134}cite}.
This is what is stipulated in the SIGS style specifications.
No other citation format is endorsed or supported.

\subsection{Tables}
Because tables cannot be split across pages, the best
placement for them is typically the top of the page
nearest their initial cite.  To
ensure this proper ``floating'' placement of tables, use the
environment \textbf{table} to enclose the table's contents and
the table caption.  The contents of the table itself must go
in the \textbf{tabular} environment, to
be aligned properly in rows and columns, with the desired
horizontal and vertical rules.  Again, detailed instructions
on \textbf{tabular} material
is found in the \textit{\LaTeX\ User's Guide}.

Immediately following this sentence is the point at which
Table 1 is included in the input file; compare the
placement of the table here with the table in the printed
dvi output of this document.

\begin{table}
\centering
\caption{Frequency of Special Characters}
\begin{tabular}{|c|c|l|} \hline
Non-English or Math&Frequency&Comments\\ \hline
\O & 1 in 1,000& For Swedish names\\ \hline
$\pi$ & 1 in 5& Common in math\\ \hline
\$ & 4 in 5 & Used in business\\ \hline
$\Psi^2_1$ & 1 in 40,000& Unexplained usage\\
\hline\end{tabular}
\end{table}

To set a wider table, which takes up the whole width of
the page's live area, use the environment
\textbf{table*} to enclose the table's contents and
the table caption.  As with a single-column table, this wide
table will ``float" to a location deemed more desirable.
Immediately following this sentence is the point at which
Table 2 is included in the input file; again, it is
instructive to compare the placement of the
table here with the table in the printed dvi
output of this document.

\begin{table*}
\centering
\caption{Some Typical Commands}
\begin{tabular}{|c|c|l|} \hline
Command&A Number&Comments\\ \hline
\texttt{{\char'134}alignauthor} & 100& Author alignment\\ \hline
\texttt{{\char'134}numberofauthors}& 200& Author enumeration\\ \hline
\texttt{{\char'134}table}& 300 & For tables\\ \hline
\texttt{{\char'134}table*}& 400& For wider tables\\ \hline\end{tabular}
\end{table*}
% end the environment with {table*}, NOTE not {table}!

\subsection{Figures}
Like tables, figures cannot be split across pages; the
best placement for them
is typically the top or the bottom of the page nearest
their initial cite.  To ensure this proper ``floating'' placement
of figures, use the environment
\textbf{figure} to enclose the figure and its caption.

This sample document contains examples of \textbf{.eps}
and \textbf{.ps} files to be displayable with \LaTeX.  More
details on each of these is found in the \textit{Author's Guide}.

\begin{figure}
\centering
\epsfig{file=fly.eps}
\caption{A sample black and white graphic (.eps format).}
\end{figure}

\begin{figure}
\centering
\epsfig{file=fly.eps, height=1in, width=1in}
\caption{A sample black and white graphic (.eps format)
that has been resized with the \texttt{epsfig} command.}
\end{figure}

As was the case with tables, you may want a figure
that spans two columns.  To do this, and still to
ensure proper ``floating'' placement of tables, use the environment
\textbf{figure*} to enclose the figure and its caption.
and don't forget to end the environment with
{figure*}, not {figure}!

\begin{figure*}
\centering
\epsfig{file=flies.eps}
\caption{A sample black and white graphic (.eps format)
that needs to span two columns of text.}
\end{figure*}

Note that either {\textbf{.ps}} or {\textbf{.eps}} formats are
used; use
the \texttt{{\char'134}epsfig} or \texttt{{\char'134}psfig}
commands as appropriate for the different file types.

\begin{figure}
\centering
\caption{A sample black and white graphic (.ps format) that has
been resized with the \texttt{psfig} command.}
\vskip -6pt
\end{figure}

\subsection{Theorem-like Constructs}
Other common constructs that may occur in your article are
the forms for logical constructs like theorems, axioms,
corollaries and proofs.  There are
two forms, one produced by the
command \texttt{{\char'134}newtheorem} and the
other by the command \texttt{{\char'134}newdef}; perhaps
the clearest and easiest way to distinguish them is
to compare the two in the output of this sample document:

This uses the \textbf{theorem} environment, created by
the\linebreak\texttt{{\char'134}newtheorem} command:
\newtheorem{theorem}{Theorem}
\begin{theorem}
Let $f$ be continuous on $[a,b]$.  If $G$ is
an antiderivative for $f$ on $[a,b]$, then
\begin{displaymath}\int^b_af(t)dt = G(b) - G(a).\end{displaymath}
\end{theorem}

The other uses the \textbf{definition} environment, created
by the \texttt{{\char'134}newdef} command:
\newdef{definition}{Definition}
\begin{definition}
If $z$ is irrational, then by $e^z$ we mean the
unique number which has
logarithm $z$: \begin{displaymath}{\log e^z = z}\end{displaymath}
\end{definition}

Two lists of constructs that use one of these
forms is given in the
\textit{Author's  Guidelines}.

There is one other similar construct environment, which is
already set up
for you; i.e. you must \textit{not} use
a \texttt{{\char'134}newdef} command to
create it: the \textbf{proof} environment.  Here
is a example of its use:
\begin{proof}
Suppose on the contrary there exists a real number $L$ such that
\begin{displaymath}
\lim_{x\rightarrow\infty} \frac{f(x)}{g(x)} = L.
\end{displaymath}
Then
\begin{displaymath}
l=\lim_{x\rightarrow c} f(x)
= \lim_{x\rightarrow c}
\left[ g{x} \cdot \frac{f(x)}{g(x)} \right ]
= \lim_{x\rightarrow c} g(x) \cdot \lim_{x\rightarrow c}
\frac{f(x)}{g(x)} = 0\cdot L = 0,
\end{displaymath}
which contradicts our assumption that $l\neq 0$.
\end{proof}

Complete rules about using these environments and using the
two different creation commands are in the
\textit{Author's Guide}; please consult it for more
detailed instructions.  If you need to use another construct,
not listed therein, which you want to have the same
formatting as the Theorem
or the Definition\cite{salas:calculus} shown above,
use the \texttt{{\char'134}newtheorem} or the
\texttt{{\char'134}newdef} command,
respectively, to create it.

\subsection*{A {\secit Caveat} for the \TeX\ Expert}
Because you have just been given permission to
use the \texttt{{\char'134}newdef} command to create a
new form, you might think you can
use \TeX's \texttt{{\char'134}def} to create a
new command: \textit{Please refrain from doing this!}
Remember that your \LaTeX\ source code is primarily intended
to create camera-ready copy, but may be converted
to other forms -- e.g. HTML. If you inadvertently omit
some or all of the \texttt{{\char'134}def}s recompilation will
be, to say the least, problematic.

\section{Conclusions}
This paragraph will end the body of this sample document.
Remember that you might still have Acknowledgments or
Appendices; brief samples of these
follow.  There is still the Bibliography to deal with; and
we will make a disclaimer about that here: with the exception
of the reference to the \LaTeX\ book, the citations in
this paper are to articles which have nothing to
do with the present subject and are used as
examples only.
%\end{document}  % This is where a 'short' article might terminate

%ACKNOWLEDGMENTS are optional
\section{Acknowledgments}
This section is optional; it is a location for you
to acknowledge grants, funding, editing assistance and
what have you.  In the present case, for example, the
authors would like to thank Gerald Murray of ACM for
his help in codifying this \textit{Author's Guide}
and the \textbf{.cls} and \textbf{.tex} files that it describes.

%
% The following two commands are all you need in the
% initial runs of your .tex file to
% produce the bibliography for the citations in your paper.
\bibliographystyle{abbrv}
\bibliography{sigproc}  % sigproc.bib is the name of the Bibliography in this case
% You must have a proper ".bib" file
%  and remember to run:
% latex bibtex latex latex
% to resolve all references
%
% ACM needs 'a single self-contained file'!
%
%APPENDICES are optional
%\balancecolumns
\appendix
%Appendix A
\section{Headings in Appendices}
The rules about hierarchical headings discussed above for
the body of the article are different in the appendices.
In the \textbf{appendix} environment, the command
\textbf{section} is used to
indicate the start of each Appendix, with alphabetic order
designation (i.e. the first is A, the second B, etc.) and
a title (if you include one).  So, if you need
hierarchical structure
\textit{within} an Appendix, start with \textbf{subsection} as the
highest level. Here is an outline of the body of this
document in Appendix-appropriate form:
\subsection{Introduction}
\subsection{The Body of the Paper}
\subsubsection{Type Changes and  Special Characters}
\subsubsection{Math Equations}
\paragraph{Inline (In-text) Equations}
\paragraph{Display Equations}
\subsubsection{Citations}
\subsubsection{Tables}
\subsubsection{Figures}
\subsubsection{Theorem-like Constructs}
\subsubsection*{A Caveat for the \TeX\ Expert}
\subsection{Conclusions}
\subsection{Acknowledgments}
\subsection{Additional Authors}
This section is inserted by \LaTeX; you do not insert it.
You just add the names and information in the
\texttt{{\char'134}additionalauthors} command at the start
of the document.
\subsection{References}
Generated by bibtex from your ~.bib file.  Run latex,
then bibtex, then latex twice (to resolve references)
to create the ~.bbl file.  Insert that ~.bbl file into
the .tex source file and comment out
the command \texttt{{\char'134}thebibliography}.
% This next section command marks the start of
% Appendix B, and does not continue the present hierarchy
\section{More Help for the Hardy}
The sig-alternate.cls file itself is chock-full of succinct
and helpful comments.  If you consider yourself a moderately
experienced to expert user of \LaTeX, you may find reading
it useful but please remember not to change it.
%\balancecolumns % GM June 2007
% That's all folks!
\end{document}

### 2015-07-29 sig-alternate.tex
Set-up + Abstract + Structure added

@@
 \documentclass{sig-alternate}
+\graphicspath{{figs/}}
 
@@
 \begin{abstract}
-This paper provides a sample of a \LaTeX\ document which conforms,
-somewhat loosely, to the formatting guidelines for
-ACM SIG Proceedings. It is an {\em alternate} style which produces
-a {\em tighter-looking} paper and was designed in response to
-concerns expressed, by authors, over page-budgets.
-It complements the document \textit{Author's (Alternate) Guide to
-Preparing ACM SIG Proceedings Using \LaTeX$2_\epsilon$\ and Bib\TeX}.
-This source file has been written with the intention of being
-compiled under \LaTeX$2_\epsilon$\ and BibTeX.
-
-The developers have tried to include every imaginable sort
-of ``bells and whistles", such as a subtitle, footnotes on
-title, subtitle and authors, as well as in the text, and
-every optional component (e.g. Acknowledgments, Additional
-Authors, Appendices), not to mention examples of
-equations, theorems, tables and figures.
-
-To make best use of this sample document, run it through \LaTeX\
-and BibTeX, and compare this source code with the printed
-output produced by the dvi file. A compiled PDF version
-is available on the web page to help you with the
-`look and feel'.
+In long term interactions between a human and a robot, it is necessary that the agent exhibits some sort of social intelligence to be able to engage the human providing the appropriate response to specific face-to-face situations.
+In a educational context, this feature becomes a must due to the fact that engagement plays a very important role in the learning process.
+Therefore, maximizing the interaction time at a high engagement level may provide learning benefits to the children.
+In this work, a real-time attention tracker based on the head pose estimation using an RGB camera is proposed to be able to identify faults during the interaction and eventually detect low engagement level situations.
+In this paper we want to show a practical approach to evaluate the engagement that arises from the interaction reporting several experiences in the field, with the robot Nao.
 \end{abstract}
@@
 \section{Introduction}
-The \textit{proceedings} are the records of a conference.
-ACM seeks to give these conference by-products a uniform,
-high-quality appearance.  To do this, ACM has some rigid
-requirements for the format of the proceedings documents: there
-is a specified format (balanced  double columns), a specified
-set of fonts (Arial or Helvetica and Times Roman) in
-certain specified sizes (for instance, 9 point for body copy),
-a specified live area (18 $\times$ 23.5 cm [7" $\times$ 9.25"]) centered on
-the page, specified size of margins (1.9 cm [0.75"]) top, (2.54 cm [1"]) bottom
-and (1.9 cm [.75"]) left and right; specified column width
-(8.45 cm [3.33"]) and gutter size (.83 cm [.33"]).
+The use of robots in the learning activity presents an opportunity for the children
+to interact with an embodied, physical agent as part of the learning experience. Furthermore,
+the use of robots in handwriting education comes with the potential to engage the child in
+meta-cognition through the learning by teaching paradigm, wherein a student takes the role of
+a teacher and experiences stronger educational benefits as a result (such as in \cite{Palinscar1984}).
 
-The good news is, with only a handful of manual
-settings\footnote{Two of these, the {\texttt{\char'134 numberofauthors}}
-and {\texttt{\char'134 alignauthor}} commands, you have
-already used; another, {\texttt{\char'134 balancecolumns}}, will
-be used in your very last run of \LaTeX\ to ensure
-balanced column heights on the last page.}, the \LaTeX\ document
-class file handles all of this for you.
+It becomes a useful to capture the significant passive information provided by the
+user during the interaction and adapt the situation to it.
 
-The remainder of this document is concerned with showing, in
-the context of an ``actual'' document, the \LaTeX\ commands
-specifically available for denoting the structure of a
-proceedings paper, rather than with giving rigorous descriptions
-or explanations of such commands.
+\subsection{Approach}
+Defining accurately the human field of view in the implementation is a must.
+Holmqvist in \cite{holmqvist2011eye} specifies that the visual human range is $ \pm  40\degree $ in the horizontal and $ \pm 25\degree $ in the vertical.
+In \cite{walker1980clinical} Spector provides a more detailed specification for each eye splitting the vertical range into $ 60\degree $ the upper region and $ 75\degree $ the lower one.
+Moreover, the horizontal range gets separated in $ 60\degree $ inwards (towards the nose) and $ 95\degree $ outwards.
+In the present implementation, the first approach explained has been chosen representing the field of view using a cone with such dimensions.
 
-\section{The {\secit Body} of The Paper}
-Typically, the body of a paper is organized
-into a hierarchical structure, with numbered or unnumbered
-headings for sections, subsections, sub-subsections, and even
-smaller sections.  The command \texttt{{\char'134}section} that
-precedes this paragraph is part of such a
-hierarchy.\footnote{This is the second footnote.  It
-starts a series of three footnotes that add nothing
-informational, but just give an idea of how footnotes work
-and look. It is a wordy one, just so you see
-how a longish one plays out.} \LaTeX\ handles the numbering
-and placement of these headings for you, when you use
-the appropriate heading commands around the titles
-of the headings.  If you want a sub-subsection or
-smaller part to be unnumbered in your output, simply append an
-asterisk to the command name.  Examples of both
-numbered and unnumbered headings will appear throughout the
-balance of this sample document.
+We used a commercial RGB camera acquiring images of 640x480 pixels at 25 fps.
+The camera was located at 5 cm from the center of Nao's feet and its holder was tied to a wood base for stability purposes.
+Thus, the camera's objective was 9 cm above and $ 40\degree $ inclination with respect to the surface of the table.
+The subjects were typically located 50 cm away from the robot with the first tablet in front and the second one 30 cm to the left of the first one.
+In the same way, the experimenter was located around 60 cm to the left of the subject.
+Finally, two observers were located far away from the interaction field to manually assess
 
-Because the entire article is contained in
-the \textbf{document} environment, you can indicate the
-start of a new paragraph with a blank line in your
-input file; that is why this sentence forms a separate paragraph.
+in the context of the CoWriter project \footnote{The primary repository is
+\url{https://github.com/chili-epfl/cowriter_letter_learning}.}
 
-\subsection{Type Changes and {\subsecit Special} Characters}
-We have already seen several typeface changes in this sample.  You
-can indicate italicized words or phrases in your text with
-the command \texttt{{\char'134}textit}; emboldening with the
-command \texttt{{\char'134}textbf}
-and typewriter-style (for instance, for computer code) with
-\texttt{{\char'134}texttt}.  But remember, you do not
-have to indicate typestyle changes when such changes are
-part of the \textit{structural} elements of your
-article; for instance, the heading of this subsection will
-be in a sans serif\footnote{A third footnote, here.
-Let's make this a rather short one to
-see how it looks.} typeface, but that is handled by the
-document class file. Take care with the use
-of\footnote{A fourth, and last, footnote.}
-the curly braces in typeface changes; they mark
-the beginning and end of
-the text that is to be in the different typeface.
+\subsection{Experimental set-up}
+Figure~\ref{drawSetup} illustrates our general experimental setup: a
+face-to-face child-robot interaction with an (autonomous) Aldebran's {\sc nao}
+robot.
 
-You can use whatever symbols, accented characters, or
-non-English characters you need anywhere in your document;
-you can find a complete list of what is
-available in the \textit{\LaTeX\
-User's Guide}\cite{Lamport:LaTeX}.
+A tactile tablet (with a custom application) is used for both the robot and the
+child to write: during a typical round, the child requests the robot to write
+something (a single letter, a number or a full word), and push the tablet
+towards the robot, the robot writes on the tablet by gesturing the writing (but
+without actually physically touching the tablet), the child then pull back the
+tablet, corrects the robot's attempt by writing him/herself on top or next to
+the robot's writing (see Figure~\ref{fig:diego}), and ``send'' his/her
+demonstration to the robot by pressing a small button on the tablet. The robot
+``learns'' from this demonstration and tries again.
 
-\subsection{Math Equations}
-You may want to display math equations in three distinct styles:
-inline, numbered or non-numbered display.  Each of
-the three are discussed in the next sections.
+Since the child is assumed to take on the role of the teacher, we had to ensure
+(s)he would be able to manage by him/herself the turn-taking and the overall
+progression of the activity (moving to the next letter or word). In our design,
+the turn-taking relies on the robot prompting for feedback once it is done with
+its writing (simple sentences like ``What do you think?''), and pressing on a
+small robot icon on the tablet once the child has finished correcting. In our
+experiments, both were easy to grasp for children.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.7\columnwidth]{drawSetup}
+    \caption{\small Our experimental setup: face-to-face interaction with a {\sc
+        nao} robot.  The robot writes on the tactile tablet, the child then
+        corrects the robot by directly overwriting its letters on the tablet
+        with a stylus. An adult (either a therapist or an experimenter,
+        depending on the studies), remains next to the child to guide the work
+        (prompting, turn taking, etc.). A second tablet allows to choose words
+        and the camera captures the subject's head. In red, the focuses of attention
+        in the scene.}
+    \label{drawSetup}
+\end{figure}
+
+Implementing such a system raises several challenges that are discussed in detail in \cite{Hood:2015}.
+
+As we observe the current set-up allow us to intuitively identify several focuses of attention along the interaction such as the two tablets, the robot, the experimenter and the observers located in the diagonal.
+
+\subsection{From face detection to Focus of Attention}
+
+In our approach we use the opensource library dlib\footnote{http://dlib.net} \cite{dlib09}, to detect and extract the skeleton of the subject's face and then, use the obtained landmarkers to calculate the head pose orientation estimation.
+In this way we can estimate where the subject was looking at.
+Compared to intrusive hardware like head mounted displays or glasses but also images acquired from an RGB-D cameras for instance, this approach has a clear advantage of simplicity but also becomes accurate enough for general purposes in face-to-face interactions.
+In addition is able to handle different number and positions of participants in the scenario.
+
+%equations
+
+In this work, head orientation is used to predict a person’s focus of attention in face-to-face interactions.
+Therefore, we assume that head pose is a reliable marker of the attention focus during a social interaction, in both cases human-human, human-robot.
+Since we estimate where the subject is looking at based on its head position, several important questions arise:
+
+1. How much the head pose contributes to the real gaze of the subject?
+2. Which accuracy can we achieve during a real case scenario in terms of head pose estimation tracking?  
+3. How accurate can we predict where the subject is looking at based only on the head pose estimation?
+4. How can we evaluate the performance of the children based on the activity?
+
+% Stats from the paper
+% Stats from me
+% Real vs captured plot
+% Real vs expected plot
+
+\begin{figure}
+    \centering
+    \includegraphics[width=1\columnwidth]{realSetup}
+    \caption{\small Typical child-robot interaction scenario during the experiments}
+
+    \label{realSetup}
+\end{figure}
+
+\subsection{Head pose contribution to the gaze}
+
+\section{Results}
+
+\section{Discussion}
+The results has shown that the percentage of accuracy in the tracking is not only conditioned by the subject's amount of movement, but also by the speed of the same movements: Restless children externalize greater levels of movement decreasing the tracker accuracy.
+Additionally, several classic computer vision challenges such as the lighting conditions, occlusions and the face angle with respect to the camera are factors that contribute to reduce the accuracy.
 
@@
 \subsection{Figures}
-Like tables, figures cannot be split across pages; the
-best placement for them
-is typically the top or the bottom of the page nearest
-their initial cite.  To ensure this proper ``floating'' placement
-of figures, use the environment
-\textbf{figure} to enclose the figure and its caption.
-
-This sample document contains examples of \textbf{.eps}
-and \textbf{.ps} files to be displayable with \LaTeX.  More
-details on each of these is found in the \textit{Author's Guide}.
-
-\begin{figure}
-\centering
-\epsfig{file=fly.eps}
-\caption{A sample black and white graphic (.eps format).}
-\end{figure}
-
-\begin{figure}
-\centering
-\epsfig{file=fly.eps, height=1in, width=1in}
-\caption{A sample black and white graphic (.eps format)
-that has been resized with the \texttt{epsfig} command.}
-\end{figure}
 
@@
 \centering
-\epsfig{file=flies.eps}
+
 \caption{A sample black and white graphic (.eps format)
@@
 
-\subsection*{A {\secit Caveat} for the \TeX\ Expert}
-Because you have just been given permission to
-use the \texttt{{\char'134}newdef} command to create a
-new form, you might think you can
-use \TeX's \texttt{{\char'134}def} to create a
-new command: \textit{Please refrain from doing this!}
-Remember that your \LaTeX\ source code is primarily intended
-to create camera-ready copy, but may be converted
-to other forms -- e.g. HTML. If you inadvertently omit
-some or all of the \texttt{{\char'134}def}s recompilation will
-be, to say the least, problematic.
-
 \section{Conclusions}
-This paragraph will end the body of this sample document.
-Remember that you might still have Acknowledgments or
-Appendices; brief samples of these
-follow.  There is still the Bibliography to deal with; and
-we will make a disclaimer about that here: with the exception
-of the reference to the \LaTeX\ book, the citations in
-this paper are to articles which have nothing to
-do with the present subject and are used as
-examples only.
-%\end{document}  % This is where a 'short' article might terminate
 
@@
 % ACM needs 'a single self-contained file'!
-%
-%APPENDICES are optional
-%\balancecolumns
-\appendix
-%Appendix A
-\section{Headings in Appendices}
-The rules about hierarchical headings discussed above for
-the body of the article are different in the appendices.
-In the \textbf{appendix} environment, the command
-\textbf{section} is used to
-indicate the start of each Appendix, with alphabetic order
-designation (i.e. the first is A, the second B, etc.) and
-a title (if you include one).  So, if you need
-hierarchical structure
-\textit{within} an Appendix, start with \textbf{subsection} as the
-highest level. Here is an outline of the body of this
-document in Appendix-appropriate form:
-\subsection{Introduction}
-\subsection{The Body of the Paper}
-\subsubsection{Type Changes and  Special Characters}
-\subsubsection{Math Equations}
-\paragraph{Inline (In-text) Equations}
-\paragraph{Display Equations}
-\subsubsection{Citations}
-\subsubsection{Tables}
-\subsubsection{Figures}
-\subsubsection{Theorem-like Constructs}
-\subsubsection*{A Caveat for the \TeX\ Expert}
-\subsection{Conclusions}
-\subsection{Acknowledgments}
-\subsection{Additional Authors}
-This section is inserted by \LaTeX; you do not insert it.
-You just add the names and information in the
-\texttt{{\char'134}additionalauthors} command at the start
-of the document.
-\subsection{References}
-Generated by bibtex from your ~.bib file.  Run latex,
-then bibtex, then latex twice (to resolve references)
-to create the ~.bbl file.  Insert that ~.bbl file into
-the .tex source file and comment out
-the command \texttt{{\char'134}thebibliography}.
-% This next section command marks the start of
-% Appendix B, and does not continue the present hierarchy
-\section{More Help for the Hardy}
-The sig-alternate.cls file itself is chock-full of succinct
-and helpful comments.  If you consider yourself a moderately
-experienced to expert user of \LaTeX, you may find reading
-it useful but please remember not to change it.
+
 %\balancecolumns % GM June 2007
-% That's all folks!
 \end{document}

### 2015-07-31 sig-alternate.tex
Set-up and Approach drafts

@@
 
+The present studies has been developed in the context of the CoWriter project \footnote{The primary repository is \url{https://github.com/chili-epfl/cowriter_letter_learning}.}.
+
 \subsection{Approach}
+
+The first step is to estimate the head pose using the pin-hole model shown in equation \ref{eq:pinHole}.
+As we mention below dlib library facilitates the task by providing the 2D points \textit{m'} that composes the face silhouette.
+However, it is enough with obtaining eight spread points, such as right and left eyes, right and left tragions, sellion or nasal depression zone, pronasale or center of the nose, stomion or center of the mouth and menton.
+
+Likewise, the 3D face model is needed.
+The proposed head pose estimation method would work best with a 3D face model of the query subject itself.
+When a 3D face model of the subject is not available, as in many practical situations, a 3D face model of a subject can be used.
+This is an estimation of the same anthropometric points \textit{M'} (for North American Caucasians) in the 3D space \cite{farkas1994anthropometry}.
+Moreover, the intrinsic camera parameters are necessary to be able to compose the camera matrix \textit{A}.
+The components are expanded in equation \ref{eq:pinHole2}.
+
+\begin{equation}
+s  \; m' = A [R|t] M'
+\label{eq:pinHole}
+\end{equation}
+\begin{equation}
+s
+\begin{bmatrix}
+x \\
+y \\
+1
+\end{bmatrix}
+=
+\begin{bmatrix}
+f_x & 0 & c_x  \\
+0 & f_y & c_y  \\
+0 & 0 & 1 
+\end{bmatrix}
+\begin{bmatrix}
+r_{11} & r_{12} & r_{13} & t_x  \\
+r_{21} & r_{22} & r_{23} & t_y  \\
+r_{31} & r_{32} & r_{33} & t_z  
+\end{bmatrix}
+\begin{bmatrix}
+X \\
+Y \\
+Z \\
+1
+\end{bmatrix}
+\label{eq:pinHole2}
+\end{equation}
+
+Finally, the matrix of extrinsic parameters [R|t] is obtained by joining the rotation matrix R (calculated from the rotation vector) and translation vector \textit{t}.
+The result is a matrix projection \textbf{P} which maps a point in the 3D space onto a point in the 2D image place solving the equation \ref{eq:p}.
+
+\begin{equation}
+s  \; m' = P \; M'
+\label{eq:p}
+\end{equation}
+
+The second step is to use the matrix [R|t] to set the correct translation and orientation of the \textit{tf} frame which represents the face in the 3D space.
+In order to achieve that it is necessary first, to set the origin defined by the vector \textit{t}.
+Second, the matrix \textit{R} needs to be reformulated into a quaternion as shown in equation \ref{eq:quat} due to the smoothly interpolation between them.
+
+\begin{equation}
+\begin{bmatrix}
+qw \\
+qx \\
+qy \\
+qz
+\end{bmatrix}
+=
+\begin{bmatrix}
+\sqrt{1 + r_{11} + r_{22} + r_{33}} /2 \\
+(r_{32} - r_{23})/( 4 \cdot qw) \\
+(r_{13} - r_{31})/( 4 \cdot qw) \\
+(r_{21} - r_{12})/( 4 \cdot qw)
+\end{bmatrix}
+\label{eq:quat}
+\end{equation}
+\\
+However, it is necessary to consider the cases where the division is performed by 0 (in 180\degree rotation about the y-axis for instance) or when \textit{qw} becomes close to zero.
+
+The third step is to generate a representation of the field of view which origin and orientation are the ones defined by the face detection \textit{tf} frame.
+
 Defining accurately the human field of view in the implementation is a must.
@@
 
-We used a commercial RGB camera acquiring images of 640x480 pixels at 25 fps.
-The camera was located at 5 cm from the center of Nao's feet and its holder was tied to a wood base for stability purposes.
-Thus, the camera's objective was 9 cm above and $ 40\degree $ inclination with respect to the surface of the table.
-The subjects were typically located 50 cm away from the robot with the first tablet in front and the second one 30 cm to the left of the first one.
-In the same way, the experimenter was located around 60 cm to the left of the subject.
-Finally, two observers were located far away from the interaction field to manually assess
+Assuming that the \textit{tf} frames representing the robotic agent, the two tablets, the experimenter and the observers are static, need to be monitored.
+This task consists on evaluating the intersection between the \textit{tf} frames and the field of view: If one focus of attention is within the region defined by the field of view, we assume the subject is looking at it.
+In order to solve such operation mathematically it is necessary to compute the transformation matrix between the coordinate frames to be able to locate the monitored frames \textit{A} in the subject's coordinate system \textit{B} (see equation \ref{eq:transform}).
 
-in the context of the CoWriter project \footnote{The primary repository is
-\url{https://github.com/chili-epfl/cowriter_letter_learning}.}
+\begin{equation}
+v' = B(A^-1)v
+\end{equation}
+
+where \textit{v} is a point in \textit{A} that becomes v' in coordinate system \textit{B}.
+However, in practice there are, at least, three different \textit{tf} frames involved in the transformation (subject-camera, camera-base, base-focus).
+Following the transformation result, the distance of the monitored frame with respect to the x-axis face \textit{tf} frame, $ d_{x_{axis}} $, representing the subject's face can be computed as simple as in equation \ref{eq:pitagoras}.
+Always considering the x-axis as the one defining the face's orientation and thus field of view's main axis.
+
+\begin{equation}
+d_{x_{axis}} = \sqrt{t_y^2 + t_z^2}
+\label{eq:pitagoras}
+\end{equation}
+
+It is necessary to compare the distance acquired $ d_{x_{axis}} $ with the radius of the field of view at the target \textit{tf} frame x-coordinate $ r_{fov} $ as shown in equation \ref{eq:fovtf}.
+
+\begin{equation}
+r_{fov} = tan\left(\frac{fov}{2}\right) \cdot t_x
+\label{eq:fovtf}
+\end{equation}
+
+where \textit{fov} is the aperture angle of the field of view. Finally, if $ d_{x_{axis}}<r_{fov} $, the target is within the subject's field of view.
+
+% Plot to clarify this??
 
 \subsection{Experimental set-up}
-Figure~\ref{drawSetup} illustrates our general experimental setup: a
+Figure~\ref{realSetup} illustrates our general experimental setup: a
 face-to-face child-robot interaction with an (autonomous) Aldebran's {\sc nao}
 robot.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=1\columnwidth]{realSetup}
+    \caption{\small Our experimental setup: face-to-face interaction with a {\sc
+            nao} robot.  The robot writes on the tactile tablet, the child then
+            corrects the robot by directly overwriting its letters on the tablet
+            with a stylus. An adult (either a therapist or an experimenter,
+            depending on the studies), remains next to the child to guide the work. 
+            A second tablet allows to choose words and the camera captures the subject's face.}
+    \label{realSetup}
+\end{figure}
 
@@
 
+Implementing such a system raises several challenges that are discussed in detail in \cite{Hood:2015}.
+
+An RGB camera has been used to acquire images of 640x480 pixels at 25 fps.
+The camera was located at 5 cm from the center of Nao's feet and its holder was tied to a wood base for stability purposes.
+Thus, the camera's objective was 9 cm above and $ 40\degree $ inclination with respect to the surface of the table.
+The subjects were typically located 50 cm away from the robot with the first tablet in front and the second one 30 cm to the left of the first one.
+In the same way, the experimenter was located around 60 cm to the left of the subject.
+Finally, two observers were located far away from the interaction field to manually assess the state of the interaction.
+
+Figure \ref{drawSetup} allows us to intuitively identify several focuses of attention along the interaction such as the two tablets, the robot, the experimenter and the observers located in the diagonal.
+
 \begin{figure}
@@
     \includegraphics[width=0.7\columnwidth]{drawSetup}
-    \caption{\small Our experimental setup: face-to-face interaction with a {\sc
-        nao} robot.  The robot writes on the tactile tablet, the child then
-        corrects the robot by directly overwriting its letters on the tablet
-        with a stylus. An adult (either a therapist or an experimenter,
-        depending on the studies), remains next to the child to guide the work
-        (prompting, turn taking, etc.). A second tablet allows to choose words
-        and the camera captures the subject's head. In red, the focuses of attention
-        in the scene.}
+    \caption{\small Graphical representation of he set-up and the possible children focuses of attention in the scene, in red.}
     \label{drawSetup}
@@
 
-Implementing such a system raises several challenges that are discussed in detail in \cite{Hood:2015}.
-
-As we observe the current set-up allow us to intuitively identify several focuses of attention along the interaction such as the two tablets, the robot, the experimenter and the observers located in the diagonal.
-
 \subsection{From face detection to Focus of Attention}
+According to \cite{stiefelhagen2002tracking}
 
@@
 
-%equations
-
 In this work, head orientation is used to predict a person’s focus of attention in face-to-face interactions.
@@
 
-1. How much the head pose contributes to the real gaze of the subject?
-2. Which accuracy can we achieve during a real case scenario in terms of head pose estimation tracking?  
-3. How accurate can we predict where the subject is looking at based only on the head pose estimation?
-4. How can we evaluate the performance of the children based on the activity?
+\begin{enumerate}
+\item How much the head pose contributes to the real gaze of the subject?
+\item Which percentage can we achieve during a real case scenario in terms of head pose estimation tracking?  
+\item How accurate can we predict where the subject is looking at based only on the head pose estimation?
+\item How can we evaluate the performance of the children based on the activity?
+\end{enumerate}
 
@@
 
-\begin{figure}
-    \centering
-    \includegraphics[width=1\columnwidth]{realSetup}
-    \caption{\small Typical child-robot interaction scenario during the experiments}
-
-    \label{realSetup}
-\end{figure}
+\section{Results}
 
@@
 
-\section{Results}
+\begin{table}
+	\centering
+	\caption{Tracking precision in face-to-face scenario}
+	\begin{tabular}{|c|c|c|c|c|c|c|c|} \hline
+	Subject & 1 & 2 & 3 & 4 & 5 & 6 & Avg\\ \hline
+	Tracked & 78\% & \% & \% & \% & \% & \% & \% \\ \hline
+	\end{tabular}
+\end{table}
 
@@
 The results has shown that the percentage of accuracy in the tracking is not only conditioned by the subject's amount of movement, but also by the speed of the same movements: Restless children externalize greater levels of movement decreasing the tracker accuracy.
-Additionally, several classic computer vision challenges such as the lighting conditions, occlusions and the face angle with respect to the camera are factors that contribute to reduce the accuracy.
-
-\subsubsection{Inline (In-text) Equations}
-A formula that appears in the running text is called an
-inline or in-text formula.  It is produced by the
-\textbf{math} environment, which can be
-invoked with the usual \texttt{{\char'134}begin. . .{\char'134}end}
-construction or with the short form \texttt{\$. . .\$}. You
-can use any of the symbols and structures,
-from $\alpha$ to $\omega$, available in
-\LaTeX\cite{Lamport:LaTeX}; this section will simply show a
-few examples of in-text equations in context. Notice how
-this equation: \begin{math}\lim_{n\rightarrow \infty}x=0\end{math},
-set here in in-line math style, looks slightly different when
-set in display style.  (See next section).
-
-\subsubsection{Display Equations}
-A numbered display equation -- one set off by vertical space
-from the text and centered horizontally -- is produced
-by the \textbf{equation} environment. An unnumbered display
-equation is produced by the \textbf{displaymath} environment.
-
-Again, in either environment, you can use any of the symbols
-and structures available in \LaTeX; this section will just
-give a couple of examples of display equations in context.
-First, consider the equation, shown as an inline equation above:
-\begin{equation}\lim_{n\rightarrow \infty}x=0\end{equation}
-Notice how it is formatted somewhat differently in
-the \textbf{displaymath}
-environment.  Now, we'll enter an unnumbered equation:
-\begin{displaymath}\sum_{i=0}^{\infty} x + 1\end{displaymath}
-and follow it with another numbered equation:
-\begin{equation}\sum_{i=0}^{\infty}x_i=\int_{0}^{\pi+2} f\end{equation}
-just to demonstrate \LaTeX's able handling of numbering.
-
-\subsection{Citations}
-Citations to articles \cite{bowman:reasoning,
-clark:pct, braams:babel, herlihy:methodology},
-conference proceedings \cite{clark:pct} or
-books \cite{salas:calculus, Lamport:LaTeX} listed
-in the Bibliography section of your
-article will occur throughout the text of your article.
-You should use BibTeX to automatically produce this bibliography;
-you simply need to insert one of several citation commands with
-a key of the item cited in the proper location in
-the \texttt{.tex} file \cite{Lamport:LaTeX}.
-The key is a short reference you invent to uniquely
-identify each work; in this sample document, the key is
-the first author's surname and a
-word from the title.  This identifying key is included
-with each item in the \texttt{.bib} file for your article.
-
-The details of the construction of the \texttt{.bib} file
-are beyond the scope of this sample document, but more
-information can be found in the \textit{Author's Guide},
-and exhaustive details in the \textit{\LaTeX\ User's
-Guide}\cite{Lamport:LaTeX}.
-
-This article shows only the plainest form
-of the citation command, using \texttt{{\char'134}cite}.
-This is what is stipulated in the SIGS style specifications.
-No other citation format is endorsed or supported.
+Additionally, several recurrent computer vision challenges such as the lighting conditions, occlusions and the face angle with respect to the camera are factors that contribute to reduce the accuracy.
 
 \subsection{Tables}
-Because tables cannot be split across pages, the best
-placement for them is typically the top of the page
-nearest their initial cite.  To
-ensure this proper ``floating'' placement of tables, use the
-environment \textbf{table} to enclose the table's contents and
-the table caption.  The contents of the table itself must go
-in the \textbf{tabular} environment, to
-be aligned properly in rows and columns, with the desired
-horizontal and vertical rules.  Again, detailed instructions
-on \textbf{tabular} material
-is found in the \textit{\LaTeX\ User's Guide}.
-
-Immediately following this sentence is the point at which
-Table 1 is included in the input file; compare the
-placement of the table here with the table in the printed
-dvi output of this document.
-
-\begin{table}
-\centering
-\caption{Frequency of Special Characters}
-\begin{tabular}{|c|c|l|} \hline
-Non-English or Math&Frequency&Comments\\ \hline
-\O & 1 in 1,000& For Swedish names\\ \hline
-$\pi$ & 1 in 5& Common in math\\ \hline
-\$ & 4 in 5 & Used in business\\ \hline
-$\Psi^2_1$ & 1 in 40,000& Unexplained usage\\
-\hline\end{tabular}
-\end{table}
 
@@
 
-\subsection{Figures}
-
-As was the case with tables, you may want a figure
-that spans two columns.  To do this, and still to
-ensure proper ``floating'' placement of tables, use the environment
-\textbf{figure*} to enclose the figure and its caption.
-and don't forget to end the environment with
-{figure*}, not {figure}!
-
-\begin{figure*}
-\centering
-
-\caption{A sample black and white graphic (.eps format)
-that needs to span two columns of text.}
-\end{figure*}
-
-Note that either {\textbf{.ps}} or {\textbf{.eps}} formats are
-used; use
-the \texttt{{\char'134}epsfig} or \texttt{{\char'134}psfig}
-commands as appropriate for the different file types.
-
-\begin{figure}
-\centering
-\caption{A sample black and white graphic (.ps format) that has
-been resized with the \texttt{psfig} command.}
-\vskip -6pt
-\end{figure}
-
-\subsection{Theorem-like Constructs}
-Other common constructs that may occur in your article are
-the forms for logical constructs like theorems, axioms,
-corollaries and proofs.  There are
-two forms, one produced by the
-command \texttt{{\char'134}newtheorem} and the
-other by the command \texttt{{\char'134}newdef}; perhaps
-the clearest and easiest way to distinguish them is
-to compare the two in the output of this sample document:
-
-This uses the \textbf{theorem} environment, created by
-the\linebreak\texttt{{\char'134}newtheorem} command:
-\newtheorem{theorem}{Theorem}
-\begin{theorem}
-Let $f$ be continuous on $[a,b]$.  If $G$ is
-an antiderivative for $f$ on $[a,b]$, then
-\begin{displaymath}\int^b_af(t)dt = G(b) - G(a).\end{displaymath}
-\end{theorem}
-
-The other uses the \textbf{definition} environment, created
-by the \texttt{{\char'134}newdef} command:
-\newdef{definition}{Definition}
-\begin{definition}
-If $z$ is irrational, then by $e^z$ we mean the
-unique number which has
-logarithm $z$: \begin{displaymath}{\log e^z = z}\end{displaymath}
-\end{definition}
-
-Two lists of constructs that use one of these
-forms is given in the
-\textit{Author's  Guidelines}.
-
-There is one other similar construct environment, which is
-already set up
-for you; i.e. you must \textit{not} use
-a \texttt{{\char'134}newdef} command to
-create it: the \textbf{proof} environment.  Here
-is a example of its use:
-\begin{proof}
-Suppose on the contrary there exists a real number $L$ such that
-\begin{displaymath}
-\lim_{x\rightarrow\infty} \frac{f(x)}{g(x)} = L.
-\end{displaymath}
-Then
-\begin{displaymath}
-l=\lim_{x\rightarrow c} f(x)
-= \lim_{x\rightarrow c}
-\left[ g{x} \cdot \frac{f(x)}{g(x)} \right ]
-= \lim_{x\rightarrow c} g(x) \cdot \lim_{x\rightarrow c}
-\frac{f(x)}{g(x)} = 0\cdot L = 0,
-\end{displaymath}
-which contradicts our assumption that $l\neq 0$.
-\end{proof}
-
-Complete rules about using these environments and using the
-two different creation commands are in the
-\textit{Author's Guide}; please consult it for more
-detailed instructions.  If you need to use another construct,
-not listed therein, which you want to have the same
-formatting as the Theorem
-or the Definition\cite{salas:calculus} shown above,
-use the \texttt{{\char'134}newtheorem} or the
-\texttt{{\char'134}newdef} command,
-respectively, to create it.
-
 \section{Conclusions}
@@
 \section{Acknowledgments}
-This section is optional; it is a location for you
-to acknowledge grants, funding, editing assistance and
-what have you.  In the present case, for example, the
-authors would like to thank Gerald Murray of ACM for
-his help in codifying this \textit{Author's Guide}
-and the \textbf{.cls} and \textbf{.tex} files that it describes.
 

### 2015-08-03 sig-alternate.tex
Results part started with comparison

@@
 \title{Alternate {\ttlit ACM} SIG Proceedings Paper in LaTeX
-Format\titlenote{(Produces the permission block, and
-copyright information). For use with
-SIG-ALTERNATE.CLS. Supported by ACM.}}
-\subtitle{[Extended Abstract]
-\titlenote{A full version of this paper is available as
-\textit{Author's Guide to Preparing ACM SIG Proceedings Using
-\LaTeX$2_\epsilon$\ and BibTeX} at
-\texttt{www.acm.org/eaddress.htm}}}
+Format}
+\author{Fernando Garcia\qquad Séverin Lemaignan \qquad Pierre Dillenbourg\\Computer-Human Interaction in Learning and Instruction Laboratory (CHILI)\\École Polytechnique Fédérale de Lausanne (EPFL)}
+
 %
@@
 
-\numberofauthors{8} %  in this sample file, there are a *total*
+%\numberofauthors{4} %  in this sample file, there are a *total*
 % of EIGHT authors. SIX appear on the 'first-page' (for formatting
@@
 %
-\author{
+%\author{
 % You can go ahead and credit any number of authors here,
@@
 % 1st. author
-\alignauthor
-Ben Trovato\titlenote{Dr.~Trovato insisted his name be first.}\\
-       \affaddr{Institute for Clarity in Documentation}\\
-       \affaddr{1932 Wallamaloo Lane}\\
-       \affaddr{Wallamaloo, New Zealand}\\
-       \email{trovato@corporation.com}
-% 2nd. author
-\alignauthor
-G.K.M. Tobin\titlenote{The secretary disavows
-any knowledge of this author's actions.}\\
-       \affaddr{Institute for Clarity in Documentation}\\
-       \affaddr{P.O. Box 1212}\\
-       \affaddr{Dublin, Ohio 43017-6221}\\
-       \email{webmaster@marysville-ohio.com}
-% 3rd. author
-\alignauthor Lars Th{\o}rv{\"a}ld\titlenote{This author is the
-one who did all the really hard work.}\\
-       \affaddr{The Th{\o}rv{\"a}ld Group}\\
-       \affaddr{1 Th{\o}rv{\"a}ld Circle}\\
-       \affaddr{Hekla, Iceland}\\
-       \email{larst@affiliation.org}
-\and  % use '\and' if you need 'another row' of author names
-% 4th. author
-\alignauthor Lawrence P. Leipuner\\
-       \affaddr{Brookhaven Laboratories}\\
-       \affaddr{Brookhaven National Lab}\\
-       \affaddr{P.O. Box 5000}\\
-       \email{lleipuner@researchlabs.org}
-% 5th. author
-\alignauthor Sean Fogarty\\
-       \affaddr{NASA Ames Research Center}\\
-       \affaddr{Moffett Field}\\
-       \affaddr{California 94035}\\
-       \email{fogartys@amesres.org}
-% 6th. author
-\alignauthor Charles Palmer\\
-       \affaddr{Palmer Research Laboratories}\\
-       \affaddr{8600 Datapoint Drive}\\
-       \affaddr{San Antonio, Texas 78229}\\
-       \email{cpalmer@prl.com}
-}
-% There's nothing stopping you putting the seventh, eighth, etc.
-% author on the opening page (as the 'third row') but we ask,
-% for aesthetic reasons that you place these 'additional authors'
-% in the \additional authors block, viz.
-\additionalauthors{Additional authors: John Smith (The Th{\o}rv{\"a}ld Group,
-email: {\texttt{jsmith@affiliation.org}}) and Julius P.~Kumquat
-(The Kumquat Consortium, email: {\texttt{jpkumquat@consortium.net}}).}
-\date{30 July 1999}
-% Just remember to make sure that the TOTAL number of authors
-% is the number that will appear on the first page PLUS the
-% number that will appear in the \additionalauthors section.
+%\alignauthor
+%Ben Trovato\titlenote{Dr.~Trovato insisted his name be first.}\\
+%       \affaddr{Institute for Clarity in Documentation}\\
+%       \affaddr{1932 Wallamaloo Lane}\\
+%       \affaddr{Wallamaloo, New Zealand}\\
+%       \email{trovato@corporation.com}
+%% 2nd. author
+%\alignauthor
+%G.K.M. Tobin\titlenote{The secretary disavows
+%any knowledge of this author's actions.}\\
+%       \affaddr{Institute for Clarity in Documentation}\\
+%       \affaddr{P.O. Box 1212}\\
+%       \affaddr{Dublin, Ohio 43017-6221}\\
+%       \email{webmaster@marysville-ohio.com}
+%% 3rd. author
+%\alignauthor Lars Th{\o}rv{\"a}ld\titlenote{This author is the
+%one who did all the really hard work.}\\
+%       \affaddr{The Th{\o}rv{\"a}ld Group}\\
+%       \affaddr{1 Th{\o}rv{\"a}ld Circle}\\
+%       \affaddr{Hekla, Iceland}\\
+%       \email{larst@affiliation.org}
+%\and  % use '\and' if you need 'another row' of author names
+%% 4th. author
+%\alignauthor Lawrence P. Leipuner\\
+%       \affaddr{Brookhaven Laboratories}\\
+%       \affaddr{Brookhaven National Lab}\\
+%       \affaddr{P.O. Box 5000}\\
+%       \email{lleipuner@researchlabs.org}
+%}
 
@@
 
-% A category with the (minimum) three required fields
-\category{H.4}{Information Systems Applications}{Miscellaneous}
-%A category including the fourth, optional field follows...
-\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]
-
-\terms{Theory}
-
-\keywords{ACM proceedings, \LaTeX, text tagging}
+%% A category with the (minimum) three required fields
+%\category{H.4}{Information Systems Applications}{Miscellaneous}
+%%A category including the fourth, optional field follows...
+%\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]
+%
+%\terms{Theory}
+%
+%\keywords{ACM proceedings, \LaTeX, text tagging}
 
@@
 
-\subsection{Approach}
+\section{Approach}
+
+In our approach we use the open-source library dlib\footnote{http://dlib.net} \cite{dlib09}, to detect and extract the skeleton of the subject's face and then, use the obtained landmarkers to calculate the head pose orientation estimation.
+In this way we can estimate where the subject was looking at.
+Compared to intrusive hardware like head mounted displays or glasses but also images acquired from an RGB-D cameras for instance, this approach has a clear advantage of simplicity but also becomes accurate enough for general purposes in face-to-face interactions.
+In addition is able to handle different number and positions of participants in the scenario.
 
 The first step is to estimate the head pose using the pin-hole model shown in equation \ref{eq:pinHole}.
-As we mention below dlib library facilitates the task by providing the 2D points \textit{m'} that composes the face silhouette.
+As we mention before dlib library facilitates the task by providing the 2D points \textit{m'} that composes the face silhouette.
 However, it is enough with obtaining eight spread points, such as right and left eyes, right and left tragions, sellion or nasal depression zone, pronasale or center of the nose, stomion or center of the mouth and menton.
@@
 \begin{equation}
-v' = B(A^-1)v
+v' = B(A^{-1})v
+\label{eq:transform}
 \end{equation}
@@
 
-\subsection{Experimental set-up}
+\subsection{Experimental Set-up}
+
+The target of this study were 6 children with ages compressed between 5 and 6 years old.
+The experiments were developed individually and consisted on a writing activity by turns and one story telling performed by the robot at a given time of the interaction.
+
 Figure~\ref{realSetup} illustrates our general experimental setup: a
@@
 
-\subsection{From face detection to Focus of Attention}
-According to \cite{stiefelhagen2002tracking}
+\section{Results}
 
-In our approach we use the opensource library dlib\footnote{http://dlib.net} \cite{dlib09}, to detect and extract the skeleton of the subject's face and then, use the obtained landmarkers to calculate the head pose orientation estimation.
-In this way we can estimate where the subject was looking at.
-Compared to intrusive hardware like head mounted displays or glasses but also images acquired from an RGB-D cameras for instance, this approach has a clear advantage of simplicity but also becomes accurate enough for general purposes in face-to-face interactions.
-In addition is able to handle different number and positions of participants in the scenario.
-
-In this work, head orientation is used to predict a person’s focus of attention in face-to-face interactions.
+In this work, head orientation is used to predict a person's focus of attention in face-to-face interactions.
 Therefore, we assume that head pose is a reliable marker of the attention focus during a social interaction, in both cases human-human, human-robot.
@@
 \begin{enumerate}
-\item How much the head pose contributes to the real gaze of the subject?
 \item Which percentage can we achieve during a real case scenario in terms of head pose estimation tracking?  
 \item How accurate can we predict where the subject is looking at based only on the head pose estimation?
-\item How can we evaluate the performance of the children based on the activity?
+\item How can we evaluate the performance of the children based on the focus of attention?
 \end{enumerate}
@@
 
-\section{Results}
+\subsection{From Face Detection to Focus of Attention}
 
-\subsection{Head pose contribution to the gaze}
+According to \cite{stiefelhagen2002tracking} the head orientation's contribution in overall gaze direction is 68.9\% showing agreement between both measurements.
+In addition, it shows that the head orientation alone can get an average accuracy of 88.7\% detecting the focus of attention in a meeting application scenario using a head-mounted display with eye and head tracking.
 
-\begin{table}
+Our results for a peer-to-peer work scenarios using an approach based on RGB camera are summarized in table \ref{tab:results} where the \textit{expected} is the focus of attention that given the state of the system, it is likely to attract the subject's gaze.
+In the same manner, the \textit{captured} is the focus of attention our approach provides and the \textit{real}, the ground truth of the subject's gaze direction based on manual annotations of two evaluators.
+
+\begin{table}[h!]
 	\centering
-	\caption{Tracking precision in face-to-face scenario}
+	\caption{Tracking accuracy in face-to-face scenario}
 	\begin{tabular}{|c|c|c|c|c|c|c|c|} \hline
-	Subject & 1 & 2 & 3 & 4 & 5 & 6 & Avg\\ \hline
-	Tracked & 78\% & \% & \% & \% & \% & \% & \% \\ \hline
+	\small Subject & 1 & 2 & 3 & 4 & 5 & 6 & Avg\\ \hline
+	\small Track(\%) & \small78.7 & \small81.1 & \small73.7 & \small75.6 & \small77.1 & \small68.7 & \small75.8 \\ \hline
+	\begin{tabular}{@{}c@{}}\small Cosine \\ \small affinity \end{tabular} & \small0.58 & \small & \small & \small & \small & \small & \small \\ \hline
+	\begin{tabular}{@{}c@{}}\small Real vs \\ \small expected(\%)\end{tabular} & \small87.3 & \small  & \small & \small & \small & \small & \small \\ \hline
+	\begin{tabular}{@{}c@{}}\small Real vs \\ \small captured(\%)\end{tabular} & \small59.4 & \small & \small & \small & \small & \small & \small \\ \hline
 	\end{tabular}
+	\label{tab:results}
 \end{table}
 
-\section{Discussion}
 The results has shown that the percentage of accuracy in the tracking is not only conditioned by the subject's amount of movement, but also by the speed of the same movements: Restless children externalize greater levels of movement decreasing the tracker accuracy.
@@
 
-\subsection{Tables}
+\subsection{Subject's Focus of Attention Analysis}
 
-To set a wider table, which takes up the whole width of
-the page's live area, use the environment
-\textbf{table*} to enclose the table's contents and
-the table caption.  As with a single-column table, this wide
-table will ``float" to a location deemed more desirable.
-Immediately following this sentence is the point at which
-Table 2 is included in the input file; again, it is
-instructive to compare the placement of the
-table here with the table in the printed dvi
-output of this document.
+% This two plots are provisional. We expect to improve them with images and boxes
+\begin{figure*}
+    \centering
+    \includegraphics[width=2\columnwidth]{realExpected}
+    \caption{\small In blue, real focus accoding to two observer assessment. In green, the expected focus based on the system state and in cyan, the overlapping between them.}
+    \label{fig:realExpected}
+\end{figure*}
 
-\begin{table*}
+\begin{figure*}
     \centering
-\caption{Some Typical Commands}
-\begin{tabular}{|c|c|l|} \hline
-Command&A Number&Comments\\ \hline
-\texttt{{\char'134}alignauthor} & 100& Author alignment\\ \hline
-\texttt{{\char'134}numberofauthors}& 200& Author enumeration\\ \hline
-\texttt{{\char'134}table}& 300 & For tables\\ \hline
-\texttt{{\char'134}table*}& 400& For wider tables\\ \hline\end{tabular}
-\end{table*}
-% end the environment with {table*}, NOTE not {table}!
+    \includegraphics[width=2\columnwidth]{realCaptured}
+    \caption{\small In blue, real focus accoding to two observer assessment. In green, the captured focus using our aproach and in cyan, the overlapping between them.}
+    \label{fig:realCaptured}
+\end{figure*}
+% % % % % % % % % % % % % % % % % % % %
+
+In several occasions during the interaction, none of the focuses of attention were within the field of view due to the time transition from one focus to another.
+In order to deal with these few cases we assume that the minimum distance between the height of the field of view (cone) and the targets corresponds to the most likely focus of attention.
+
+However, we also want to distinguish such cases when the subject is directly not looking where the defined targets are located, but towards a region far from the interaction field.
+These cases are not common but they can be easily identified by setting a horizontal threshold to the field of view.
+
+Several results can extracted from figure \ref{fig:realExpected}.
+First, the matching curve shows the overlapping between the expected and the real focuses of attention.
+As we can see there is a fluctuation at the beginning of the interaction suggesting a novelty effect.
+It suggests that the adaptation time is around the first 2 minutes in average.
+
+Second, once the turn-taking is established, a pattern becomes visible during the subject's feedback, when a word is being written by the children: Such pattern shows a high intermittent gaze frequency from the tablet to the selection.
+It shows that the subject is using the word model in the selection tablet to provide a better answer to the robot.
+But not only that, the subjects also decrease the response time after each shift as well as the gaze frequency towards the model suggesting and improvement over time.
+Moreover, when a new word is selected for playing, the same behavior is replicated.
+
+A third trend that can be observed is the continuous 'look for approval' situation where the subjects look at the experimenter in order to provide feedback about the robot's response before providing their own one.
+Such transition tablet-facilitator becomes more frequent in several cases, for instance: An unexpected situation is faced or the facilitator make a suggestion.
+
+Finally, the subjects set their attention more often to the tablet displaying the generated hand-writing during the robot feedback rather than to the robot itself.
 
@@
 %ACKNOWLEDGMENTS are optional
-\section{Acknowledgments}
+\section*{Acknowledgments}
+This research was partially supported by the Funda\c{c}\~{a}o para a Ci\^{e}ncia
+e a Tecnologia (FCT) with reference UID/CEC/ 50021/2013, and by the Swiss
+National Science Foundation through the National Centre of Competence in
+Research Robotics.
 
-%
-% The following two commands are all you need in the
-% initial runs of your .tex file to
-% produce the bibliography for the citations in your paper.
 \bibliographystyle{abbrv}
-\bibliography{sigproc}  % sigproc.bib is the name of the Bibliography in this case
-% You must have a proper ".bib" file
-%  and remember to run:
-% latex bibtex latex latex
-% to resolve all references
-%
-% ACM needs 'a single self-contained file'!
-
-%\balancecolumns % GM June 2007
+\bibliography{sigproc}
 \end{document}

### 2015-08-08 sig-alternate.tex
Introduction draft done

@@
 % --- Author Metadata here ---
-\conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}
+\conferenceinfo{HRI}{'15 Chrischurch, New Zealand.}
 %\CopyrightYear{2007} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
@@
 Format}
-\author{Fernando Garcia\qquad Séverin Lemaignan \qquad Pierre Dillenbourg\\Computer-Human Interaction in Learning and Instruction Laboratory (CHILI)\\École Polytechnique Fédérale de Lausanne (EPFL)}
+\author{Fernando Garcia \qquad Séverin Lemaignan \qquad Alexis Jacq \qquad Pierre Dillenbourg\\Computer-Human Interaction in Learning and Instruction Laboratory (CHILI)\\École Polytechnique Fédérale de Lausanne (EPFL)}
 
@@
 \begin{abstract}
-In long term interactions between a human and a robot, it is necessary that the agent exhibits some sort of social intelligence to be able to engage the human providing the appropriate response to specific face-to-face situations.
-In a educational context, this feature becomes a must due to the fact that engagement plays a very important role in the learning process.
-Therefore, maximizing the interaction time at a high engagement level may provide learning benefits to the children.
-In this work, a real-time attention tracker based on the head pose estimation using an RGB camera is proposed to be able to identify faults during the interaction and eventually detect low engagement level situations.
-In this paper we want to show a practical approach to evaluate the engagement that arises from the interaction reporting several experiences in the field, with the robot Nao.
+In long term interactions between humans and robots, it is necessary that the agent exhibits some sort of social intelligence to be able to engage the human providing the appropriate response to specific face-to-face situations.
+In a educational context, this feature becomes a must due to the fact that engagement plays a relevant role in the learning process.
+Therefore, maximizing the interaction time at a high engagement level provides learning benefits.
+In this work, a visual focus of attention based on head pose estimation using a camera is proposed to be able to monitor and assess the quality of the interaction.
+We want to show a practical approach to evaluate the engagement that arises from the interaction reporting several experiences in the field, with the robot Nao and discussing the experimental results.
 \end{abstract}
@@
 %\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]
-%
-%\terms{Theory}
-%
-%\keywords{ACM proceedings, \LaTeX, text tagging}
+
+\terms{Experimentation, Human Factors.}
+
+\keywords{Human-Robot Interaction; Social Robotics; Children.}
 
 \section{Introduction}
-The use of robots in the learning activity presents an opportunity for the children
-to interact with an embodied, physical agent as part of the learning experience. Furthermore,
-the use of robots in handwriting education comes with the potential to engage the child in
-meta-cognition through the learning by teaching paradigm, wherein a student takes the role of
+
+In the recent years, research is pushing forward in building intelligent systems and environments that try to maximize the benefits of cooperation with humans based on verbal and non-verbal communication.
+The interaction between social robots and humans provide a rich set of behaviors, and reveals a need of capturing the significant passive information provided by the user during the interaction process.
+Such information should be used to adapt the system to the specific user's needs in order to be able to provide a social or emotional intelligence to the agents.
+
+While new algorithms and robotic technologies ensure better cognitive and interactive abilities, a deeper study about the benefits in the social interaction needs to be done: Is not enough to exhibit human-like capabilities in social tasks, but to do it in a human-like manner.
+For instance, several studies showed that humans prefer to interact with a robots whose task outcome is delayed, shows uncertainty or an undesired output \cite{Admoni,Short}.
+In fact, the ability to adapt has been proved to be beneficial in long-term interactions \cite{Tielman:2014, Lim:2014}, specifically with children, manifesting a more positive attitude over time.
+
+In order to build intelligent agents, it is necessary to equip them with perceptive capabilities based on sensors an algorithms: face and object tracking and recognition \cite{Zhao:2003, Jafri:2014}, path planning \cite{Galceran:2013}, speech recognition \cite{brick2007incremental} or task learning \cite{calinon2007learning} are some of the current research topics.
+However, each algorithm is evaluated in their own metric space of its specific problem: efficiency, accuracy or time are some few examples.
+In \cite{anzalone} is exposed the need to evaluate such processes in terms of their capability to obtain the desired effect in a human-robot interaction context through several proposed metrics around engagement as quality experience indicator.
+
+Engagement describes a cognitive, affective (specifically intrinsic motivation), and behavioral state of interaction with a computer application that ``makes the user want to be there" \cite{OBrien:2010}.
+Engaging inter-actions were thought to involve attention, intrinsic interest, interactivity, perceived control and choice, functionality and motivation.
+
+In this way, in situ Visual Focus of Attention (VFOA) estimation is advantageous to human-robot interaction and a major contribution towards a better understanding of the interaction quality.
+\cite{yarbus1967eye, barber1976perception} suggest that human's interest is highly correlated with what they look at and the existing relationship between gaze and attention during social interaction has been exposed in \cite{argyle1969social}.
+
+One of the most interesting fields where to develop this research is in a learning context since children's engagement sustainability over time is one of the most important variables to make
+the process of learning effective \cite{Umbach}.
+It is well known that the role of the teacher \cite{smith2005pedagogies}, as well as the way the activity is formulated do matters in the learning efficiency.
+Moreover, the use of robots in the learning activity presents an opportunity for children
+to interact with an embodied, physical agent as part of the learning experience.
+
+Furthermore, while \cite{zhao2012learning} recommends the development of a teachable agent that is actively, rather than passively, engaged in the learning by teaching interaction, the increased learning gain results attained in \cite{okita2006observation} were done so even with the pupil being taught having been instructed to not contribute much information.
+This suggests that it may not be necessary, from a learning by teaching point of view, for the teachable agent to participate actively in the interaction to incite educational benefits.
+However, from an engagement point of view, this may not necessarily be true.
+
+In fact, the use of robots in education comes with the potential to engage the child in meta-cognition through the learning by teaching paradigm, wherein a student takes the role of
 a teacher and experiences stronger educational benefits as a result (such as in \cite{Palinscar1984}).
 
-It becomes a useful to capture the significant passive information provided by the
-user during the interaction and adapt the situation to it.
+The focus of this research is to evaluate the quality of the interaction using the focus of attention as an engagement indicator in a complex scenario using an RGB camera.
+The present studies has been developed in the context of the CoWriter project \footnote{The primary repository is \url{https://github.com/chili-epfl/cowriter_letter_learning}.}.
+However, in a child-robot interaction context, avoiding 'invasive' technologies such as wearable devices is a must to keep the experience as natural as possible.
 
-The present studies has been developed in the context of the CoWriter project \footnote{The primary repository is \url{https://github.com/chili-epfl/cowriter_letter_learning}.}.
+%Needs to be redifined
+\section{Related Work}
+
+It is a model based approach to head pose estimation that finds correspondences between facial landmarks points and their respective locations in a face model -> plug and play
+
+In this research, we address the problem of tracking the children's focus of attention during a long term child-robot interaction in a learning context, first.
+And second, we provide qualitative and quantitative evaluation of the outcome provided using a first person RGB camera approach.
+
+In neural networks requires a previous image acquisition and training as well as a fixed point of view correspondent to the data acquired.
+-> The systems only works for a user standing within a certain region in front of an inmobile robot and probabilisty maps the users head orientation to some fixed focus target.
+
+The evaluation of response times to provide an estimation of the user's engagement based on the correctness has been proposed in a learning context \cite{Beck} before.
+However, several aspects like the inter-subject variability or the environment complexity make it hardly applicable in a child-robot interaction
+
+PAIVA curvature
+
+%Needs to be redifined
+\section{Research Questions}
+Measurement
+The participants' VFOA acquisition was the only dependent variable. This included their gaze times and patterns of looking or not looking to the targets. Data from a robot head-
+mounted Microsoft Kinect camera was coded post-hoc to quantify these measures.
+
+Hypothesis and Predictions
+It is hypothesized that there will be no generalizable pattern of participant VFOA behavior, necessitating online, in situ VFOA estimation capabilities to enable robots to communicate intentionally.
+
+%The identification of the patterns can lead to a better understanding to the children progression. Less time means better understanding a move to next level.
 
@@
 In this way we can estimate where the subject was looking at.
-Compared to intrusive hardware like head mounted displays or glasses but also images acquired from an RGB-D cameras for instance, this approach has a clear advantage of simplicity but also becomes accurate enough for general purposes in face-to-face interactions.
-In addition is able to handle different number and positions of participants in the scenario.
+Compared to intrusive hardware like head mounted displays, eye-trackers, wearables or RGB-D cameras for instance, this approach has a clear advantage of simplicity but also becomes accurate enough for general purposes in face-to-face interactions.
+In addition, it is able to handle different number and positions of participants in the scenario.
 
+\subsection{Estimating Head Pose}
 The first step is to estimate the head pose using the pin-hole model shown in equation \ref{eq:pinHole}.
@@
 
+\subsection{Setting the \textit{tf} Frame}
 The second step is to use the matrix [R|t] to set the correct translation and orientation of the \textit{tf} frame which represents the face in the 3D space.
@@
 
+\subsection{Defining the Field of View}
 The third step is to generate a representation of the field of view which origin and orientation are the ones defined by the face detection \textit{tf} frame.
@@
 
+\section{Methodology}
 \subsection{Experimental Set-up}
@@
 The experiments were developed individually and consisted on a writing activity by turns and one story telling performed by the robot at a given time of the interaction.
+All this performed in a over all 20 minutes interaction average time.
 
@@
 \begin{enumerate}
-\item Which percentage can we achieve during a real case scenario in terms of head pose estimation tracking?  
-\item How accurate can we predict where the subject is looking at based only on the head pose estimation?
+\item Which head pose estimation tracking rate can we achieve during a real case scenario?  
+\item How accurate can we predict where the subject is looking at based on the head pose estimation?
 \item How can we evaluate the performance of the children based on the focus of attention?
@@
 Our results for a peer-to-peer work scenarios using an approach based on RGB camera are summarized in table \ref{tab:results} where the \textit{expected} is the focus of attention that given the state of the system, it is likely to attract the subject's gaze.
-In the same manner, the \textit{captured} is the focus of attention our approach provides and the \textit{real}, the ground truth of the subject's gaze direction based on manual annotations of two evaluators.
+In the same manner, the \textit{captured} is the focus of attention our approach provides and the \textit{real}, the ground truth of the subject's gaze direction based on manual annotations from two evaluators with an average reliability of $ \alpha--------\% $.
+
+In order to calculate the inter-rater reliability between the two measurements (real-captured and real-expected), the overlap percentage between the two curves has been computed.
+A more reliable measurement between the real and captured focuses of attention is provided using the cosine similarity.
+These results are shown in table \ref{tab:results}.
 
@@
 	\small Track(\%) & \small78.7 & \small81.1 & \small73.7 & \small75.6 & \small77.1 & \small68.7 & \small75.8 \\ \hline
-	\begin{tabular}{@{}c@{}}\small Cosine \\ \small affinity \end{tabular} & \small0.58 & \small & \small & \small & \small & \small & \small \\ \hline
-	\begin{tabular}{@{}c@{}}\small Real vs \\ \small expected(\%)\end{tabular} & \small87.3 & \small  & \small & \small & \small & \small & \small \\ \hline
+	\begin{tabular}{@{}c@{}}\small Real vs \\ \small expected(\%)\end{tabular} & \small93.1 & \small  & \small & \small & \small & \small & \small \\ \hline
 	\begin{tabular}{@{}c@{}}\small Real vs \\ \small captured(\%)\end{tabular} & \small59.4 & \small & \small & \small & \small & \small & \small \\ \hline
+	\begin{tabular}{@{}c@{}}\small Cosine \\ \small affinity \end{tabular} & \small0.72 & \small & \small & \small & \small & \small & \small \\ \hline
 	\end{tabular}
@@
 
-% This two plots are provisional. We expect to improve them with images and boxes
 \begin{figure*}
@@
     \includegraphics[width=2\columnwidth]{realExpected}
-    \caption{\small In blue, real focus accoding to two observer assessment. In green, the expected focus based on the system state and in cyan, the overlapping between them.}
+    \caption{\small In blue, real focus accoding to two observer assessment.
+In cyan, the overlap between the real and the expected focus of attentions.
+In yellow, the repeated pattern in turn-taking and orange is the experimenter's suggestion to switch to a difficult word}
     \label{fig:realExpected}
@@
     \centering
-    \includegraphics[width=2\columnwidth]{realCaptured}
-    \caption{\small In blue, real focus accoding to two observer assessment. In green, the captured focus using our aproach and in cyan, the overlapping between them.}
+    \includegraphics[width=2\columnwidth]{bitmap_zoom}
+    \caption{\small At the top in blue, the real focus accoding to two observer assessment.
+In green, the captured focus using our aproach and in cyan, the overlap between them.
+At the bottom, the cosine similarity matrix between both curves with a window size of 10 seconds.}
     \label{fig:realCaptured}
 \end{figure*}
-% % % % % % % % % % % % % % % % % % % %
 
@@
 But not only that, the subjects also decrease the response time after each shift as well as the gaze frequency towards the model suggesting and improvement over time.
-Moreover, when a new word is selected for playing, the same behavior is replicated.
+Moreover, when a new word is selected for playing, the same behavior is replicated (see yellow boxes in figure \ref{fig:realExpected}).
 
-A third trend that can be observed is the continuous 'look for approval' situation where the subjects look at the experimenter in order to provide feedback about the robot's response before providing their own one.
-Such transition tablet-facilitator becomes more frequent in several cases, for instance: An unexpected situation is faced or the facilitator make a suggestion.
+A third trend that can be observed is the continuous 'look for approval' situation where the subjects look at the experimenter in order to provide feedback about the robot's response before providing their own one (see yellow boxes in figure \ref{fig:realExpected}).
+Such transition tablet-experimenter becomes more frequent in several cases, for instance: An unexpected situation is faced (before the story telling) or the experimenter makes a suggestion (orange box in figure \ref{fig:realExpected}).
 
@@
 
+%what do you think? -> tablet -> robot -> facilitator -> pattern
+
+%Too much gaze to selection: the challenge may be too big, let's adapt response from robot
+
+%Use time to account quality of interaction?
+
 \section{Conclusions}
+The rate of attention shift can predict retained learning.
+%Support this with a plot averaging all interactions of all subjects 
+The solution presented can be adapted to work in new locations
+Thanks to their easiness of use in real world scenario, due to employment of non-intrusive sensors
+
+%Future work
+Possible focuses of attention has to be known in advance.  
+Our system could be extended to allow both user and robot to move freely.
+develop the concept of gaze-aware robot Woud be nice to do it in an unconstrained environment.
+
+It is necessary to take into account that eye gaze information is neglected which induces a certain amount of error, as well as the noise induced by the 3D pose estimation computed from face landmarkers.
+
+From first person videos -> we put the camera there to maximize the detection during the writing time but it could installed in the head of the robot as tf even with movement constantly updating the tf frame matrix.
 

### 2015-08-09 sig-alternate.tex
Related Work and Research Question drafts

@@
 
-The focus of this research is to evaluate the quality of the interaction using the focus of attention as an engagement indicator in a complex scenario using an RGB camera.
+The focus of this research is to evaluate the quality of the interaction using the VFOA as an engagement indicator and the current performed task in a complex scenario using an RGB camera.
 The present studies has been developed in the context of the CoWriter project \footnote{The primary repository is \url{https://github.com/chili-epfl/cowriter_letter_learning}.}.
@@
 
-%Needs to be redifined
 \section{Related Work}
 
-It is a model based approach to head pose estimation that finds correspondences between facial landmarks points and their respective locations in a face model -> plug and play
-
-In this research, we address the problem of tracking the children's focus of attention during a long term child-robot interaction in a learning context, first.
-And second, we provide qualitative and quantitative evaluation of the outcome provided using a first person RGB camera approach.
-
-In neural networks requires a previous image acquisition and training as well as a fixed point of view correspondent to the data acquired.
--> The systems only works for a user standing within a certain region in front of an inmobile robot and probabilisty maps the users head orientation to some fixed focus target.
+The notion of engagement has actively been studied in a diverse set of domains, and several variables and social signals have been proposed in the literature to quantify and study it in social interactions.
 
 The evaluation of response times to provide an estimation of the user's engagement based on the correctness has been proposed in a learning context \cite{Beck} before.
-However, several aspects like the inter-subject variability or the environment complexity make it hardly applicable in a child-robot interaction
+However, several aspects like the inter-subject variability or the environment complexity make it hardly applicable in a child-robot interaction.
+Additionally, other techniques \cite{Chipman07postureas} like the use of the weight distribution using a pressure sensitive chair has reported results.
+But it is in the computer vision field where major contributions have been done.
 
-PAIVA curvature
+In \cite{peters2010investigating} approach, a model of user's interest and engagement with a virtual agent displaying shared attention behavior, by using eye gaze and head direction information is proposed.
+Similarly, in \cite{nakano2010estimating} an approach to estimate the user's conversational engagement with a conversational agent based on analysis of gaze patterns was implemented.
+Other significant work \cite{Castellano:2009} focus on feature evaluation related to the user's non-verbal behavior, the task and the companion's affective reactions are identified to predict the children's level of engagement using a Bayesian model.
 
-%Needs to be redifined
+Some few studies have proposed approaches for the detection of engagement with a robot. In \cite{Rich:2010} a computational model based on the recognition of connection events such
+as directed gaze, mutual facial gaze is proposed.
+In this context, some other \cite{Sanghvi:2011} focus only on the back or trunk posture as a determined factor for the assessment.
+A recent study \cite{anzalone} with social robots in face to face scenarios proposed a set of metrics based on non-verbal cues but manifesting some possible limitations in long-term scenarios.
+
+In this research, we address the problem of tracking the children's engagement based on their VFOA and current performed task during a long term child-robot interaction in a learning context, first.
+And second, we provide qualitative and quantitative evaluation of the outcome provided using a first person RGB camera approach.
+Moreover, this monitoring activity has been supporting our research in child-robot interaction for learning and instruction.
+\\\\
+
 \section{Research Questions}
-Measurement
-The participants' VFOA acquisition was the only dependent variable. This included their gaze times and patterns of looking or not looking to the targets. Data from a robot head-
-mounted Microsoft Kinect camera was coded post-hoc to quantify these measures.
 
-Hypothesis and Predictions
-It is hypothesized that there will be no generalizable pattern of participant VFOA behavior, necessitating online, in situ VFOA estimation capabilities to enable robots to communicate intentionally.
+The participants' VFOA acquisition is the only dependent variable.
+This includes their gaze times and patterns of looking or not looking to the targets and the synchrony on the movements.
+This data extracted from an RGB camera was also coded post-hoc to quantify some of these measurements.
 
-%The identification of the patterns can lead to a better understanding to the children progression. Less time means better understanding a move to next level.
+We predict that there will be no generalizable pattern of participant's VFOA behavior for different scenarios such as multiple subjects, necessitating in situ VFOA estimation capabilities to enable robots to adapt in the specific scenarios.
+However, we believe that off-line pattern identification can lead to a better understanding of the children progression as well as engagement level assessment in face-to-face long term interactions.
 
@@
 
-In our approach we use the open-source library dlib\footnote{http://dlib.net} \cite{dlib09}, to detect and extract the skeleton of the subject's face and then, use the obtained landmarkers to calculate the head pose orientation estimation.
-In this way we can estimate where the subject was looking at.
-Compared to intrusive hardware like head mounted displays, eye-trackers, wearables or RGB-D cameras for instance, this approach has a clear advantage of simplicity but also becomes accurate enough for general purposes in face-to-face interactions.
-In addition, it is able to handle different number and positions of participants in the scenario.
+We propose a adaptive model based solution to approach head pose estimation that finds correspondences between facial landmarks points and their respective locations in a face model.
+We use the open-source library dlib\footnote{http://dlib.net} \cite{dlib09}, to detect and extract the skeleton of the subject's face and then, use the obtained landmarkers to calculate the head pose orientation estimation.
+In this way we can estimate where the subject was looking at and obtain a 3D scene representation as it is executed as a ROS (Robot Operating System) node that can be seen in figure \ref{rviz}.
+Compared to intrusive hardware like head mounted displays, eye-trackers, wearables for instance, this approach has a clear advantage of simplicity and integration with robotic systems, but also becomes accurate enough for general purposes in face-to-face interactions.
+In addition, it is able to handle different number and positions of participants in the scene.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=1\columnwidth]{rviz}
+    \caption{\small Rviz screenshot with Nao, camera (Nao's feet), face marker with VFOA (gray cone) and several static marker representations in the scene.}
+    \label{rviz}
+\end{figure}
 
@@
 
+\newpage
 Likewise, the 3D face model is needed.
@@
 This is an estimation of the same anthropometric points \textit{M'} (for North American Caucasians) in the 3D space \cite{farkas1994anthropometry}.
-Moreover, the intrinsic camera parameters are necessary to be able to compose the camera matrix \textit{A}.
+Moreover, the intrinsic camera parameters are necessary to be able to compose its matrix \textit{A}.
 The components are expanded in equation \ref{eq:pinHole2}.
@@
 
-% Plot to clarify this??
+% Plot to clarify this?
 
@@
 
+%. Less time means better understanding a move to next level.
 %Future work

### 2015-08-10 sig-alternate.tex
Discussion draft added

@@
 We use the open-source library dlib\footnote{http://dlib.net} \cite{dlib09}, to detect and extract the skeleton of the subject's face and then, use the obtained landmarkers to calculate the head pose orientation estimation.
-In this way we can estimate where the subject was looking at and obtain a 3D scene representation as it is executed as a ROS (Robot Operating System) node that can be seen in figure \ref{rviz}.
+In this way, we can estimate what the subject was looking at and obtain a 3D scene representation as it is executed as a ROS (Robot Operating System) node that can be seen in figure \ref{rviz}.
 Compared to intrusive hardware like head mounted displays, eye-trackers, wearables for instance, this approach has a clear advantage of simplicity and integration with robotic systems, but also becomes accurate enough for general purposes in face-to-face interactions.
-In addition, it is able to handle different number and positions of participants in the scene.
+In addition, it allows as to model complex environments, handling different number and positions of participants in the scene.
 
@@
     \includegraphics[width=1\columnwidth]{rviz}
-    \caption{\small Rviz screenshot with Nao, camera (Nao's feet), face marker with VFOA (gray cone) and several static marker representations in the scene.}
+    \caption{\small Rviz screenshot with Nao, camera (Nao's feet), face marker with VFOA (gray cone) and blue marker intersecting the field of view.}
     \label{rviz}
@@
 
-% Plot to clarify this?
+% Schema to clarify this?
 
-\section{Methodology}
+\section{Case Study}
+The presented method have been successfully employed in a real child-robot interaction situation where the robot tries to engage the child in handwriting tasks using the paradigm learning by teaching.
 \subsection{Experimental Set-up}
@@
             depending on the studies), remains next to the child to guide the work. 
-            A second tablet allows to choose words and the camera captures the subject's face.}
+            A second tablet allows to choose demonstrations and the camera captures the subject's face.}
     \label{realSetup}
@@
 
-A tactile tablet (with a custom application) is used for both the robot and the
-child to write: during a typical round, the child requests the robot to write
-something (a single letter, a number or a full word), and push the tablet
-towards the robot, the robot writes on the tablet by gesturing the writing (but
-without actually physically touching the tablet), the child then pull back the
-tablet, corrects the robot's attempt by writing him/herself on top or next to
-the robot's writing (see Figure~\ref{fig:diego}), and ``send'' his/her
-demonstration to the robot by pressing a small button on the tablet. The robot
-``learns'' from this demonstration and tries again.
+A tactile tablet (with a custom application) is used for both the robot and the child to write: during a typical round, the child requests the robot to write something (a single letter, a number or a full word), and push the tablet towards the robot, the robot writes on the tablet by gesturing the writing (but without actually physically touching the tablet), the child then pull back the
+tablet, corrects the robot's attempt by writing him/herself on top or next to the robot's writing (see Figure~\ref{fig:diego}), and ``send'' his/her demonstration to the robot by pressing a small button on the tablet.
+The robot ``learns'' from this demonstration and tries again.
 
-Since the child is assumed to take on the role of the teacher, we had to ensure
-(s)he would be able to manage by him/herself the turn-taking and the overall
-progression of the activity (moving to the next letter or word). In our design,
-the turn-taking relies on the robot prompting for feedback once it is done with
-its writing (simple sentences like ``What do you think?''), and pressing on a
-small robot icon on the tablet once the child has finished correcting. In our
-experiments, both were easy to grasp for children.
+Since the child is assumed to take on the role of the teacher, we had to ensure (s)he would be able to manage by him/herself the turn-taking and the overall progression of the activity (moving to the next letter or word).
+In our design, the turn-taking relies on the robot prompting for feedback once it is done with its writing (simple sentences like ``What do you think?''), and pressing on a
+small robot icon on the tablet once the child has finished correcting. In our experiments, both were easy to grasp for children.
 
@@
 The camera was located at 5 cm from the center of Nao's feet and its holder was tied to a wood base for stability purposes.
-Thus, the camera's objective was 9 cm above and $ 40\degree $ inclination with respect to the surface of the table.
+Thus, the camera's objective was 9 cm above and $ 40\degree $ inclination with respect to the surface of the table to maximize the detection during the writing time.
+However, Nao's camera could have been used, always considering the \textit{tf} frame matrix update according to the head orientation.
+
 The subjects were typically located 50 cm away from the robot with the first tablet in front and the second one 30 cm to the left of the first one.
@@
 In order to calculate the inter-rater reliability between the two measurements (real-captured and real-expected), the overlap percentage between the two curves has been computed.
-A more reliable measurement between the real and captured focuses of attention is provided using the cosine similarity.
+A more reliable measurement in terms of time shifting between the real and captured focuses of attention is provided using the cosine similarity.
 These results are shown in table \ref{tab:results}.
@@
     \label{fig:realExpected}
+
+    % Maybe we should add the expected VFOA
 \end{figure*}
@@
 
-In several occasions during the interaction, none of the focuses of attention were within the field of view due to the time transition from one focus to another.
-In order to deal with these few cases we assume that the minimum distance between the height of the field of view (cone) and the targets corresponds to the most likely focus of attention.
+According to figure \ref{fig:realExpected} several findings can be analyzed in detail.
+The matching curve (cyan color) shows the overlapping between the expected and the real focuses of attention.
+As can be noticed there is a fluctuation at the beginning of the interaction as well as just before the activity switching.
+These chaotic gaze manifestation is what we call an automatic discovery of events, when the user experiences an uncertainty in the present situation.
 
+The first one of the previous two situations presented above is due to the novelty effect the subjects experience for the first time in front of the system and according to the rest of participants suggests that the adaptation time is around the first 2 minutes of interaction.
+In the second case, an unexpected robot behavior produces such reaction due to an activity switching, the story telling.
+
+It is important to consider that in some occasions during the interaction, none of the focuses of attention were within the field of view due to the time transition from one focus to another resulting in an ambiguity.
+In order to deal with these few cases it is assumed that the minimum distance between the height of the field of view (cone) and the targets corresponds to the most likely focus of attention.
 However, we also want to distinguish such cases when the subject is directly not looking where the defined targets are located, but towards a region far from the interaction field.
@@
 
-Several results can extracted from figure \ref{fig:realExpected}.
-First, the matching curve shows the overlapping between the expected and the real focuses of attention.
-As we can see there is a fluctuation at the beginning of the interaction suggesting a novelty effect.
-It suggests that the adaptation time is around the first 2 minutes in average.
+A relevant trend that can be observed is the continuous `look for approval' situation where the subjects look at the experimenter in order to provide feedback about the robot's response before providing their own one (see yellow boxes in figure \ref{fig:realExpected}).
+Such transition tablet-experimenter becomes more frequent in several cases, for instance when the experimenter makes a suggestion (orange box in figure \ref{fig:realExpected}), in this case, switch to a more complicated word.
 
-Second, once the turn-taking is established, a pattern becomes visible during the subject's feedback, when a word is being written by the children: Such pattern shows a high intermittent gaze frequency from the tablet to the selection.
-It shows that the subject is using the word model in the selection tablet to provide a better answer to the robot.
-But not only that, the subjects also decrease the response time after each shift as well as the gaze frequency towards the model suggesting and improvement over time.
-Moreover, when a new word is selected for playing, the same behavior is replicated (see yellow boxes in figure \ref{fig:realExpected}).
+Once the turn-taking is established, a pattern becomes visible during the subject's feedback, when an attempt is performed: Such pattern shows a high intermittent gaze frequency from the tablet to the selection.
+It also reveals that the subject is using the demonstration model in the selection tablet to provide a better answer to the robot.
+In fact, subjects decrease the response time after each shift as well as the gaze frequency towards the model suggesting an assimilation over time.
+Moreover, when a new demonstration is selected for playing, the same behavior is replicated (see yellow boxes in figure \ref{fig:realExpected}).
 
-A third trend that can be observed is the continuous 'look for approval' situation where the subjects look at the experimenter in order to provide feedback about the robot's response before providing their own one (see yellow boxes in figure \ref{fig:realExpected}).
-Such transition tablet-experimenter becomes more frequent in several cases, for instance: An unexpected situation is faced (before the story telling) or the experimenter makes a suggestion (orange box in figure \ref{fig:realExpected}).
+If we consider the previous case, evaluating the subject's progression together with the time on task may be an indicator of engagement: If the subject answers so quickly with respect to the previous attempts with a poorer handwriting may be the case she is not engaged anymore.
+In such hypothetical case an activity switch can be easily implemented in a robot architecture.
+It confirms the findings reported in \cite{anzalone} where time response to event is proposed to account quality of interaction.
+
+Therefore, a high amount of gaze to selection towards the demonstration provided can reveal that the challenge is too big or no gaze, the exactly the opposite.
+In this case, the robot response can be adapted to the situation.
+
+It is important to notice that the initiative of the robot into the learning process in a turn-taking activity produces an effect in the engagement of the human partner.
+However, experience suggests that a robot that learns evoke a positive action in the partner.
+But the opposite, a robot that is not able to progress over time, makes the opposite effect.
 
@@
 
-%what do you think? -> tablet -> robot -> facilitator -> pattern
-
-%Too much gaze to selection: the challenge may be too big, let's adapt response from robot
-
-%Use time to account quality of interaction?
-
+% Still under construction
 \section{Conclusions}
 The rate of attention shift can predict retained learning.
-%Support this with a plot averaging all interactions of all subjects 
-The solution presented can be adapted to work in new locations
-Thanks to their easiness of use in real world scenario, due to employment of non-intrusive sensors
+If we consider the gaze shift within the feedback episodes, it is possible to distinguish between long and short based on the number of previous repetitions playing with the same word.
+%Plot -> Support this with a plot averaging all interactions of all subjects
+%Plot -> Time gets reduced at the same time correctness improves
 
-%. Less time means better understanding a move to next level.
-%Future work
-Possible focuses of attention has to be known in advance.  
-Our system could be extended to allow both user and robot to move freely.
-develop the concept of gaze-aware robot Woud be nice to do it in an unconstrained environment.
+Furthermore, the solution presented can be extended to work in different platforms for simulating complex interaction scenarios monitoring several VFOA without employment of intrusive sensors.
+Although it has not been tested in a real scenario with several subjects at the same time, the approach allows to handle such situations too.
 
-It is necessary to take into account that eye gaze information is neglected which induces a certain amount of error, as well as the noise induced by the 3D pose estimation computed from face landmarkers.
+%Future work & Limitations
+On one hand, several future work need to be considered in the future.
+The most relevant one, the fact that the positions of the VFOA need to be known in advance and defined as static \textit{tf} frames.
+Moreover the gaze information acquired during the interactions can also be useful to develop actively the concept gaze-aware robot, where not only the robot knows where you look at, but shares the same subject's VFOA when the activity state allows it.
 
-From first person videos -> we put the camera there to maximize the detection during the writing time but it could installed in the head of the robot as tf even with movement constantly updating the tf frame matrix.
+On the other hand, the main limitations of this approach are based on the fact that the eye gaze information is neglected which induces a certain amount of error, as well as the noise induced by the 3D pose estimation computed from face landmarkers.
 

### 2015-08-15 sig-alternate.tex
Heat map added

@@
     \centering
-    \includegraphics[width=1\columnwidth]{rviz}
-    \caption{\small Rviz screenshot with Nao, camera (Nao's feet), face marker with VFOA (gray cone) and blue marker intersecting the field of view.}
+    \includegraphics[width=0.9\columnwidth]{rviz_camera}
+    \caption{\small Rviz screenshot showing Nao, camera (Nao's feet), face detected with its field of view (gray cone) and blue marker intersecting as current VFOA.}
     \label{rviz}
@@
 
-\newpage
 Likewise, the 3D face model is needed.
@@
 
-\begin{figure}
+\begin{figure}[h!]
     \centering
@@
 Our results for a peer-to-peer work scenarios using an approach based on RGB camera are summarized in table \ref{tab:results} where the \textit{expected} is the focus of attention that given the state of the system, it is likely to attract the subject's gaze.
-In the same manner, the \textit{captured} is the focus of attention our approach provides and the \textit{real}, the ground truth of the subject's gaze direction based on manual annotations from two evaluators with an average reliability of $ \alpha--------\% $.
+In the same manner, the \textit{captured} is the focus of attention our approach provides and the \textit{real}, the ground truth of the subject's gaze direction based on manual annotations from two evaluators with an average reliability of $ \alpha = .89 $.
 
@@
     \centering
-    \includegraphics[width=2\columnwidth]{realExpected}
-    \caption{\small In blue, real focus accoding to two observer assessment.
-In cyan, the overlap between the real and the expected focus of attentions.
-In yellow, the repeated pattern in turn-taking and orange is the experimenter's suggestion to switch to a difficult word}
+    \includegraphics[width=1.8\columnwidth]{realExpected}
+    \caption{\small In blue, real VFOA accoding to two observer's assessment.
+In cyan, the matching between the real and the obtained VFOA.
+In yellow, the repeated pattern in turn-taking and in orange, the experimenter's suggestion to choose a new word.}
     \label{fig:realExpected}
@@
     % Maybe we should add the expected VFOA
-\end{figure*}
-
-\begin{figure*}
-    \centering
-    \includegraphics[width=2\columnwidth]{bitmap_zoom}
-    \caption{\small At the top in blue, the real focus accoding to two observer assessment.
-In green, the captured focus using our aproach and in cyan, the overlap between them.
-At the bottom, the cosine similarity matrix between both curves with a window size of 10 seconds.}
-    \label{fig:realCaptured}
 \end{figure*}
@@
 These chaotic gaze manifestation is what we call an automatic discovery of events, when the user experiences an uncertainty in the present situation.
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.8\columnwidth]{bitmap}
+    \caption{\small Cosine similarity matrix between real and obtained VFOA using a 10 sec. window size computation. The correspondent numbers are shown in figure \ref{fig:realCaptured}.}
+    \label{fig:bitmap}
+\end{figure}
 
@@
 
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.9\columnwidth]{heatmap}
+    \caption{\small 2D histogram of a subject's head movements.}
+    \label{heatmap}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.9\columnwidth]{kmeans}
+    \caption{\small K-means clustering of the VFOA.}
+    \label{kmeans}
+\end{figure}
+
 Therefore, a high amount of gaze to selection towards the demonstration provided can reveal that the challenge is too big or no gaze, the exactly the opposite.
@@
 But the opposite, a robot that is not able to progress over time, makes the opposite effect.
+
+\begin{figure*}
+    \centering
+    \includegraphics[width=1.81\columnwidth]{realCaptured}
+    \caption{\small Real VFOA accoding to two observer's assessment.
+In green, the obtained VFOA using our aproach and in cyan, the mathing between them.
+Finally, the correspondent numbers with figure \ref{fig:bitmap}.}
+    \label{fig:realCaptured}
+\end{figure*}
 

### 2015-08-25 sig-alternate.tex
Gaze and PCA-error results & System diagram

@@
 We propose a adaptive model based solution to approach head pose estimation that finds correspondences between facial landmarks points and their respective locations in a face model.
-We use the open-source library dlib\footnote{http://dlib.net} \cite{dlib09}, to detect and extract the skeleton of the subject's face and then, use the obtained landmarkers to calculate the head pose orientation estimation.
+We use the open-source library dlib\footnote{http://dlib.net} \cite{dlib09}, to detect and extract the skeleton of the subject's face and then, use the obtained landmarkers to calculate the head pose orientation estimation (see figure \ref{system}).
 In this way, we can estimate what the subject was looking at and obtain a 3D scene representation as it is executed as a ROS (Robot Operating System) node that can be seen in figure \ref{rviz}.
@@
 In addition, it allows as to model complex environments, handling different number and positions of participants in the scene.
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.9\columnwidth]{system}
+    \caption{\small System overview showing the most relevant components involved in the estimation.}
+    \label{system}
+\end{figure}
 
@@
 
-Figure~\ref{realSetup} illustrates our general experimental setup: a
+Figure~\ref{fig:realSetup} illustrates our general experimental setup: a
 face-to-face child-robot interaction with an (autonomous) Aldebran's {\sc nao}
@@
             A second tablet allows to choose demonstrations and the camera captures the subject's face.}
-    \label{realSetup}
+    \label{fig:realSetup}
 \end{figure}
@@
 A tactile tablet (with a custom application) is used for both the robot and the child to write: during a typical round, the child requests the robot to write something (a single letter, a number or a full word), and push the tablet towards the robot, the robot writes on the tablet by gesturing the writing (but without actually physically touching the tablet), the child then pull back the
-tablet, corrects the robot's attempt by writing him/herself on top or next to the robot's writing (see Figure~\ref{fig:diego}), and ``send'' his/her demonstration to the robot by pressing a small button on the tablet.
+tablet, corrects the robot's attempt by writing him/herself on top or next to the robot's writing, and ``send'' his/her demonstration to the robot by pressing a small button on the tablet.
 The robot ``learns'' from this demonstration and tries again.
@@
 	\small Subject & 1 & 2 & 3 & 4 & 5 & 6 & Avg\\ \hline
-	\small Track(\%) & \small78.7 & \small81.1 & \small73.7 & \small75.6 & \small77.1 & \small68.7 & \small75.8 \\ \hline
-	\begin{tabular}{@{}c@{}}\small Real vs \\ \small expected(\%)\end{tabular} & \small93.1 & \small  & \small & \small & \small & \small & \small \\ \hline
-	\begin{tabular}{@{}c@{}}\small Real vs \\ \small captured(\%)\end{tabular} & \small59.4 & \small & \small & \small & \small & \small & \small \\ \hline
-	\begin{tabular}{@{}c@{}}\small Cosine \\ \small affinity \end{tabular} & \small0.72 & \small & \small & \small & \small & \small & \small \\ \hline
+	\small Track(\%) & \small88.2 & \small83.5 & \small90.5 & \small83.1 & \small87.9 & \small85.0 & \small \\ \hline
+	\begin{tabular}{@{}c@{}}\small Real vs \\ \small expected(\%)\end{tabular} & \small92.6 & \small88.8  & \small95.8 & \small93.1 & \small96.5 & \small89.6 & \small \\ \hline
+	\begin{tabular}{@{}c@{}}\small Real vs \\ \small captured(\%)\end{tabular} & \small61.1 & \small85.5 & \small80.8 & \small62.2 & \small63.4 & \small83.9 & \small \\ \hline
+	\begin{tabular}{@{}c@{}}\small Cosine \\ \small affinity \end{tabular} & \small0.73 & \small0.82 & \small0.87 & \small0.73 & \small0.77 & \small0.92 & \small \\ \hline
 	\end{tabular}
@@
     \includegraphics[width=1.8\columnwidth]{realExpected}
-    \caption{\small In blue, real VFOA accoding to two observer's assessment.
+    \caption{\small In blue, real VFOA of subject 4 accoding to two observer's assessment.
 In cyan, the matching between the real and the obtained VFOA.
@@
     \includegraphics[width=0.9\columnwidth]{kmeans}
-    \caption{\small K-means clustering of the VFOA.}
+    \caption{\small K-means clustering of the head movements.}
     \label{kmeans}
@@
     \includegraphics[width=1.81\columnwidth]{realCaptured}
-    \caption{\small Real VFOA accoding to two observer's assessment.
+    \caption{\small Real VFOA of subject 4 accoding to two observer's assessment.
 In green, the obtained VFOA using our aproach and in cyan, the mathing between them.
@@
 
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.9\columnwidth]{errorR}
+    \caption{\small .}
+    \label{errorR}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.9\columnwidth]{gazeR}
+    \caption{\small .}
+    \label{gazeR}
+\end{figure}
+
 % Still under construction
@@
 %Plot -> Support this with a plot averaging all interactions of all subjects
-%Plot -> Time gets reduced at the same time correctness improves
+%Plot -> Time gets reduced at the same time correctness improves -> Did not show any good results
+%Plot -> Difference in gaze during feedback from a subject's whose learning is assimilated vs a one is not: The second shows more intermitent tablet-selection gaze directions.
+%Diagram dlib/ROS/launch/nodes/params_ini/txt_model_face
 

### 2015-08-26 sig-alternate.tex
Updated results values in table

@@
 	\small Subject & 1 & 2 & 3 & 4 & 5 & 6 & Avg\\ \hline
-	\small Track(\%) & \small88.2 & \small83.5 & \small90.5 & \small83.1 & \small87.9 & \small85.0 & \small \\ \hline
-	\begin{tabular}{@{}c@{}}\small Real vs \\ \small expected(\%)\end{tabular} & \small92.6 & \small88.8  & \small95.8 & \small93.1 & \small96.5 & \small89.6 & \small \\ \hline
-	\begin{tabular}{@{}c@{}}\small Real vs \\ \small captured(\%)\end{tabular} & \small61.1 & \small85.5 & \small80.8 & \small62.2 & \small63.4 & \small83.9 & \small \\ \hline
-	\begin{tabular}{@{}c@{}}\small Cosine \\ \small affinity \end{tabular} & \small0.73 & \small0.82 & \small0.87 & \small0.73 & \small0.77 & \small0.92 & \small \\ \hline
+	\small Track(\%) & \small88.2 & \small83.5 & \small90.5 & \small83.1 & \small87.9 & \small85.0 & \small86.4 \\ \hline
+	\begin{tabular}{@{}c@{}}\small Real vs \\ \small expected(\%)\end{tabular} & \small92.6 & \small88.8  & \small95.8 & \small93.1 & \small96.5 & \small89.6 & \small92.7 \\ \hline
+	\begin{tabular}{@{}c@{}}\small Real vs \\ \small captured(\%)\end{tabular} & \small61.1 & \small85.5 & \small80.8 & \small62.2 & \small63.4 & \small83.9 & \small72.8 \\ \hline
+	\begin{tabular}{@{}c@{}}\small Cosine \\ \small affinity \end{tabular} & \small0.73 & \small0.82 & \small0.87 & \small0.73 & \small0.77 & \small0.89 & \small0.8 \\ \hline
 	\end{tabular}
@@
     \includegraphics[width=0.9\columnwidth]{errorR}
-    \caption{\small .}
+    \caption{\small Linear regression of all subject attempts against the word correctness, measured as the average euclidean distance of each leter in its corresponding eigenspace with respect to a reference leter.
+The gray region defines the confidence interval.}
     \label{errorR}
@@
     \includegraphics[width=0.9\columnwidth]{gazeR}
-    \caption{\small .}
+    \caption{\small Linear regression of all subject attempts against the occurrences of a gaze shift from the writing tablet towards the model provided.
+The gray region defines the confidence interval.}
     \label{gazeR}
@@
 
+%If less than a 0.5 seconds without tracking -> Precission  
+% Maker-less technology decrease the complexity of the system and increase the naturally of the interaction
 % Still under construction
@@
 If we consider the gaze shift within the feedback episodes, it is possible to distinguish between long and short based on the number of previous repetitions playing with the same word.
-%Plot -> Support this with a plot averaging all interactions of all subjects
+%Plot -> Support this with a plot averaging all interactions of all subjects -> Done
 %Plot -> Time gets reduced at the same time correctness improves -> Did not show any good results
 %Plot -> Difference in gaze during feedback from a subject's whose learning is assimilated vs a one is not: The second shows more intermitent tablet-selection gaze directions.
-%Diagram dlib/ROS/launch/nodes/params_ini/txt_model_face
+%Diagram dlib/ROS/launch/nodes/params_ini/txt_model_face -
+%Need of more subjects to test the hypothesis
 

### 2015-08-26 paper.tex
{sig-alternate -> paper}.tex & added a Makefile

--- 2015-08-10 sig-alternate.tex
@@
-% This is "sig-alternate.tex" V2.0 May 2012
-% This file should be compiled with V2.5 of "sig-alternate.cls" May 2012
-%
-% This example file demonstrates the use of the 'sig-alternate.cls'
-% V2.5 LaTeX2e document class file. It is for those submitting
-% articles to ACM Conference Proceedings WHO DO NOT WISH TO
-% STRICTLY ADHERE TO THE SIGS (PUBS-BOARD-ENDORSED) STYLE.
-% The 'sig-alternate.cls' file will produce a similar-looking,
-% albeit, 'tighter' paper resulting in, invariably, fewer pages.
-%
-% ----------------------------------------------------------------------------------------------------------------
-% This .tex file (and associated .cls V2.5) produces:
-%       1) The Permission Statement
-%       2) The Conference (location) Info information
-%       3) The Copyright Line with ACM data
-%       4) NO page numbers
-%
-% as against the acm_proc_article-sp.cls file which
-% DOES NOT produce 1) thru' 3) above.
-%
-% Using 'sig-alternate.cls' you have control, however, from within
-% the source .tex file, over both the CopyrightYear
-% (defaulted to 200X) and the ACM Copyright Data
-% (defaulted to X-XXXXX-XX-X/XX/XX).
-% e.g.
-% \CopyrightYear{2007} will cause 2007 to appear in the copyright line.
-% \crdata{0-12345-67-8/90/12} will cause 0-12345-67-8/90/12 to appear in the copyright line.
-%
-% ---------------------------------------------------------------------------------------------------------------
-% This .tex source is an example which *does* use
-% the .bib file (from which the .bbl file % is produced).
-% REMEMBER HOWEVER: After having produced the .bbl file,
-% and prior to final submission, you *NEED* to 'insert'
-% your .bbl file into your source .tex file so as to provide
-% ONE 'self-contained' source file.
-%
-% ================= IF YOU HAVE QUESTIONS =======================
-% Questions regarding the SIGS styles, SIGS policies and
-% procedures, Conferences etc. should be sent to
-% Adrienne Griscti (griscti@acm.org)
-%
-% Technical questions _only_ to
-% Gerald Murray (murray@hq.acm.org)
-% ===============================================================
-%
-% For tracking purposes - this is V2.0 - May 2012
+\documentclass{sig-alternate}
 
-\documentclass{sig-alternate}
 \graphicspath{{figs/}}
@@
 % --- Author Metadata here ---
-\conferenceinfo{HRI}{'15 Chrischurch, New Zealand.}
+\conferenceinfo{HRI}{'16 Chrischurch, New Zealand.}
 %\CopyrightYear{2007} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
@@
 
-\title{Alternate {\ttlit ACM} SIG Proceedings Paper in LaTeX
-Format}
+\title{On-line Vision-based Engagement Assessment for Interactive Robots}
 \author{Fernando Garcia \qquad Séverin Lemaignan \qquad Alexis Jacq \qquad Pierre Dillenbourg\\Computer-Human Interaction in Learning and Instruction Laboratory (CHILI)\\École Polytechnique Fédérale de Lausanne (EPFL)}
-
-%
-% You need the command \numberofauthors to handle the 'placement
-% and alignment' of the authors beneath the title.
-%
-% For aesthetic reasons, we recommend 'three authors at a time'
-% i.e. three 'name/affiliation blocks' be placed beneath the title.
-%
-% NOTE: You are NOT restricted in how many 'rows' of
-% "name/affiliations" may appear. We just ask that you restrict
-% the number of 'columns' to three.
-%
-% Because of the available 'opening page real-estate'
-% we ask you to refrain from putting more than six authors
-% (two rows with three columns) beneath the article title.
-% More than six makes the first-page appear very cluttered indeed.
-%
-% Use the \alignauthor commands to handle the names
-% and affiliations for an 'aesthetic maximum' of six authors.
-% Add names, affiliations, addresses for
-% the seventh etc. author(s) as the argument for the
-% \additionalauthors command.
-% These 'additional authors' will be output/set for you
-% without further effort on your part as the last section in
-% the body of your article BEFORE References or any Appendices.
-
-%\numberofauthors{4} %  in this sample file, there are a *total*
-% of EIGHT authors. SIX appear on the 'first-page' (for formatting
-% reasons) and the remaining two appear in the \additionalauthors section.
-%
-%\author{
-% You can go ahead and credit any number of authors here,
-% e.g. one 'row of three' or two rows (consisting of one row of three
-% and a second row of one, two or three).
-%
-% The command \alignauthor (no curly braces needed) should
-% precede each author name, affiliation/snail-mail address and
-% e-mail address. Additionally, tag each line of
-% affiliation/address with \affaddr, and tag the
-% e-mail address with \email.
-%
-% 1st. author
-%\alignauthor
-%Ben Trovato\titlenote{Dr.~Trovato insisted his name be first.}\\
-%       \affaddr{Institute for Clarity in Documentation}\\
-%       \affaddr{1932 Wallamaloo Lane}\\
-%       \affaddr{Wallamaloo, New Zealand}\\
-%       \email{trovato@corporation.com}
-%% 2nd. author
-%\alignauthor
-%G.K.M. Tobin\titlenote{The secretary disavows
-%any knowledge of this author's actions.}\\
-%       \affaddr{Institute for Clarity in Documentation}\\
-%       \affaddr{P.O. Box 1212}\\
-%       \affaddr{Dublin, Ohio 43017-6221}\\
-%       \email{webmaster@marysville-ohio.com}
-%% 3rd. author
-%\alignauthor Lars Th{\o}rv{\"a}ld\titlenote{This author is the
-%one who did all the really hard work.}\\
-%       \affaddr{The Th{\o}rv{\"a}ld Group}\\
-%       \affaddr{1 Th{\o}rv{\"a}ld Circle}\\
-%       \affaddr{Hekla, Iceland}\\
-%       \email{larst@affiliation.org}
-%\and  % use '\and' if you need 'another row' of author names
-%% 4th. author
-%\alignauthor Lawrence P. Leipuner\\
-%       \affaddr{Brookhaven Laboratories}\\
-%       \affaddr{Brookhaven National Lab}\\
-%       \affaddr{P.O. Box 5000}\\
-%       \email{lleipuner@researchlabs.org}
-%}
 
@@
 
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.9\columnwidth]{system}
+    \caption{\small System overview showing the most relevant components involved in the estimation.}
+    \label{system}
+\end{figure}
+
 \begin{figure}
     \centering
-    \includegraphics[width=1\columnwidth]{rviz}
-    \caption{\small Rviz screenshot with Nao, camera (Nao's feet), face marker with VFOA (gray cone) and blue marker intersecting the field of view.}
+    \includegraphics[width=0.9\columnwidth]{rviz_camera}
+    \caption{\small Rviz screenshot showing Nao, camera (Nao's feet), face detected with its field of view (gray cone) and blue marker intersecting as current VFOA.}
     \label{rviz}
@@
 	\small Subject & 1 & 2 & 3 & 4 & 5 & 6 & Avg\\ \hline
-	\small Track(\%) & \small78.7 & \small81.1 & \small73.7 & \small75.6 & \small77.1 & \small68.7 & \small75.8 \\ \hline
-	\begin{tabular}{@{}c@{}}\small Real vs \\ \small expected(\%)\end{tabular} & \small93.1 & \small  & \small & \small & \small & \small & \small \\ \hline
-	\begin{tabular}{@{}c@{}}\small Real vs \\ \small captured(\%)\end{tabular} & \small59.4 & \small & \small & \small & \small & \small & \small \\ \hline
-	\begin{tabular}{@{}c@{}}\small Cosine \\ \small affinity \end{tabular} & \small0.72 & \small & \small & \small & \small & \small & \small \\ \hline
+	\small Track(\%) & \small88.2 & \small83.5 & \small90.5 & \small83.1 & \small87.9 & \small85.0 & \small86.4 \\ \hline
+	\begin{tabular}{@{}c@{}}\small Real vs \\ \small expected(\%)\end{tabular} & \small92.6 & \small88.8  & \small95.8 & \small93.1 & \small96.5 & \small89.6 & \small92.7 \\ \hline
+	\begin{tabular}{@{}c@{}}\small Real vs \\ \small captured(\%)\end{tabular} & \small61.1 & \small85.5 & \small80.8 & \small62.2 & \small63.4 & \small83.9 & \small72.8 \\ \hline
+	\begin{tabular}{@{}c@{}}\small Cosine \\ \small affinity \end{tabular} & \small0.73 & \small0.82 & \small0.87 & \small0.73 & \small0.77 & \small0.89 & \small0.8 \\ \hline
 	\end{tabular}
@@
     \centering
-    \includegraphics[width=2\columnwidth]{realExpected}
-    \caption{\small In blue, real focus accoding to two observer assessment.
-In cyan, the overlap between the real and the expected focus of attentions.
-In yellow, the repeated pattern in turn-taking and orange is the experimenter's suggestion to switch to a difficult word}
+    \includegraphics[width=1.8\columnwidth]{realExpected}
+    \caption{\small In blue, real VFOA of subject 4 accoding to two observer's assessment.
+In cyan, the matching between the real and the obtained VFOA.
+In yellow, the repeated pattern in turn-taking and in orange, the experimenter's suggestion to choose a new word.}
     \label{fig:realExpected}
@@
 
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.9\columnwidth]{heatmap}
+    \caption{\small 2D histogram of a subject's head movements.}
+    \label{heatmap}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.9\columnwidth]{kmeans}
+    \caption{\small K-means clustering of the head movements.}
+    \label{kmeans}
+\end{figure}
+
 Therefore, a high amount of gaze to selection towards the demonstration provided can reveal that the challenge is too big or no gaze, the exactly the opposite.
@@
 
+\begin{figure*}
+    \centering
+    \includegraphics[width=1.81\columnwidth]{realCaptured}
+    \caption{\small Real VFOA of subject 4 accoding to two observer's assessment.
+In green, the obtained VFOA using our aproach and in cyan, the mathing between them.
+Finally, the correspondent numbers with figure \ref{fig:bitmap}.}
+    \label{fig:realCaptured}
+\end{figure*}
+
 Finally, the subjects set their attention more often to the tablet displaying the generated hand-writing during the robot feedback rather than to the robot itself.
 
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.9\columnwidth]{errorR}
+    \caption{\small Linear regression of all subject attempts against the word correctness, measured as the average euclidean distance of each leter in its corresponding eigenspace with respect to a reference leter.
+The gray region defines the confidence interval.}
+    \label{errorR}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.9\columnwidth]{gazeR}
+    \caption{\small Linear regression of all subject attempts against the occurrences of a gaze shift from the writing tablet towards the model provided.
+The gray region defines the confidence interval.}
+    \label{gazeR}
+\end{figure}
+
+%If less than a 0.5 seconds without tracking -> Precission  
+% Maker-less technology decrease the complexity of the system and increase the naturally of the interaction
 % Still under construction
@@
 If we consider the gaze shift within the feedback episodes, it is possible to distinguish between long and short based on the number of previous repetitions playing with the same word.
-%Plot -> Support this with a plot averaging all interactions of all subjects
-%Plot -> Time gets reduced at the same time correctness improves
+%Plot -> Support this with a plot averaging all interactions of all subjects -> Done
+%Plot -> Time gets reduced at the same time correctness improves -> Did not show any good results
+%Plot -> Difference in gaze during feedback from a subject's whose learning is assimilated vs a one is not: The second shows more intermitent tablet-selection gaze directions.
+%Diagram dlib/ROS/launch/nodes/params_ini/txt_model_face -
+%Need of more subjects to test the hypothesis
 

### 2015-08-26 paper.tex
Conclusions draft added

@@
 
-% Stats from the paper
-% Stats from me
-% Real vs captured plot
-% Real vs expected plot
-
 \subsection{From Face Detection to Focus of Attention}
@@
 
-Therefore, a high amount of gaze to selection towards the demonstration provided can reveal that the challenge is too big or no gaze, the exactly the opposite.
+However, a high amount of gaze shift \ref{fig:gazeR} towards the demonstration provided can reveal that the challenge is too big or no gaze, the exactly the opposite.
 In this case, the robot response can be adapted to the situation.
+Similar, but least relevant results are obtained using the correctness \ref{fig:errorR} of the child's feedback after each attempt.
 
-It is important to notice that the initiative of the robot into the learning process in a turn-taking activity produces an effect in the engagement of the human partner.
+This information can be used to estimate the child's state during the interaction.
+For instance, if we consider that a low level of engagement is produced by extremely high or low demanding situations, it is possible to assess the engagement level during the interaction based on gaze shift and word correctness.
+
+It is also important to notice that the initiative of the robot into the learning process in a turn-taking activity produces an effect in the engagement of the human partner.
 However, experience suggests that a robot that learns evoke a positive action in the partner.
-But the opposite, a robot that is not able to progress over time, makes the opposite effect.
+But the opposite, a robot that is not able to progress over time, produces the opposite effect.
 
@@
 
-Finally, the subjects set their attention more often to the tablet displaying the generated hand-writing during the robot feedback rather than to the robot itself.
+%Finally, the subjects set their attention more often to the tablet displaying the generated hand-writing during the robot feedback rather than to the robot itself.
 
@@
 The gray region defines the confidence interval.}
-    \label{errorR}
+    \label{fig:errorR}
 \end{figure}
@@
 The gray region defines the confidence interval.}
-    \label{gazeR}
+    \label{fig:gazeR}
 \end{figure}
 
-%If less than a 0.5 seconds without tracking -> Precission  
-% Maker-less technology decrease the complexity of the system and increase the naturally of the interaction
-% Still under construction
 \section{Conclusions}
-The rate of attention shift can predict retained learning.
-If we consider the gaze shift within the feedback episodes, it is possible to distinguish between long and short based on the number of previous repetitions playing with the same word.
-%Plot -> Support this with a plot averaging all interactions of all subjects -> Done
-%Plot -> Time gets reduced at the same time correctness improves -> Did not show any good results
-%Plot -> Difference in gaze during feedback from a subject's whose learning is assimilated vs a one is not: The second shows more intermitent tablet-selection gaze directions.
-%Diagram dlib/ROS/launch/nodes/params_ini/txt_model_face -
-%Need of more subjects to test the hypothesis
+In this work we have presented an on-line vision-based engagement assessment solution that can be extended to work in different platforms for simulating complex interaction scenarios monitoring several VFOA with a maker-less technology.
+It provides some benefits like the decrement of the system complexity as well as the increment of the naturalness of the interaction.
 
-Furthermore, the solution presented can be extended to work in different platforms for simulating complex interaction scenarios monitoring several VFOA without employment of intrusive sensors.
-Although it has not been tested in a real scenario with several subjects at the same time, the approach allows to handle such situations too.
+Furthermore, although it has not been tested in a real scenario with several subjects at the same time, the approach also allows to handle such situations too.
 
-%Future work & Limitations
-On one hand, several future work need to be considered in the future.
+During the case study, results show that it is possible to distinguish between long and short feedback episodes based on the gaze shift and depending on the previous attempts performed by the subject.
+As a result, gaze-shift together with the number of attempts becomes a reliable indicator candidate in the assessment of the engagement during the activity.
+
+Additionally, both the gaze shift and the correctness of the feedback seem to be strictly correlated with the learning gain during the activity as they go in decrement after each attempt provided by the subjects.
+Despite the fact that the results are encouraging, there is a need of validation with a greater amount of subject's.
+
+Regarding to the approach used, some limitations need to be considered.
 The most relevant one, the fact that the positions of the VFOA need to be known in advance and defined as static \textit{tf} frames.
-Moreover the gaze information acquired during the interactions can also be useful to develop actively the concept gaze-aware robot, where not only the robot knows where you look at, but shares the same subject's VFOA when the activity state allows it.
+Moreover the gaze information acquired during the interactions can also be useful to develop actively the concept gaze-aware robot, where not only the robot knows where the subject looks at, but shares the same subject's VFOA when the activity state allows it.
 
-On the other hand, the main limitations of this approach are based on the fact that the eye gaze information is neglected which induces a certain amount of error, as well as the noise induced by the 3D pose estimation computed from face landmarkers.
+Moreover, other limitations are based on the fact that the eye gaze information is neglected which induces a certain amount of error, as well as the noise induced by the 3D pose estimation computed from face landmarkers.
 

### 2015-08-28 paper.tex
First paper draft

@@
 While new algorithms and robotic technologies ensure better cognitive and interactive abilities, a deeper study about the benefits in the social interaction needs to be done: Is not enough to exhibit human-like capabilities in social tasks, but to do it in a human-like manner.
-For instance, several studies showed that humans prefer to interact with a robots whose task outcome is delayed, shows uncertainty or an undesired output \cite{Admoni,Short}.
+For instance, several studies showed that humans prefer to interact with robots whose task outcome is delayed, shows uncertainty or an undesired output \cite{Admoni,Short}.
 In fact, the ability to adapt has been proved to be beneficial in long-term interactions \cite{Tielman:2014, Lim:2014}, specifically with children, manifesting a more positive attitude over time.
@@
 Engagement describes a cognitive, affective (specifically intrinsic motivation), and behavioral state of interaction with a computer application that ``makes the user want to be there" \cite{OBrien:2010}.
-Engaging inter-actions were thought to involve attention, intrinsic interest, interactivity, perceived control and choice, functionality and motivation.
+Engaging interactions were thought to involve attention, intrinsic interest, interactivity, perceived control and choice, functionality and motivation.
 
@@
 
-In fact, the use of robots in education comes with the potential to engage the child in meta-cognition through the learning by teaching paradigm, wherein a student takes the role of
-a teacher and experiences stronger educational benefits as a result (such as in \cite{Palinscar1984}).
+In fact, the use of robots in education comes with the potential to engage the child in meta-cognition through the learning by teaching paradigm, wherein a student takes the role of a teacher and experiences stronger educational benefits as a result (such as in \cite{Palinscar1984}).
 
-The focus of this research is to evaluate the quality of the interaction using the VFOA as an engagement indicator and the current performed task in a complex scenario using an RGB camera.
-The present studies has been developed in the context of the CoWriter project \footnote{The primary repository is \url{https://github.com/chili-epfl/cowriter_letter_learning}.}.
-However, in a child-robot interaction context, avoiding 'invasive' technologies such as wearable devices is a must to keep the experience as natural as possible.
+The focus of this research is to evaluate the quality of the interaction using the VFOA as well as the correctness in the performed task as engagement indicators in the context of the CoWriter project \footnote{The primary repository is \url{https://github.com/chili-epfl/cowriter_letter_learning}.}.
+In order to do that avoiding the use of 'invasive' technologies such as wearable devices becomes a must to keep the experience as natural as possible.
 
@@
 In this research, we address the problem of tracking the children's engagement based on their VFOA and current performed task during a long term child-robot interaction in a learning context, first.
-And second, we provide qualitative and quantitative evaluation of the outcome provided using a first person RGB camera approach.
-Moreover, this monitoring activity has been supporting our research in child-robot interaction for learning and instruction.
-\\\\
+And second, we provide qualitative and quantitative evaluation of the outcome provided using a first person RGB sensor approach.
+Moreover, this monitoring activity has been supporting our research in child-robot interaction for learning and instruction purposes.
+\vfill
 
@@
 We use the open-source library dlib\footnote{http://dlib.net} \cite{dlib09}, to detect and extract the skeleton of the subject's face and then, use the obtained landmarkers to calculate the head pose orientation estimation (see figure \ref{system}).
-In this way, we can estimate what the subject was looking at and obtain a 3D scene representation as it is executed as a ROS (Robot Operating System) node that can be seen in figure \ref{rviz}.
-Compared to intrusive hardware like head mounted displays, eye-trackers, wearables for instance, this approach has a clear advantage of simplicity and integration with robotic systems, but also becomes accurate enough for general purposes in face-to-face interactions.
-In addition, it allows as to model complex environments, handling different number and positions of participants in the scene.
+In this way, we can estimate what the subject was looking at and obtain an \textit{rviz} 3D scene representation as it is executed as a ROS (Robot Operating System) node that can be seen in figure \ref{rviz}.
+Compared to intrusive hardware like head mounted displays, eye-trackers or wearables for instance, this approach has a clear advantage of simplicity and integration with robotic systems, but also becomes accurate enough for general purposes in face-to-face interactions.
+In addition, it allows us to model complex environments, handling different number and positions of participants in the scene.
 
@@
 The first step is to estimate the head pose using the pin-hole model shown in equation \ref{eq:pinHole}.
-As we mention before dlib library facilitates the task by providing the 2D points \textit{m'} that composes the face silhouette.
+As mentioned above, dlib library facilitates the task by providing the 2D points \textit{m'} that composes the face silhouette.
 However, it is enough with obtaining eight spread points, such as right and left eyes, right and left tragions, sellion or nasal depression zone, pronasale or center of the nose, stomion or center of the mouth and menton.
@@
 The proposed head pose estimation method would work best with a 3D face model of the query subject itself.
-When a 3D face model of the subject is not available, as in many practical situations, a 3D face model of a subject can be used.
+When a 3D face model of the subject is not available, as in many practical situations, a generic 3D face model can be used.
 This is an estimation of the same anthropometric points \textit{M'} (for North American Caucasians) in the 3D space \cite{farkas1994anthropometry}.
@@
 Finally, the matrix of extrinsic parameters [R|t] is obtained by joining the rotation matrix R (calculated from the rotation vector) and translation vector \textit{t}.
-The result is a matrix projection \textbf{P} which maps a point in the 3D space onto a point in the 2D image place solving the equation \ref{eq:p}.
+The result is a matrix projection \textbf{P} which maps a point in the 3D space onto a point in the 2D image plane solving equation \ref{eq:p}.
 
@@
 \subsection{Defining the Field of View}
-The third step is to generate a representation of the field of view which origin and orientation are the ones defined by the face detection \textit{tf} frame.
+The third step is to generate a representation of the field of view whose origin and orientation are the ones defined by the face detection \textit{tf} frame.
 
@@
 
-Assuming that the \textit{tf} frames representing the robotic agent, the two tablets, the experimenter and the observers are static, need to be monitored.
+Assuming that the \textit{tf} frames representing the robotic agent and the VFOA candidates are static, need to be monitored.
 This task consists on evaluating the intersection between the \textit{tf} frames and the field of view: If one focus of attention is within the region defined by the field of view, we assume the subject is looking at it.
@@
 \section{Case Study}
-The presented method have been successfully employed in a real child-robot interaction situation where the robot tries to engage the child in handwriting tasks using the paradigm learning by teaching.
+The presented method have been successfully employed in a real child-robot interaction situation where the robot tries to engage the child in handwriting tasks using learning by teaching paradigm.
 \subsection{Experimental Set-up}
@@
 The experiments were developed individually and consisted on a writing activity by turns and one story telling performed by the robot at a given time of the interaction.
-All this performed in a over all 20 minutes interaction average time.
+All this performed in a over all 18 minutes interaction average time.
 
@@
     \includegraphics[width=0.7\columnwidth]{drawSetup}
-    \caption{\small Graphical representation of he set-up and the possible children focuses of attention in the scene, in red.}
+    \caption{\small Graphical representation of the set-up and the possible children focuses of attention in the scene, in red.}
     \label{drawSetup}
@@
 \item How accurate can we predict where the subject is looking at based on the head pose estimation?
-\item How can we evaluate the performance of the children based on the focus of attention?
+\item How can we evaluate the engagement of the children based on the focus of attention?\newline
 \end{enumerate}
 
-\subsection{From Face Detection to Focus of Attention}
+\subsection{From Face Detection to Focus of Attention} \label{fromFaceTo}
 
@@
 In order to calculate the inter-rater reliability between the two measurements (real-captured and real-expected), the overlap percentage between the two curves has been computed.
-A more reliable measurement in terms of time shifting between the real and captured focuses of attention is provided using the cosine similarity.
-These results are shown in table \ref{tab:results}.
+A more reliable measurement in terms of time shifting between the real and captured focuses of attention is provided using the cosine similarity also shown in table \ref{tab:results}.
 
@@
 
-The results has shown that the percentage of accuracy in the tracking is not only conditioned by the subject's amount of movement, but also by the speed of the same movements: Restless children externalize greater levels of movement decreasing the tracker accuracy.
-Additionally, several recurrent computer vision challenges such as the lighting conditions, occlusions and the face angle with respect to the camera are factors that contribute to reduce the accuracy.
+The results has proved that the VFOA prediction accuracy in the tracking is not only conditioned by the subject's amount of movement, but also by the speed of the same movements: Restless children externalize greater levels of movement decreasing the VFOA prediction accuracy.
+
+For instance, subjects 1,4 and 5 despite the fact that the tracking accuracy is in the average, the bad precision in the captured VFOA suggests that such children are highly likely to modify the gaze direction without head move intervention reducing dramatically the overlap between captured and real VFOA.
+
+Additionally, several recurrent computer vision challenges such as the lighting conditions, occlusions with the hands and the face angle with respect to the camera, looking to the experimenter for instance, are factors that contribute to reduce the tracking accuracy.
 
@@
 
+\begin{figure*}
+    \centering
+    \includegraphics[width=1.81\columnwidth]{realCaptured}
+    \caption{\small Real VFOA of subject 4 accoding to two observer's assessment.
+In green, the obtained VFOA using our aproach and in cyan, the mathing between them.
+Finally, the correspondent numbers with figure \ref{fig:bitmap}.}
+    \label{fig:realCaptured}
+\end{figure*}
+
 According to figure \ref{fig:realExpected} several findings can be analyzed in detail.
@@
 These chaotic gaze manifestation is what we call an automatic discovery of events, when the user experiences an uncertainty in the present situation.
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.8\columnwidth]{bitmap}
-    \caption{\small Cosine similarity matrix between real and obtained VFOA using a 10 sec. window size computation. The correspondent numbers are shown in figure \ref{fig:realCaptured}.}
-    \label{fig:bitmap}
-\end{figure}
 
@@
 
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.8\columnwidth]{bitmap}
+    \caption{\small Cosine similarity matrix between real and cap-tured VFOA using a 10 sec. window size computation. The correspondent numbers are shown in figure \ref{fig:realCaptured}.}
+    \label{fig:bitmap}
+\end{figure}
+
 A relevant trend that can be observed is the continuous `look for approval' situation where the subjects look at the experimenter in order to provide feedback about the robot's response before providing their own one (see yellow boxes in figure \ref{fig:realExpected}).
@@
 Moreover, when a new demonstration is selected for playing, the same behavior is replicated (see yellow boxes in figure \ref{fig:realExpected}).
-
-If we consider the previous case, evaluating the subject's progression together with the time on task may be an indicator of engagement: If the subject answers so quickly with respect to the previous attempts with a poorer handwriting may be the case she is not engaged anymore.
-In such hypothetical case an activity switch can be easily implemented in a robot architecture.
-It confirms the findings reported in \cite{anzalone} where time response to event is proposed to account quality of interaction.
+\\
+\subsection{Focus of attention distribution}
+Additional information of the head pose can be obtained for the evaluation of the VFOA distribution.
+Figure \ref{heatmap} show the most salient focuses of attention where the peaks are located.
+These peaks are mainly projected between the triangle composed by the robot, the writing tablet and the selection tablet using an unsupervised K-means clustering method as figure \ref{kmeans} reveals, considering the observers as negligible.
 
@@
     \includegraphics[width=0.9\columnwidth]{heatmap}
-    \caption{\small 2D histogram of a subject's head movements.}
+    \caption{\small 2D histogram of subject 4 head movements. (head yaw on x-axis, pitch on y-axis). In red, peaks representing the most salient zones.}
     \label{heatmap}
@@
 
-However, a high amount of gaze shift \ref{fig:gazeR} towards the demonstration provided can reveal that the challenge is too big or no gaze, the exactly the opposite.
+\subsection{Engagement assessment}
+If we consider the case shown in section \ref{fromFaceTo}, evaluating the subject's progression together with the time on task may be an indicator of engagement: If the subject answers so quickly with respect to the previous attempts with a poorer handwriting may be the case she is not engaged anymore.
+In such hypothetical case an activity switch can be easily implemented in a robot architecture.
+It confirms the findings reported in \cite{anzalone} where time response to event is proposed to account quality of interaction.
+
+However, a high amount of gaze shift (figure \ref{fig:gazeR}) towards the demonstration provided can reveal that the challenge is too big or no gaze, the exactly the opposite.
 In this case, the robot response can be adapted to the situation.
-Similar, but least relevant results are obtained using the correctness \ref{fig:errorR} of the child's feedback after each attempt.
+Similar, but least relevant results are obtained using the correctness (figure \ref{fig:errorR}) of the child's feedback after each attempt.
 
-This information can be used to estimate the child's state during the interaction.
-For instance, if we consider that a low level of engagement is produced by extremely high or low demanding situations, it is possible to assess the engagement level during the interaction based on gaze shift and word correctness.
+\begin{figure}
+    \centering
+    \includegraphics[width=0.9\columnwidth]{gazeR}
+    \caption{\small Linear regression of all subject attempts against the occurrences of a gaze shift from the writing tablet towards the model provided.
+The gray region defines the confidence interval.}
+    \label{fig:gazeR}
+\end{figure}
 
-It is also important to notice that the initiative of the robot into the learning process in a turn-taking activity produces an effect in the engagement of the human partner.
-However, experience suggests that a robot that learns evoke a positive action in the partner.
-But the opposite, a robot that is not able to progress over time, produces the opposite effect.
-
-\begin{figure*}
-    \centering
-    \includegraphics[width=1.81\columnwidth]{realCaptured}
-    \caption{\small Real VFOA of subject 4 accoding to two observer's assessment.
-In green, the obtained VFOA using our aproach and in cyan, the mathing between them.
-Finally, the correspondent numbers with figure \ref{fig:bitmap}.}
-    \label{fig:realCaptured}
-\end{figure*}
-
-%Finally, the subjects set their attention more often to the tablet displaying the generated hand-writing during the robot feedback rather than to the robot itself.
-
-\begin{figure}[h!]
+\begin{figure}
     \centering
@@
 
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.9\columnwidth]{gazeR}
-    \caption{\small Linear regression of all subject attempts against the occurrences of a gaze shift from the writing tablet towards the model provided.
-The gray region defines the confidence interval.}
-    \label{fig:gazeR}
-\end{figure}
+This information can be used to estimate the child's state during the interaction.
+For instance, if we consider that a low level of engagement is produced by extremely high or low demanding situations, it is possible to assess the engagement level during the interaction based on gaze shift and word correctness.
+
+It is also important to notice that the initiative of the robot into the learning process in a turn-taking activity produces an effect in the engagement of the human partner.
+However, experience suggests that a robot that learns evoke a positive action in the partner.
+But the opposite, a robot that is not able to progress over time, produces the contrary effect.
+
+%Finally, the subjects set their attention more often to the tablet displaying the generated hand-writing during the robot feedback rather than to the robot itself.
 
@@
 In this work we have presented an on-line vision-based engagement assessment solution that can be extended to work in different platforms for simulating complex interaction scenarios monitoring several VFOA with a maker-less technology.
-It provides some benefits like the decrement of the system complexity as well as the increment of the naturalness of the interaction.
+It provides some benefits like the decrement of the system complexity as well as the increment of the naturalness of the interaction due to the use of an RGB sensor.
 
@@
 Additionally, both the gaze shift and the correctness of the feedback seem to be strictly correlated with the learning gain during the activity as they go in decrement after each attempt provided by the subjects.
-Despite the fact that the results are encouraging, there is a need of validation with a greater amount of subject's.
+Despite the fact that these results are encouraging, there is a need of validation with a greater amount of subjects.
 
 Regarding to the approach used, some limitations need to be considered.
-The most relevant one, the fact that the positions of the VFOA need to be known in advance and defined as static \textit{tf} frames.
-Moreover the gaze information acquired during the interactions can also be useful to develop actively the concept gaze-aware robot, where not only the robot knows where the subject looks at, but shares the same subject's VFOA when the activity state allows it.
+The most relevant one, the fact that the positions of the VFOA need to be known in advance and defined as static \textit{tf} frames in the system.
+Moreover, it is important to notice that eye gaze information is neglected which induces a certain amount of error, as well as the noise induced by the 3D pose estimation computed from face landmarkers.
 
-Moreover, other limitations are based on the fact that the eye gaze information is neglected which induces a certain amount of error, as well as the noise induced by the 3D pose estimation computed from face landmarkers.
+Gaze information acquired during the interactions is not only useful to detect and predict engagement, but also allows the possibility to develop actively the concept gaze-aware robot.
+In other words, make the system able to estimate where the subject looks at, but also imitate the same subject's VFOA when the activity state allows it.
 

### 2015-08-29 paper.tex
Updated title
Rewritten abstract

@@
 
-\title{On-line Vision-based Engagement Assessment for Interactive Robots}
+\title{Real-time Attention Assessment for Social Robots}
 \author{Fernando Garcia \qquad Séverin Lemaignan \qquad Alexis Jacq \qquad Pierre Dillenbourg\\Computer-Human Interaction in Learning and Instruction Laboratory (CHILI)\\École Polytechnique Fédérale de Lausanne (EPFL)}
@@
 \begin{abstract}
-In long term interactions between humans and robots, it is necessary that the agent exhibits some sort of social intelligence to be able to engage the human providing the appropriate response to specific face-to-face situations.
-In a educational context, this feature becomes a must due to the fact that engagement plays a relevant role in the learning process.
-Therefore, maximizing the interaction time at a high engagement level provides learning benefits.
-In this work, a visual focus of attention based on head pose estimation using a camera is proposed to be able to monitor and assess the quality of the interaction.
-We want to show a practical approach to evaluate the engagement that arises from the interaction reporting several experiences in the field, with the robot Nao and discussing the experimental results.
+Successfully sustaining face-to-face interaction between humans and robots mandates fine perception of each other attentional focus.
+Along this line, we present a novel technique to assess the visual focus of attention (VFOA) based on accurate, real-time head pose estimation with monocular camera.
+The software is fully open-source and can readily be adopted on a wide range of social robots.
+We propose an in-depth evaluation of the technique, using manual video annotations as ground truth, and show that our system reaches high level of accuracy in a real-world face-to-face interaction scenario.
+We finally show how to derive a metric of engagement from the monitoring of the VFOA, allowing the robot to account and react to the user's changing engagement in real-time.
 \end{abstract}
@@
 
-\keywords{Human-Robot Interaction; Social Robotics; Children.}
+\keywords{Human-Robot Interaction; Social Robotics; Attention Assessment; Engagement; Computer Vision.}
 

### 2015-08-29 paper.tex
Work on Introduction
A few unresolved issues, marked as 'fix me'

@@
 \documentclass{sig-alternate}
+
+\usepackage[draft,nomargin,footnote]{fixme}
+
+\usepackage{xspace}
+\newcommand{\eg}{\textit{e.g.}\xspace}
+\newcommand{\etal}{\textit{et al.}\xspace}
+\newcommand{\ie}{\textit{i.e.}\xspace}
+\newcommand{\etc}{\textit{etc.}\xspace}
+\newcommand{\vs}{\textit{vs.}\xspace}
 
@@
 In the recent years, research is pushing forward in building intelligent systems and environments that try to maximize the benefits of cooperation with humans based on verbal and non-verbal communication.
-The interaction between social robots and humans provide a rich set of behaviors, and reveals a need of capturing the significant passive information provided by the user during the interaction process.
-Such information should be used to adapt the system to the specific user's needs in order to be able to provide a social or emotional intelligence to the agents.
+The interaction between social robots and humans entails a broad set of situations, and evidences the need for capturing significant backchannel information generated by the user during the interaction process.
+Such information is used to adapt the system to the specific user's needs to eventually provide a social or emotional intelligence to the agents.
 
-While new algorithms and robotic technologies ensure better cognitive and interactive abilities, a deeper study about the benefits in the social interaction needs to be done: Is not enough to exhibit human-like capabilities in social tasks, but to do it in a human-like manner.
+It is indeed not enough to exhibit human-like capabilities in social tasks: a robot needs to achieve human-like manners.
 For instance, several studies showed that humans prefer to interact with robots whose task outcome is delayed, shows uncertainty or an undesired output \cite{Admoni,Short}.
@@
 
-In order to build intelligent agents, it is necessary to equip them with perceptive capabilities based on sensors an algorithms: face and object tracking and recognition \cite{Zhao:2003, Jafri:2014}, path planning \cite{Galceran:2013}, speech recognition \cite{brick2007incremental} or task learning \cite{calinon2007learning} are some of the current research topics.
-However, each algorithm is evaluated in their own metric space of its specific problem: efficiency, accuracy or time are some few examples.
-In \cite{anzalone} is exposed the need to evaluate such processes in terms of their capability to obtain the desired effect in a human-robot interaction context through several proposed metrics around engagement as quality experience indicator.
+In order to build intelligent agents, it is necessary to equip them with a range of perceptual capabilities: while face and object tracking and recognition~\cite{Zhao:2003, Jafri:2014}, path planning~\cite{Galceran:2013}, speech recognition~\cite{brick2007incremental} or task learning~\cite{calinon2007learning} are amongst the current major research directions, existing literature tends to evaluate each of these algorithms in their own metric space, without considering the overall interaction quality.
+Along this line, Anzalone \etal~\cite{anzalone} have recently argued in favor of such a global evaluation, and propose to assess these algorithms in term of their capability to obtain the desired effect in a human-robot interaction context, and they correspondingly propose metrics build around the measurement of \emph{engagement} as indicator of the quality of the experience.
 
@@
 
-In this way, in situ Visual Focus of Attention (VFOA) estimation is advantageous to human-robot interaction and a major contribution towards a better understanding of the interaction quality.
-\cite{yarbus1967eye, barber1976perception} suggest that human's interest is highly correlated with what they look at and the existing relationship between gaze and attention during social interaction has been exposed in \cite{argyle1969social}.
+In this way, in situ Visual Focus of Attention (VFOA) estimation is advantageous to human-robot interaction and a major contribution towards a better understanding of the interaction quality.\cite{yarbus1967eye, barber1976perception} suggest that human's interest is highly correlated with what they look at and the existing relationship between gaze and attention during social interaction has been exposed in~\cite{argyle1969social}.\fxfatal{This paragraph is not well connected to the previous ones}
 
-One of the most interesting fields where to develop this research is in a learning context since children's engagement sustainability over time is one of the most important variables to make
-the process of learning effective \cite{Umbach}.
-It is well known that the role of the teacher \cite{smith2005pedagogies}, as well as the way the activity is formulated do matters in the learning efficiency.
-Moreover, the use of robots in the learning activity presents an opportunity for children
-to interact with an embodied, physical agent as part of the learning experience.
+\subsection{Importance in learning environments}
 
-Furthermore, while \cite{zhao2012learning} recommends the development of a teachable agent that is actively, rather than passively, engaged in the learning by teaching interaction, the increased learning gain results attained in \cite{okita2006observation} were done so even with the pupil being taught having been instructed to not contribute much information.
+Education and learning environments are interesting application fields for research on attention assessment, since sustaining children's engagement over time is a key variable to ensure
+effective learning (see for instance~\cite{Umbach}), and they indeed motivate this research.
+Besides, while~\cite{zhao2012learning} recommends the development of a teachable agent that is actively, rather than passively, engaged in the learning by teaching interaction, the increased learning gain results attained in~\cite{okita2006observation} were done so even with the pupil being taught having been instructed to not contribute much information.
 This suggests that it may not be necessary, from a learning by teaching point of view, for the teachable agent to participate actively in the interaction to incite educational benefits.
-However, from an engagement point of view, this may not necessarily be true.
+However, from an engagement point of view, this may not necessarily be true.\fxfatal{rephrase, and be careful: 'learning by teaching' is only introduced in the next paragraph}
 
-In fact, the use of robots in education comes with the potential to engage the child in meta-cognition through the learning by teaching paradigm, wherein a student takes the role of a teacher and experiences stronger educational benefits as a result (such as in \cite{Palinscar1984}).
+In fact, the use of robots in education comes with the potential to engage the child in meta-cognition through the learning by teaching paradigm, wherein a student takes the role of a teacher and experiences stronger educational benefits as a result (such as in \cite{Palinscar1984}).\fxnote{to be merged with previous paragraph?}
 
-The focus of this research is to evaluate the quality of the interaction using the VFOA as well as the correctness in the performed task as engagement indicators in the context of the CoWriter project \footnote{The primary repository is \url{https://github.com/chili-epfl/cowriter_letter_learning}.}.
-In order to do that avoiding the use of 'invasive' technologies such as wearable devices becomes a must to keep the experience as natural as possible.
+The goal of this research is therefore to \textbf{evaluate the quality of an interaction} using the \textbf{visual focus of attention as an engagement indicator}.
+We validate our approach with field experiments conducted in the educational context of the CoWriter project~\cite{Hood:2015}, where our technique, being based on a regular, low-end RGB camera, avoids the use of otherwise typically invasive technologies such as eyetrackers, that would negatively impact the natural of the interaction.
 
-\section{Related Work}
+\subsection{Related Work}
 
@@
 
-\section{Research Questions}
+\subsection{Research Questions}
 

### 2015-08-31 paper.tex
Minor format changes

--- 2015-08-28 paper.tex
@@
 The participants' VFOA acquisition is the only dependent variable.
-This includes their gaze times and patterns of looking or not looking to the targets and the synchrony on the movements.
+This includes their gaze times and patterns of looking or not looking to the targets.
 This data extracted from an RGB camera was also coded post-hoc to quantify some of these measurements.
@@
     \centering
-    \includegraphics[width=0.8\columnwidth]{bitmap}
+    \includegraphics[width=0.7\columnwidth]{bitmap}
     \caption{\small Cosine similarity matrix between real and cap-tured VFOA using a 10 sec. window size computation. The correspondent numbers are shown in figure \ref{fig:realCaptured}.}
@@
 Additional information of the head pose can be obtained for the evaluation of the VFOA distribution.
-Figure \ref{heatmap} show the most salient focuses of attention where the peaks are located.
-These peaks are mainly projected between the triangle composed by the robot, the writing tablet and the selection tablet using an unsupervised K-means clustering method as figure \ref{kmeans} reveals, considering the observers as negligible.
+Figure \ref{fig:heatmap} shows the most salient focuses of attention where the peaks are located.
+These peaks are mainly projected between the triangle composed by the robot, the writing tablet and the selection tablet using an unsupervised K-means clustering method reveals, considering the observers as negligible VFOA.
 
-\begin{figure}[h!]
+% Workaround to do subfigure since the packege looks not to work
+\begin{figure}
   \centering
-    \includegraphics[width=0.9\columnwidth]{heatmap}
-    \caption{\small 2D histogram of subject 4 head movements. (head yaw on x-axis, pitch on y-axis). In red, peaks representing the most salient zones.}
-    \label{heatmap}
-\end{figure}
+  \begin{tabular}{@{}c@{}}
+    \includegraphics[width=.7\linewidth,height=100pt]{heatmap} \\
+    \small (a)  2D histogram.
+  \end{tabular}
 
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.9\columnwidth]{kmeans}
-    \caption{\small K-means clustering of the head movements.}
-    \label{kmeans}
+  \vspace{\floatsep}
+
+  \begin{tabular}{@{}c@{}}
+    \includegraphics[width=.7\linewidth,height=100pt]{kmeans} \\
+    \small (b) K-means clustering.
+  \end{tabular}
+
+  \caption{ \small Subject 4 head movements. (head yaw on x-axis, pitch on y-axis).}
+  \label{fig:heatmap}
 \end{figure}
@@
   \centering
-    \includegraphics[width=0.9\columnwidth]{gazeR}
-    \caption{\small Linear regression of all subject attempts against the occurrences of a gaze shift from the writing tablet towards the model provided.
+  \begin{tabular}{@{}c@{}}
+    \includegraphics[width=.7\linewidth,height=100pt]{gazeR} \\
+    \small (a) Occurrences of a gaze shift.
+  \end{tabular}\label{fig:gazeR}
+
+  \vspace{\floatsep}
+
+  \begin{tabular}{@{}c@{}}
+    \includegraphics[width=.7\linewidth,height=100pt]{errorR} \\
+    \small (b) Word correctness.
+  \end{tabular}\label{fig:errorR}
+
+  \caption{ \small Linear regression of all subject attempts against the gaze shift tablet-model (a) and the correctness as the average euclidean distance of each letter in its corresponding eigenspace with respect to a reference letter (b).
 The gray region defines the confidence interval.}
-    \label{fig:gazeR}
-\end{figure}
 
-\begin{figure}
-    \centering
-    \includegraphics[width=0.9\columnwidth]{errorR}
-    \caption{\small Linear regression of all subject attempts against the word correctness, measured as the average euclidean distance of each leter in its corresponding eigenspace with respect to a reference leter.
-The gray region defines the confidence interval.}
-    \label{fig:errorR}
 \end{figure}

### 2015-09-01 paper.tex
Fixed comments

@@
 Engagement describes a cognitive, affective (specifically intrinsic motivation), and behavioral state of interaction with a computer application that ``makes the user want to be there" \cite{OBrien:2010}.
-Engaging interactions were thought to involve attention, intrinsic interest, interactivity, perceived control and choice, functionality and motivation.
-
-In this way, in situ Visual Focus of Attention (VFOA) estimation is advantageous to human-robot interaction and a major contribution towards a better understanding of the interaction quality.\cite{yarbus1967eye, barber1976perception} suggest that human's interest is highly correlated with what they look at and the existing relationship between gaze and attention during social interaction has been exposed in~\cite{argyle1969social}.\fxfatal{This paragraph is not well connected to the previous ones}
+Engaging interactions were thought to involve intrinsic interest, interactivity, perceived control and choice, functionality, motivation and attention.
+In fact, \cite{yarbus1967eye,barber1976perception} suggest that human's attention is highly correlated with what they look at, and the existing relationship between gaze and attention during social interaction has been exposed in~\cite{argyle1969social}.
+In this way, in situ Visual Focus of Attention (VFOA) estimation is advantageous to human-robot interaction and a major contribution towards a better understanding of the interaction quality.
 
@@
 
-Education and learning environments are interesting application fields for research on attention assessment, since sustaining children's engagement over time is a key variable to ensure
-effective learning (see for instance~\cite{Umbach}), and they indeed motivate this research.
-Besides, while~\cite{zhao2012learning} recommends the development of a teachable agent that is actively, rather than passively, engaged in the learning by teaching interaction, the increased learning gain results attained in~\cite{okita2006observation} were done so even with the pupil being taught having been instructed to not contribute much information.
+Education and learning environments are interesting application fields for research on attention assessment, since sustaining children's engagement over time is a key variable to ensure effective learning (see for instance~\cite{Umbach}), and they indeed motivate this research.
+In fact, the use of robots in education comes with the potential to engage the child in meta-cognition through the learning by teaching paradigm, wherein a student takes the role of a teacher and experiences stronger educational benefits as a result (such as in \cite{Palinscar1984}).
+
+While~\cite{zhao2012learning} recommends the development of a teachable agent that is actively, rather than passively, engaged in the learning by teaching interaction, the increased learning gain results attained in~\cite{okita2006observation} were done so even with the pupil being taught having been instructed to not contribute much information.
 This suggests that it may not be necessary, from a learning by teaching point of view, for the teachable agent to participate actively in the interaction to incite educational benefits.
-However, from an engagement point of view, this may not necessarily be true.\fxfatal{rephrase, and be careful: 'learning by teaching' is only introduced in the next paragraph}
-
-In fact, the use of robots in education comes with the potential to engage the child in meta-cognition through the learning by teaching paradigm, wherein a student takes the role of a teacher and experiences stronger educational benefits as a result (such as in \cite{Palinscar1984}).\fxnote{to be merged with previous paragraph?}
+However, from an engagement point of view, this may not necessarily be true.
 
@@
 Moreover, this monitoring activity has been supporting our research in child-robot interaction for learning and instruction purposes.
-\vfill
 

### 2015-09-14 paper.tex
Wrapped the whole paper at 80 cols
This will ease diffing

@@
 \begin{abstract}
-Successfully sustaining face-to-face interaction between humans and robots mandates fine perception of each other attentional focus.
-Along this line, we present a novel technique to assess the visual focus of attention (VFOA) based on accurate, real-time head pose estimation with monocular camera.
-The software is fully open-source and can readily be adopted on a wide range of social robots.
-We propose an in-depth evaluation of the technique, using manual video annotations as ground truth, and show that our system reaches high level of accuracy in a real-world face-to-face interaction scenario.
-We finally show how to derive a metric of engagement from the monitoring of the VFOA, allowing the robot to account and react to the user's changing engagement in real-time.
+
+Successfully sustaining face-to-face interaction between humans and robots
+mandates fine perception of each other attentional focus. Along this line, we
+present a novel technique to assess the visual focus of attention (VFOA) based
+on accurate, real-time head pose estimation with monocular camera. The software
+is fully open-source and can readily be adopted on a wide range of social
+robots. We propose an in-depth evaluation of the technique, using manual video
+annotations as ground truth, and show that our system reaches high level of
+accuracy in a real-world face-to-face interaction scenario. We finally show how
+to derive a metric of engagement from the monitoring of the VFOA, allowing the
+robot to account and react to the user's changing engagement in real-time.
+
 \end{abstract}
@@
 
-In the recent years, research is pushing forward in building intelligent systems and environments that try to maximize the benefits of cooperation with humans based on verbal and non-verbal communication.
-The interaction between social robots and humans entails a broad set of situations, and evidences the need for capturing significant backchannel information generated by the user during the interaction process.
-Such information is used to adapt the system to the specific user's needs to eventually provide a social or emotional intelligence to the agents.
+In the recent years, research is pushing forward in building intelligent systems
+and environments that try to maximize the benefits of cooperation with humans
+based on verbal and non-verbal communication. The interaction between social
+robots and humans entails a broad set of situations, and evidences the need for
+capturing significant backchannel information generated by the user during the
+interaction process. Such information is used to adapt the system to the
+specific user's needs to eventually provide a social or emotional intelligence
+to the agents.
 
-It is indeed not enough to exhibit human-like capabilities in social tasks: a robot needs to achieve human-like manners.
-For instance, several studies showed that humans prefer to interact with robots whose task outcome is delayed, shows uncertainty or an undesired output \cite{Admoni,Short}.
-In fact, the ability to adapt has been proved to be beneficial in long-term interactions \cite{Tielman:2014, Lim:2014}, specifically with children, manifesting a more positive attitude over time.
+It is indeed not enough to exhibit human-like capabilities in social tasks: a
+robot needs to achieve human-like manners. For instance, several studies showed
+that humans prefer to interact with robots whose task outcome is delayed, shows
+uncertainty or an undesired output \cite{Admoni,Short}. In fact, the ability to
+adapt has been proved to be beneficial in long-term interactions
+\cite{Tielman:2014, Lim:2014}, specifically with children, manifesting a more
+positive attitude over time.
 
-In order to build intelligent agents, it is necessary to equip them with a range of perceptual capabilities: while face and object tracking and recognition~\cite{Zhao:2003, Jafri:2014}, path planning~\cite{Galceran:2013}, speech recognition~\cite{brick2007incremental} or task learning~\cite{calinon2007learning} are amongst the current major research directions, existing literature tends to evaluate each of these algorithms in their own metric space, without considering the overall interaction quality.
-Along this line, Anzalone \etal~\cite{anzalone} have recently argued in favor of such a global evaluation, and propose to assess these algorithms in term of their capability to obtain the desired effect in a human-robot interaction context, and they correspondingly propose metrics build around the measurement of \emph{engagement} as indicator of the quality of the experience.
+In order to build intelligent agents, it is necessary to equip them with a range
+of perceptual capabilities: while face and object tracking and
+recognition~\cite{Zhao:2003, Jafri:2014}, path planning~\cite{Galceran:2013},
+speech recognition~\cite{brick2007incremental} or task
+learning~\cite{calinon2007learning} are amongst the current major research
+directions, existing literature tends to evaluate each of these algorithms in
+their own metric space, without considering the overall interaction quality.
+Along this line, Anzalone \etal~\cite{anzalone} have recently argued in favor of
+such a global evaluation, and propose to assess these algorithms in term of
+their capability to obtain the desired effect in a human-robot interaction
+context, and they correspondingly propose metrics build around the measurement
+of \emph{engagement} as indicator of the quality of the experience.
 
-Engagement describes a cognitive, affective (specifically intrinsic motivation), and behavioral state of interaction with a computer application that ``makes the user want to be there" \cite{OBrien:2010}.
-Engaging interactions were thought to involve intrinsic interest, interactivity, perceived control and choice, functionality, motivation and attention.
-In fact, \cite{yarbus1967eye,barber1976perception} suggest that human's attention is highly correlated with what they look at, and the existing relationship between gaze and attention during social interaction has been exposed in~\cite{argyle1969social}.
-In this way, in situ Visual Focus of Attention (VFOA) estimation is advantageous to human-robot interaction and a major contribution towards a better understanding of the interaction quality.
+Engagement describes a cognitive, affective (specifically intrinsic motivation),
+and behavioral state of interaction with a computer application that ``makes the
+user want to be there" \cite{OBrien:2010}. Engaging interactions were thought to
+involve intrinsic interest, interactivity, perceived control and choice,
+functionality, motivation and attention. In fact,
+\cite{yarbus1967eye,barber1976perception} suggest that human's attention is
+highly correlated with what they look at, and the existing relationship between
+gaze and attention during social interaction has been exposed
+in~\cite{argyle1969social}. In this way, in situ Visual Focus of Attention
+(VFOA) estimation is advantageous to human-robot interaction and a major
+contribution towards a better understanding of the interaction quality.
 
@@
 
-Education and learning environments are interesting application fields for research on attention assessment, since sustaining children's engagement over time is a key variable to ensure effective learning (see for instance~\cite{Umbach}), and they indeed motivate this research.
-In fact, the use of robots in education comes with the potential to engage the child in meta-cognition through the learning by teaching paradigm, wherein a student takes the role of a teacher and experiences stronger educational benefits as a result (such as in \cite{Palinscar1984}).
+Education and learning environments are interesting application fields for
+research on attention assessment, since sustaining children's engagement over
+time is a key variable to ensure effective learning (see for
+instance~\cite{Umbach}), and they indeed motivate this research. In fact, the
+use of robots in education comes with the potential to engage the child in
+meta-cognition through the learning by teaching paradigm, wherein a student
+takes the role of a teacher and experiences stronger educational benefits as a
+result (such as in \cite{Palinscar1984}).
 
-While~\cite{zhao2012learning} recommends the development of a teachable agent that is actively, rather than passively, engaged in the learning by teaching interaction, the increased learning gain results attained in~\cite{okita2006observation} were done so even with the pupil being taught having been instructed to not contribute much information.
-This suggests that it may not be necessary, from a learning by teaching point of view, for the teachable agent to participate actively in the interaction to incite educational benefits.
-However, from an engagement point of view, this may not necessarily be true.
+While~\cite{zhao2012learning} recommends the development of a teachable agent
+that is actively, rather than passively, engaged in the learning by teaching
+interaction, the increased learning gain results attained
+in~\cite{okita2006observation} were done so even with the pupil being taught
+having been instructed to not contribute much information. This suggests that it
+may not be necessary, from a learning by teaching point of view, for the
+teachable agent to participate actively in the interaction to incite educational
+benefits. However, from an engagement point of view, this may not necessarily be
+true.
 
-The goal of this research is therefore to \textbf{evaluate the quality of an interaction} using the \textbf{visual focus of attention as an engagement indicator}.
-We validate our approach with field experiments conducted in the educational context of the CoWriter project~\cite{Hood:2015}, where our technique, being based on a regular, low-end RGB camera, avoids the use of otherwise typically invasive technologies such as eyetrackers, that would negatively impact the natural of the interaction.
+The goal of this research is therefore to \textbf{evaluate the quality of an
+interaction} using the \textbf{visual focus of attention as an engagement
+indicator}. We validate our approach with field experiments conducted in the
+educational context of the CoWriter project~\cite{Hood:2015}, where our
+technique, being based on a regular, low-end RGB camera, avoids the use of
+otherwise typically invasive technologies such as eyetrackers, that would
+negatively impact the natural of the interaction.
 
@@
 
-The notion of engagement has actively been studied in a diverse set of domains, and several variables and social signals have been proposed in the literature to quantify and study it in social interactions.
+The notion of engagement has actively been studied in a diverse set of domains,
+and several variables and social signals have been proposed in the literature to
+quantify and study it in social interactions.
 
-The evaluation of response times to provide an estimation of the user's engagement based on the correctness has been proposed in a learning context \cite{Beck} before.
-However, several aspects like the inter-subject variability or the environment complexity make it hardly applicable in a child-robot interaction.
-Additionally, other techniques \cite{Chipman07postureas} like the use of the weight distribution using a pressure sensitive chair has reported results.
-But it is in the computer vision field where major contributions have been done.
+The evaluation of response times to provide an estimation of the user's
+engagement based on the correctness has been proposed in a learning context
+\cite{Beck} before. However, several aspects like the inter-subject variability
+or the environment complexity make it hardly applicable in a child-robot
+interaction. Additionally, other techniques \cite{Chipman07postureas} like the
+use of the weight distribution using a pressure sensitive chair has reported
+results. But it is in the computer vision field where major contributions have
+been done.
 
-In \cite{peters2010investigating} approach, a model of user's interest and engagement with a virtual agent displaying shared attention behavior, by using eye gaze and head direction information is proposed.
-Similarly, in \cite{nakano2010estimating} an approach to estimate the user's conversational engagement with a conversational agent based on analysis of gaze patterns was implemented.
-Other significant work \cite{Castellano:2009} focus on feature evaluation related to the user's non-verbal behavior, the task and the companion's affective reactions are identified to predict the children's level of engagement using a Bayesian model.
+In \cite{peters2010investigating} approach, a model of user's interest and
+engagement with a virtual agent displaying shared attention behavior, by using
+eye gaze and head direction information is proposed. Similarly, in
+\cite{nakano2010estimating} an approach to estimate the user's conversational
+engagement with a conversational agent based on analysis of gaze patterns was
+implemented. Other significant work \cite{Castellano:2009} focus on feature
+evaluation related to the user's non-verbal behavior, the task and the
+companion's affective reactions are identified to predict the children's level
+of engagement using a Bayesian model.
 
-Some few studies have proposed approaches for the detection of engagement with a robot. In \cite{Rich:2010} a computational model based on the recognition of connection events such
-as directed gaze, mutual facial gaze is proposed.
-In this context, some other \cite{Sanghvi:2011} focus only on the back or trunk posture as a determined factor for the assessment.
-A recent study \cite{anzalone} with social robots in face to face scenarios proposed a set of metrics based on non-verbal cues but manifesting some possible limitations in long-term scenarios.
+Some few studies have proposed approaches for the detection of engagement with a
+robot. In \cite{Rich:2010} a computational model based on the recognition of
+connection events such as directed gaze, mutual facial gaze is proposed. In this
+context, some other \cite{Sanghvi:2011} focus only on the back or trunk posture
+as a determined factor for the assessment. A recent study \cite{anzalone} with
+social robots in face to face scenarios proposed a set of metrics based on
+non-verbal cues but manifesting some possible limitations in long-term
+scenarios.
 
-In this research, we address the problem of tracking the children's engagement based on their VFOA and current performed task during a long term child-robot interaction in a learning context, first.
-And second, we provide qualitative and quantitative evaluation of the outcome provided using a first person RGB sensor approach.
-Moreover, this monitoring activity has been supporting our research in child-robot interaction for learning and instruction purposes.
+In this research, we address the problem of tracking the children's engagement
+based on their VFOA and current performed task during a long term child-robot
+interaction in a learning context, first. And second, we provide qualitative and
+quantitative evaluation of the outcome provided using a first person RGB sensor
+approach. Moreover, this monitoring activity has been supporting our research in
+child-robot interaction for learning and instruction purposes.
 
@@
 
-The participants' VFOA acquisition is the only dependent variable.
-This includes their gaze times and patterns of looking or not looking to the targets.
-This data extracted from an RGB camera was also coded post-hoc to quantify some of these measurements.
+The participants' VFOA acquisition is the only dependent variable. This includes
+their gaze times and patterns of looking or not looking to the targets. This
+data extracted from an RGB camera was also coded post-hoc to quantify some of
+these measurements.
 
-We predict that there will be no generalizable pattern of participant's VFOA behavior for different scenarios such as multiple subjects, necessitating in situ VFOA estimation capabilities to enable robots to adapt in the specific scenarios.
-However, we believe that off-line pattern identification can lead to a better understanding of the children progression as well as engagement level assessment in face-to-face long term interactions.
+We predict that there will be no generalizable pattern of participant's VFOA
+behavior for different scenarios such as multiple subjects, necessitating in
+situ VFOA estimation capabilities to enable robots to adapt in the specific
+scenarios. However, we believe that off-line pattern identification can lead to
+a better understanding of the children progression as well as engagement level
+assessment in face-to-face long term interactions.
 
@@
 
-We propose a adaptive model based solution to approach head pose estimation that finds correspondences between facial landmarks points and their respective locations in a face model.
-We use the open-source library dlib\footnote{http://dlib.net} \cite{dlib09}, to detect and extract the skeleton of the subject's face and then, use the obtained landmarkers to calculate the head pose orientation estimation (see figure \ref{system}).
-In this way, we can estimate what the subject was looking at and obtain an \textit{rviz} 3D scene representation as it is executed as a ROS (Robot Operating System) node that can be seen in figure \ref{rviz}.
-Compared to intrusive hardware like head mounted displays, eye-trackers or wearables for instance, this approach has a clear advantage of simplicity and integration with robotic systems, but also becomes accurate enough for general purposes in face-to-face interactions.
-In addition, it allows us to model complex environments, handling different number and positions of participants in the scene.
+We propose a adaptive model based solution to approach head pose estimation that
+finds correspondences between facial landmarks points and their respective
+locations in a face model. We use the open-source library
+dlib\footnote{http://dlib.net} \cite{dlib09}, to detect and extract the skeleton
+of the subject's face and then, use the obtained landmarkers to calculate the
+head pose orientation estimation (see figure \ref{system}). In this way, we can
+estimate what the subject was looking at and obtain an \textit{rviz} 3D scene
+representation as it is executed as a ROS (Robot Operating System) node that can
+be seen in figure \ref{rviz}. Compared to intrusive hardware like head mounted
+displays, eye-trackers or wearables for instance, this approach has a clear
+advantage of simplicity and integration with robotic systems, but also becomes
+accurate enough for general purposes in face-to-face interactions. In addition,
+it allows us to model complex environments, handling different number and
+positions of participants in the scene.
 
@@
     \centering
-    \includegraphics[width=0.9\columnwidth]{system}
+    %\includegraphics[width=0.9\columnwidth]{system}
     \caption{\small System overview showing the most relevant components involved in the estimation.}
@@
 \subsection{Estimating Head Pose}
-The first step is to estimate the head pose using the pin-hole model shown in equation \ref{eq:pinHole}.
-As mentioned above, dlib library facilitates the task by providing the 2D points \textit{m'} that composes the face silhouette.
-However, it is enough with obtaining eight spread points, such as right and left eyes, right and left tragions, sellion or nasal depression zone, pronasale or center of the nose, stomion or center of the mouth and menton.
+The first step is to estimate the head pose using the pin-hole model shown in
+equation \ref{eq:pinHole}. As mentioned above, dlib library facilitates the task
+by providing the 2D points \textit{m'} that composes the face silhouette.
+However, it is enough with obtaining eight spread points, such as right and left
+eyes, right and left tragions, sellion or nasal depression zone, pronasale or
+center of the nose, stomion or center of the mouth and menton.
 
-Likewise, the 3D face model is needed.
-The proposed head pose estimation method would work best with a 3D face model of the query subject itself.
-When a 3D face model of the subject is not available, as in many practical situations, a generic 3D face model can be used.
-This is an estimation of the same anthropometric points \textit{M'} (for North American Caucasians) in the 3D space \cite{farkas1994anthropometry}.
-Moreover, the intrinsic camera parameters are necessary to be able to compose its matrix \textit{A}.
-The components are expanded in equation \ref{eq:pinHole2}.
+Likewise, the 3D face model is needed. The proposed head pose estimation method
+would work best with a 3D face model of the query subject itself. When a 3D face
+model of the subject is not available, as in many practical situations, a
+generic 3D face model can be used. This is an estimation of the same
+anthropometric points \textit{M'} (for North American Caucasians) in the 3D
+space \cite{farkas1994anthropometry}. Moreover, the intrinsic camera parameters
+are necessary to be able to compose its matrix \textit{A}. The components are
+expanded in equation \ref{eq:pinHole2}.
 
@@
 
-Finally, the matrix of extrinsic parameters [R|t] is obtained by joining the rotation matrix R (calculated from the rotation vector) and translation vector \textit{t}.
-The result is a matrix projection \textbf{P} which maps a point in the 3D space onto a point in the 2D image plane solving equation \ref{eq:p}.
+Finally, the matrix of extrinsic parameters [R|t] is obtained by joining the
+rotation matrix R (calculated from the rotation vector) and translation vector
+\textit{t}. The result is a matrix projection \textbf{P} which maps a point in
+the 3D space onto a point in the 2D image plane solving equation \ref{eq:p}.
 
@@
 \subsection{Setting the \textit{tf} Frame}
-The second step is to use the matrix [R|t] to set the correct translation and orientation of the \textit{tf} frame which represents the face in the 3D space.
-In order to achieve that it is necessary first, to set the origin defined by the vector \textit{t}.
-Second, the matrix \textit{R} needs to be reformulated into a quaternion as shown in equation \ref{eq:quat} due to the smoothly interpolation between them.
+The second step is to use the matrix [R|t] to set the correct translation and
+orientation of the \textit{tf} frame which represents the face in the 3D space.
+In order to achieve that it is necessary first, to set the origin defined by the
+vector \textit{t}. Second, the matrix \textit{R} needs to be reformulated into a
+quaternion as shown in equation \ref{eq:quat} due to the smoothly interpolation
+between them.
 
@@
 \\
-However, it is necessary to consider the cases where the division is performed by 0 (in 180\degree rotation about the y-axis for instance) or when \textit{qw} becomes close to zero.
+However, it is necessary to consider the cases where the division is performed
+by 0 (in 180\degree rotation about the y-axis for instance) or when \textit{qw}
+becomes close to zero.
 
 \subsection{Defining the Field of View}
-The third step is to generate a representation of the field of view whose origin and orientation are the ones defined by the face detection \textit{tf} frame.
+The third step is to generate a representation of the field of view whose origin
+and orientation are the ones defined by the face detection \textit{tf} frame.
 
 Defining accurately the human field of view in the implementation is a must.
-Holmqvist in \cite{holmqvist2011eye} specifies that the visual human range is $ \pm  40\degree $ in the horizontal and $ \pm 25\degree $ in the vertical.
-In \cite{walker1980clinical} Spector provides a more detailed specification for each eye splitting the vertical range into $ 60\degree $ the upper region and $ 75\degree $ the lower one.
-Moreover, the horizontal range gets separated in $ 60\degree $ inwards (towards the nose) and $ 95\degree $ outwards.
-In the present implementation, the first approach explained has been chosen representing the field of view using a cone with such dimensions.
+Holmqvist in \cite{holmqvist2011eye} specifies that the visual human range is $
+\pm  40\degree $ in the horizontal and $ \pm 25\degree $ in the vertical. In
+\cite{walker1980clinical} Spector provides a more detailed specification for
+each eye splitting the vertical range into $ 60\degree $ the upper region and $
+75\degree $ the lower one. Moreover, the horizontal range gets separated in $
+60\degree $ inwards (towards the nose) and $ 95\degree $ outwards. In the
+present implementation, the first approach explained has been chosen
+representing the field of view using a cone with such dimensions.
 
-Assuming that the \textit{tf} frames representing the robotic agent and the VFOA candidates are static, need to be monitored.
-This task consists on evaluating the intersection between the \textit{tf} frames and the field of view: If one focus of attention is within the region defined by the field of view, we assume the subject is looking at it.
-In order to solve such operation mathematically it is necessary to compute the transformation matrix between the coordinate frames to be able to locate the monitored frames \textit{A} in the subject's coordinate system \textit{B} (see equation \ref{eq:transform}).
+Assuming that the \textit{tf} frames representing the robotic agent and the VFOA
+candidates are static, need to be monitored. This task consists on evaluating
+the intersection between the \textit{tf} frames and the field of view: If one
+focus of attention is within the region defined by the field of view, we assume
+the subject is looking at it. In order to solve such operation mathematically it
+is necessary to compute the transformation matrix between the coordinate frames
+to be able to locate the monitored frames \textit{A} in the subject's coordinate
+system \textit{B} (see equation \ref{eq:transform}).
 
@@
 
-where \textit{v} is a point in \textit{A} that becomes v' in coordinate system \textit{B}.
-However, in practice there are, at least, three different \textit{tf} frames involved in the transformation (subject-camera, camera-base, base-focus).
-Following the transformation result, the distance of the monitored frame with respect to the x-axis face \textit{tf} frame, $ d_{x_{axis}} $, representing the subject's face can be computed as simple as in equation \ref{eq:pitagoras}.
-Always considering the x-axis as the one defining the face's orientation and thus field of view's main axis.
+where \textit{v} is a point in \textit{A} that becomes v' in coordinate system
+\textit{B}. However, in practice there are, at least, three different
+\textit{tf} frames involved in the transformation (subject-camera, camera-base,
+base-focus). Following the transformation result, the distance of the monitored
+frame with respect to the x-axis face \textit{tf} frame, $ d_{x_{axis}} $,
+representing the subject's face can be computed as simple as in equation
+\ref{eq:pitagoras}. Always considering the x-axis as the one defining the face's
+orientation and thus field of view's main axis.
 
@@
 
-It is necessary to compare the distance acquired $ d_{x_{axis}} $ with the radius of the field of view at the target \textit{tf} frame x-coordinate $ r_{fov} $ as shown in equation \ref{eq:fovtf}.
+It is necessary to compare the distance acquired $ d_{x_{axis}} $ with the
+radius of the field of view at the target \textit{tf} frame x-coordinate $
+r_{fov} $ as shown in equation \ref{eq:fovtf}.
 
@@
 
-where \textit{fov} is the aperture angle of the field of view. Finally, if $ d_{x_{axis}}<r_{fov} $, the target is within the subject's field of view.
+where \textit{fov} is the aperture angle of the field of view. Finally, if $
+d_{x_{axis}}<r_{fov} $, the target is within the subject's field of view.
 
@@
 \section{Case Study}
-The presented method have been successfully employed in a real child-robot interaction situation where the robot tries to engage the child in handwriting tasks using learning by teaching paradigm.
+
+The presented method have been successfully employed in a real child-robot
+interaction situation where the robot tries to engage the child in handwriting
+tasks using learning by teaching paradigm.
+
 \subsection{Experimental Set-up}
 
-The target of this study were 6 children with ages compressed between 5 and 6 years old.
-The experiments were developed individually and consisted on a writing activity by turns and one story telling performed by the robot at a given time of the interaction.
-All this performed in a over all 18 minutes interaction average time.
+The target of this study were 6 children with ages compressed between 5 and 6
+years old. The experiments were developed individually and consisted on a
+writing activity by turns and one story telling performed by the robot at a
+given time of the interaction. All this performed in a over all 18 minutes
+interaction average time.
 
@@
 
-A tactile tablet (with a custom application) is used for both the robot and the child to write: during a typical round, the child requests the robot to write something (a single letter, a number or a full word), and push the tablet towards the robot, the robot writes on the tablet by gesturing the writing (but without actually physically touching the tablet), the child then pull back the
-tablet, corrects the robot's attempt by writing him/herself on top or next to the robot's writing, and ``send'' his/her demonstration to the robot by pressing a small button on the tablet.
-The robot ``learns'' from this demonstration and tries again.
+A tactile tablet (with a custom application) is used for both the robot and the
+child to write: during a typical round, the child requests the robot to write
+something (a single letter, a number or a full word), and push the tablet
+towards the robot, the robot writes on the tablet by gesturing the writing (but
+without actually physically touching the tablet), the child then pull back the
+tablet, corrects the robot's attempt by writing him/herself on top or next to
+the robot's writing, and ``send'' his/her demonstration to the robot by pressing
+a small button on the tablet. The robot ``learns'' from this demonstration and
+tries again.
 
-Since the child is assumed to take on the role of the teacher, we had to ensure (s)he would be able to manage by him/herself the turn-taking and the overall progression of the activity (moving to the next letter or word).
-In our design, the turn-taking relies on the robot prompting for feedback once it is done with its writing (simple sentences like ``What do you think?''), and pressing on a
-small robot icon on the tablet once the child has finished correcting. In our experiments, both were easy to grasp for children.
+Since the child is assumed to take on the role of the teacher, we had to ensure
+(s)he would be able to manage by him/herself the turn-taking and the overall
+progression of the activity (moving to the next letter or word). In our design,
+the turn-taking relies on the robot prompting for feedback once it is done with
+its writing (simple sentences like ``What do you think?''), and pressing on a
+small robot icon on the tablet once the child has finished correcting. In our
+experiments, both were easy to grasp for children.
 
-Implementing such a system raises several challenges that are discussed in detail in \cite{Hood:2015}.
+Implementing such a system raises several challenges that are discussed in
+detail in \cite{Hood:2015}.
 
-An RGB camera has been used to acquire images of 640x480 pixels at 25 fps.
-The camera was located at 5 cm from the center of Nao's feet and its holder was tied to a wood base for stability purposes.
-Thus, the camera's objective was 9 cm above and $ 40\degree $ inclination with respect to the surface of the table to maximize the detection during the writing time.
-However, Nao's camera could have been used, always considering the \textit{tf} frame matrix update according to the head orientation.
+An RGB camera has been used to acquire images of 640x480 pixels at 25 fps. The
+camera was located at 5 cm from the center of Nao's feet and its holder was tied
+to a wood base for stability purposes. Thus, the camera's objective was 9 cm
+above and $ 40\degree $ inclination with respect to the surface of the table to
+maximize the detection during the writing time. However, Nao's camera could have
+been used, always considering the \textit{tf} frame matrix update according to
+the head orientation.
 
-The subjects were typically located 50 cm away from the robot with the first tablet in front and the second one 30 cm to the left of the first one.
-In the same way, the experimenter was located around 60 cm to the left of the subject.
-Finally, two observers were located far away from the interaction field to manually assess the state of the interaction.
+The subjects were typically located 50 cm away from the robot with the first
+tablet in front and the second one 30 cm to the left of the first one. In the
+same way, the experimenter was located around 60 cm to the left of the subject.
+Finally, two observers were located far away from the interaction field to
+manually assess the state of the interaction.
 
-Figure \ref{drawSetup} allows us to intuitively identify several focuses of attention along the interaction such as the two tablets, the robot, the experimenter and the observers located in the diagonal.
+Figure \ref{drawSetup} allows us to intuitively identify several focuses of
+attention along the interaction such as the two tablets, the robot, the
+experimenter and the observers located in the diagonal.
 
@@
 
-In this work, head orientation is used to predict a person's focus of attention in face-to-face interactions.
-Therefore, we assume that head pose is a reliable marker of the attention focus during a social interaction, in both cases human-human, human-robot.
-Since we estimate where the subject is looking at based on its head position, several important questions arise:
+In this work, head orientation is used to predict a person's focus of attention
+in face-to-face interactions. Therefore, we assume that head pose is a reliable
+marker of the attention focus during a social interaction, in both cases
+human-human, human-robot. Since we estimate where the subject is looking at
+based on its head position, several important questions arise:
 
@@
 
-According to \cite{stiefelhagen2002tracking} the head orientation's contribution in overall gaze direction is 68.9\% showing agreement between both measurements.
-In addition, it shows that the head orientation alone can get an average accuracy of 88.7\% detecting the focus of attention in a meeting application scenario using a head-mounted display with eye and head tracking.
+According to \cite{stiefelhagen2002tracking} the head orientation's contribution
+in overall gaze direction is 68.9\% showing agreement between both measurements.
+In addition, it shows that the head orientation alone can get an average
+accuracy of 88.7\% detecting the focus of attention in a meeting application
+scenario using a head-mounted display with eye and head tracking.
 
-Our results for a peer-to-peer work scenarios using an approach based on RGB camera are summarized in table \ref{tab:results} where the \textit{expected} is the focus of attention that given the state of the system, it is likely to attract the subject's gaze.
-In the same manner, the \textit{captured} is the focus of attention our approach provides and the \textit{real}, the ground truth of the subject's gaze direction based on manual annotations from two evaluators with an average reliability of $ \alpha = .89 $.
+Our results for a peer-to-peer work scenarios using an approach based on RGB
+camera are summarized in table \ref{tab:results} where the \textit{expected} is
+the focus of attention that given the state of the system, it is likely to
+attract the subject's gaze. In the same manner, the \textit{captured} is the
+focus of attention our approach provides and the \textit{real}, the ground truth
+of the subject's gaze direction based on manual annotations from two evaluators
+with an average reliability of $ \alpha = .89 $.
 
-In order to calculate the inter-rater reliability between the two measurements (real-captured and real-expected), the overlap percentage between the two curves has been computed.
-A more reliable measurement in terms of time shifting between the real and captured focuses of attention is provided using the cosine similarity also shown in table \ref{tab:results}.
+In order to calculate the inter-rater reliability between the two measurements
+(real-captured and real-expected), the overlap percentage between the two curves
+has been computed. A more reliable measurement in terms of time shifting between
+the real and captured focuses of attention is provided using the cosine
+similarity also shown in table \ref{tab:results}.
 
@@
 
-The results has proved that the VFOA prediction accuracy in the tracking is not only conditioned by the subject's amount of movement, but also by the speed of the same movements: Restless children externalize greater levels of movement decreasing the VFOA prediction accuracy.
+The results has proved that the VFOA prediction accuracy in the tracking is not
+only conditioned by the subject's amount of movement, but also by the speed of
+the same movements: Restless children externalize greater levels of movement
+decreasing the VFOA prediction accuracy.
 
-For instance, subjects 1,4 and 5 despite the fact that the tracking accuracy is in the average, the bad precision in the captured VFOA suggests that such children are highly likely to modify the gaze direction without head move intervention reducing dramatically the overlap between captured and real VFOA.
+For instance, subjects 1,4 and 5 despite the fact that the tracking accuracy is
+in the average, the bad precision in the captured VFOA suggests that such
+children are highly likely to modify the gaze direction without head move
+intervention reducing dramatically the overlap between captured and real VFOA.
 
-Additionally, several recurrent computer vision challenges such as the lighting conditions, occlusions with the hands and the face angle with respect to the camera, looking to the experimenter for instance, are factors that contribute to reduce the tracking accuracy.
+Additionally, several recurrent computer vision challenges such as the lighting
+conditions, occlusions with the hands and the face angle with respect to the
+camera, looking to the experimenter for instance, are factors that contribute to
+reduce the tracking accuracy.
 
@@
     \includegraphics[width=1.8\columnwidth]{realExpected}
-    \caption{\small In blue, real VFOA of subject 4 accoding to two observer's assessment.
-In cyan, the matching between the real and the obtained VFOA.
-In yellow, the repeated pattern in turn-taking and in orange, the experimenter's suggestion to choose a new word.}
+    \caption{\small In blue, real VFOA of subject 4 accoding to two observer's
+    assessment. In cyan, the matching between the real and the obtained VFOA. In
+    yellow, the repeated pattern in turn-taking and in orange, the experimenter's
+    suggestion to choose a new word.}
+
     \label{fig:realExpected}
@@
     \includegraphics[width=1.81\columnwidth]{realCaptured}
-    \caption{\small Real VFOA of subject 4 accoding to two observer's assessment.
-In green, the obtained VFOA using our aproach and in cyan, the mathing between them.
-Finally, the correspondent numbers with figure \ref{fig:bitmap}.}
+    \caption{\small Real VFOA of subject 4 accoding to two observer's
+    assessment. In green, the obtained VFOA using our aproach and in cyan, the
+    mathing between them. Finally, the correspondent numbers with figure \ref{fig:bitmap}.}
+
     \label{fig:realCaptured}
@@
 
-According to figure \ref{fig:realExpected} several findings can be analyzed in detail.
-The matching curve (cyan color) shows the overlapping between the expected and the real focuses of attention.
-As can be noticed there is a fluctuation at the beginning of the interaction as well as just before the activity switching.
-These chaotic gaze manifestation is what we call an automatic discovery of events, when the user experiences an uncertainty in the present situation.
+According to figure \ref{fig:realExpected} several findings can be analyzed in
+detail. The matching curve (cyan color) shows the overlapping between the
+expected and the real focuses of attention. As can be noticed there is a
+fluctuation at the beginning of the interaction as well as just before the
+activity switching. These chaotic gaze manifestation is what we call an
+automatic discovery of events, when the user experiences an uncertainty in the
+present situation.
 
-The first one of the previous two situations presented above is due to the novelty effect the subjects experience for the first time in front of the system and according to the rest of participants suggests that the adaptation time is around the first 2 minutes of interaction.
-In the second case, an unexpected robot behavior produces such reaction due to an activity switching, the story telling.
+The first one of the previous two situations presented above is due to the
+novelty effect the subjects experience for the first time in front of the system
+and according to the rest of participants suggests that the adaptation time is
+around the first 2 minutes of interaction. In the second case, an unexpected
+robot behavior produces such reaction due to an activity switching, the story
+telling.
 
-It is important to consider that in some occasions during the interaction, none of the focuses of attention were within the field of view due to the time transition from one focus to another resulting in an ambiguity.
-In order to deal with these few cases it is assumed that the minimum distance between the height of the field of view (cone) and the targets corresponds to the most likely focus of attention.
-However, we also want to distinguish such cases when the subject is directly not looking where the defined targets are located, but towards a region far from the interaction field.
-These cases are not common but they can be easily identified by setting a horizontal threshold to the field of view.
+It is important to consider that in some occasions during the interaction, none
+of the focuses of attention were within the field of view due to the time
+transition from one focus to another resulting in an ambiguity. In order to deal
+with these few cases it is assumed that the minimum distance between the height
+of the field of view (cone) and the targets corresponds to the most likely focus
+of attention. However, we also want to distinguish such cases when the subject
+is directly not looking where the defined targets are located, but towards a
+region far from the interaction field. These cases are not common but they can
+be easily identified by setting a horizontal threshold to the field of view.
 
@@
     \includegraphics[width=0.7\columnwidth]{bitmap}
-    \caption{\small Cosine similarity matrix between real and cap-tured VFOA using a 10 sec. window size computation. The correspondent numbers are shown in figure \ref{fig:realCaptured}.}
+    \caption{\small Cosine similarity matrix between real and cap-tured VFOA
+    using a 10 sec. window size computation. The correspondent numbers are shown
+    in figure \ref{fig:realCaptured}.}
     \label{fig:bitmap}
@@
 
-A relevant trend that can be observed is the continuous `look for approval' situation where the subjects look at the experimenter in order to provide feedback about the robot's response before providing their own one (see yellow boxes in figure \ref{fig:realExpected}).
-Such transition tablet-experimenter becomes more frequent in several cases, for instance when the experimenter makes a suggestion (orange box in figure \ref{fig:realExpected}), in this case, switch to a more complicated word.
+A relevant trend that can be observed is the continuous `look for approval'
+situation where the subjects look at the experimenter in order to provide
+feedback about the robot's response before providing their own one (see yellow
+boxes in figure \ref{fig:realExpected}). Such transition tablet-experimenter
+becomes more frequent in several cases, for instance when the experimenter makes
+a suggestion (orange box in figure \ref{fig:realExpected}), in this case, switch
+to a more complicated word.
 
-Once the turn-taking is established, a pattern becomes visible during the subject's feedback, when an attempt is performed: Such pattern shows a high intermittent gaze frequency from the tablet to the selection.
-It also reveals that the subject is using the demonstration model in the selection tablet to provide a better answer to the robot.
-In fact, subjects decrease the response time after each shift as well as the gaze frequency towards the model suggesting an assimilation over time.
-Moreover, when a new demonstration is selected for playing, the same behavior is replicated (see yellow boxes in figure \ref{fig:realExpected}).
-\\
+Once the turn-taking is established, a pattern becomes visible during the
+subject's feedback, when an attempt is performed: Such pattern shows a high
+intermittent gaze frequency from the tablet to the selection. It also reveals
+that the subject is using the demonstration model in the selection tablet to
+provide a better answer to the robot. In fact, subjects decrease the response
+time after each shift as well as the gaze frequency towards the model suggesting
+an assimilation over time. Moreover, when a new demonstration is selected for
+playing, the same behavior is replicated (see yellow boxes in figure
+\ref{fig:realExpected}).
+
 \subsection{Focus of attention distribution}
-Additional information of the head pose can be obtained for the evaluation of the VFOA distribution.
-Figure \ref{fig:heatmap} shows the most salient focuses of attention where the peaks are located.
-These peaks are mainly projected between the triangle composed by the robot, the writing tablet and the selection tablet using an unsupervised K-means clustering method reveals, considering the observers as negligible VFOA.
+
+Additional information of the head pose can be obtained for the evaluation of
+the VFOA distribution. Figure \ref{fig:heatmap} shows the most salient focuses
+of attention where the peaks are located. These peaks are mainly projected
+between the triangle composed by the robot, the writing tablet and the selection
+tablet using an unsupervised K-means clustering method reveals, considering the
+observers as negligible VFOA.
 
@@
 \subsection{Engagement assessment}
-If we consider the case shown in section \ref{fromFaceTo}, evaluating the subject's progression together with the time on task may be an indicator of engagement: If the subject answers so quickly with respect to the previous attempts with a poorer handwriting may be the case she is not engaged anymore.
-In such hypothetical case an activity switch can be easily implemented in a robot architecture.
-It confirms the findings reported in \cite{anzalone} where time response to event is proposed to account quality of interaction.
 
-However, a high amount of gaze shift (figure \ref{fig:gazeR}) towards the demonstration provided can reveal that the challenge is too big or no gaze, the exactly the opposite.
-In this case, the robot response can be adapted to the situation.
-Similar, but least relevant results are obtained using the correctness (figure \ref{fig:errorR}) of the child's feedback after each attempt.
+If we consider the case shown in section \ref{fromFaceTo}, evaluating the
+subject's progression together with the time on task may be an indicator of
+engagement: If the subject answers so quickly with respect to the previous
+attempts with a poorer handwriting may be the case she is not engaged anymore.
+In such hypothetical case an activity switch can be easily implemented in a
+robot architecture. It confirms the findings reported in \cite{anzalone} where
+time response to event is proposed to account quality of interaction.
+
+However, a high amount of gaze shift (figure \ref{fig:gazeR}) towards the
+demonstration provided can reveal that the challenge is too big or no gaze, the
+exactly the opposite. In this case, the robot response can be adapted to the
+situation. Similar, but least relevant results are obtained using the
+correctness (figure \ref{fig:errorR}) of the child's feedback after each
+attempt.
 
@@
 
-  \caption{ \small Linear regression of all subject attempts against the gaze shift tablet-model (a) and the correctness as the average euclidean distance of each letter in its corresponding eigenspace with respect to a reference letter (b).
-The gray region defines the confidence interval.}
+  \caption{ \small Linear regression of all subject attempts against the gaze
+  shift tablet-model (a) and the correctness as the average euclidean distance
+  of each letter in its corresponding eigenspace with respect to a reference
+  letter (b). The gray region defines the confidence interval.}
 
@@
 
-This information can be used to estimate the child's state during the interaction.
-For instance, if we consider that a low level of engagement is produced by extremely high or low demanding situations, it is possible to assess the engagement level during the interaction based on gaze shift and word correctness.
+This information can be used to estimate the child's state during the
+interaction. For instance, if we consider that a low level of engagement is
+produced by extremely high or low demanding situations, it is possible to assess
+the engagement level during the interaction based on gaze shift and word
+correctness.
 
-It is also important to notice that the initiative of the robot into the learning process in a turn-taking activity produces an effect in the engagement of the human partner.
-However, experience suggests that a robot that learns evoke a positive action in the partner.
-But the opposite, a robot that is not able to progress over time, produces the contrary effect.
+It is also important to notice that the initiative of the robot into the
+learning process in a turn-taking activity produces an effect in the engagement
+of the human partner. However, experience suggests that a robot that learns
+evoke a positive action in the partner. But the opposite, a robot that is not
+able to progress over time, produces the contrary effect.
 
@@
 \section{Conclusions}
-In this work we have presented an on-line vision-based engagement assessment solution that can be extended to work in different platforms for simulating complex interaction scenarios monitoring several VFOA with a maker-less technology.
-It provides some benefits like the decrement of the system complexity as well as the increment of the naturalness of the interaction due to the use of an RGB sensor.
+In this work we have presented an on-line vision-based engagement assessment
+solution that can be extended to work in different platforms for simulating
+complex interaction scenarios monitoring several VFOA with a maker-less
+technology. It provides some benefits like the decrement of the system
+complexity as well as the increment of the naturalness of the interaction due to
+the use of an RGB sensor.
 
-Furthermore, although it has not been tested in a real scenario with several subjects at the same time, the approach also allows to handle such situations too.
+Furthermore, although it has not been tested in a real scenario with several
+subjects at the same time, the approach also allows to handle such situations
+too.
 
-During the case study, results show that it is possible to distinguish between long and short feedback episodes based on the gaze shift and depending on the previous attempts performed by the subject.
-As a result, gaze-shift together with the number of attempts becomes a reliable indicator candidate in the assessment of the engagement during the activity.
+During the case study, results show that it is possible to distinguish between
+long and short feedback episodes based on the gaze shift and depending on the
+previous attempts performed by the subject. As a result, gaze-shift together
+with the number of attempts becomes a reliable indicator candidate in the
+assessment of the engagement during the activity.
 
-Additionally, both the gaze shift and the correctness of the feedback seem to be strictly correlated with the learning gain during the activity as they go in decrement after each attempt provided by the subjects.
-Despite the fact that these results are encouraging, there is a need of validation with a greater amount of subjects.
+Additionally, both the gaze shift and the correctness of the feedback seem to be
+strictly correlated with the learning gain during the activity as they go in
+decrement after each attempt provided by the subjects. Despite the fact that
+these results are encouraging, there is a need of validation with a greater
+amount of subjects.
 
-Regarding to the approach used, some limitations need to be considered.
-The most relevant one, the fact that the positions of the VFOA need to be known in advance and defined as static \textit{tf} frames in the system.
-Moreover, it is important to notice that eye gaze information is neglected which induces a certain amount of error, as well as the noise induced by the 3D pose estimation computed from face landmarkers.
+Regarding to the approach used, some limitations need to be considered. The most
+relevant one, the fact that the positions of the VFOA need to be known in
+advance and defined as static \textit{tf} frames in the system. Moreover, it is
+important to notice that eye gaze information is neglected which induces a
+certain amount of error, as well as the noise induced by the 3D pose estimation
+computed from face landmarkers.
 
-Gaze information acquired during the interactions is not only useful to detect and predict engagement, but also allows the possibility to develop actively the concept gaze-aware robot.
-In other words, make the system able to estimate where the subject looks at, but also imitate the same subject's VFOA when the activity state allows it.
+Gaze information acquired during the interactions is not only useful to detect
+and predict engagement, but also allows the possibility to develop actively the
+concept gaze-aware robot. In other words, make the system able to estimate where
+the subject looks at, but also imitate the same subject's VFOA when the activity
+state allows it.
 
@@
 \section*{Acknowledgments}
+
 This research was partially supported by the Funda\c{c}\~{a}o para a Ci\^{e}ncia

### 2015-09-14 paper.tex
Significant change in the paper sectionning
This idea is to organize the paper into:
1 - visual attention assessment
2 - experimental validation
3 - from attention to engagement

@@
 
-\section{Introduction}
-
+\section{Introduction: Attention and Engagement}
 In the recent years, research is pushing forward in building intelligent systems
@@
 
-\subsection{Importance in learning environments}
+\paragraph{Motivation: Impact on Learning}
 
@@
 
-\subsection{Related Work}
-
+\subsection*{Engagement and Attention: Related Work}
 The notion of engagement has actively been studied in a diverse set of domains,
@@
 
-\subsection{Research Questions}
+\fixme{Litterature review on visual attention...?}
 
+According to \cite{stiefelhagen2002tracking} the head orientation's contribution
+in overall gaze direction is 68.9\% showing agreement between both measurements.
+In addition, it shows that the head orientation alone can get an average
+accuracy of 88.7\% detecting the focus of attention in a meeting application
+scenario using a head-mounted display with eye and head tracking.
+
+\subsection*{Overview of the Article}
+
+\paragraph{Real-Time Assessment of Visual Focus of Attention}
 The participants' VFOA acquisition is the only dependent variable. This includes
@@
 
-\section{Approach}
+\paragraph{Experimental Validation of the Approach}
+
+\fixme{TBD}
+
+\paragraph{From Attention to Engagement}
+
+\fixme{TBD}
+
+\section{Attention Assessment}
 
@@
 
-\subsection{Setting the \textit{tf} Frame}
+\subsection{Acquiring the View of Attention}
+
+\paragraph{Setting the \textit{tf} Frame}\fixme{tf is ROS specific: we should
+not write the article in a ROS specific way}
 The second step is to use the matrix [R|t] to set the correct translation and
@@
 
-\subsection{Defining the Field of View}
+\paragraph{Computing the Field of Attention}
+
 The third step is to generate a representation of the field of view whose origin
@@
 
-\section{Case Study}
+\paragraph{Acquiring the Focus of Attention}
+
+Explain that: other modules have to provide the sitation assessment (what
+is where?) in terms of published frames, and a last module compute the
+intersection of the VFOA and these frames\fixme{TBD}
+
+\section{Experimental Validation}
 
@@
 
-\subsection{Experimental Set-up}
+\subsection{Experimental Procedure}
 
@@
 
-\section{Results}
+\subsection{Results}
 
@@
 
-\subsection{From Face Detection to Focus of Attention} \label{fromFaceTo}
-
-According to \cite{stiefelhagen2002tracking} the head orientation's contribution
-in overall gaze direction is 68.9\% showing agreement between both measurements.
-In addition, it shows that the head orientation alone can get an average
-accuracy of 88.7\% detecting the focus of attention in a meeting application
-scenario using a head-mounted display with eye and head tracking.
+\subsubsection{From Face Detection to Focus of Attention} \label{fromFaceTo}
 
@@
 
-\subsection{Subject's Focus of Attention Analysis}
+\subsubsection{Subject's Focus of Attention Analysis}
 
@@
 
-\subsection{Focus of attention distribution}
+\subsubsection{Focus of attention distribution}
 
@@
 
-\subsection{Engagement assessment}
+\section{Discussion and Conclusion}
+\subsection{From Attention to Engagement Assessment}
 
@@
 
-\section{Conclusions}
+\subsection{Conclusion}
+
 In this work we have presented an on-line vision-based engagement assessment

### 2015-09-14 paper.tex
Commented out for now out-of-scope parts of intro

@@
 \section{Introduction: Attention and Engagement}
-In the recent years, research is pushing forward in building intelligent systems
-and environments that try to maximize the benefits of cooperation with humans
-based on verbal and non-verbal communication. The interaction between social
-robots and humans entails a broad set of situations, and evidences the need for
-capturing significant backchannel information generated by the user during the
-interaction process. Such information is used to adapt the system to the
-specific user's needs to eventually provide a social or emotional intelligence
-to the agents.
-
-It is indeed not enough to exhibit human-like capabilities in social tasks: a
-robot needs to achieve human-like manners. For instance, several studies showed
-that humans prefer to interact with robots whose task outcome is delayed, shows
-uncertainty or an undesired output \cite{Admoni,Short}. In fact, the ability to
-adapt has been proved to be beneficial in long-term interactions
-\cite{Tielman:2014, Lim:2014}, specifically with children, manifesting a more
-positive attitude over time.
+%In the recent years, research is pushing forward in building intelligent systems
+%and environments that try to maximize the benefits of cooperation with humans
+%based on verbal and non-verbal communication. The interaction between social
+%robots and humans entails a broad set of situations, and evidences the need for
+%capturing significant backchannel information generated by the user during the
+%interaction process. Such information is used to adapt the system to the
+%specific user's needs to eventually provide a social or emotional intelligence
+%to the agents.
+%
+%It is indeed not enough to exhibit human-like capabilities in social tasks: a
+%robot needs to achieve human-like manners. For instance, several studies showed
+%that humans prefer to interact with robots whose task outcome is delayed, shows
+%uncertainty or an undesired output \cite{Admoni,Short}. In fact, the ability to
+%adapt has been proved to be beneficial in long-term interactions
+%\cite{Tielman:2014, Lim:2014}, specifically with children, manifesting a more
+%positive attitude over time.
 

### 2015-09-14 paper.tex
Minor change to abstract

@@
 annotations as ground truth, and show that our system reaches high level of
-accuracy in a real-world face-to-face interaction scenario. We finally show how
+accuracy in a real-world face-to-face interaction scenario. We finally suggest how
 to derive a metric of engagement from the monitoring of the VFOA, allowing the

### 2015-09-15 paper.tex

@@
 a better understanding of the children progression as well as engagement level
-assessment in face-to-face long term interactions.
+assessment in face-to-face long term interactions.\fixme{I did not understand
+this paragraph}
+
+\fixme{Underline difference between field of attention and focus of attention}
+
+\fixme{State the hypothesis: head-pose estimation provides a good-enough
+approximation of the gaze direction, and in turn of the field of attention, and
+in turn of the focus of attention}
 
@@
 
-\section{Attention Assessment}
+\section{Visual Attention Assessment}
+
+\subsection{Overview of the Technique}
 
@@
 
-The presented method have been successfully employed in a real child-robot
-interaction situation where the robot tries to engage the child in handwriting
-tasks using learning by teaching paradigm.
+As presented above, we use the (6D) head pose as an approximation of the
+actual gaze direction, and we approximate from here the participant's field of
+attention. The assumption that such an approximation of the field of attention
+allows to derive the actual focus of attention is the core hypothesis that we
+ought to validate.
+
+This section presents such an experimental validation, set in the context of a
+real child-robot interaction scenario where the robot tries to engage a child in
+handwriting tasks using learning by teaching paradigm.
 
@@
 \subsection{Results}
-
-In this work, head orientation is used to predict a person's focus of attention
-in face-to-face interactions. Therefore, we assume that head pose is a reliable
-marker of the attention focus during a social interaction, in both cases
-human-human, human-robot. Since we estimate where the subject is looking at
-based on its head position, several important questions arise:
-
-\begin{enumerate}
-\item Which head pose estimation tracking rate can we achieve during a real case scenario?  
-\item How accurate can we predict where the subject is looking at based on the head pose estimation?
-\item How can we evaluate the engagement of the children based on the focus of attention?\newline
-\end{enumerate}
 

### 2015-09-15 paper.tex
Added a few very relevant references

@@
 \subsection*{Engagement and Attention: Related Work}
+
+\paragraph{Head Tracking and Visual Attention}
+
+Technique for head pose estimation: based on depth
+sensing~\cite{fanelli2012real}.
+
+In our work, we rely on recent advances on template-based face
+alignement~\cite{kazemi2014one} that allows fast (in the order of
+a few milliseconds) face features extraction on 2D images, that we found to be fairly
+robust to real-world lighting conditions and occlusions naturally occuring in child-robot
+interactions.
+
+\cite{kennedy2015head} -> it does not work!
+
+\paragraph{Engagement}
+
 The notion of engagement has actively been studied in a diverse set of domains,
@@
 scenarios.
+
+In~\cite{baxter2014tracking}, Baxter~\etal posit that the measure of the
+direction and timing of gaze in child-robot interactions is a proxy for
+engagement and attribution of social agency, However, they conduct these
+measures as post-hoc analyses. This work suggests a path forward for real-time
+assessment of engagement.
 
@@
 certain amount of error, as well as the noise induced by the 3D pose estimation
-computed from face landmarkers.
+computed from face features.
+
+In that regards, it is interesting to put this work on human-robot interaction
+in perspective with previous research conducted on ape-human interaction by
+Tomasello~\etal~\cite{tomasello2007reliance}: they show that apes rely primarily
+on the head pose to estimate the focus of attention of a human partner, in to contrast
+to human children who primarily rely on eye direction. From this perspective,
+our robot mimics attention tracking mechanism comparable to the ones found in apes.
 

### 2015-09-15 paper.tex
Minor formatting

@@
 mandates fine perception of each other attentional focus. Along this line, we
-present a novel technique to assess the visual focus of attention (VFOA) based
+present a novel technique to assess the visual focus of attention (VFoA) based
 on accurate, real-time head pose estimation with monocular camera. The software
@@
 accuracy in a real-world face-to-face interaction scenario. We finally suggest how
-to derive a metric of engagement from the monitoring of the VFOA, allowing the
+to derive a metric of engagement from the monitoring of the VFoA, allowing the
 robot to account and react to the user's changing engagement in real-time.
@@
 in~\cite{argyle1969social}. In this way, in situ Visual Focus of Attention
-(VFOA) estimation is advantageous to human-robot interaction and a major
+(VFoA) estimation is advantageous to human-robot interaction and a major
 contribution towards a better understanding of the interaction quality.
@@
 In this research, we address the problem of tracking the children's engagement
-based on their VFOA and current performed task during a long term child-robot
+based on their VFoA and current performed task during a long term child-robot
 interaction in a learning context, first. And second, we provide qualitative and
@@
 \paragraph{Real-Time Assessment of Visual Focus of Attention}
-The participants' VFOA acquisition is the only dependent variable. This includes
+
+The participants' VFoA acquisition is the only dependent variable. This includes
 their gaze times and patterns of looking or not looking to the targets. This
@@
 
-We predict that there will be no generalizable pattern of participant's VFOA
+We predict that there will be no generalizable pattern of participant's VFoA
 behavior for different scenarios such as multiple subjects, necessitating in
-situ VFOA estimation capabilities to enable robots to adapt in the specific
+situ VFoA estimation capabilities to enable robots to adapt in the specific
 scenarios. However, we believe that off-line pattern identification can lead to
@@
 dlib\footnote{http://dlib.net} \cite{dlib09}, to detect and extract the skeleton
-of the subject's face and then, use the obtained landmarkers to calculate the
+of the subject's face and then, use the obtained features to calculate the
 head pose orientation estimation (see figure \ref{system}). In this way, we can
@@
     \centering
-    %\includegraphics[width=0.9\columnwidth]{system}
+    \includegraphics[width=0.9\columnwidth]{system}
     \caption{\small System overview showing the most relevant components involved in the estimation.}
@@
     \includegraphics[width=0.9\columnwidth]{rviz_camera}
-    \caption{\small Rviz screenshot showing Nao, camera (Nao's feet), face detected with its field of view (gray cone) and blue marker intersecting as current VFOA.}
+    \caption{\small Rviz screenshot showing Nao, camera (Nao's feet), face detected with its field of view (gray cone) and blue marker intersecting as current VFoA.}
     \label{rviz}
@@
 
-Assuming that the \textit{tf} frames representing the robotic agent and the VFOA
+Assuming that the \textit{tf} frames representing the robotic agent and the VFoA
 candidates are static, need to be monitored. This task consists on evaluating
@@
 is where?) in terms of published frames, and a last module compute the
-intersection of the VFOA and these frames\fixme{TBD}
+intersection of the VFoA and these frames\fixme{TBD}
 
@@
 
-The results has proved that the VFOA prediction accuracy in the tracking is not
+The results has proved that the VFoA prediction accuracy in the tracking is not
 only conditioned by the subject's amount of movement, but also by the speed of
 the same movements: Restless children externalize greater levels of movement
-decreasing the VFOA prediction accuracy.
+decreasing the VFoA prediction accuracy.
 
 For instance, subjects 1,4 and 5 despite the fact that the tracking accuracy is
-in the average, the bad precision in the captured VFOA suggests that such
+in the average, the bad precision in the captured VFoA suggests that such
 children are highly likely to modify the gaze direction without head move
-intervention reducing dramatically the overlap between captured and real VFOA.
+intervention reducing dramatically the overlap between captured and real VFoA.
 
@@
     \includegraphics[width=1.8\columnwidth]{realExpected}
-    \caption{\small In blue, real VFOA of subject 4 accoding to two observer's
-    assessment. In cyan, the matching between the real and the obtained VFOA. In
+    \caption{\small In blue, real VFoA of subject 4 accoding to two observer's
+    assessment. In cyan, the matching between the real and the obtained VFoA. In
     yellow, the repeated pattern in turn-taking and in orange, the experimenter's
@@
 
-    % Maybe we should add the expected VFOA
+    % Maybe we should add the expected VFoA
 \end{figure*}
@@
     \includegraphics[width=1.81\columnwidth]{realCaptured}
-    \caption{\small Real VFOA of subject 4 accoding to two observer's
-    assessment. In green, the obtained VFOA using our aproach and in cyan, the
+    \caption{\small Real VFoA of subject 4 accoding to two observer's
+    assessment. In green, the obtained VFoA using our aproach and in cyan, the
     mathing between them. Finally, the correspondent numbers with figure \ref{fig:bitmap}.}
@@
     \includegraphics[width=0.7\columnwidth]{bitmap}
-    \caption{\small Cosine similarity matrix between real and cap-tured VFOA
+    \caption{\small Cosine similarity matrix between real and cap-tured VFoA
     using a 10 sec. window size computation. The correspondent numbers are shown
@@
 Additional information of the head pose can be obtained for the evaluation of
-the VFOA distribution. Figure \ref{fig:heatmap} shows the most salient focuses
+the VFoA distribution. Figure \ref{fig:heatmap} shows the most salient focuses
 of attention where the peaks are located. These peaks are mainly projected
@@
 tablet using an unsupervised K-means clustering method reveals, considering the
-observers as negligible VFOA.
+observers as negligible VFoA.
 
 % Workaround to do subfigure since the packege looks not to work
-\begin{figure}
+\begin{figure}[th!]
   \centering
@@
 solution that can be extended to work in different platforms for simulating
-complex interaction scenarios monitoring several VFOA with a maker-less
+complex interaction scenarios monitoring several VFoA with a maker-less
 technology. It provides some benefits like the decrement of the system
@@
 Regarding to the approach used, some limitations need to be considered. The most
-relevant one, the fact that the positions of the VFOA need to be known in
+relevant one, the fact that the positions of the VFoA need to be known in
 advance and defined as static \textit{tf} frames in the system. Moreover, it is
@@
 concept gaze-aware robot. In other words, make the system able to estimate where
-the subject looks at, but also imitate the same subject's VFOA when the activity
+the subject looks at, but also imitate the same subject's VFoA when the activity
 state allows it.

### 2015-09-15 paper.tex
Working on the intro/state of art

@@
 
-\section{Introduction: Attention and Engagement}
+\section{Introduction}
+
 %In the recent years, research is pushing forward in building intelligent systems
@@
 their own metric space, without considering the overall interaction quality.
-Along this line, Anzalone \etal~\cite{anzalone} have recently argued in favor of
+Along this line, Anzalone~\etal~\cite{anzalone} have recently argued in favor of
 such a global evaluation, and propose to assess these algorithms in term of
@@
 involve intrinsic interest, interactivity, perceived control and choice,
-functionality, motivation and attention. In fact,
-\cite{yarbus1967eye,barber1976perception} suggest that human's attention is
-highly correlated with what they look at, and the existing relationship between
-gaze and attention during social interaction has been exposed
-in~\cite{argyle1969social}. In this way, in situ Visual Focus of Attention
-(VFoA) estimation is advantageous to human-robot interaction and a major
-contribution towards a better understanding of the interaction quality.
+functionality, motivation and attention.
 
@@
 
-\subsection*{Engagement and Attention: Related Work}
+\subsection*{Attention and Engagement: Related Work}
 
@@
 
-Technique for head pose estimation: based on depth
-sensing~\cite{fanelli2012real}.
+The relations between one's focus of attention and what he/she is looking at has
+long been established~\cite{yarbus1967eye,barber1976perception}, and more
+specifically, the existing relationship between gaze and attention during social
+interaction and the related gaze patterns, belongs to classic textbooks
+like~\cite{argyle1969social}. As such, measuring the direction of gaze is
+a natural proxy to estimate the (visual) focus of attention of a social agent.
+
+However, performing such a measure on a robot, in real-time, and in ecologically
+valid environments (which rules out bulky or invasive apparatus like
+eye-trackers, or techniques requiring fine calibration and/or static
+interactions) remains a key challenge in the field of Human-Robot Interaction.
+
+The current state-of-the-art techniques found in HRI that operate both on-line
+and during field experiments rely on head pose
+estimation, most often based on depth sensors (RGB-D). Fanelli~\etal provides a
+overview of these approaches in~\cite{fanelli2012real}. While fairly robust, these
+techniques remain CPU intensive for a limited accuracy.
 
 In our work, we rely on recent advances on template-based face
-alignement~\cite{kazemi2014one} that allows fast (in the order of
-a few milliseconds) face features extraction on 2D images, that we found to be fairly
-robust to real-world lighting conditions and occlusions naturally occuring in child-robot
-interactions.
+alignment~\cite{kazemi2014one} that allows fast (in the order of a few
+milliseconds) face features extraction on 2D images, that we found to be fairly
+robust to real-world lighting conditions and occlusions naturally occurring in
+child-robot interactions.
+
+In this way, in situ Visual Focus of Attention
+(VFoA) estimation is advantageous to human-robot interaction and a major
+contribution towards a better understanding of the interaction quality.
 

### 2015-09-16 paper.tex
Work on literature + technique + added/updated figures

@@
 specifically, the existing relationship between gaze and attention during social
-interaction and the related gaze patterns, belongs to classic textbooks
-like~\cite{argyle1969social}. As such, measuring the direction of gaze is
-a natural proxy to estimate the (visual) focus of attention of a social agent.
+interaction, and the related gaze patterns, has been part of classic textbooks
+like~\cite{argyle1969social} since decades. As such, there is little doubt that
+measuring the direction of gaze is a useful proxy to estimate the (visual) focus
+of attention of a social agent, and indeed this is one of the basic tool used in
+social psychology.
+
+Estimating attention using gaze is certainly not new to robotics either. A
+recent survey by Ruhland~\etal~\cite{ruhland2015review} presents in a synthetic
+way eye gaze research in HCI and social robotics, and is an essential read in
+that regard. It remains however an active field of research, as illustrated by
+several recent publications~\cite{baxter2014tracking,anzalone,kennedy2015head}:
+as attention tracking is essential in human-robot interaction
 
@@
 
-\subsection{Overview of the Technique}
+\subsection{Head Pose Estimation}
 
-We propose a adaptive model based solution to approach head pose estimation that
-finds correspondences between facial landmarks points and their respective
-locations in a face model. We use the open-source library
-dlib\footnote{http://dlib.net} \cite{dlib09}, to detect and extract the skeleton
+As explained, we derive the visual field of attention from the head pose. Our
+technique only involves a single monocular RGB camera for used for facial
+feature extraction, and a static simplified 3D mesh of a human head
+(Figure~\ref{head_pose}). 68 facial features are extracted using a fast
+template-based face alignment algorithm by Kazemi and
+Sullivan~\cite{kazemi2014one}, as implemented in the open-source {\tt dlib}
+library\footnote{http://dlib.net}~\cite{dlib09}.  Eight of these features are
+matched to their 3D counterparts and we rely on an iterative $PnP$ algorithm
+(OpenCV's implementation) to compute the translation and rotation of the head
+with respect to the camera frame. With this approach, knowing the intrinsic
+parameters of the camera (calibrated camera) is required for an accurate
+estimation of the absolute 3D localization of the head.
+
+Besides being fast, the face alignment algorithm has been found to perform well,
+including in a range of difficult situations (occlusions, head angle) as
+encountered in field experiments.  Figure~\ref{head_pose_real_world} shows a
+selection of such difficult scenes on one child.
+
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=0.9\linewidth]{head_pose}
+    \caption{The 6D head pose is estimated by fitting a 3D model of an
+        adult head (left) onto the detected 2D features of the face (right). We
+        rely on an iterative $PnP$ algorithm, using 8 correspondence pairs
+        (three are depicted: the sellion -- the nasal depression --, the left
+        tragion and the menton).}
+
+    \label{head_pose}
+\end{figure}
+
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=\linewidth]{head_pose_real_world}
+    \caption{Head pose results on images captured during a field experiment.
+    Detection of face features (and therefore, estimation of the pose) is robust
+    to significant occlusions and face rotations.}
+    \label{head_pose_real_world}
+\end{figure}
+
+\subsection{Field of Attention}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.9\columnwidth]{field_of_attention}
+    \caption{\small The visual field of attention is approximated to a
+        20$\degree$ cone, spanning from the head's sellion. The objects whose 3D
+    pose intersect with this cone are considered \emph{in focus}.}
+    \label{rviz}
+\end{figure}
+
+, to detect and extract the skeleton
 of the subject's face and then, use the obtained features to calculate the
@@
 positions of participants in the scene.
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.9\columnwidth]{system}
-    \caption{\small System overview showing the most relevant components involved in the estimation.}
-    \label{system}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=0.9\columnwidth]{rviz_camera}
-    \caption{\small Rviz screenshot showing Nao, camera (Nao's feet), face detected with its field of view (gray cone) and blue marker intersecting as current VFoA.}
-    \label{rviz}
-\end{figure}
-
-\subsection{Estimating Head Pose}
-The first step is to estimate the head pose using the pin-hole model shown in
-equation \ref{eq:pinHole}. As mentioned above, dlib library facilitates the task
-by providing the 2D points \textit{m'} that composes the face silhouette.
-However, it is enough with obtaining eight spread points, such as right and left
-eyes, right and left tragions, sellion or nasal depression zone, pronasale or
-center of the nose, stomion or center of the mouth and menton.
-
-Likewise, the 3D face model is needed. The proposed head pose estimation method
-would work best with a 3D face model of the query subject itself. When a 3D face
-model of the subject is not available, as in many practical situations, a
-generic 3D face model can be used. This is an estimation of the same
-anthropometric points \textit{M'} (for North American Caucasians) in the 3D
-space \cite{farkas1994anthropometry}. Moreover, the intrinsic camera parameters
-are necessary to be able to compose its matrix \textit{A}. The components are
-expanded in equation \ref{eq:pinHole2}.
-
-\begin{equation}
-s  \; m' = A [R|t] M'
-\label{eq:pinHole}
-\end{equation}
-\begin{equation}
-s
-\begin{bmatrix}
-x \\
-y \\
-1
-\end{bmatrix}
-=
-\begin{bmatrix}
-f_x & 0 & c_x  \\
-0 & f_y & c_y  \\
-0 & 0 & 1 
-\end{bmatrix}
-\begin{bmatrix}
-r_{11} & r_{12} & r_{13} & t_x  \\
-r_{21} & r_{22} & r_{23} & t_y  \\
-r_{31} & r_{32} & r_{33} & t_z  
-\end{bmatrix}
-\begin{bmatrix}
-X \\
-Y \\
-Z \\
-1
-\end{bmatrix}
-\label{eq:pinHole2}
-\end{equation}
-
-Finally, the matrix of extrinsic parameters [R|t] is obtained by joining the
-rotation matrix R (calculated from the rotation vector) and translation vector
-\textit{t}. The result is a matrix projection \textbf{P} which maps a point in
-the 3D space onto a point in the 2D image plane solving equation \ref{eq:p}.
-
-\begin{equation}
-s  \; m' = P \; M'
-\label{eq:p}
-\end{equation}
-
-\subsection{Acquiring the View of Attention}
 
@@
 
-Defining accurately the human field of view in the implementation is a must.
-Holmqvist in \cite{holmqvist2011eye} specifies that the visual human range is $
+The breadth of the human field of attention has been discussed in the
+literature.
+Holmqvist in~\cite{holmqvist2011eye} specifies that the visual human
+range\fixme{field of view or field of attention?} is $
 \pm  40\degree $ in the horizontal and $ \pm 25\degree $ in the vertical. In
@@
 75\degree $ the lower one. Moreover, the horizontal range gets separated in $
-60\degree $ inwards (towards the nose) and $ 95\degree $ outwards. In the
+60\degree $ inwards (towards the nose) and $ 95\degree $ outwards. In robotics,
+Sisbot\etal~\cite{sisbot2011situation} use a cone of 30$\degree$. In the
 present implementation, the first approach explained has been chosen
-representing the field of view using a cone with such dimensions.
+representing the field of view using a cone with such dimensions\fixme{which
+one? 40deg? 25deg?}.
 
@@
     \label{drawSetup}
+\end{figure}
+
+\subsection{System Implementation}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.9\columnwidth]{system}
+    \caption{\small System overview showing the most relevant components involved in the estimation.}
+    \label{system}
 \end{figure}

### 2015-09-17 paper.tex
Added a paragraph mentioning no treatment of visual occlusions in focus of attention assessment

@@
 
+Besides, our approach does not account for visual occlusions when computing the
+focus of attention: any frame within the cone of the visual field of attention
+is considered as \emph{in focus}. In reality, some of the frames may be visually
+occluded and thus not seen by the subject. Computing actual visibility requires
+more general situation assessment capabilities, and in particular (approximate)
+3D models of the surrounding objects so that visibility testing may be
+performed, as done in~\cite{sisbot2011situation} for instance. Our experimental
+setting did not lead to any specific visual occlusions, and we did not developed
+this capability.
+
+Compared to intrusive hardware like head mounted
+displays, eye-trackers or wearables for instance, this approach has a clear
+advantage of simplicity and integration with robotic systems, but also becomes
+accurate enough for general purposes in face-to-face interactions. In addition,
+it allows us to model complex environments, handling different number and
+positions of participants in the scene.
+
 In that regards, it is interesting to put this work on human-robot interaction

### 2015-09-17 paper.tex
Shuffle around pieces of text, new figure to illustrate matching between ground truth and computed VFoA

@@
 \keywords{Human-Robot Interaction; Social Robotics; Attention Assessment; Engagement; Computer Vision.}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-The current state-of-the-art techniques found in HRI that operate both on-line
-and during field experiments rely on head pose
-estimation, most often based on depth sensors (RGB-D). Fanelli~\etal provides a
-overview of these approaches in~\cite{fanelli2012real}. While fairly robust, these
-techniques remain CPU intensive for a limited accuracy.
+Looking at techniques that both operate on-line and have been deployed in field
+experiments, one finds that most approaches rely on head pose estimation alone
+(no eye gaze tracking) and are most often based on depth sensors (RGB-D) .
+Fanelli~\etal provides a overview of these approaches in~\cite{fanelli2012real},
+and recent examples include~\cite{baxter2014tracking} or~\cite{anzalone}.
 
-In our work, we rely on recent advances on template-based face
-alignment~\cite{kazemi2014one} that allows fast (in the order of a few
-milliseconds) face features extraction on 2D images, that we found to be fairly
-robust to real-world lighting conditions and occlusions naturally occurring in
-child-robot interactions.
+Approaches based on monocular 2D vision have been explored as
+well~\cite{peters2010investigating}, with however limited robustness to
+occlusions or lightning conditions, and over-reliance on tracking to maintain
+real-time performances.  Our work relies on recent advances on template-based
+face alignment~\cite{kazemi2014one} that allows fast (in the order of a few
+milliseconds) facial features extraction on 2D images, combined with 3D model
+fitting, to obtain a fast, robust and stable 6D head pose estimate, that was
+successfully deployed in field experiments involving child-robot interactions.
 
-In this way, in situ Visual Focus of Attention
-(VFoA) estimation is advantageous to human-robot interaction and a major
-contribution towards a better understanding of the interaction quality.
+We then derive the field of attention from the head pose: according
+to~\cite{stiefelhagen2002tracking} the head orientation's contribution in
+overall gaze direction is 68.9\% showing agreement between both measurements.
+In addition, it shows that the head orientation alone can get an average
+accuracy of 88.7\% detecting the focus of attention in a meeting application
+scenario using a head-mounted display with eye and head tracking.
 
-\cite{kennedy2015head} -> it does not work!
+While previous preliminary research in HRI seemed to indicate that deriving
+attentional focus from head pose alone would not be accurate
+enough~\cite{kennedy2015head}, we found in our case acceptable levels of
+agreement between the robot observation and manual post-hoc annotations.
 
@@
 
-In \cite{peters2010investigating} approach, a model of user's interest and
+In~\cite{peters2010investigating} approach, a model of user's interest and
 engagement with a virtual agent displaying shared attention behavior, by using
@@
 
-\fixme{Litterature review on visual attention...?}
-
-According to \cite{stiefelhagen2002tracking} the head orientation's contribution
-in overall gaze direction is 68.9\% showing agreement between both measurements.
-In addition, it shows that the head orientation alone can get an average
-accuracy of 88.7\% detecting the focus of attention in a meeting application
-scenario using a head-mounted display with eye and head tracking.
-
 \subsection*{Overview of the Article}
@@
 \fixme{TBD}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \end{figure}
-
-, to detect and extract the skeleton
-of the subject's face and then, use the obtained features to calculate the
-head pose orientation estimation (see figure \ref{system}). In this way, we can
-estimate what the subject was looking at and obtain an \textit{rviz} 3D scene
-representation as it is executed as a ROS (Robot Operating System) node that can
-be seen in figure \ref{rviz}. Compared to intrusive hardware like head mounted
-displays, eye-trackers or wearables for instance, this approach has a clear
-advantage of simplicity and integration with robotic systems, but also becomes
-accurate enough for general purposes in face-to-face interactions. In addition,
-it allows us to model complex environments, handling different number and
-positions of participants in the scene.
-
-\paragraph{Setting the \textit{tf} Frame}\fixme{tf is ROS specific: we should
-not write the article in a ROS specific way}
-The second step is to use the matrix [R|t] to set the correct translation and
-orientation of the \textit{tf} frame which represents the face in the 3D space.
-In order to achieve that it is necessary first, to set the origin defined by the
-vector \textit{t}. Second, the matrix \textit{R} needs to be reformulated into a
-quaternion as shown in equation \ref{eq:quat} due to the smoothly interpolation
-between them.
-
-\begin{equation}
-\begin{bmatrix}
-qw \\
-qx \\
-qy \\
-qz
-\end{bmatrix}
-=
-\begin{bmatrix}
-\sqrt{1 + r_{11} + r_{22} + r_{33}} /2 \\
-(r_{32} - r_{23})/( 4 \cdot qw) \\
-(r_{13} - r_{31})/( 4 \cdot qw) \\
-(r_{21} - r_{12})/( 4 \cdot qw)
-\end{bmatrix}
-\label{eq:quat}
-\end{equation}
-\\
-However, it is necessary to consider the cases where the division is performed
-by 0 (in 180\degree rotation about the y-axis for instance) or when \textit{qw}
-becomes close to zero.
-
-\paragraph{Computing the Field of Attention}
 
@@
 intersection of the VFoA and these frames\fixme{TBD}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
     \includegraphics[width=0.7\columnwidth]{drawSetup}
-    \caption{\small Graphical representation of the set-up and the possible children focuses of attention in the scene, in red.}
+    \caption{\small \textbf{Experimental setting} Graphical representation of the set-up and the possible children focuses of attention in the scene, in red.}
     \label{drawSetup}
@@
 
+, to detect and extract the skeleton
+of the subject's face and then, use the obtained features to calculate the
+head pose orientation estimation (see figure \ref{system}). In this way, we can
+estimate what the subject was looking at and obtain an \textit{rviz} 3D scene
+representation as it is executed as a ROS (Robot Operating System) node that can
+be seen in figure \ref{rviz}.
+
 \subsection{Results}
@@
 of the subject's gaze direction based on manual annotations from two evaluators
-with an average reliability of $ \alpha = .89 $.
+with an average reliability of $ \alpha = .89 $\fixme{Is that Cronbach alpha or
+Krippendorff's alpha?}.
 
 In order to calculate the inter-rater reliability between the two measurements
-(real-captured and real-expected), the overlap percentage between the two curves
+(real-captured and real-expected), the overlap percentage between the two
+curves\fixme{which curve?}
 has been computed. A more reliable measurement in terms of time shifting between
 the real and captured focuses of attention is provided using the cosine
-similarity also shown in table \ref{tab:results}.
+similarity\fixme{this needs to be better justified: why not a Cronbach alpha?} also shown in table~\ref{tab:results}.
 
@@
     \centering
-    \includegraphics[width=1.8\columnwidth]{realExpected}
-    \caption{\small In blue, real VFoA of subject 4 accoding to two observer's
-    assessment. In cyan, the matching between the real and the obtained VFoA. In
-    yellow, the repeated pattern in turn-taking and in orange, the experimenter's
-    suggestion to choose a new word.}
+    \includegraphics[width=\linewidth]{matches}
+    \caption{\small \textbf{Comparison of computed focus of attention \vs ground
+        truth} during a 20+ min face-to-face child-robot interaction.
+        In orange (top line), the focus of attention as computed by the robot;
+        in blue (bottom line), the focus of attention as manually annotated. The
+    bottom graph shows agreements.}
 
@@
 \end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2015-09-17 paper.tex
Tentative rewrite of the abstract with the concept of with-me-ness

@@
 
-\title{Real-time Attention Assessment for Social Robots}
+\title{From Real-time Attention Assessment to ``With-me-ness'' in Human-Robot Interaction}
 \author{Fernando Garcia \qquad Séverin Lemaignan \qquad Alexis Jacq \qquad Pierre Dillenbourg\\Computer-Human Interaction in Learning and Instruction Laboratory (CHILI)\\École Polytechnique Fédérale de Lausanne (EPFL)}
@@
 Successfully sustaining face-to-face interaction between humans and robots
-mandates fine perception of each other attentional focus. Along this line, we
-present a novel technique to assess the visual focus of attention (VFoA) based
-on accurate, real-time head pose estimation with monocular camera. The software
-is fully open-source and can readily be adopted on a wide range of social
-robots. We propose an in-depth evaluation of the technique, using manual video
-annotations as ground truth, and show that our system reaches high level of
-accuracy in a real-world face-to-face interaction scenario. We finally suggest how
-to derive a metric of engagement from the monitoring of the VFoA, allowing the
-robot to account and react to the user's changing engagement in real-time.
+mandates fine perception of each other attentional focus. This article presents
+a novel open-source technique to assess the visual focus of attention (VFoA)
+based on accurate, real-time head pose estimation with a single monocular
+camera.  Using manual video annotations as ground truth, we show that our system
+reaches satisfactory levels of accuracy in a real-world face-to-face interaction
+scenario.
+
+We then propose to introduce the concept of ``with-me-ness'', borrowed from the
+field of {\it Computer-Supported Collaborative Learning}, as an
+\emph{in-the-moment} measurement of the quality of interaction. We argue that
+this new metric, more specific than the broad concept of engagement, may be well
+suited for operational, real-time assessment of an on-going interaction.
 

### 2015-09-17 paper.tex
Rewritten introduction, re-organized the paper one more time
Starts to look better IMHO :-)

@@
 
-In order to build intelligent agents, it is necessary to equip them with a range
+Building capable social agents requires to equip them with a range
 of perceptual capabilities: while face and object tracking and
-recognition~\cite{Zhao:2003, Jafri:2014}, path planning~\cite{Galceran:2013},
-speech recognition~\cite{brick2007incremental} or task
-learning~\cite{calinon2007learning} are amongst the current major research
+recognition, path planning, speech recognition or task learning are some of the current major research
 directions, existing literature tends to evaluate each of these algorithms in
-their own metric space, without considering the overall interaction quality.
-Along this line, Anzalone~\etal~\cite{anzalone} have recently argued in favor of
-such a global evaluation, and propose to assess these algorithms in term of
-their capability to obtain the desired effect in a human-robot interaction
-context, and they correspondingly propose metrics build around the measurement
+their own metric space, without considering the interaction quality at a global
+level: Anzalone~\etal~\cite{anzalone} have recently argued in favor of
+such a global evaluation, and they propose to assess these algorithms in term of
+their capability to \emph{obtain the desired effect} in a human-robot interaction
+context. They correspondingly propose metrics build around the measurement
 of \emph{engagement} as indicator of the quality of the experience.
 
-Engagement describes a cognitive, affective (specifically intrinsic motivation),
-and behavioral state of interaction with a computer application that ``makes the
-user want to be there" \cite{OBrien:2010}. Engaging interactions were thought to
-involve intrinsic interest, interactivity, perceived control and choice,
-functionality, motivation and attention.
+``Engagement'', the cognitive, affective and behavioral state of interaction
+with a computer application that ``makes the user want to be
+there"~\cite{OBrien:2010}, has actively been studied in a diverse set of
+domains, and specifically in robotics, several variables and social signals have been proposed in the
+literature to quantify it. A recent review of these is in~\cite{ivaldi2015towards}.
 
-\paragraph{Motivation: Impact on Learning}
+%In a learning context, the evaluation of response times in relation with
+%correctness has been proposed~\cite{Beck} to provide estimate the user's 
+%engagement, with however large inter-subject and environment variability.
+%Biometric measurements have also been proposed, like the use the weight
+%distribution of the user on a pressure-sensitive
+%chair~\cite{Chipman07postureas}.
 
-Education and learning environments are interesting application fields for
-research on attention assessment, since sustaining children's engagement over
-time is a key variable to ensure effective learning (see for
-instance~\cite{Umbach}), and they indeed motivate this research. In fact, the
-use of robots in education comes with the potential to engage the child in
-meta-cognition through the learning by teaching paradigm, wherein a student
-takes the role of a teacher and experiences stronger educational benefits as a
-result (such as in \cite{Palinscar1984}).
+For instance, \cite{Castellano:2009} proposes to predict children's level of
+engagement by integrating in a Bayesian model non-verbal cues (gaze and smiles)
+with the current state of the interaction. While they report high level of
+accuracy, their approach requires post-hoc video annotations, and is not
+applicable to on-line engagement assessment. Similarly,
+Baxter~\etal~\cite{baxter2014tracking} posit that the measure of the direction
+and timing of gaze in child-robot interactions is a proxy for engagement and
+attribution of social agency. However, they also conduct these measures as
+post-hoc analyses.  \cite{peters2010investigating} model the user's interest and
+engagement with a virtual agent by tracking eye gaze and head direction.
+Similarly, \cite{ishii2011combining} estimates the user's engagement with a
+conversational agent based on the analysis of gaze patterns.
+In~\cite{Rich:2010}, a computational model based on the recognition of
+connection events such as directed gaze, mutual facial gaze is proposed. Not
+relying on gaze, \cite{Sanghvi:2011} focus only on the back or trunk posture as
+a determining factor for the assessment. A recent study~\cite{anzalone} with
+social robots in face-to-face scenarios proposed a set of metrics based on
+non-verbal cues but manifesting some possible limitations in long-term
+scenarios.
 
-While~\cite{zhao2012learning} recommends the development of a teachable agent
-that is actively, rather than passively, engaged in the learning by teaching
-interaction, the increased learning gain results attained
-in~\cite{okita2006observation} were done so even with the pupil being taught
-having been instructed to not contribute much information. This suggests that it
-may not be necessary, from a learning by teaching point of view, for the
-teachable agent to participate actively in the interaction to incite educational
-benefits. However, from an engagement point of view, this may not necessarily be
-true.
+The variety of these approaches reflects the fact that \emph{engagement} remains
+a broad concept, fairly ill-defined and thus difficult to operationalize.
+Therefore, instead of introducing ``yet another metric of engagement'', we
+propose in this paper to introduce the more specific concept of
+``\emph{with-me-ness}'': to what extend the human remains ``with me'', the
+robot, during our interaction.
 
-The goal of this research is therefore to \textbf{evaluate the quality of an
-interaction} using the \textbf{visual focus of attention as an engagement
-indicator}. We validate our approach with field experiments conducted in the
-educational context of the CoWriter project~\cite{Hood:2015}, where our
-technique, being based on a regular, low-end RGB camera, avoids the use of
-otherwise typically invasive technologies such as eyetrackers, that would
-negatively impact the natural of the interaction.
+We assess this level of ``with-me-ness'' by combining real-time estimation of the
+attentional focus of the human with the expected, {\it a priori} focus
+requested by the task, as decided by the robot.
 
-\subsection*{Attention and Engagement: Related Work}
+As such, this article is organized in three sections: we first present a novel
+method for on-line estimation of the focus of attention based on fast 6D head
+pose estimation. We then present an extensive validation of this approach based on
+a real-world field experiment with 6 children. We finally show how to compute in
+real-time ``with-me-ness'' to provide a new \emph{in-the-moment} measurement of
+the quality of interaction, and compare levels computed by the robot with a
+ground truth based on manual post-hoc annotations.
 
-\paragraph{Head Tracking and Visual Attention}
+%\paragraph{Motivation: Impact on Learning}
+%
+%Education and learning environments are interesting application fields for
+%research on attention assessment, since sustaining children's engagement over
+%time is a key variable to ensure effective learning (see for
+%instance~\cite{Umbach}), and they indeed motivate this research. In fact, the
+%use of robots in education comes with the potential to engage the child in
+%meta-cognition through the learning by teaching paradigm, wherein a student
+%takes the role of a teacher and experiences stronger educational benefits as a
+%result (such as in \cite{Palinscar1984}).
+%
+%While~\cite{zhao2012learning} recommends the development of a teachable agent
+%that is actively, rather than passively, engaged in the learning by teaching
+%interaction, the increased learning gain results attained
+%in~\cite{okita2006observation} were done so even with the pupil being taught
+%having been instructed to not contribute much information. This suggests that it
+%may not be necessary, from a learning by teaching point of view, for the
+%teachable agent to participate actively in the interaction to incite educational
+%benefits. However, from an engagement point of view, this may not necessarily be
+%true.
+%
+%The goal of this research is therefore to \textbf{evaluate the quality of an
+%interaction} using the \textbf{visual focus of attention as an engagement
+%indicator}. We validate our approach with field experiments conducted in the
+%educational context of the CoWriter project~\cite{Hood:2015}, where our
+%technique, being based on a regular, low-end RGB camera, avoids the use of
+%otherwise typically invasive technologies such as eyetrackers, that would
+%negatively impact the natural of the interaction.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Visual Attention Assessment}
+
+\subsection{Related Work}
 
@@
 agreement between the robot observation and manual post-hoc annotations.
-
-\paragraph{Engagement}
-
-The notion of engagement has actively been studied in a diverse set of domains,
-and several variables and social signals have been proposed in the literature to
-quantify and study it in social interactions.
-
-The evaluation of response times to provide an estimation of the user's
-engagement based on the correctness has been proposed in a learning context
-\cite{Beck} before. However, several aspects like the inter-subject variability
-or the environment complexity make it hardly applicable in a child-robot
-interaction. Additionally, other techniques \cite{Chipman07postureas} like the
-use of the weight distribution using a pressure sensitive chair has reported
-results. But it is in the computer vision field where major contributions have
-been done.
-
-In~\cite{peters2010investigating} approach, a model of user's interest and
-engagement with a virtual agent displaying shared attention behavior, by using
-eye gaze and head direction information is proposed. Similarly, in
-\cite{nakano2010estimating} an approach to estimate the user's conversational
-engagement with a conversational agent based on analysis of gaze patterns was
-implemented. Other significant work \cite{Castellano:2009} focus on feature
-evaluation related to the user's non-verbal behavior, the task and the
-companion's affective reactions are identified to predict the children's level
-of engagement using a Bayesian model.
-
-Some few studies have proposed approaches for the detection of engagement with a
-robot. In \cite{Rich:2010} a computational model based on the recognition of
-connection events such as directed gaze, mutual facial gaze is proposed. In this
-context, some other \cite{Sanghvi:2011} focus only on the back or trunk posture
-as a determined factor for the assessment. A recent study \cite{anzalone} with
-social robots in face to face scenarios proposed a set of metrics based on
-non-verbal cues but manifesting some possible limitations in long-term
-scenarios.
-
-In~\cite{baxter2014tracking}, Baxter~\etal posit that the measure of the
-direction and timing of gaze in child-robot interactions is a proxy for
-engagement and attribution of social agency, However, they conduct these
-measures as post-hoc analyses. This work suggests a path forward for real-time
-assessment of engagement.
-
-In this research, we address the problem of tracking the children's engagement
-based on their VFoA and current performed task during a long term child-robot
-interaction in a learning context, first. And second, we provide qualitative and
-quantitative evaluation of the outcome provided using a first person RGB sensor
-approach. Moreover, this monitoring activity has been supporting our research in
-child-robot interaction for learning and instruction purposes.
-
-\subsection*{Overview of the Article}
-
-\paragraph{Real-Time Assessment of Visual Focus of Attention}
-
-The participants' VFoA acquisition is the only dependent variable. This includes
-their gaze times and patterns of looking or not looking to the targets. This
-data extracted from an RGB camera was also coded post-hoc to quantify some of
-these measurements.
-
-We predict that there will be no generalizable pattern of participant's VFoA
-behavior for different scenarios such as multiple subjects, necessitating in
-situ VFoA estimation capabilities to enable robots to adapt in the specific
-scenarios. However, we believe that off-line pattern identification can lead to
-a better understanding of the children progression as well as engagement level
-assessment in face-to-face long term interactions.\fixme{I did not understand
-this paragraph}
-
-\fixme{Underline difference between field of attention and focus of attention}
-
-\fixme{State the hypothesis: head-pose estimation provides a good-enough
-approximation of the gaze direction, and in turn of the field of attention, and
-in turn of the focus of attention}
-
-\paragraph{Experimental Validation of the Approach}
-
-\fixme{TBD}
-
-\paragraph{From Attention to Engagement}
-
-\fixme{TBD}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Visual Attention Assessment}
 
@@
 
+\paragraph{Real-Time Assessment of Visual Focus of Attention}
+
+The participants' VFoA acquisition is the only dependent variable. This includes
+their gaze times and patterns of looking or not looking to the targets. This
+data extracted from an RGB camera was also coded post-hoc to quantify some of
+these measurements.
+
+We predict that there will be no generalizable pattern of participant's VFoA
+behavior for different scenarios such as multiple subjects, necessitating in
+situ VFoA estimation capabilities to enable robots to adapt in the specific
+scenarios. However, we believe that off-line pattern identification can lead to
+a better understanding of the children progression as well as engagement level
+assessment in face-to-face long term interactions.\fixme{I did not understand
+this paragraph}
+
+\fixme{Underline difference between field of attention and focus of attention}
+
+\fixme{State the hypothesis: head-pose estimation provides a good-enough
+approximation of the gaze direction, and in turn of the field of attention, and
+in turn of the focus of attention}
+
 \subsection{Experimental Procedure}
@@
 	\centering
-	\caption{Tracking accuracy in face-to-face scenario}
+    \caption{Tracking accuracy in face-to-face scenario\fixme{std dev needed!}}
 	\begin{tabular}{|c|c|c|c|c|c|c|c|} \hline
@@
     % Maybe we should add the expected VFoA
-\end{figure*}
-
-\begin{figure*}
-    \centering
-    \includegraphics[width=1.81\columnwidth]{realCaptured}
-    \caption{\small Real VFoA of subject 4 accoding to two observer's
-    assessment. In green, the obtained VFoA using our aproach and in cyan, the
-    mathing between them. Finally, the correspondent numbers with figure \ref{fig:bitmap}.}
-
-    \label{fig:realCaptured}
 \end{figure*}
@@
 
-\section{Discussion and Conclusion}
-\subsection{From Attention to Engagement Assessment}
+\section{With-me-ness}
+
+\begin{figure*}
+    \centering
+    \includegraphics[width=1.81\columnwidth]{realCaptured}
+    \caption{\small Real VFoA of subject 4 accoding to two observer's
+    assessment. In green, the obtained VFoA using our aproach and in cyan, the
+    mathing between them. Finally, the correspondent numbers with figure \ref{fig:bitmap}.}
+
+    \label{fig:realCaptured}
+\end{figure*}
 
@@
 
-\subsection{Conclusion}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Discussion and Conclusion}
 

### 2015-09-17 paper.tex
Updated experimental setting figure
Formally introduce with-me-ness

@@
 \documentclass{sig-alternate}
+
+\usepackage{algorithm}
+\usepackage{algpseudocode}
 
@@
     \centering
-    \includegraphics[width=0.7\columnwidth]{drawSetup}
-    \caption{\small \textbf{Experimental setting} Graphical representation of the set-up and the possible children focuses of attention in the scene, in red.}
+    \includegraphics[width=0.8\columnwidth]{experimental_setup}
+    \caption{\small \textbf{Experimental setting}. The areas of interest (AOI)
+    considered of potential focuses of attention are circled in red.}
     \label{drawSetup}
@@
 
+The concept of \emph{with-me-ness} has been introduced in the field of
+\emph{Computer Supported Collaborative Learning} (CSCL) by Sharma~\etal
+in~\cite{sharma2014me}. Sharma~\etal introduce this concept in an attempt to
+answer a recurrent teacher's question: {\it ``how much are the students with
+me?''}. They distinguish what they call \emph{perceptual with-me-ness} (the
+student follows what the teacher refers to with deictic gestures) from
+\emph{conceptual with-me-ness} (the student follows what the teacher refers to
+verbally), and they show in an eye-tracking study that \emph{conceptual
+with-me-ness} in particular correlates with better learning performance. This
+also relates to the concept of gaze cross-recurrence that has been shown to
+reflect the quality of the interaction~\cite{jermann2012effects} in
+collaborative learning tasks.
+
+Sharma~\etal simply define \emph{conceptual with-me-ness} as the normalized
+percentage of time during which the student's gaze overlapped the areas of
+teaching slides currently referred to by the teacher.
+In order to apply it to human-robot interactions, we propose to extend this
+concept, and to define \emph{conceptual with-me-ness} as the normalized
+ratio of time that the human interactant focuses its attention on the
+attentional target expected by the robot for the current task (or sub-task).
+
+\begin{algorithm}[h!]
+    \centering
+
+    \begin{algorithmic}[1]
+    \Procedure{Compute With-me-ness} {}
+    \State $d_w, d_e \gets 0$
+    \State $t \gets t_{start}$
+    \Repeat
+    \If{$task(t) \neq$ nil \textbf{and} \par
+    \hskip\algorithmicindent$F(task(t)) \neq \varnothing$ \textbf{and} \par
+    \hskip\algorithmicindent $f(t) \neq$ nil } \label{algline:skiplosttrack}
+    \If{$f(t) \in F(task(t))$}
+            \State $d_w \gets d_w + 1$
+        \EndIf
+        \State $d_e \gets d_e + 1$
+    \EndIf
+    \State $t \gets t + 1$
+    \Until{$t = t_{end}$}
+
+    \State $\mathcal{W}_{[start, end]} \gets \frac{d_w}{d_e}$
+    \State \Return $\mathcal{W}_{[start, end]}$
+    \EndProcedure
+
+    \end{algorithmic}
+
+    \caption{\textbf{Computation of \emph{with-me-ness}}. $d_w$ stands for the duration
+        the human is \emph{with} the robot, $d_e$ stands for the duration where
+        the human would be \emph{expected to be with} the robot, $task(t)$
+        represents the task performed by the robot at time $t$ (possibly none),
+        $F(task)$ represents the (possibly empty) set of expected attentional
+        targets associated to task $task$, $f(t)$ represents the actual focus of
+        attention measured at time $t$. $\mathcal{W}_{[start, end]}$ represents
+        the level of \emph{with-me-ness} from $t_{start}$ to $t_{end}$.}
+    \label{alg:with-me-ness}
+\end{algorithm}
+
+Algorithm~\ref{alg:with-me-ness} provides a formal way of computing the level of
+with-me-ness $\mathcal{W}$ between two time points $[t_{start}, t_{end}]$. A
+notable difference with the original definition by Sharma~\etal is that, at a
+given time $t$, the task $task(t)$ performed by the robot may elicit more than
+one attentional target; thus, at a given time, more than one location can be
+regarded as possible \emph{expected} focuses of attention for the human. For
+example, a robot which is writing, could typically elicit gazes to its hand as
+well as to its head. A human looking at either of these locations would be
+considered to be \emph{with} the robot in terms of interaction. Also notable, we
+exclude from the computation of $\mathcal{W}$ all of the periods of time where
+the user's focus of attention can not be estimated (Algorithm~\ref{alg:with-me-ness},
+line~\ref{algline:skiplosttrack}, third condition), typically because the user's
+face is not visible at those times.
+
 \begin{figure*}
@@
     \caption{\small Real VFoA of subject 4 accoding to two observer's
-    assessment. In green, the obtained VFoA using our aproach and in cyan, the
-    mathing between them. Finally, the correspondent numbers with figure \ref{fig:bitmap}.}
+    assessment. In green, the obtained VFoA using our approach and in cyan, the
+    matching between them. Finally, the correspondent numbers with figure \ref{fig:bitmap}.}
 

### 2015-09-17 paper.tex
Minor change to make the 'matches' picture more readble

@@
     \caption{\small \textbf{Experimental setting}. The areas of interest (AOI)
-    considered of potential focuses of attention are circled in red.}
+    of potential focuses of attention are circled in red.}
     \label{drawSetup}
@@
     \caption{\small \textbf{Comparison of computed focus of attention \vs ground
-        truth} during a 20+ min face-to-face child-robot interaction.
-        In orange (top line), the focus of attention as computed by the robot;
-        in blue (bottom line), the focus of attention as manually annotated. The
-    bottom graph shows agreements.}
+        truth} during a 20+ min face-to-face child-robot interaction (subject 4
+        in table~\ref{tab:results}).
+        In blue (top line), the focus of attention as computed by the robot;
+        in orange (bottom line), the focus of attention as manually annotated
+        (ground truth). The bottom graph shows agreements.}
 

### 2015-09-17 paper.tex
Split the results tables into attentions results and with-me-ness results
+ added a placeholder for the with-me-ness figure

@@
 	\centering
-    \caption{Tracking accuracy in face-to-face scenario\fixme{std dev needed!}}
-	\begin{tabular}{|c|c|c|c|c|c|c|c|} \hline
-	\small Subject & 1 & 2 & 3 & 4 & 5 & 6 & Avg\\ \hline
-	\small Track(\%) & \small88.2 & \small83.5 & \small90.5 & \small83.1 & \small87.9 & \small85.0 & \small86.4 \\ \hline
-	\begin{tabular}{@{}c@{}}\small Real vs \\ \small expected(\%)\end{tabular} & \small92.6 & \small88.8  & \small95.8 & \small93.1 & \small96.5 & \small89.6 & \small92.7 \\ \hline
-	\begin{tabular}{@{}c@{}}\small Real vs \\ \small captured(\%)\end{tabular} & \small61.1 & \small85.5 & \small80.8 & \small62.2 & \small63.4 & \small83.9 & \small72.8 \\ \hline
-	\begin{tabular}{@{}c@{}}\small Cosine \\ \small affinity \end{tabular} & \small0.73 & \small0.82 & \small0.87 & \small0.73 & \small0.77 & \small0.89 & \small0.8 \\ \hline
+    \caption{\textbf{Results of attention tracking accuracy}. \emph{Head pose} is the percentage of total time during which the head pose of the participant was successfully estimated;}
+	\begin{tabular}{|c|c|c|c|c|c|c|c|c|} \hline
+        \small \textbf{Subject} & 1 & 2 & 3 & 4 & 5 & 6 & Avg. & Std dev\\ \hline
+        \small Head pose (\%) & \small88.2 & \small83.5 & \small90.5 & \small83.1 & \small87.9 & \small85.0 & \small86.4 & \\ \hline
+            \begin{tabular}{@{}c@{}}\small Groundtruth \vs \\ \small robot (\%)\end{tabular} & \small61.1 & \small85.5 & \small80.8 & \small62.2 & \small63.4 & \small83.9 & \small72.8 & \\ \hline
+        \begin{tabular}{@{}c@{}}\small Agreement ($\alpha$)\end{tabular} &  & & & & & & & \\ \hline
+            \begin{tabular}{@{}c@{}}\small Cosine \\ \small affinity \end{tabular} & \small0.73 & \small0.82 & \small0.87 & \small0.73 & \small0.77 & \small0.89 & \small0.8 & \\ \hline
 	\end{tabular}
@@
 
+\begin{table}[h!]
+	\centering
+    \caption{\textbf{Results of attention tracking accuracy}. \emph{Head pose} is the percentage of total time during which the head pose of the participant was successfully estimated;}
+	\begin{tabular}{|c|c|c|c|c|c|c|c|c|} \hline
+        \small \textbf{Subject} & 1 & 2 & 3 & 4 & 5 & 6 & Avg. & Std dev\\ \hline
+        \begin{tabular}{@{}c@{}}\small Groundtruth vs \\ \small expected(\%)\end{tabular} & \small92.6 & \small88.8  & \small95.8 & \small93.1 & \small96.5 & \small89.6 & \small92.7 & \\ \hline
+        \begin{tabular}{@{}c@{}}\small Groundtruth \\ \small Agreement ($\alpha$)\end{tabular} &  & & & & & & & \\ \hline
+            \begin{tabular}{@{}c@{}}\small Robot vs \\ \small expected(\%)\end{tabular} &  & & & & & & & \\ \hline
+        \begin{tabular}{@{}c@{}}\small Robot \\ \small Agreement ($\alpha$)\end{tabular} &  & & & & & & & \\ \hline
+	\end{tabular}
+	\label{tab:results}
+\end{table}
+
 \begin{figure*}
     \centering
-    \includegraphics[width=1.81\columnwidth]{realCaptured}
-    \caption{\small Real VFoA of subject 4 accoding to two observer's
-    assessment. In green, the obtained VFoA using our approach and in cyan, the
-    matching between them. Finally, the correspondent numbers with figure \ref{fig:bitmap}.}
+    \includegraphics[width=\linewidth]{with-me-ness}
+    \caption{\small \textbf{With-me-ness}. Evolution of the level of
+    \emph{with-me-ness} over a 20min long child-robot interaction. The diagram
+    is identical to Figure~\ref{fig:realExpected} with the \emph{expected}
+    attentional targets added in green. The bottom curves represent the levels
+    of with-me-ness (since the beginning of the interaction), computed following
+    Algorithm~\ref{alg:with-me-ness}. The blue line is with-me-ness as estimated by
+    the robot, the orange line is with-me-ness computed from manually annotated
+    focuses of attention. The ``holes'' in the with-me-ness computed by the
+    robot correspond to period of time where the child's face is not detected.}
 
-    \label{fig:realCaptured}
+    \label{fig:with-me-ness}
 \end{figure*}

### 2015-09-18 paper.tex
Cleaned up the tables

@@
 \usepackage{algpseudocode}
+
+\usepackage{booktabs}
 
@@
 
-\begin{table}[h!]
+\begin{table*}[h!]
     \centering
-    \caption{\textbf{Results of attention tracking accuracy}. \emph{Head pose} is the percentage of total time during which the head pose of the participant was successfully estimated;}
-	\begin{tabular}{|c|c|c|c|c|c|c|c|c|} \hline
-        \small \textbf{Subject} & 1 & 2 & 3 & 4 & 5 & 6 & Avg. & Std dev\\ \hline
-        \small Head pose (\%) & \small88.2 & \small83.5 & \small90.5 & \small83.1 & \small87.9 & \small85.0 & \small86.4 & \\ \hline
-            \begin{tabular}{@{}c@{}}\small Groundtruth \vs \\ \small robot (\%)\end{tabular} & \small61.1 & \small85.5 & \small80.8 & \small62.2 & \small63.4 & \small83.9 & \small72.8 & \\ \hline
-        \begin{tabular}{@{}c@{}}\small Agreement ($\alpha$)\end{tabular} &  & & & & & & & \\ \hline
-            \begin{tabular}{@{}c@{}}\small Cosine \\ \small affinity \end{tabular} & \small0.73 & \small0.82 & \small0.87 & \small0.73 & \small0.77 & \small0.89 & \small0.8 & \\ \hline
+    \caption{\textbf{Results of attention tracking accuracy}. \emph{Head pose
+    tracking} is the percentage of total time of successful detection of the
+    head pose; \emph{Ground-truth/robot match} is the percentage of matching time
+    between manually annotated focus of attention (ground-truth) and robot's
+    computed focus of attention.}
+
+    \begin{tabular}{p{5.5cm}cccccccc}
+        \toprule
+        {\bf Subject} & 1 & 2 & 3 & 4 & 5 & 6 & {\bf Avg.} & {\it SD} \\
+        \midrule
+        {\bf Head pose tracking} (\%) & 88.2 & 83.5 & 90.5 & 83.1 & 87.9 & 85.0 & {\bf 86.4} & {\it 3.0} \\ 
+        \midrule
+        {\bf Ground-truth/robot match} (\%) & 61.1 & 85.5 & 80.8 & 62.2 & 63.4 & 83.9 & {\bf 72.8} & {\it 11.7}\\
+        {\bf Cronbach's $\alpha$} &  & & & & & & & \\
+        {\bf Cosine affinity} & 0.73 & 0.82 & 0.87 & 0.73 & 0.77 & 0.89 & {\bf 0.8} & {\it 0.07} \\
+        \bottomrule
     \end{tabular}
     \label{tab:results}
-\end{table}
+\end{table*}
 
@@
     \centering
-    \caption{\textbf{Results of attention tracking accuracy}. \emph{Head pose} is the percentage of total time during which the head pose of the participant was successfully estimated;}
-	\begin{tabular}{|c|c|c|c|c|c|c|c|c|} \hline
-        \small \textbf{Subject} & 1 & 2 & 3 & 4 & 5 & 6 & Avg. & Std dev\\ \hline
-        \begin{tabular}{@{}c@{}}\small Groundtruth vs \\ \small expected(\%)\end{tabular} & \small92.6 & \small88.8  & \small95.8 & \small93.1 & \small96.5 & \small89.6 & \small92.7 & \\ \hline
-        \begin{tabular}{@{}c@{}}\small Groundtruth \\ \small Agreement ($\alpha$)\end{tabular} &  & & & & & & & \\ \hline
-            \begin{tabular}{@{}c@{}}\small Robot vs \\ \small expected(\%)\end{tabular} &  & & & & & & & \\ \hline
-        \begin{tabular}{@{}c@{}}\small Robot \\ \small Agreement ($\alpha$)\end{tabular} &  & & & & & & & \\ \hline
+    \caption{\textbf{Levels of with-me-ness}. For each subject, the with-me-ness level is reported over the whole interaction.}
+    \begin{tabular}{p{2cm}cccccccc}
+        \toprule
+        \textbf{Subject} & 1 & 2 & 3 & 4 & 5 & 6 & Avg. & SD \\
+        \midrule
+        %Ground-truth \vs expected (\%) & 92.6 & 88.8  & 95.8 & 93.1 & 96.5 & 89.6 & 92.7 & \\ \hline
+        $\mathcal{W}_{groundtruth}$ &  & & & & & & & \\ 
+        {\it Cronbach's $\alpha$} &  & & & & & & & \\
+        \midrule
+        $\mathcal{W}_{robot}$ &  & & & & & & & \\
+        {\it Cronbach's $\alpha$} &  & & & & & & & \\
+        \bottomrule
     \end{tabular}
-	\label{tab:results}
+    \label{tab:results-with-me-ness}
 \end{table}

### 2015-09-18 paper.tex
Done with a first pass on section 2

@@
 
-Besides being fast, the face alignment algorithm has been found to perform well,
-including in a range of difficult situations (occlusions, head angle) as
-encountered in field experiments.  Figure~\ref{head_pose_real_world} shows a
-selection of such difficult scenes on one child.
+Besides being fast, the face alignment algorithm has been found to perform well
+in terms of robustness, including in a range of difficult situations
+(occlusions, head angle) as encountered in field experiments.
+Figure~\ref{head_pose_real_world} shows a selection of such difficult scenes with
+one child.
 
@@
         (three are depicted: the sellion -- the nasal depression --, the left
-        tragion and the menton).}
+        tragion and the menton). The 3D origin of the head is set at the sellion.}
 
@@
 
-\subsection{Field of Attention}
+\subsection{Field \& Focus of Attention}
 
@@
     pose intersect with this cone are considered \emph{in focus}.}
-    \label{rviz}
+    \label{fig:vfoa}
 \end{figure}
 
-The third step is to generate a representation of the field of view whose origin
-and orientation are the ones defined by the face detection \textit{tf} frame.
+We model the field of attention as the central region of the field of view.
+The field of view itself is approximated to a cone spanned from the nasal
+depression (sellion) of the human face.
 
-The breadth of the human field of attention has been discussed in the
-literature.
-Holmqvist in~\cite{holmqvist2011eye} specifies that the visual human
-range\fixme{field of view or field of attention?} is $
-\pm  40\degree $ in the horizontal and $ \pm 25\degree $ in the vertical. In
-\cite{walker1980clinical} Spector provides a more detailed specification for
-each eye splitting the vertical range into $ 60\degree $ the upper region and $
-75\degree $ the lower one. Moreover, the horizontal range gets separated in $
-60\degree $ inwards (towards the nose) and $ 95\degree $ outwards. In robotics,
-Sisbot\etal~\cite{sisbot2011situation} use a cone of 30$\degree$. In the
-present implementation, the first approach explained has been chosen
-representing the field of view using a cone with such dimensions\fixme{which
-one? 40deg? 25deg?}.
+The width of the human field of attention has been discussed in the literature.
+Holmqvist~\cite{holmqvist2011eye} specifies that the visual human
+range\fixme{field of view or field of attention?} is $ \pm  40\degree $ in the
+horizontal and $ \pm 25\degree $ in the vertical.
+In~\cite{walker1980clinical}, Spector provides a more detailed specification
+for each eye splitting the vertical range into $ 60\degree $ the upper region
+and $ 75\degree $ the lower one. Moreover, the horizontal range gets separated
+in $ 60\degree $ inwards (towards the nose) and $ 95\degree $ outwards.
+Previous work on visual perspective taking for social robotics
+application~\cite{sisbot2011situation} propose to use a cone of 30$\degree$ of
+aperture. In the present implementation, the first approach explained has been
+chosen representing the field of view using a cone with such
+dimensions\fixme{which one? 40deg? 25deg?}.
 
-Assuming that the \textit{tf} frames representing the robotic agent and the VFoA
-candidates are static, need to be monitored. This task consists on evaluating
-the intersection between the \textit{tf} frames and the field of view: If one
-focus of attention is within the region defined by the field of view, we assume
-the subject is looking at it. In order to solve such operation mathematically it
-is necessary to compute the transformation matrix between the coordinate frames
-to be able to locate the monitored frames \textit{A} in the subject's coordinate
-system \textit{B} (see equation \ref{eq:transform}).
+We then approximate the visual focus of attention (VFoA) of the human to the
+objects which lie inside the field of attention (Figure~\ref{fig:vfoa}). At a
+given time, more than one object can therefore be \emph{in focus}.
 
-\begin{equation}
-v' = B(A^{-1})v
-\label{eq:transform}
-\end{equation}
+Our implementation has two limitations: objects are approximated to their origin
+(\ie objects are either in or out of the field of attention), and we do not
+check actual visibility: one object could be hidden by another, it would still
+be considered as in focus. We did not address these limitations since our
+experimental setup (involving relatively small objects with no occlusions) did
+not require it. Techniques for a more complete assessment of the visual
+perspective can be found in~\cite{sisbot2011situation} for instance.
 
-where \textit{v} is a point in \textit{A} that becomes v' in coordinate system
-\textit{B}. However, in practice there are, at least, three different
-\textit{tf} frames involved in the transformation (subject-camera, camera-base,
-base-focus). Following the transformation result, the distance of the monitored
-frame with respect to the x-axis face \textit{tf} frame, $ d_{x_{axis}} $,
-representing the subject's face can be computed as simple as in equation
-\ref{eq:pitagoras}. Always considering the x-axis as the one defining the face's
-orientation and thus field of view's main axis.
+Within these limitations, computing if object $A(x_A,y_A,z_A)$ is in the field
+of attention of the human requires first to transform the coordinates
+$A(X_A,Y_A,Z_A)$ into the frame of the face, and then to verify the following
+inequality (with $fov$ the aperture, and assuming that the main axis of the
+field of attention is aligned on the face's $\vec{x}$):
 
-\begin{equation}
-d_{x_{axis}} = \sqrt{t_y^2 + t_z^2}
-\label{eq:pitagoras}
-\end{equation}
+\begin{align}
+    \sqrt{Y_A^2 + Z_A^2} < tan\left(\frac{fov}{2}\right) \cdot X_A
+\label{eq:fov}
+\end{align}
 
-It is necessary to compare the distance acquired $ d_{x_{axis}} $ with the
-radius of the field of view at the target \textit{tf} frame x-coordinate $
-r_{fov} $ as shown in equation \ref{eq:fovtf}.
-
-\begin{equation}
-r_{fov} = tan\left(\frac{fov}{2}\right) \cdot t_x
-\label{eq:fovtf}
-\end{equation}
-
-where \textit{fov} is the aperture angle of the field of view. Finally, if $
-d_{x_{axis}}<r_{fov} $, the target is within the subject's field of view.
-
-% Schema to clarify this?
-
-\paragraph{Acquiring the Focus of Attention}
-
-Explain that: other modules have to provide the sitation assessment (what
-is where?) in terms of published frames, and a last module compute the
-intersection of the VFoA and these frames\fixme{TBD}
+Our approach assumes that the poses of the objects of interest are available to
+the system: as described in section~\ref{sec:system}, our implementation relies
+on the ROS {\sc tf} framework to manage and make available to all software
+modules the list of poses of existing objects (represented as {\it frames}), and
+dedicated perception modules are in charged of publishing up-to-date
+informations regarding the location of the points of interest (the so-called
+\emph{situation assessment}). Due to the nature of the experiment, most of the
+points of interest considered for the experimental validation presented
+hereafter are actually taken as static with respect to the robot.
 
@@
 \subsection{System Implementation}
+\label{sec:system}
 

### 2015-09-18 paper.tex
Results in table are actually with-me-ness values
First pass on section 3-Experimental validation, including a new, tikz-based system overview

@@
 \documentclass{sig-alternate}
+
+\usepackage{hyperref}
 
@@
 \usepackage[draft,nomargin,footnote]{fixme}
+
+\usepackage[usenames,dvipsnames]{xcolor}
+\usepackage{tikz}
+\usetikzlibrary{positioning, calc}
 
@@
 \section{Visual Attention Assessment}
+\label{sec:vfoa}
 
@@
 attention. The assumption that such an approximation of the field of attention
-allows to derive the actual focus of attention is the core hypothesis that we
-ought to validate.
-
-This section presents such an experimental validation, set in the context of a
-real child-robot interaction scenario where the robot tries to engage a child in
-handwriting tasks using learning by teaching paradigm.
-
-\paragraph{Real-Time Assessment of Visual Focus of Attention}
-
-The participants' VFoA acquisition is the only dependent variable. This includes
-their gaze times and patterns of looking or not looking to the targets. This
-data extracted from an RGB camera was also coded post-hoc to quantify some of
-these measurements.
-
-We predict that there will be no generalizable pattern of participant's VFoA
-behavior for different scenarios such as multiple subjects, necessitating in
-situ VFoA estimation capabilities to enable robots to adapt in the specific
-scenarios. However, we believe that off-line pattern identification can lead to
-a better understanding of the children progression as well as engagement level
-assessment in face-to-face long term interactions.\fixme{I did not understand
-this paragraph}
-
-\fixme{Underline difference between field of attention and focus of attention}
-
-\fixme{State the hypothesis: head-pose estimation provides a good-enough
-approximation of the gaze direction, and in turn of the field of attention, and
-in turn of the focus of attention}
+allows to derive the actual focus of attention forms the hypothesis that we
+want to validate experimentally. The experiment involves child-robot
+interactions in the context of handwriting remediation.
+This section details the experimental procedure and presents our results.
 
@@
 
-The target of this study were 6 children with ages compressed between 5 and 6
-years old. The experiments were developed individually and consisted on a
-writing activity by turns and one story telling performed by the robot at a
-given time of the interaction. All this performed in a over all 18 minutes
-interaction average time.
+The experiment, part of the CoWriter project~\cite{Hood:2015}, involves a robot
+which tries to engage a child in handwriting tasks using a \emph{learning by teaching}
+paradigm (\ie the child is the teacher, and he/she attempts to improve the
+robot's handwriting). A tactile tablet is used as writing support.
 
-Figure~\ref{fig:realSetup} illustrates our general experimental setup: a
-face-to-face child-robot interaction with an (autonomous) Aldebran's {\sc nao}
-robot.
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.8\columnwidth]{experimental_setup}
+    \caption{\small \textbf{Experimental setup}: face-to-face interaction with a {\sc
+            nao} robot. The robot writes on the tactile tablet, the child then
+            corrects the robot by directly overwriting its letters on the tablet
+            with a stylus. The facilitator remains next to the child to guide the work. 
+            The secondary tablet allows the child to tell the robot what to
+            write. The areas of interest -- corresponding to potential target of
+            attention -- are circled in red with their name.}
+    \label{fig:setup}
+\end{figure}
+
+Figures~\ref{fig:setup} and~\ref{fig:realSetup} illustrate the experimental setup: a face-to-face
+child-robot interaction with an (autonomous) Aldebran's {\sc nao} robot, in
+presence of a facilitator (one of the researchers).
 
@@
     \includegraphics[width=1\columnwidth]{realSetup}
-    \caption{\small Our experimental setup: face-to-face interaction with a {\sc
-            nao} robot.  The robot writes on the tactile tablet, the child then
-            corrects the robot by directly overwriting its letters on the tablet
-            with a stylus. An adult (either a therapist or an experimenter,
-            depending on the studies), remains next to the child to guide the work. 
-            A second tablet allows to choose demonstrations and the camera captures the subject's face.}
+    \caption{\small Picture of the interaction with one of the children.}
     \label{fig:realSetup}
@@
 
-A tactile tablet (with a custom application) is used for both the robot and the
-child to write: during a typical round, the child requests the robot to write
-something (a single letter, a number or a full word), and push the tablet
-towards the robot, the robot writes on the tablet by gesturing the writing (but
-without actually physically touching the tablet), the child then pull back the
-tablet, corrects the robot's attempt by writing him/herself on top or next to
-the robot's writing, and ``send'' his/her demonstration to the robot by pressing
-a small button on the tablet. The robot ``learns'' from this demonstration and
-tries again.
+The dependent variable is the measurement of the participants' VFoA, assessed in
+terms of what are the attentional targets of the child over time. The face of the child
+is acquired through a fixed webcam (Logitech {\sf c920}), placed on the table
+(see Figure~\ref{fig:setup}), and the attentional targets are then computed as
+presented in section~\ref{sec:vfoa}.
 
-Since the child is assumed to take on the role of the teacher, we had to ensure
-(s)he would be able to manage by him/herself the turn-taking and the overall
-progression of the activity (moving to the next letter or word). In our design,
-the turn-taking relies on the robot prompting for feedback once it is done with
-its writing (simple sentences like ``What do you think?''), and pressing on a
-small robot icon on the tablet once the child has finished correcting. In our
-experiments, both were easy to grasp for children.
+%An RGB camera has been used to acquire images of 640x480 pixels at 25 fps. The
+%camera was located at 5 cm from the center of Nao's feet and its holder was tied
+%to a wood base for stability purposes. Thus, the camera's objective was 9 cm
+%above and $ 40\degree $ inclination with respect to the surface of the table to
+%maximize the detection during the writing time. However, Nao's camera could have
+%been used, always considering the \textit{tf} frame matrix update according to
+%the head orientation.
 
-Implementing such a system raises several challenges that are discussed in
-detail in \cite{Hood:2015}.
+The subjects were typically located 50 cm away from the robot with the primary
+(writing) tablet in front and the secondary one 30 cm to the left of the first
+one.  The facilitator was located about 60 cm to the left of the subject.
+Finally, two observers were located further away from the interaction field.
+Figure~\ref{fig:setup} indicates accordingly the location of main areas of
+interest (the two tablets, the robot, the facilitator and the observers).
 
-An RGB camera has been used to acquire images of 640x480 pixels at 25 fps. The
-camera was located at 5 cm from the center of Nao's feet and its holder was tied
-to a wood base for stability purposes. Thus, the camera's objective was 9 cm
-above and $ 40\degree $ inclination with respect to the surface of the table to
-maximize the detection during the writing time. However, Nao's camera could have
-been used, always considering the \textit{tf} frame matrix update according to
-the head orientation.
+\subsection{Experimental Procedure}
 
-The subjects were typically located 50 cm away from the robot with the first
-tablet in front and the second one 30 cm to the left of the first one. In the
-same way, the experimenter was located around 60 cm to the left of the subject.
-Finally, two observers were located far away from the interaction field to
-manually assess the state of the interaction.
+Six children (ages 5 to 6\fixme{boys? girls?}) were enrolled for this study.
+The study took place at school, in a separate room (the computer
+lab). The participants were chosen by the teacher, and would come one after the
+other for a face-to-face interaction with a {\sc Nao} robot (average duration:
+18 min, $SD:$ \fixme{stddev?}).
 
-Figure \ref{drawSetup} allows us to intuitively identify several focuses of
-attention along the interaction such as the two tablets, the robot, the
-experimenter and the observers located in the diagonal.
+The experiment is organized in rounds of writing: during a typical round, the
+child requests the robot to write something (a single letter, a number or a full
+word), and pushes a tactile tablet (equipped with a custom application that we
+developed) towards the robot; the robot writes on the tablet by mimicking the
+writing gesture (but without actually physically touching the tablet); the child
+then pull back the tablet, corrects the robot's attempt by writing him/herself
+on top or next to the robot's writing, and ``send'' his/her demonstration to the
+robot by pressing a small button on the tablet. The robot learns from this
+demonstration and tries again. The child continues the turn-taking until he/she
+decides to train the robot on another word. Details on the rationale and
+implementation of this study can be found in~\cite{Hood:2015}.
 
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.8\columnwidth]{experimental_setup}
-    \caption{\small \textbf{Experimental setting}. The areas of interest (AOI)
-    of potential focuses of attention are circled in red.}
-    \label{drawSetup}
-\end{figure}
+\fixme{how many rounds for each child?}
+
+Besides, once per interaction, the robot did interrupt the handwriting task to
+propose to tell a funny story. The story-telling lasted in average \fixme{how
+long?} min, and the turn-based hand-writing task continued afterwards. The
+intended purpose of the story-telling episode was to locally trigger a different
+set of gaze behaviors.
 
 \subsection{System Implementation}
+
 \label{sec:system}
@@
     \centering
-    \includegraphics[width=0.9\columnwidth]{system}
-    \caption{\small System overview showing the most relevant components involved in the estimation.}
-    \label{system}
+
+\resizebox{0.9\linewidth}{!}{%
+
+\begin{tikzpicture}[
+    >=latex,
+    node distance=2cm,
+    every edge/.style={draw, very thick},
+    redarrow/.style={fill=Red, draw=Red},
+    greenarrow/.style={fill=GreenYellow, draw=GreenYellow},
+    yellowarrow/.style={fill=BurntOrange, draw=BurntOrange},
+    cmpt/.style={draw, align=center, rounded corners, inner sep=5pt, font=\sf, fill=black!20},
+    label/.style={midway, align=left, font=\scriptsize\sf, fill=white, above,opacity=0,text opacity=1}]
+
+%    \node at (0,0)[cmpt, fill=Cyan!50, text width=2.7cm] (tactile) {tactile interaction \\ manager};
+
+    \node at (0,0) [cmpt, fill=Green!50] (tf) {tf};
+    \node [cmpt, fill=YellowOrange!50, left=of tf] (attention) {attention tracking};
+    \node [above=0.6cm of attention, text width=2.7cm,align=center] (input) {\tt /image \\ /camera\_info \\ dlib {\sf faces model} \\ {\sf 3D face template}};
+    \node [cmpt, right=of tf] (static) {static pose publishers};
+    \node [above=0.6cm of static, text width=2.7cm,align=center] (staticposes) {\sf locations of \\areas of interest};
+    \node [cmpt, above=0.6cm of tf] (pose) {robot state publisher};
+    \node [cmpt, fill=YellowOrange!50, below=0.6cm of tf] (focus) {focus estimation};
+    \node [below=0.6cm of focus] (output) {\tt /frames\_in\_focus};
+
+    \path (staticposes) edge [->] (static);
+    \path (static) edge [->] (tf);
+    \path (input) edge [->] (attention);
+    \path (attention) edge [->] (tf);
+    \path (pose) edge [->]  (tf);
+    \path (tf) edge [->]  (focus);
+    \path (focus) edge [->]  (output);
+
+\end{tikzpicture}
+}
+
+    \caption{\small ROS nodes involved in the VFoA estimation (yellow nodes were developed for this work).}
+    \label{fig:system}
 \end{figure}
 
-, to detect and extract the skeleton
-of the subject's face and then, use the obtained features to calculate the
-head pose orientation estimation (see figure \ref{system}). In this way, we can
-estimate what the subject was looking at and obtain an \textit{rviz} 3D scene
-representation as it is executed as a ROS (Robot Operating System) node that can
-be seen in figure \ref{rviz}.
+The experiment was carried with an Aldebaran's {\sc Nao} robot, using ROS as a
+middleware to build the attention estimation pipeline
+(Figure~\ref{fig:system}). Head pose estimation builds on the {\tt dlib} and
+OpenCV libraries, while pose transformations are handled by the {\sc tf}
+library. The implementation is open-source and available at
+\url{https://github.com/chili-epfl/attention-tracker}.
+
+Please refer to~\cite{Hood:2015} for the implementation of the hand-writing
+activity.
 
@@
 \subsubsection{From Face Detection to Focus of Attention} \label{fromFaceTo}
+(note that the focus of attention
+was separately hand-coded, {\it post-hoc}, from the video recording of the
+interaction).
 
@@
     \includegraphics[width=0.7\columnwidth]{bitmap}
-    \caption{\small Cosine similarity matrix between real and cap-tured VFoA
-    using a 10 sec. window size computation. The correspondent numbers are shown
+    \caption{\small Cosine similarity matrix between ground-truth and robot-measured VFoA
+    using a 10 sec. window size. The correspondent numbers are shown
     in figure \ref{fig:realCaptured}.}
@@
     \centering
-    \caption{\textbf{Levels of with-me-ness}. For each subject, the with-me-ness level is reported over the whole interaction.}
-    \begin{tabular}{p{2cm}cccccccc}
+    \caption{\textbf{Levels of with-me-ness}.
+For each subject, the with-me-ness level is reported over the whole interaction, either based on the ground-truth (\ie \emph{ground-truth with-me-ness}), or based on the focus of attention measured by the robot.}
+    \begin{tabular}{p{1cm}cccccccc}
         \toprule
-        \textbf{Subject} & 1 & 2 & 3 & 4 & 5 & 6 & Avg. & SD \\
+        Subject & 1 & 2 & 3 & 4 & 5 & 6 & Avg. & SD \\
         \midrule
-        %Ground-truth \vs expected (\%) & 92.6 & 88.8  & 95.8 & 93.1 & 96.5 & 89.6 & 92.7 & \\ \hline
-        $\mathcal{W}_{groundtruth}$ &  & & & & & & & \\ 
-        {\it Cronbach's $\alpha$} &  & & & & & & & \\
-        \midrule
+        $\mathcal{W}_{g.truth}$ & 92.6 & 88.8  & 95.8 & 93.1 & 96.5 & 89.6 & 92.7 & 3.1 \\ 
+        %{\it Cronbach's $\alpha$} &  & & & & & & & \\
+        %\midrule
         $\mathcal{W}_{robot}$ &  & & & & & & & \\
-        {\it Cronbach's $\alpha$} &  & & & & & & & \\
+        %{\it Cronbach's $\alpha$} &  & & & & & & & \\
         \bottomrule

### 2015-09-18 paper.tex
Minor changes

@@
 
-Please refer to~\cite{Hood:2015} for the implementation of the hand-writing
-activity.
+The implementation of the hand-writing activity itself has been previously
+presented in~\cite{Hood:2015}.
 
@@
 
+We performed a {\it post-hoc} coding the video records to manually of the
+interaction were (note that the focus of attention was separately hand-coded,
+{\it post-hoc}, from the video recording of the interaction).
+
 \subsubsection{From Face Detection to Focus of Attention} \label{fromFaceTo}
-(note that the focus of attention
-was separately hand-coded, {\it post-hoc}, from the video recording of the
-interaction).
 

### 2015-09-20 paper.tex
Fixed FoV fixMes

@@
 
-The width of the human field of attention has been discussed in the literature.
+The width of the human field of view has been discussed in the literature.
 Holmqvist~\cite{holmqvist2011eye} specifies that the visual human
-range\fixme{field of view or field of attention?} is $ \pm  40\degree $ in the
+range is $ \pm  40\degree $ in the
 horizontal and $ \pm 25\degree $ in the vertical.
@@
 aperture. In the present implementation, the first approach explained has been
-chosen representing the field of view using a cone with such
-dimensions\fixme{which one? 40deg? 25deg?}.
+chosen representing the field of view using a cone with an elliptical cross section 
+and the dimensions stated by Holmqvist.
 

### 2015-09-20 paper.tex
Fixed mean, std and times

@@
 
-Six children (ages 5 to 6\fixme{boys? girls?}) were enrolled for this study.
+Six children (3 boys and 3 girls with ages 5 to 6) were enrolled for this study.
 The study took place at school, in a separate room (the computer
 lab). The participants were chosen by the teacher, and would come one after the
-other for a face-to-face interaction with a {\sc Nao} robot (average duration:
-18 min, $SD:$ \fixme{stddev?}).
+other for a face-to-face interaction with a {\sc Nao} robot (the duration has $\mu:
+19.59$, $SD:1.58$).
 
@@
 
-\fixme{how many rounds for each child?}
+In this case, the number of rounds performed by each subject ($\mu: 12,16$, $SD:2.61$) 
+varies depending on the interaction time, their ability writing and the easiness of use.
+The number of rounds has a .
 
 Besides, once per interaction, the robot did interrupt the handwriting task to
-propose to tell a funny story. The story-telling lasted in average \fixme{how
-long?} min, and the turn-based hand-writing task continued afterwards. The
+propose to tell a funny story. The story-telling lasted for 1 min. 58 sec., 
+and the turn-based hand-writing task continued afterwards. The
 intended purpose of the story-telling episode was to locally trigger a different

### 2015-09-20 paper.tex
Cosine similarity figure removed gray blur

@@
 of the subject's gaze direction based on manual annotations from two evaluators
-with an average reliability of $ \alpha = .89 $\fixme{Is that Cronbach alpha or
-Krippendorff's alpha?}.
+with an average reliability of $ \alpha = .89 $ computed using Cronbach's alpha.
 

### 2015-09-20 paper.tex
Chronbach`s alpha test results added

@@
         {\bf Ground-truth/robot match} (\%) & 61.1 & 85.5 & 80.8 & 62.2 & 63.4 & 83.9 & {\bf 72.8} & {\it 11.7}\\
-        {\bf Cronbach's $\alpha$} &  & & & & & & & \\
+        {\bf Cronbach's $\alpha$} & 0.63 & 0.81 & 0.82 & 0.59 & 0.64 & 0.78 & {\bf0.71} & {\it0.10}\\
         {\bf Cosine affinity} & 0.73 & 0.82 & 0.87 & 0.73 & 0.77 & 0.89 & {\bf 0.8} & {\it 0.07} \\
@@
 in the average, the bad precision in the captured VFoA suggests that such
-children are highly likely to modify the gaze direction without head move
-intervention reducing dramatically the overlap between captured and real VFoA.
+children are highly likely to modify the gaze direction without head movement
+intervention reducing dramatically the overlap between the manually annotated 
+focus of attention and robot's computed focus of attention.
 

### 2015-09-21 paper.tex
Minnor changes
Formatting

@@
 
-Six children (3 boys and 3 girls with ages 5 to 6) were enrolled for this study.
+Six children (ages 5 to 6, 3 boys, 3 girls) were enrolled for this study.
 The study took place at school, in a separate room (the computer
 lab). The participants were chosen by the teacher, and would come one after the
-other for a face-to-face interaction with a {\sc Nao} robot (the duration has $\mu:
-19.59$, $SD:1.58$).
+other for a face-to-face interaction with a {\sc Nao} robot (average duration:
+19.6 min, $SD:1.58$).
 
@@
 demonstration and tries again. The child continues the turn-taking until he/she
-decides to train the robot on another word. Details on the rationale and
-implementation of this study can be found in~\cite{Hood:2015}.
+decides to train the robot on another word. In total, the children performed in
+average 12.16 ($SD:2.61$) rounds of writing.
 
-In this case, the number of rounds performed by each subject ($\mu: 12,16$, $SD:2.61$) 
-varies depending on the interaction time, their ability writing and the easiness of use.
-The number of rounds has a .
+Besides, once per interaction, the robot interrupts the handwriting task to
+propose to tell a story (that lasted for about 2 min), and the turn-based
+hand-writing task continued afterwards. The intended purpose of the
+story-telling episode was to locally trigger a different set of gaze behaviors.
 
-Besides, once per interaction, the robot did interrupt the handwriting task to
-propose to tell a funny story. The story-telling lasted for 1 min. 58 sec., 
-and the turn-based hand-writing task continued afterwards. The
-intended purpose of the story-telling episode was to locally trigger a different
-set of gaze behaviors.
+Details on the rationale and implementation of this experiment can be found
+in~\cite{Hood:2015}.
 
@@
 intervention reducing dramatically the overlap between the manually annotated 
-focus of attention and robot's computed focus of attention.
+focus of attention and robot's computed focus of attention. And in consequence,
+increasing the standard deviation.
 

### 2015-09-21 paper.tex
Re-organization, formatting, starting to work on 4.2

@@
 \section{Experimental Validation}
+\label{sec:expe}
 
@@
 other for a face-to-face interaction with a {\sc Nao} robot (average duration:
-19.6 min, $SD:1.58$).
+19.6 min, $SD=1.58$).
 
@@
 decides to train the robot on another word. In total, the children performed in
-average 12.16 ($SD:2.61$) rounds of writing.
+average 12.16 ($SD=2.61$) rounds of writing.
 
@@
 The experiment was carried with an Aldebaran's {\sc Nao} robot, using ROS as a
-middleware to build the attention estimation pipeline
-(Figure~\ref{fig:system}). Head pose estimation builds on the {\tt dlib} and
-OpenCV libraries, while pose transformations are handled by the {\sc tf}
-library. The implementation is open-source and available at
+middleware to build the attention estimation pipeline (Figure~\ref{fig:system}).
+Head pose estimation, presented in section~\ref{sec:vfoa}, builds on the {\tt
+dlib} and OpenCV libraries; the pose transformations are handled by the ROS {\sc
+tf} library. The implementation is open-source and available at
 \url{https://github.com/chili-epfl/attention-tracker}.
 
-The implementation of the hand-writing activity itself has been previously
-presented in~\cite{Hood:2015}.
+The implementation of the hand-writing activity itself has been presented by
+Hood~\etal in~\cite{Hood:2015}.
+
+\subsection{Data Collection \& Analysis}
+
+Successful detection of the head, and, when detected, the attentional target of
+the children as estimated by the robot, were logged during the experiment (in
+total, $6\times19.6=117.6$ min of interaction).\fixme{What it the case that 2
+focuses of attention were sometimes detected? what did we do in such a
+case?}
+
+Besides, we video-recorded the interactions, and performed a {\it post-hoc}
+manual coding of the focus of attention by two annotators (Cronbach's
+$\alpha=.89$)\fixme{did the 2nd annotator annotated the 6 whole interactions?}.
+The manual coding forms our attentional \emph{ground-truth}.
+
+To assess the accuracy of the attention estimation by the robot, we computed
+over time (excluding the periods when the head is not detected) the
+overlap between the ground-truth and the robot's estimate, the
+inter-judge agreement (Cronbach's $\alpha$) and the cosine similarity
+matrix.
+
+The cosine similarity is a more reliable measurement between the ground-truth
+and the robot measure of the focus of attention as it accounts for temporal
+delays.\fixme{this needs to be better justified/explained.}.
 
@@
 
-We performed a {\it post-hoc} coding the video records to manually of the
-interaction were (note that the focus of attention was separately hand-coded,
-{\it post-hoc}, from the video recording of the interaction).
-
-\subsubsection{From Face Detection to Focus of Attention} \label{fromFaceTo}
-
-Our results for a peer-to-peer work scenarios using an approach based on RGB
-camera are summarized in table \ref{tab:results} where the \textit{expected} is
-the focus of attention that given the state of the system, it is likely to
-attract the subject's gaze. In the same manner, the \textit{captured} is the
-focus of attention our approach provides and the \textit{real}, the ground truth
-of the subject's gaze direction based on manual annotations from two evaluators
-with an average reliability of $ \alpha = .89 $ computed using Cronbach's alpha.
-
-In order to calculate the inter-rater reliability between the two measurements
-(real-captured and real-expected), the overlap percentage between the two
-curves\fixme{which curve?}
-has been computed. A more reliable measurement in terms of time shifting between
-the real and captured focuses of attention is provided using the cosine
-similarity\fixme{this needs to be better justified: why not a Cronbach alpha?} also shown in table~\ref{tab:results}.
+The main results are reported in Table~\ref{tab:results} and
+Figure~\ref{fig:cosinematrix} (cosine similarity matrix);
+Figure~\ref{fig:realExpected} further gives a concrete picture of the
+ground-truth \vs computed attentional targets for subject 4 (the subject with
+the least successful tracking).
 
@@
     \centering
-    \caption{\textbf{Results of attention tracking accuracy}. \emph{Head pose
+    \caption{\textbf{Attention tracking accuracy}. \emph{Head pose
     tracking} is the percentage of total time of successful detection of the
@@
     between manually annotated focus of attention (ground-truth) and robot's
-    computed focus of attention.}
+    computed focus of attention. Total duration: 117.6 min.}
 
@@
 
-The results has proved that the VFoA prediction accuracy in the tracking is not
-only conditioned by the subject's amount of movement, but also by the speed of
-the same movements: Restless children externalize greater levels of movement
-decreasing the VFoA prediction accuracy.
-
-For instance, subjects 1,4 and 5 despite the fact that the tracking accuracy is
-in the average, the bad precision in the captured VFoA suggests that such
-children are highly likely to modify the gaze direction without head movement
-intervention reducing dramatically the overlap between the manually annotated 
-focus of attention and robot's computed focus of attention. And in consequence,
-increasing the standard deviation.
-
-Additionally, several recurrent computer vision challenges such as the lighting
-conditions, occlusions with the hands and the face angle with respect to the
-camera, looking to the experimenter for instance, are factors that contribute to
-reduce the tracking accuracy.
-
-\subsubsection{Subject's Focus of Attention Analysis}
+\begin{figure}[th!]
+    \centering
+    \includegraphics[width=0.7\columnwidth]{bitmap}
+    \caption{\small Cosine similarity matrix between ground-truth and robot-measured VFoA
+    using a 10 sec. window size. The correspondent numbers are shown
+    in figure \ref{fig:realCaptured}.}
+    \label{fig:cosinematrix}
+\end{figure}
 
@@
     \centering
-    \includegraphics[width=\linewidth]{matches}
+    \includegraphics[width=\linewidth]{matches-excerpt}
     \caption{\small \textbf{Comparison of computed focus of attention \vs ground
-        truth} during a 20+ min face-to-face child-robot interaction (subject 4
-        in table~\ref{tab:results}).
+        truth} during a face-to-face child-robot interaction (subject 4
+        in table~\ref{tab:results}, 3.5min-long excerpt).
         In blue (top line), the focus of attention as computed by the robot;
         in orange (bottom line), the focus of attention as manually annotated
-        (ground truth). The bottom graph shows agreements.}
+        (ground-truth). The bottom graph shows agreement between both (whenever
+        the head is detected).}
 
@@
 
-According to figure \ref{fig:realExpected} several findings can be analyzed in
-detail. The matching curve (cyan color) shows the overlapping between the
-expected and the real focuses of attention. As can be noticed there is a
-fluctuation at the beginning of the interaction as well as just before the
-activity switching. These chaotic gaze manifestation is what we call an
-automatic discovery of events, when the user experiences an uncertainty in the
-present situation.
+\subsubsection{Relevance of Head Pose Estimation for Attention Tracking}
+\label{sec:fromFaceTo}
 
-The first one of the previous two situations presented above is due to the
-novelty effect the subjects experience for the first time in front of the system
-and according to the rest of participants suggests that the adaptation time is
-around the first 2 minutes of interaction. In the second case, an unexpected
-robot behavior produces such reaction due to an activity switching, the story
-telling.
+During the whole interaction, the head pose of the children was satisfactorily
+and consistently tracked (86\% of the time in average, $SD=3.0$), which is
+expected for a face-to-face interaction with a static head-tracking camera
+(meaning that the child head would remain in the field of view of the camera
+most of the time). This is however comforting in terms of suitability of our
+approach for head pose estimation with children in field experiments.
 
-It is important to consider that in some occasions during the interaction, none
-of the focuses of attention were within the field of view due to the time
-transition from one focus to another resulting in an ambiguity. In order to deal
-with these few cases it is assumed that the minimum distance between the height
-of the field of view (cone) and the targets corresponds to the most likely focus
-of attention. However, we also want to distinguish such cases when the subject
-is directly not looking where the defined targets are located, but towards a
-region far from the interaction field. These cases are not common but they can
-be easily identified by setting a horizontal threshold to the field of view.
+In our experiment, the primary causes of lost head pose were occlusions with the
+hands (similar to the middle-bottom picture in
+Figure~\ref{head_pose_real_world}), close proximity with the tablet while
+writing, and gaze directed to the facilitator (sitting directly on the left of
+the child, Figure~\ref{fig:realSetup}).
 
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.7\columnwidth]{bitmap}
-    \caption{\small Cosine similarity matrix between ground-truth and robot-measured VFoA
-    using a 10 sec. window size. The correspondent numbers are shown
-    in figure \ref{fig:realCaptured}.}
-    \label{fig:bitmap}
-\end{figure}
-
-A relevant trend that can be observed is the continuous `look for approval'
-situation where the subjects look at the experimenter in order to provide
-feedback about the robot's response before providing their own one (see yellow
-boxes in figure \ref{fig:realExpected}). Such transition tablet-experimenter
-becomes more frequent in several cases, for instance when the experimenter makes
-a suggestion (orange box in figure \ref{fig:realExpected}), in this case, switch
-to a more complicated word.
-
-Once the turn-taking is established, a pattern becomes visible during the
-subject's feedback, when an attempt is performed: Such pattern shows a high
-intermittent gaze frequency from the tablet to the selection. It also reveals
-that the subject is using the demonstration model in the selection tablet to
-provide a better answer to the robot. In fact, subjects decrease the response
-time after each shift as well as the gaze frequency towards the model suggesting
-an assimilation over time. Moreover, when a new demonstration is selected for
-playing, the same behavior is replicated (see yellow boxes in figure
-\ref{fig:realExpected}).
+In terms of attention tracking, two groups of children stand out of
+Table~\ref{tab:results}: the robot's estimate of the focus of attention for
+subjects 1, 4 and 5 has questionable agreement with the ground-truth (Cronbach's
+$\alpha$ close to 0.6), while for subjects 2, 3 and 6, we find a good agreement
+($\alpha$ close to 0.8). This does not correlate with the success rate of the
+head tracking itself, and, after further analysis of the videos, it rather
+reflects that the first group of children rely more on their eyes to direct
+their gaze than the second group. However, and notwithstanding the small sample
+size, it is rather intriguing that the two groups appear so clearly separated.
+This certainly calls for further experimental investigation.
 
@@
 \section{With-me-ness}
+
+\subsection{Concept \& Calculation}
 
@@
 
+\fixme{mention somewhere probabilistic extensions of withmeness}
+
+\subsection{Experimental Interpretation}
+
+During the experiment presented above (section~\ref{sec:expe}), the robot
+controller associated a (set of) expected attentional targets to the task being
+performed. For instance, while the robot was waiting for the child's handwriting
+demonstration, the expected attentional target was the tablet (since the child
+was supposed to write there), and while the robot was writing itself, the
+attentional targets we expected to be either on the tablet or on the robot.
+These expected targets, forming together the robot's attentional {\it a priori},
+are depicted on Figure~\ref{fig:with-me-ness} as the green lines.
+
+\begin{figure*}
+    \centering
+    \includegraphics[width=\linewidth]{with-me-ness}
+    \caption{\small \textbf{With-me-ness}. Evolution of the level of
+        \emph{with-me-ness} over the whole $\approx$20min long interaction of
+        subject 4. \emph{Interaction Episodes} describe the distinct phases of
+        the interaction, and are discussed in the text. The central chart is
+        similar to Figure~\ref{fig:realExpected} with the \emph{expected}
+        attentional targets added in green. The bottom diagram represents the
+        levels of with-me-ness over successive windows of 30 seconds. The blue
+        line is with-me-ness as estimated by the robot, the orange line is
+        with-me-ness computed from manually annotated attentional targets. The
+        ``holes'' in the with-me-ness computed by the robot correspond to period of time
+        where the child's face is not detected.}
+
+    \label{fig:with-me-ness}
+\end{figure*}
+
 \begin{table}[h!]
@@
         $\mathcal{W}_{g.truth}$ & 92.6 & 88.8  & 95.8 & 93.1 & 96.5 & 89.6 & 92.7 & 3.1 \\ 
-        %{\it Cronbach's $\alpha$} &  & & & & & & & \\
-        %\midrule
         $\mathcal{W}_{robot}$ &  & & & & & & & \\
-        %{\it Cronbach's $\alpha$} &  & & & & & & & \\
         \bottomrule
@@
 
-\begin{figure*}
-    \centering
-    \includegraphics[width=\linewidth]{with-me-ness}
-    \caption{\small \textbf{With-me-ness}. Evolution of the level of
-    \emph{with-me-ness} over a 20min long child-robot interaction. The diagram
-    is identical to Figure~\ref{fig:realExpected} with the \emph{expected}
-    attentional targets added in green. The bottom curves represent the levels
-    of with-me-ness (since the beginning of the interaction), computed following
-    Algorithm~\ref{alg:with-me-ness}. The blue line is with-me-ness as estimated by
-    the robot, the orange line is with-me-ness computed from manually annotated
-    focuses of attention. The ``holes'' in the with-me-ness computed by the
-    robot correspond to period of time where the child's face is not detected.}
+According to figure \ref{fig:with-me-ness} several findings can be analyzed in
+detail. The matching curve (cyan color) shows the overlapping between the
+expected and the real focuses of attention. As can be noticed there is a
+fluctuation at the beginning of the interaction as well as just before the
+activity switching. These chaotic gaze manifestation is what we call an
+automatic discovery of events, when the user experiences an uncertainty in the
+present situation.
 
-    \label{fig:with-me-ness}
-\end{figure*}
+The first one of the previous two situations presented above is due to the
+novelty effect the subjects experience for the first time in front of the system
+and according to the rest of participants suggests that the adaptation time is
+around the first 2 minutes of interaction. In the second case, an unexpected
+robot behavior produces such reaction due to an activity switching, the story
+telling.
 
-If we consider the case shown in section \ref{fromFaceTo}, evaluating the
+It is important to consider that in some occasions during the interaction, none
+of the focuses of attention were within the field of view due to the time
+transition from one focus to another resulting in an ambiguity. In order to deal
+with these few cases it is assumed that the minimum distance between the height
+of the field of view (cone) and the targets corresponds to the most likely focus
+of attention. However, we also want to distinguish such cases when the subject
+is directly not looking where the defined targets are located, but towards a
+region far from the interaction field. These cases are not common but they can
+be easily identified by setting a horizontal threshold to the field of view.
+
+A relevant trend that can be observed is the continuous `look for approval'
+situation where the subjects look at the experimenter in order to provide
+feedback about the robot's response before providing their own one (see yellow
+boxes in figure \ref{fig:realExpected}). Such transition tablet-experimenter
+becomes more frequent in several cases, for instance when the experimenter makes
+a suggestion (orange box in figure \ref{fig:realExpected}), in this case, switch
+to a more complicated word.
+
+Once the turn-taking is established, a pattern becomes visible during the
+subject's feedback, when an attempt is performed: Such pattern shows a high
+intermittent gaze frequency from the tablet to the selection. It also reveals
+that the subject is using the demonstration model in the selection tablet to
+provide a better answer to the robot. In fact, subjects decrease the response
+time after each shift as well as the gaze frequency towards the model suggesting
+an assimilation over time. Moreover, when a new demonstration is selected for
+playing, the same behavior is replicated (see yellow boxes in figure
+\ref{fig:realExpected}).
+
+If we consider the case shown in section \ref{sec:fromFaceTo}, evaluating the
 subject's progression together with the time on task may be an indicator of

### 2015-09-22 data/analysis/process.py
Added the EAF for subject 4 + a basic python script to plot it

@@ -1,0 +1,79 @@
+from collections import OrderedDict
+import pympi
+
+from simplesvg import Scene, Line
+
+
+############################################################################
+#     Processing of the EAF file
+############################################################################
+
+elan_file = "../4.eaf"
+
+tier = "Attention focus"
+#tier = "Robot state"
+
+eaf = pympi.Elan.Eaf(elan_file)
+
+annotation2foa = {"Waiting for feedback": ("Tablet", "Secondary tablet", "Facilitator"),
+        "Writing word": ("Tablet", "Robot"),
+        "Waiting for word to write": ("Secondary tablet",),
+        "Story telling": ("Robot",),
+        "Presentation": ("Robot",),
+        "robot": ("Robot",),
+        "tablet": ("Tablet",),
+        "facilitator": ("Facilitator",),
+        "selection": ("Secondary tablet",),
+        "observer": ("Observer",),
+        "paper": ("Other",), # is that correct?
+        "other": ("Other",)}
+
+
+def get_time(ts):
+    """ Returns the timestamp of an EAF 'timeslot', in seconds
+    """
+    return eaf.timeslots[ts] / 1000.
+
+
+# First, convert all EAF timeslots into timestamp in seconds and 'normalize'
+# the focus of attention using the mappings in annotation2foa
+rawevents = {}
+for _, entry in eaf.tiers[tier][0].items():
+    ts, te, ann, _ = entry
+
+    if ann not in annotation2foa:
+        print("Unknown annotation: %s" % ann)
+        continue
+
+    rawevents[get_time(ts)] = (annotation2foa[ann], get_time(te))
+
+# Order the 'events' by timestamps
+events = OrderedDict(sorted(rawevents.items(), key=lambda t: t[0]))
+
+############################################################################
+#     SVG rendering
+############################################################################
+
+plotting_order = {"Observer": 0,
+                  "Facilitator": 1,
+                  "Secondary tablet": 2,
+                  "Robot": 3,
+                  "Tablet": 4,
+                  "Other": 5}
+
+
+svg = Scene(tier)
+
+for ts, evt in events.items():
+
+    targets, te = evt
+    for t in targets:
+        print("%s: %d to %d" % (t, ts, te))
+
+        # a different offset for each 'target'
+        offset = plotting_order[t] * 10
+
+        # simply draw a plain line between the two timestamps
+        svg.add(Line((ts, offset), (te,offset)))
+
+svg.write_svg()

### 2015-09-22 data/analysis/simplesvg.py

@@ -1,0 +1,109 @@
+#taken from http://code.activestate.com/recipes/325823-draw-svg-images-in-python/
+import os
+display_prog = 'display' # Command to execute to display images.
+      
+class Scene:
+    def __init__(self,name="svg",height=400,width=400):
+        self.name = name
+        self.items = []
+        self.height = height
+        self.width = width
+        return
+
+    def add(self,item): self.items.append(item)
+
+    def strarray(self):
+        var = ["<?xml version=\"1.0\"?>\n",
+               "<svg height=\"%d\" width=\"%d\" >\n" % (self.height,self.width),
+               " <g style=\"fill-opacity:1.0; stroke:black;\n",
+               "  stroke-width:1;\">\n"]
+        for item in self.items: var += item.strarray()            
+        var += [" </g>\n</svg>\n"]
+        return var
+
+    def write_svg(self,filename=None):
+        if filename:
+            self.svgname = filename
+        else:
+            self.svgname = self.name + ".svg"
+        file = open(self.svgname,'w')
+        file.writelines(self.strarray())
+        file.close()
+        return
+
+    def display(self,prog=display_prog):
+        os.system("%s %s" % (prog,self.svgname))
+        return        
+        
+
+class Line:
+    def __init__(self,start,end):
+        self.start = start #xy tuple
+        self.end = end     #xy tuple
+        return
+
+    def strarray(self):
+        return ["  <line x1=\"%d\" y1=\"%d\" x2=\"%d\" y2=\"%d\" />\n" %\
+                (self.start[0],self.start[1],self.end[0],self.end[1])]
+
+
+class Circle:
+    def __init__(self,center,radius,color):
+        self.center = center #xy tuple
+        self.radius = radius #xy tuple
+        self.color = color   #rgb tuple in range(0,256)
+        return
+
+    def strarray(self):
+        return ["  <circle cx=\"%d\" cy=\"%d\" r=\"%d\"\n" %\
+                (self.center[0],self.center[1],self.radius),
+                "    style=\"fill:%s;\"  />\n" % colorstr(self.color)]
+
+class Rectangle:
+    def __init__(self,origin,height,width,color):
+        self.origin = origin
+        self.height = height
+        self.width = width
+        self.color = color
+        return
+
+    def strarray(self):
[…]

### 2015-09-22 paper.tex
Robot with-me-ness computed
Commented out the heat-map for now

@@
 
-\subsubsection{Focus of attention distribution}
-
-Additional information of the head pose can be obtained for the evaluation of
-the VFoA distribution. Figure \ref{fig:heatmap} shows the most salient focuses
-of attention where the peaks are located. These peaks are mainly projected
-between the triangle composed by the robot, the writing tablet and the selection
-tablet using an unsupervised K-means clustering method reveals, considering the
-observers as negligible VFoA.
-
-% Workaround to do subfigure since the packege looks not to work
-\begin{figure}[th!]
-  \centering
-  \begin{tabular}{@{}c@{}}
-    \includegraphics[width=.7\linewidth,height=100pt]{heatmap} \\
-    \small (a)  2D histogram.
-  \end{tabular}
-
-  \vspace{\floatsep}
-
-  \begin{tabular}{@{}c@{}}
-    \includegraphics[width=.7\linewidth,height=100pt]{kmeans} \\
-    \small (b) K-means clustering.
-  \end{tabular}
-
-  \caption{ \small Subject 4 head movements. (head yaw on x-axis, pitch on y-axis).}
-  \label{fig:heatmap}
-\end{figure}
+%\subsubsection{Focus of attention distribution}
+%
+%Additional information of the head pose can be obtained for the evaluation of
+%the VFoA distribution. Figure \ref{fig:heatmap} shows the most salient focuses
+%of attention where the peaks are located. These peaks are mainly projected
+%between the triangle composed by the robot, the writing tablet and the selection
+%tablet using an unsupervised K-means clustering method reveals, considering the
+%observers as negligible VFoA.
+%
+%% Workaround to do subfigure since the packege looks not to work
+%\begin{figure}[th!]
+%  \centering
+%  \begin{tabular}{@{}c@{}}
+%    \includegraphics[width=.7\linewidth,height=100pt]{heatmap} \\
+%    \small (a)  2D histogram.
+%  \end{tabular}
+%
+%  \vspace{\floatsep}
+%
+%  \begin{tabular}{@{}c@{}}
+%    \includegraphics[width=.7\linewidth,height=100pt]{kmeans} \\
+%    \small (b) K-means clustering.
+%  \end{tabular}
+%
+%  \caption{ \small Subject 4 head movements. (head yaw on x-axis, pitch on y-axis).}
+%  \label{fig:heatmap}
+%\end{figure}
 
@@
         $\mathcal{W}_{g.truth}$ & 92.6 & 88.8  & 95.8 & 93.1 & 96.5 & 89.6 & 92.7 & 3.1 \\ 
-        $\mathcal{W}_{robot}$ &  & & & & & & & \\
+        $\mathcal{W}_{robot}$ & 89.6 & 84.4 & 86.6 & 72.8 & 81.9 & 79.7 & 82.5 & 5.4 \\
         \bottomrule

### 2015-09-22 paper.tex
Minor formatting
Commented out the 'gaze-shift' part for now

@@
 
-\subsubsection{Relevance of Head Pose Estimation for Attention Tracking}
-\label{sec:fromFaceTo}
-
 During the whole interaction, the head pose of the children was satisfactorily
@@
     \centering
-    \caption{\textbf{Levels of with-me-ness}.
-For each subject, the with-me-ness level is reported over the whole interaction, either based on the ground-truth (\ie \emph{ground-truth with-me-ness}), or based on the focus of attention measured by the robot.}
+    \caption{\textbf{Levels of with-me-ness}. For each subject, the with-me-ness
+    level is reported over the whole interaction, either based on the
+    ground-truth (\ie \emph{ground-truth with-me-ness}), or based on the focus of
+    attention measured by the robot.}
+
     \begin{tabular}{p{1cm}cccccccc}
@@
 time response to event is proposed to account quality of interaction. 
-
-However, a high amount of gaze shift (figure \ref{fig:gazeR}) towards the
-demonstration provided can reveal that the challenge is too big or no gaze, the
-exactly the opposite. In this case, the robot response can be adapted to the
-situation. Similar, but least relevant results are obtained using the
-correctness (figure \ref{fig:errorR}) of the child's feedback after each
-attempt.
-
-\begin{figure}
-  \centering
-  \begin{tabular}{@{}c@{}}
-    \includegraphics[width=.7\linewidth,height=100pt]{gazeR} \\
-    \small (a) Occurrences of a gaze shift.
-  \end{tabular}\label{fig:gazeR}
-
-  \vspace{\floatsep}
-
-  \begin{tabular}{@{}c@{}}
-    \includegraphics[width=.7\linewidth,height=100pt]{errorR} \\
-    \small (b) Word correctness.
-  \end{tabular}\label{fig:errorR}
-
-  \caption{ \small Linear regression of all subject attempts against the gaze
-  shift tablet-model (a) and the correctness as the average euclidean distance
-  of each letter in its corresponding eigenspace with respect to a reference
-  letter (b). The gray region defines the confidence interval.}
-
-\end{figure}
-
-This information can be used to estimate the child's state during the
-interaction. For instance, if we consider that a low level of engagement is
-produced by extremely high or low demanding situations, it is possible to assess
-the engagement level during the interaction based on gaze shift and word
-correctness.
+%
+%However, a high amount of gaze shift (figure \ref{fig:gazeR}) towards the
+%demonstration provided can reveal that the challenge is too big or no gaze, the
+%exactly the opposite. In this case, the robot response can be adapted to the
+%situation. Similar, but least relevant results are obtained using the
+%correctness (figure \ref{fig:errorR}) of the child's feedback after each
+%attempt. 
+%
+%\begin{figure}
+%  \centering
+%  \begin{tabular}{@{}c@{}}
+%    \includegraphics[width=.7\linewidth,height=100pt]{gazeR} \\
+%    \small (a) Occurrences of a gaze shift.
+%  \end{tabular}\label{fig:gazeR}
+%
+%  \vspace{\floatsep}
+%
+%  \begin{tabular}{@{}c@{}}
+%    \includegraphics[width=.7\linewidth,height=100pt]{errorR} \\
+%    \small (b) Word correctness.
+%  \end{tabular}\label{fig:errorR}
+%
+%  \caption{ \small Linear regression of all subject attempts against the gaze
+%  shift tablet-model (a) and the correctness as the average euclidean distance
+%  of each letter in its corresponding eigenspace with respect to a reference
+%  letter (b). The gray region defines the confidence interval.}
+%  
+%\end{figure}
+%
+%
+%This information can be used to estimate the child's state during the
+%interaction. For instance, if we consider that a low level of engagement is
+%produced by extremely high or low demanding situations, it is possible to assess
+%the engagement level during the interaction based on gaze shift and word
+%correctness.
 
@@
 
-During the case study, results show that it is possible to distinguish between
-long and short feedback episodes based on the gaze shift and depending on the
-previous attempts performed by the subject. As a result, gaze-shift together
-with the number of attempts becomes a reliable indicator candidate in the
-assessment of the engagement during the activity.
-
-Additionally, both the gaze shift and the correctness of the feedback seem to be
-strictly correlated with the learning gain during the activity as they go in
-decrement after each attempt provided by the subjects. Despite the fact that
-these results are encouraging, there is a need of validation with a greater
-amount of subjects.
+%During the case study, results show that it is possible to distinguish between
+%long and short feedback episodes based on the gaze shift and depending on the
+%previous attempts performed by the subject. As a result, gaze-shift together
+%with the number of attempts becomes a reliable indicator candidate in the
+%assessment of the engagement during the activity.
+%
+%Additionally, both the gaze shift and the correctness of the feedback seem to be
+%strictly correlated with the learning gain during the activity as they go in
+%decrement after each attempt provided by the subjects. Despite the fact that
+%these results are encouraging, there is a need of validation with a greater
+%amount of subjects.
 

### 2015-09-22 paper.tex
Working on conclusion

@@
 
-In this work we have presented an on-line vision-based engagement assessment
-solution that can be extended to work in different platforms for simulating
-complex interaction scenarios monitoring several VFoA with a maker-less
-technology. It provides some benefits like the decrement of the system
-complexity as well as the increment of the naturalness of the interaction due to
-the use of an RGB sensor.
+As hinted by the title of this article, we have presented here how a robot can
+effectively assess in real-time and with a regular camera the focus of attention
+of its interactant, and how we can, in turn, combine it with the robot's {\it a priori}
+about the interaction to build a straight-forward (yet novel in HRI) metric of
+\emph{with-me-ness} over the course of the interaction.
+
+\paragraph{Practicality and Relevance of 2D Methods for Head Pose Estimation}
 
@@
 
-Gaze information acquired during the interactions is not only useful to detect
-and predict engagement, but also allows the possibility to develop actively the
-concept gaze-aware robot. In other words, make the system able to estimate where
-the subject looks at, but also imitate the same subject's VFoA when the activity
-state allows it.
+%Gaze information acquired during the interactions is not only useful to detect
+%and predict engagement, but also allows the possibility to develop actively the
+%concept gaze-aware robot. In other words, make the system able to estimate where
+%the subject looks at, but also imitate the same subject's VFoA when the activity
+%state allows it.
 
-%ACKNOWLEDGMENTS are optional
+\paragraph{With-me-ness: yet another metric of engagement?}
+
+Borrowing the neologism from the field of CSCL, we have also introduced here
+\emph{with-me-ness} as a measure of ``how much the user in \emph{with} the robot during
+a task''. Interestingly, this can be measured over the course the interaction,
+thus providing the robot with a relatively high-level social construct,
+undoubtedly related to engagement.
+
+One may reasonably wonder how different is \emph{with-me-ness} from \emph{joint
+attention} on one hand, and from \emph{engagement} on the other hand.
+With-me-ness is related to both, with however noteworthy nuances: (Triadic) \emph{joint
+attention} is understood as the cognitive realization of a shared attention to
+an object, itself building on a shared perception of that object (\ie joint attention
+builds on a \emph{perceptual} alignment of two agents). \emph{Conceptual
+with-me-ness} as proposed by Sharma~\etal is on the other hand \emph{referential}: ``you are
+\emph{with} me if you focus on what I refer to, either explicitly or
+implicitly''. We propose to understand it here in a slightly broader way, that
+better reflect the \emph{interaction}: ``you are \emph{with} me if you focus on
+what is important for the interactive task at hand.''
+
+On the other hand, with-me-ness is only a precursor of engagement: it
+does not say much about the cognitive commitment of a user to an interaction. A
+user may closely adhere to the injunctions of the robot (or, actually, of the
+experimenters) without being \emph{engaged} into the interaction. This is
+typically seen in child-robot interaction: children will attempt to closely
+follow what they are asked to do -- which may \emph{look like} they are engaged
+into the interaction -- while they merely \emph{obey orders}.
+
+One of the strength of with-me-ness is its objectivity: we can formalize it, and
+as such, it is valuable to assess and compare how users are willing or able to
+interact with a robot. Besides, we have hopefully demonstrated in this article
+that with-me-ness is an operational \emph{in-the-moment} metric that can be also
+used as feed-back to the robot controller to build richer, more adaptive
+interactive behaviors for our robots.
+
 \section*{Acknowledgments}

### 2015-09-22 paper.tex
WIP on the discussion

@@
 
-\paragraph{Practicality and Relevance of 2D Methods for Head Pose Estimation}
-
-Furthermore, although it has not been tested in a real scenario with several
-subjects at the same time, the approach also allows to handle such situations
-too.
+\paragraph{Head Pose to Assess Attention: is it Relevant?}
 
@@
 
-Regarding to the approach used, some limitations need to be considered. The most
-relevant one, the fact that the positions of the VFoA need to be known in
-advance and defined as static \textit{tf} frames in the system. Moreover, it is
-important to notice that eye gaze information is neglected which induces a
-certain amount of error, as well as the noise induced by the 3D pose estimation
-computed from face features.
+We already stated the main limitations of our approach to estimation of the
+focus of attention: eye gaze information is neglected and we do not perform
+actual visibility check of the in focus objects (we simply approximate them to
+their origins, ignoring possible occlusions).
 
-Besides, our approach does not account for visual occlusions when computing the
-focus of attention: any frame within the cone of the visual field of attention
-is considered as \emph{in focus}. In reality, some of the frames may be visually
-occluded and thus not seen by the subject. Computing actual visibility requires
-more general situation assessment capabilities, and in particular (approximate)
-3D models of the surrounding objects so that visibility testing may be
-performed, as done in~\cite{sisbot2011situation} for instance. Our experimental
-setting did not lead to any specific visual occlusions, and we did not developed
-this capability.
-
-Compared to intrusive hardware like head mounted
-displays, eye-trackers or wearables for instance, this approach has a clear
-advantage of simplicity and integration with robotic systems, but also becomes
-accurate enough for general purposes in face-to-face interactions. In addition,
-it allows us to model complex environments, handling different number and
-positions of participants in the scene.
+The first issue is shared with most of the other vision (2D or 3D) or motion
+capture techniques for gaze estimation. In that regard, our approach does not
+perform worst or better than others. In fact, based on our results, relying
+purely on head pose estimation to estimate gaze direction leads to real-world
+results that are worth being considered and used: they may not match manual
+annotations, but they are definitely a valuable \emph{in-the-moment} input for
+the robot, and we hope to see more works relying on those to build adaptive
+robot controllers in the coming years.
 
@@
 our robot mimics attention tracking mechanism comparable to the ones found in apes.
+
+Besides, our approach relies on a simple, non-intrusive sensor (a RGB camera on
+the robot) and an open-source, fast pose estimation algorithm\fixme{test on Nao!
+how fast is it actually?}: we also hope that this may contribute to a widespread
+adoption of such a technique on a range of robots, including the common {\sc
+Nao} platform.
 

### 2015-09-22 paper.tex
Various minor changes

@@
 
-We then propose to introduce the concept of ``with-me-ness'', borrowed from the
+We secondly propose to introduce the concept of ``with-me-ness'', borrowed from the
 field of {\it Computer-Supported Collaborative Learning}, as an
@@
 Fanelli~\etal provides a overview of these approaches in~\cite{fanelli2012real},
-and recent examples include~\cite{baxter2014tracking} or~\cite{anzalone}.
+and recent examples include~\cite{baxter2014tracking, anzalone}.
 
@@
 
-The width of the human field of view has been discussed in the literature.
-Holmqvist~\cite{holmqvist2011eye} specifies that the visual human
-range is $ \pm  40\degree $ in the
-horizontal and $ \pm 25\degree $ in the vertical.
-In~\cite{walker1980clinical}, Spector provides a more detailed specification
-for each eye splitting the vertical range into $ 60\degree $ the upper region
-and $ 75\degree $ the lower one. Moreover, the horizontal range gets separated
-in $ 60\degree $ inwards (towards the nose) and $ 95\degree $ outwards.
-Previous work on visual perspective taking for social robotics
-application~\cite{sisbot2011situation} propose to use a cone of 30$\degree$ of
-aperture. In the present implementation, the first approach explained has been
-chosen representing the field of view using a cone with an elliptical cross section 
-and the dimensions stated by Holmqvist.
+The dimensions of the human field of view have been measured at different level
+of accuracy. Holmqvist~\cite{holmqvist2011eye} models it with an horizontal
+aperture of $ \pm 40\degree $ and a vertical aperture of $ \pm 25\degree $,
+while Spector~\cite{walker1980clinical} for instance suggests $60\degree$ up,
+$75\degree$ down, $60\degree$ inwards (towards the nose) and $95\degree$
+outwards.  Previous work on visual perspective taking for social
+robotics~\cite{sisbot2011situation} simply use a cone of $30\degree$.  In this
+work, we follow Holmqvist and thus model the field of view as a cone with an elliptical
+cross-section of $80\degree$ by $50\degree$.
 
 We then approximate the visual focus of attention (VFoA) of the human to the
-objects which lie inside the field of attention (Figure~\ref{fig:vfoa}). At a
+objects which lay inside this field of view (Figure~\ref{fig:vfoa}). At a
 given time, more than one object can therefore be \emph{in focus}.
@@
 lab). The participants were chosen by the teacher, and would come one after the
-other for a face-to-face interaction with a {\sc Nao} robot (average duration:
-19.6 min, $SD=1.58$).
+other for a face-to-face interaction with a {\sc Nao} robot ($M=19.6$ min, $SD=1.58$).
 
@@
 dlib} and OpenCV libraries; the pose transformations are handled by the ROS {\sc
-tf} library. The implementation is open-source and available at
+tf} library. The same {\sc tf} library is used to represent as individual frames
+the possible point of interests: an object is considered to be in focus when its
+frame lays within the field of attention of the participant
+(Figure~\ref{fig:vfoa}).  The implementation is open-source and available at
 \url{https://github.com/chili-epfl/attention-tracker}.

### 2015-09-22 paper.tex
Refining presentation of result tables

@@
         \toprule
-        {\bf Subject} & 1 & 2 & 3 & 4 & 5 & 6 & {\bf Avg.} & {\it SD} \\
+        {\bf Subject} & 1 & 2 & 3 & 4 & 5 & 6 & {\bf M} & {\it SD} \\
         \midrule
@@
 well as to its head. A human looking at either of these locations would be
-considered to be \emph{with} the robot in terms of interaction. Also notable, we
-exclude from the computation of $\mathcal{W}$ all of the periods of time where
-the user's focus of attention can not be estimated (Algorithm~\ref{alg:with-me-ness},
+considered to be \emph{with} the robot in terms of interaction\footnote{As
+future work, this could easily be extended into a probabilistic model of
+attention expectations.}. Also notable, we exclude from the computation of
+$\mathcal{W}$ all of the periods of time where the user's focus of attention can
+not be estimated (Algorithm~\ref{alg:with-me-ness},
 line~\ref{algline:skiplosttrack}, third condition), typically because the user's
 face is not visible at those times.
-
-\fixme{mention somewhere probabilistic extensions of withmeness}
 
@@
 attentional targets we expected to be either on the tablet or on the robot.
-These expected targets, forming together the robot's attentional {\it a priori},
-are depicted on Figure~\ref{fig:with-me-ness} as the green lines.
+These expected targets, forming altogether the robot's attentional {\it a
+priori}, are depicted on Figure~\ref{fig:with-me-ness} as the green lines. They
+are used to compute the with-me-ness. With-me-ness can be calculated over the
+whole interaction (Table~\ref{tab:results-with-me-ness}) or over shorter time
+windows (bottom graph on Figure~\ref{fig:with-me-ness}). Shorter time windows
+are interesting for two purposes: to analyse the level of with-me-ness in
+relation to specific interaction episodes; to allow a measurement of
+with-me-ness by the robot \emph{over the course} of the interaction
+(\emph{in-the-moment} measurement). In this later case, one may typically want
+to consider a sliding time window.
 
@@
         \toprule
-        Subject & 1 & 2 & 3 & 4 & 5 & 6 & Avg. & SD \\
+        Subject & 1 & 2 & 3 & 4 & 5 & 6 & {\bf M} & {\it SD} \\
         \midrule
-        $\mathcal{W}_{g.truth}$ & 92.6 & 88.8  & 95.8 & 93.1 & 96.5 & 89.6 & 92.7 & 3.1 \\ 
-        $\mathcal{W}_{robot}$ & 89.6 & 84.4 & 86.6 & 72.8 & 81.9 & 79.7 & 82.5 & 5.4 \\
+        $\mathcal{W}_{g.truth}$ & 92.6 & 88.8  & 95.8 & 93.1 & 96.5 & 89.6 & {\bf 92.7} & {\it 3.1} \\ 
+        $\mathcal{W}_{robot}$ & 89.6 & 84.4 & 86.6 & 72.8 & 81.9 & 79.7 & {\bf 82.5} & {\it 5.4} \\
         \bottomrule
@@
 
-According to figure \ref{fig:with-me-ness} several findings can be analyzed in
-detail. The matching curve (cyan color) shows the overlapping between the
+Figure~\ref{fig:with-me-ness} can be further analyzed:\fixme{This paragraph
+needs to be adapted}
+The matching curve (cyan color) shows the overlapping between the
 expected and the real focuses of attention. As can be noticed there is a

### 2015-09-22 data/analysis/process.py
Data analysis: 'paper' maps to 'secondary tablet'

@@ -10,8 +10,8 @@ from simplesvg import Scene, Line
 
 elan_file = "../4.eaf"
 
-tier = "Attention focus"
-#tier = "Robot state"
+#tier = "Attention focus"
+tier = "Robot state"
 
 eaf = pympi.Elan.Eaf(elan_file)
 
@@ -25,7 +25,7 @@ annotation2foa = {"Waiting for feedback": ("Tablet", "Secondary tablet", "Facili
         "facilitator": ("Facilitator",),
         "selection": ("Secondary tablet",),
         "observer": ("Observer",),
-        "paper": ("Other",), # is that correct?
+        "paper": ("Secondary tablet",), # is that correct?
         "other": ("Other",)}
 
 

### 2015-09-22 data/analysis/process.py
Data analysis: Added computation of with-me-ness

@@ -10,8 +10,8 @@ from simplesvg import Scene, Line
 
 elan_file = "../4.eaf"
 
-#tier = "Attention focus"
-tier = "Robot state"
+tier_observed = "Attention focus"
+tier_expected = "Robot state"
 
 eaf = pympi.Elan.Eaf(elan_file)
 
@@ -37,18 +37,22 @@ def get_time(ts):
 
 # First, convert all EAF timeslots into timestamp in seconds and 'normalize'
 # the focus of attention using the mappings in annotation2foa
-rawevents = {}
-for _, entry in eaf.tiers[tier][0].items():
-    ts, te, ann, _ = entry
+def prepare(tier):
+    rawevents = {}
+    for _, entry in eaf.tiers[tier][0].items():
+        ts, te, ann, _ = entry
 
-    if ann not in annotation2foa:
-        print("Unknown annotation: %s" % ann)
-        continue
+        if ann not in annotation2foa:
+            print("Unknown annotation: %s" % ann)
+            continue
 
-    rawevents[get_time(ts)] = (annotation2foa[ann], get_time(te))
+        rawevents[get_time(ts)] = (annotation2foa[ann], get_time(te))
 
-# Order the 'events' by timestamps
-events = OrderedDict(sorted(rawevents.items(), key=lambda t: t[0]))
+    # Order the 'events' by timestamps
+    return OrderedDict(sorted(rawevents.items(), key=lambda t: t[0]))
+
+events_observed = prepare(tier_observed)
+events_expected = prepare(tier_expected)
 
 ############################################################################
 #     SVG rendering
@@ -62,18 +66,77 @@ plotting_order = {"Observer": 0,
                   "Other": 5}
 
 
-svg = Scene(tier)
+def plot(name, events):
+    svg = Scene(name)
 
-for ts, evt in events.items():
+    for ts, evt in events.items():
 
-    targets, te = evt
-    for t in targets:
-        print("%s: %d to %d" % (t, ts, te))
+        targets, te = evt
+        for t in targets:
+            print("%s: %d to %d" % (t, ts, te))
 
-        # a different offset for each 'target'
-        offset = plotting_order[t] * 10
+            # a different offset for each 'target'
+            offset = plotting_order[t] * 10
 
-        # simply draw a plain line between the two timestamps
-        svg.add(Line((ts, offset), (te,offset)))
+            # simply draw a plain line between the two timestamps
+            svg.add(Line((ts, offset), (te,offset)))
 
-svg.write_svg()
[…]

### 2015-09-22 paper.tex
Cosine Similarity fixme `fixed`

@@
 
-The cosine similarity is a more reliable measurement between the ground-truth
-and the robot measure of the focus of attention as it accounts for temporal
-delays.\fixme{this needs to be better justified/explained.}.
+The Cronbach's $\alpha$ can be defined as an average of the correlations between 
+the items or values at the same time step \textit{t}. Therefore, results can vary 
+significantly if both sets of values are time shifted. However,
+the cosine similarity becomes a useful measurement in this case as it accounts 
+for temporal delays. Specifically, the ground-truth values acquired though human 
+annotations may not match the measurements of the focuses of attention computed 
+by the robot at the exactly same time step \textit{t} even though the correspondent
+curves may look very similar in shape.
 

### 2015-09-23 data/analysis/process.py
Different algo to compute with-me-ness, much closer to the paper

@@ -90,49 +90,38 @@ plot("groundtruth", events_observed)
 #     With-me-ness
 ############################################################################
 
-def get_previous_observation_start(start, timestamps):
-    for idx, t in enumerate(timestamps):
-        if start < t:
-            yield timestamps[idx-1] if (idx-1) >= 0 else timestamps[0]
+from numpy import arange
+
+dt = 0.1 # seconds
+
+t_end = events_expected[events_expected.keys()[-1]][1]
+
+def find_interval(t, intervals):
+    for ts, te in intervals:
+        if ts <= t and te >= t:
+            return ts, te
+    return None, None
 
 def withmeness(observed, expected):
 
     total_time = 0.
     total_time_matching = 0.
 
-    # we iterate primarily on the *expected* targets of attention
-    for idx, t_expected_start in enumerate(expected.keys()):
+    expected_intervals = [(k, v[1]) for k,v in events_expected.items()]
+    observed_intervals = [(k, v[1]) for k,v in events_observed.items()]
 
-        targets, t_expected_end = expected[t_expected_start]
+    for t in arange(0, t_end, dt):
+            expected_start, _ = find_interval(t, expected_intervals)
+            observed_start, _ = find_interval(t, observed_intervals)
 
-        total_time += t_expected_end - t_expected_start
+            if expected_start and observed_start:
+                total_time += dt
 
-        # then, for each time period, we iterate over the actual observation,
-        # starting at the last observation *before* the considered period.
-        for t_obs_start in get_previous_observation_start(t_expected_start, observed.keys()):
+                target = events_observed[observed_start][0][0]
+                expected_targets = events_expected[expected_start][0]
 
-            target, t_obs_end = observed[t_obs_start]
-
-            # if the observation starts *after* the considered period, we move
-            # to the next period, ie the next 'expected' targets
-            if t_obs_start > t_expected_end: 
-                break
-
-            # if the observation ends *before* the considered period (it may be
-            # the case, since we start with the last observation which started
-            # before the period), we skip it.
-            if t_obs_end < t_expected_start:
-                continue
-
-            # if the observed target is one of the expected one, add the actual
-            # observation time *for this period* to total_time_matching
-            if target[0] in targets:
-                print("Match at %s!" % t_obs_start)
-                start = max(t_expected_start, t_obs_start)
-                end = min(t_expected_end, t_obs_end)
-                total_time_matching += end - start
-            else:
[…]

### 2015-09-23 data/analysis/process.py
Sligthly stricter list of 'expected FoA' -> with-me-ness for subject 4 ~> 0.87

@@ -15,17 +15,21 @@ tier_expected = "Robot state"
 
 eaf = pympi.Elan.Eaf(elan_file)
 
-annotation2foa = {"Waiting for feedback": ("Tablet", "Secondary tablet", "Facilitator"),
+annotation2foa = {
+        ##### Robot state #####
+        "Waiting for feedback": ("Tablet", "Secondary tablet"),
         "Writing word": ("Tablet", "Robot"),
         "Waiting for word to write": ("Secondary tablet",),
         "Story telling": ("Robot",),
         "Presentation": ("Robot",),
+        "Bye": ("Robot",),
+        ##### Attention focus #####
         "robot": ("Robot",),
         "tablet": ("Tablet",),
         "facilitator": ("Facilitator",),
         "selection": ("Secondary tablet",),
         "observer": ("Observer",),
-        "paper": ("Secondary tablet",), # is that correct?
+        "paper": ("Secondary tablet",), 
         "other": ("Other",)}
 
 

### 2015-09-24 paper.tex
Global proof-reading (except sec 4.2)

--- 2015-09-22 paper.tex
@@
 there"~\cite{OBrien:2010}, has actively been studied in a diverse set of
-domains, and specifically in robotics, several variables and social signals have been proposed in the
-literature to quantify it. A recent review of these is in~\cite{ivaldi2015towards}.
+domains. Specifically in robotics, several variables and social signals have been proposed in the
+literature to quantify it. A recent review of these is presented in~\cite{ivaldi2015towards}.
 
@@
 In~\cite{Rich:2010}, a computational model based on the recognition of
-connection events such as directed gaze, mutual facial gaze is proposed. Not
+\emph{connection} events such as directed gaze, mutual facial gaze is proposed. Not
 relying on gaze, \cite{Sanghvi:2011} focus only on the back or trunk posture as
-a determining factor for the assessment. A recent study~\cite{anzalone} with
-social robots in face-to-face scenarios proposed a set of metrics based on
-non-verbal cues but manifesting some possible limitations in long-term
+a determining factor for the assessment. Dinally, a recent study with
+social robots in face-to-face scenarios~\cite{anzalone} explores a set of metrics based on
+non-verbal cues but they also underline possible limitations in long-term
 scenarios.
@@
 propose in this paper to introduce the more specific concept of
-``\emph{with-me-ness}'': to what extend the human remains ``with me'', the
-robot, during our interaction.
+``\emph{with-me-ness}'': to what extend the human is ``with me'', the
+robot, during the interaction.
 
@@
 attentional focus of the human with the expected, {\it a priori} focus
-requested by the task, as decided by the robot.
+requested by the task, as estimated by the robot.
 
-As such, this article is organized in three sections: we first present a novel
+This article is organized in three sections: we first present a novel
 method for on-line estimation of the focus of attention based on fast 6D head
-pose estimation. We then present an extensive validation of this approach based on
-a real-world field experiment with 6 children. We finally show how to compute in
-real-time ``with-me-ness'' to provide a new \emph{in-the-moment} measurement of
+pose estimation. We then propose to validate this approach with 
+a real-world field experiment with children. We finally show how to compute
+``with-me-ness'' over the course of the interaction to provide a new \emph{in-the-moment} measurement of
 the quality of interaction, and compare levels computed by the robot with a
@@
 
-The relations between one's focus of attention and what he/she is looking at has
+The relation between one's focus of attention and what he/she is looking at has
 long been established~\cite{yarbus1967eye,barber1976perception}, and more
@@
 
-Estimating attention using gaze is certainly not new to robotics either. A
+Estimating attention using gaze is not new to robotics either. A
 recent survey by Ruhland~\etal~\cite{ruhland2015review} presents in a synthetic
-way eye gaze research in HCI and social robotics, and is an essential read in
-that regard. It remains however an active field of research, as illustrated by
+way eye gaze research in HCI and social robotics. It remains however an active field of research, as illustrated by
 several recent publications~\cite{baxter2014tracking,anzalone,kennedy2015head}:
-as attention tracking is essential in human-robot interaction
-
-However, performing such a measure on a robot, in real-time, and in ecologically
+performing such a measure on a robot, in real-time, and in ecologically
 valid environments (which rules out bulky or invasive apparatus like
 eye-trackers, or techniques requiring fine calibration and/or static
-interactions) remains a key challenge in the field of Human-Robot Interaction.
+interactions) remains a challenge in HRI.
 
@@
 experiments, one finds that most approaches rely on head pose estimation alone
-(no eye gaze tracking) and are most often based on depth sensors (RGB-D) .
+(no eye gaze tracking) and are generally based on depth sensors (RGB-D) .
 Fanelli~\etal provides a overview of these approaches in~\cite{fanelli2012real},
@@
 milliseconds) facial features extraction on 2D images, combined with 3D model
-fitting, to obtain a fast, robust and stable 6D head pose estimate, that was
+fitting, to obtain a fast, robust and stable 6D head pose estimate, that we
 successfully deployed in field experiments involving child-robot interactions.
 
-We then derive the field of attention from the head pose: according
-to~\cite{stiefelhagen2002tracking} the head orientation's contribution in
-overall gaze direction is 68.9\% showing agreement between both measurements.
-In addition, it shows that the head orientation alone can get an average
-accuracy of 88.7\% detecting the focus of attention in a meeting application
-scenario using a head-mounted display with eye and head tracking.
+We derive the field of attention from the head pose: this is supported by
+previous work, like~\cite{stiefelhagen2002tracking} that shows that the head orientation's contribution in
+overall gaze direction is 68.9\%, which further translates into a 88.7\%
+accuracy in estimating the focus of attention from head pose only in a
+particular meeting scenario (using eye and head tracking).
 
-While previous preliminary research in HRI seemed to indicate that deriving
+While previous preliminary research in HRI seemed on the contrary to indicate that deriving
 attentional focus from head pose alone would not be accurate
 enough~\cite{kennedy2015head}, we found in our case acceptable levels of
-agreement between the robot observation and manual post-hoc annotations.
+agreement between the robot observations and manual post-hoc annotations, as
+detailled hereafter.
 
@@
 As explained, we derive the visual field of attention from the head pose. Our
-technique only involves a single monocular RGB camera for used for facial
-feature extraction, and a static simplified 3D mesh of a human head
-(Figure~\ref{head_pose}). 68 facial features are extracted using a fast
+technique only involves a single monocular RGB camera used for facial
+feature extraction, and a static simplified 3D mesh of a human head.
+68 facial features are extracted using a fast
 template-based face alignment algorithm by Kazemi and
 Sullivan~\cite{kazemi2014one}, as implemented in the open-source {\tt dlib}
-library\footnote{http://dlib.net}~\cite{dlib09}.  Eight of these features are
-matched to their 3D counterparts and we rely on an iterative $PnP$ algorithm
+library~\cite{dlib09}.  Eight of these features are
+then matched to their 3D counterparts (Figure~\ref{head_pose}) and we rely on an iterative $PnP$ algorithm
 (OpenCV's implementation) to compute the translation and rotation of the head
@@
 in terms of robustness, including in a range of difficult situations
-(occlusions, head angle) as encountered in field experiments.
+(occlusions, large head rotations) as encountered in field experiments.
 Figure~\ref{head_pose_real_world} shows a selection of such difficult scenes with
@@
     \caption{\small The visual field of attention is approximated to a
-        20$\degree$ cone, spanning from the head's sellion. The objects whose 3D
+        40$\degree$ cone, spanning from the head's sellion. The objects whose 3D
     pose intersect with this cone are considered \emph{in focus}.}
@@
 
-We model the field of attention as the central region of the field of view.
-The field of view itself is approximated to a cone spanned from the nasal
-depression (sellion) of the human face.
+We model the field of attention as the central region of the field of view.  The
+field of view itself is approximated to a cone spanned from the nasal depression
+(sellion) of the human face. Different dimensions for the human field of view
+can be found in the litterature: Holmqvist~\cite{holmqvist2011eye} models it
+with an horizontal aperture of $ \pm 40\degree $ and a vertical aperture of $
+\pm 25\degree $, while Spector~\cite{walker1980clinical} for instance suggests
+$60\degree$ up, $75\degree$ down, $60\degree$ inwards (towards the nose) and
+$95\degree$ outwards.  Previous work on visual perspective taking for social
+robotics~\cite{sisbot2011situation} model the field of attention as a cone of
+$30\degree$. We retained in this work a slightly wider aperture of $40\degree$.
 
-The dimensions of the human field of view have been measured at different level
-of accuracy. Holmqvist~\cite{holmqvist2011eye} models it with an horizontal
-aperture of $ \pm 40\degree $ and a vertical aperture of $ \pm 25\degree $,
-while Spector~\cite{walker1980clinical} for instance suggests $60\degree$ up,
-$75\degree$ down, $60\degree$ inwards (towards the nose) and $95\degree$
-outwards.  Previous work on visual perspective taking for social
-robotics~\cite{sisbot2011situation} simply use a cone of $30\degree$.  In this
-work, we follow Holmqvist and thus model the field of view as a cone with an elliptical
-cross-section of $80\degree$ by $50\degree$.
-
-We then approximate the visual focus of attention (VFoA) of the human to the
-objects which lay inside this field of view (Figure~\ref{fig:vfoa}). At a
+We then approximate the visual \emph{focus} of attention (VFoA) of the human to the
+objects which lay inside this field of attention (Figure~\ref{fig:vfoa}). At a
 given time, more than one object can therefore be \emph{in focus}.
 
-Our implementation has two limitations: objects are approximated to their origin
-(\ie objects are either in or out of the field of attention), and we do not
+Our implementation has two limitations: objects are approximated to points
+(their origin), and we do not
 check actual visibility: one object could be hidden by another, it would still
@@
 experimental setup (involving relatively small objects with no occlusions) did
-not require it. Techniques for a more complete assessment of the visual
-perspective can be found in~\cite{sisbot2011situation} for instance.
+not mandate it. Techniques for more accurate assessment of the visual
+perspective of the human peer can be found in~\cite{sisbot2011situation} for instance.
 
@@
 $A(X_A,Y_A,Z_A)$ into the frame of the face, and then to verify the following
-inequality (with $fov$ the aperture, and assuming that the main axis of the
-field of attention is aligned on the face's $\vec{x}$):
+simple inequality (with $fov$ the aperture, and assuming that the main axis
+$\vec{x}$ of the field of attention points forward):
 
@@
 
-Our approach assumes that the poses of the objects of interest are available to
+Our approach assumes that the pose of the objects of interest are available to
 the system: as described in section~\ref{sec:system}, our implementation relies
@@
 dedicated perception modules are in charged of publishing up-to-date
-informations regarding the location of the points of interest (the so-called
+informations regarding the location of the objects of interest (the so-called
 \emph{situation assessment}). Due to the nature of the experiment, most of the
 points of interest considered for the experimental validation presented
-hereafter are actually taken as static with respect to the robot.
+hereafter are static with respect to the robot, thus simplifying the scene
+perception.
 
@@
 
-As presented above, we use the (6D) head pose as an approximation of the
-actual gaze direction, and we approximate from here the participant's field of
-attention. The assumption that such an approximation of the field of attention
-allows to derive the actual focus of attention forms the hypothesis that we
-want to validate experimentally. The experiment involves child-robot
+As presented above, we use the 6D head pose as an approximation of the
+actual gaze direction, and we further approximate from here the participant's field of
+attention. The assumption that such an double approximation of the field of attention
+allows to derive the actual focus of attention needs to validated experimentally.
+Our proposed experiment involves child-robot
 interactions in the context of handwriting remediation.
@@
             nao} robot. The robot writes on the tactile tablet, the child then
-            corrects the robot by directly overwriting its letters on the tablet
+            corrects the robot by directly overwriting the words on the tablet
             with a stylus. The facilitator remains next to the child to guide the work. 
@@
 
+The subjects were typically located 50 cm away from the robot with the primary
+(writing) tablet in front and the secondary one 30 cm to the left of the first
+one.  The facilitator was located about 60 cm to the left of the subject.
+Finally, two observers were located further away from the interaction field.
+Figure~\ref{fig:setup} indicates accordingly the location of main areas of
+interest (the two tablets, the robot, the facilitator and the observers).
+
 The dependent variable is the measurement of the participants' VFoA, assessed in
@@
 %An RGB camera has been used to acquire images of 640x480 pixels at 25 fps. The
-%camera was located at 5 cm from the center of Nao's feet and its holder was tied
+%camera was located at 5 cm from the center of nao's feet and its holder was tied
 %to a wood base for stability purposes. Thus, the camera's objective was 9 cm
 %above and $ 40\degree $ inclination with respect to the surface of the table to
-%maximize the detection during the writing time. However, Nao's camera could have
+%maximize the detection during the writing time. However, nao's camera could have
 %been used, always considering the \textit{tf} frame matrix update according to
 %the head orientation.
-
-The subjects were typically located 50 cm away from the robot with the primary
-(writing) tablet in front and the secondary one 30 cm to the left of the first
-one.  The facilitator was located about 60 cm to the left of the subject.
-Finally, two observers were located further away from the interaction field.
-Figure~\ref{fig:setup} indicates accordingly the location of main areas of
-interest (the two tablets, the robot, the facilitator and the observers).
 
@@
 lab). The participants were chosen by the teacher, and would come one after the
-other for a face-to-face interaction with a {\sc Nao} robot ($M=19.6$ min, $SD=1.58$).
+other to interact with the robot (duration: $M=19.6$ min, $SD=1.58$).
 
-The experiment is organized in rounds of writing: during a typical round, the
+The interaction is organized in rounds of writing: during a typical round, the
 child requests the robot to write something (a single letter, a number or a full
-word), and pushes a tactile tablet (equipped with a custom application that we
-developed) towards the robot; the robot writes on the tablet by mimicking the
+word), and presents a tactile tablet (equipped with a custom writing
+application) to the robot; the robot writes on the tablet by mimicking the
 writing gesture (but without actually physically touching the tablet); the child
@@
 decides to train the robot on another word. In total, the children performed in
-average 12.16 ($SD=2.61$) rounds of writing.
+average 12.16 ($SD=2.61$) rounds of writing (complete details on the rationale
+and implementation of this experiment can be found in~\cite{Hood:2015}).
 
 Besides, once per interaction, the robot interrupts the handwriting task to
-propose to tell a story (that lasted for about 2 min), and the turn-based
-hand-writing task continued afterwards. The intended purpose of the
-story-telling episode was to locally trigger a different set of gaze behaviors.
-
-Details on the rationale and implementation of this experiment can be found
-in~\cite{Hood:2015}.
+propose to tell a story (taking about 2 min), and the turn-based
+hand-writing task continues afterwards. The intended purpose of the
+story-telling episode is to locally trigger a different set of attention behaviors.
 
@@
     \path (staticposes) edge [->] (static);
-    \path (static) edge [->] (tf);
+    \path (static) edge [->] node[label,below] {objects poses} (tf);
     \path (input) edge [->] (attention);
-    \path (attention) edge [->] (tf);
-    \path (pose) edge [->]  (tf);
+    \path (attention) edge [->] node[label,below] {child face} (tf);
+    \path (pose) edge [->] node[label,right] {robot pose} (tf);
     \path (tf) edge [->]  (focus);
@@
 
-The experiment was carried with an Aldebaran's {\sc Nao} robot, using ROS as a
+The experiment was carried with an Aldebaran's {\sc nao} robot, using ROS as a
 middleware to build the attention estimation pipeline (Figure~\ref{fig:system}).
@@
 over time (excluding the periods when the head is not detected) the
-overlap between the ground-truth and the robot's estimate, the
-inter-judge agreement (Cronbach's $\alpha$) and the cosine similarity
-matrix.
+overlap between the ground-truth and the robot's estimate and the
+inter-judge agreement (Cronbach's $\alpha$).
+%and the cosine similarity
+%matrix.
 
-The cosine similarity is a more reliable measurement between the ground-truth
-and the robot measure of the focus of attention as it accounts for temporal
-delays.\fixme{this needs to be better justified/explained.}.
+%The Cronbach's $\alpha$ can be defined as an average of the correlations between 
+%the items or values at the same time step \textit{t}. Therefore, results can vary 
+%significantly if both sets of values are time shifted. However,
+%the cosine similarity becomes a useful measurement in this case as it accounts 
+%for temporal delays. Specifically, the ground-truth values acquired though human 
+%annotations may not match the measurements of the focuses of attention computed 
+%by the robot at the exactly same time step \textit{t} even though the correspondent
+%curves may look very similar in shape.
 
@@
 
-The main results are reported in Table~\ref{tab:results} and
-Figure~\ref{fig:cosinematrix} (cosine similarity matrix);
-Figure~\ref{fig:realExpected} further gives a concrete picture of the
-ground-truth \vs computed attentional targets for subject 4 (the subject with
-the least successful tracking).
-
-\begin{table*}[h!]
+\begin{table*}[ht!]
     \centering
@@
     tracking} is the percentage of total time of successful detection of the
-    head pose; \emph{Ground-truth/robot match} is the percentage of matching time
+    head pose; \emph{Agreement} is the percentage of matching time
     between manually annotated focus of attention (ground-truth) and robot's
@@
         \midrule
-        {\bf Ground-truth/robot match} (\%) & 61.1 & 85.5 & 80.8 & 62.2 & 63.4 & 83.9 & {\bf 72.8} & {\it 11.7}\\
+        {\bf Agreement} (\%) & 61.1 & 85.5 & 80.8 & 62.2 & 63.4 & 83.9 & {\bf 72.8} & {\it 11.7}\\
         {\bf Cronbach's $\alpha$} & 0.63 & 0.81 & 0.82 & 0.59 & 0.64 & 0.78 & {\bf0.71} & {\it0.10}\\
-        {\bf Cosine affinity} & 0.73 & 0.82 & 0.87 & 0.73 & 0.77 & 0.89 & {\bf 0.8} & {\it 0.07} \\
+        %{\bf Cosine affinity} & 0.73 & 0.82 & 0.87 & 0.73 & 0.77 & 0.89 & {\bf 0.8} & {\it 0.07} \\
         \bottomrule
@@
 
-\begin{figure}[th!]
-    \centering
-    \includegraphics[width=0.7\columnwidth]{bitmap}
-    \caption{\small Cosine similarity matrix between ground-truth and robot-measured VFoA
-    using a 10 sec. window size. The correspondent numbers are shown
-    in figure \ref{fig:realCaptured}.}
-    \label{fig:cosinematrix}
-\end{figure}
+The main results are reported in Table~\ref{tab:results}
+%and Figure~\ref{fig:cosinematrix} (cosine similarity matrix);
+Figure~\ref{fig:realExpected} further gives a concrete picture of the
+ground-truth \vs computed attentional targets for subject 4 (the subject with
+the \emph{least} successful tracking).
 
-\begin{figure*}
+%
+%\begin{figure}[th!]
+%    \centering
+%    \includegraphics[width=0.7\columnwidth]{bitmap}
+%    \caption{\small Cosine similarity matrix between ground-truth and robot-measured VFoA
+%    using a 10 sec. window size. The correspondent numbers are shown
+%    in figure \ref{fig:realCaptured}.}
+%    \label{fig:cosinematrix}
+%\end{figure}
+
+\begin{figure*}[ht!]
     \centering
@@
 During the whole interaction, the head pose of the children was satisfactorily
-and consistently tracked (86\% of the time in average, $SD=3.0$), which is
-expected for a face-to-face interaction with a static head-tracking camera
+and consistently tracked (86\% of the time in average, $SD=3.0$). While this
+high score is expected for a face-to-face interaction with a static head-tracking camera
 (meaning that the child head would remain in the field of view of the camera
-most of the time). This is however comforting in terms of suitability of our
-approach for head pose estimation with children in field experiments.
-
-In our experiment, the primary causes of lost head pose were occlusions with the
+most of the time), this is still comforting in terms of suitability of our
+approach for head pose estimation with children in field experiments of this
+kind. Expectedly, the primary causes of lost head pose were occlusions with the
 hands (similar to the middle-bottom picture in
 Figure~\ref{head_pose_real_world}), close proximity with the tablet while
-writing, and gaze directed to the facilitator (sitting directly on the left of
+writing, and gaze directed to the facilitator (who was sitting directly on the left of
 the child, Figure~\ref{fig:realSetup}).
@@
 $\alpha$ close to 0.6), while for subjects 2, 3 and 6, we find a good agreement
-($\alpha$ close to 0.8). This does not correlate with the success rate of the
-head tracking itself, and, after further analysis of the videos, it rather
+($\alpha$ close to 0.8). After further analysis of the videos, it
 reflects that the first group of children rely more on their eyes to direct
-their gaze than the second group. However, and notwithstanding the small sample
-size, it is rather intriguing that the two groups appear so clearly separated.
-This certainly calls for further experimental investigation.
+their gaze than the second group. Note that, notwithstanding the small sample
+size, it is rather intriguing that the two groups appear so well distinct.
+This calls for further experimental investigation.
 
@@
     \If{$f(t) \in F(task(t))$}
-            \State $d_w \gets d_w + 1$
+            \State $d_w \gets d_w + \delta_t$
         \EndIf
-        \State $d_e \gets d_e + 1$
+        \State $d_e \gets d_e + \delta_t$
     \EndIf
-    \State $t \gets t + 1$
+    \State $t \gets t + \delta_t$
     \Until{$t = t_{end}$}
@@
     \caption{\textbf{Computation of \emph{with-me-ness}}. $d_w$ stands for the duration
-        the human is \emph{with} the robot, $d_e$ stands for the duration where
-        the human would be \emph{expected to be with} the robot, $task(t)$
+        the human is actually \emph{with} the robot, while $d_e$ stands for the
+        total time where the human would be \emph{expected to be with} the robot, $task(t)$
         represents the task performed by the robot at time $t$ (possibly none),
@@
         targets associated to task $task$, $f(t)$ represents the actual focus of
-        attention measured at time $t$. $\mathcal{W}_{[start, end]}$ represents
+        attention of the human measured at time $t$. $\mathcal{W}_{[start, end]}$ represents
         the level of \emph{with-me-ness} from $t_{start}$ to $t_{end}$.}
@@
 well as to its head. A human looking at either of these locations would be
-considered to be \emph{with} the robot in terms of interaction\footnote{As
-future work, this could easily be extended into a probabilistic model of
-attention expectations.}. Also notable, we exclude from the computation of
+considered to be \emph{with} the robot in terms of
+interaction\footnote{Considering a probabilistic model of
+attention expectations would be an interesting extension of this metric.}. Also notable, we exclude from the computation of
 $\mathcal{W}$ all of the periods of time where the user's focus of attention can
-not be estimated (Algorithm~\ref{alg:with-me-ness},
-line~\ref{algline:skiplosttrack}, third condition), typically because the user's
-face is not visible at those times.
+not be estimated (typically because the user's
+face is not visible at those times).
 
@@
 
-During the experiment presented above (section~\ref{sec:expe}), the robot
-controller associated a (set of) expected attentional targets to the task being
+Over the course of the experiment presented at section~\ref{sec:expe}, the robot
+controller would associate a set of expected attentional targets to the task being
 performed. For instance, while the robot was waiting for the child's handwriting
@@
 was supposed to write there), and while the robot was writing itself, the
-attentional targets we expected to be either on the tablet or on the robot.
+attentional targets were expected to be either on the tablet or on the robot.
 These expected targets, forming altogether the robot's attentional {\it a
@@
 with-me-ness by the robot \emph{over the course} of the interaction
-(\emph{in-the-moment} measurement). In this later case, one may typically want
+(\emph{in-the-moment} measurement) -- in this later case, one may typically want
 to consider a sliding time window.
@@
 of its interactant, and how we can, in turn, combine it with the robot's {\it a priori}
-about the interaction to build a straight-forward (yet novel in HRI) metric of
+about the interaction to build a metric of
 \emph{with-me-ness} over the course of the interaction.
@@
 
-The first issue is shared with most of the other vision (2D or 3D) or motion
-capture techniques for gaze estimation. In that regard, our approach does not
-perform worst or better than others. In fact, based on our results, relying
+While the first issue is shared with most of the other vision (2D or 3D) or motion
+capture techniques for real-time gaze estimation found in robotics,
+our results are positive: we shows that relying
 purely on head pose estimation to estimate gaze direction leads to real-world
-results that are worth being considered and used: they may not match manual
+measures that are worth being considered and used. They may not match manual
 annotations, but they are definitely a valuable \emph{in-the-moment} input for
-the robot, and we hope to see more works relying on those to build adaptive
-robot controllers in the coming years.
+the robot, reaching for certain children levels of accuracy traditionally
+considered as very good.
 
-In that regards, it is interesting to put this work on human-robot interaction
-in perspective with previous research conducted on ape-human interaction by
-Tomasello~\etal~\cite{tomasello2007reliance}: they show that apes rely primarily
+Interestingly, previous research conducted on ape-human interaction by
+Tomasello~\etal~\cite{tomasello2007reliance} shows that apes rely primarily
 on the head pose to estimate the focus of attention of a human partner, in to contrast
@@
 
-Besides, our approach relies on a simple, non-intrusive sensor (a RGB camera on
-the robot) and an open-source, fast pose estimation algorithm\fixme{test on Nao!
-how fast is it actually?}: we also hope that this may contribute to a widespread
+Our approach relies on a simple, non-intrusive sensor (a RGB camera on
+the robot) and an open-source, fast pose estimation algorithm\fixme{test on nao!
+how fast is it actually?}: we hope that this may contribute to a widespread
 adoption of such a technique on a range of robots, including the common {\sc
-Nao} platform.
+nao} platform.
 
@@
 
-Borrowing the neologism from the field of CSCL, we have also introduced here
+Borrowing the neologism from the field of CSCL, we have also introduced in this
+article
 \emph{with-me-ness} as a measure of ``how much the user in \emph{with} the robot during
-a task''. Interestingly, this can be measured over the course the interaction,
-thus providing the robot with a relatively high-level social construct,
+a task''. This can be acquired over the course the interaction,
+thus providing the robot with a real-time metric for relatively high-level social construct,
 undoubtedly related to engagement.
@@
 builds on a \emph{perceptual} alignment of two agents). \emph{Conceptual
-with-me-ness} as proposed by Sharma~\etal is on the other hand \emph{referential}: ``you are
+with-me-ness} as proposed by Sharma~\etal is on the contrary \emph{referential}: ``you are
 \emph{with} me if you focus on what I refer to, either explicitly or
-implicitly''. We propose to understand it here in a slightly broader way, that
+implicitly''. We propose to understand it here in an even slightly broader sense, that
 better reflect the \emph{interaction}: ``you are \emph{with} me if you focus on
@@
 On the other hand, with-me-ness is only a precursor of engagement: it
-does not say much about the cognitive commitment of a user to an interaction. A
+does not say much about the \emph{cognitive} commitment of a user to an interaction. A
 user may closely adhere to the injunctions of the robot (or, actually, of the
-experimenters) without being \emph{engaged} into the interaction. This is
+experimenters), with thus high levels of with-me-ness,  without being \emph{engaged} into the interaction. This is
 typically seen in child-robot interaction: children will attempt to closely
@@
 
-One of the strength of with-me-ness is its objectivity: we can formalize it, and
+Compared to engagement, one of the strength of with-me-ness is its specificity:
+it is well-defined, we can formalize it, and
 as such, it is valuable to assess and compare how users are willing or able to

### 2015-09-24 data/analysis/process.py
Added FoA datasets for all subjects + updated scripts to process them
Also added a very simple R script to compute agreements

@@ -1,135 +1,0 @@
-from collections import OrderedDict
-import pympi
-
-from simplesvg import Scene, Line
-
-
-############################################################################
-#     Processing of the EAF file
-############################################################################
-
-elan_file = "../4.eaf"
-
-tier_observed = "Attention focus"
-tier_expected = "Robot state"
-
-eaf = pympi.Elan.Eaf(elan_file)
-
-annotation2foa = {
-        ##### Robot state #####
-        "Waiting for feedback": ("Tablet", "Secondary tablet"),
-        "Writing word": ("Tablet", "Robot"),
-        "Waiting for word to write": ("Secondary tablet",),
-        "Story telling": ("Robot",),
-        "Presentation": ("Robot",),
-        "Bye": ("Robot",),
-        ##### Attention focus #####
-        "robot": ("Robot",),
-        "tablet": ("Tablet",),
-        "facilitator": ("Facilitator",),
-        "selection": ("Secondary tablet",),
-        "observer": ("Observer",),
-        "paper": ("Secondary tablet",), 
-        "other": ("Other",)}
-
-
-def get_time(ts):
-    """ Returns the timestamp of an EAF 'timeslot', in seconds
-    """
-    return eaf.timeslots[ts] / 1000.
-
-
-# First, convert all EAF timeslots into timestamp in seconds and 'normalize'
-# the focus of attention using the mappings in annotation2foa
-def prepare(tier):
-    rawevents = {}
-    for _, entry in eaf.tiers[tier][0].items():
-        ts, te, ann, _ = entry
-
-        if ann not in annotation2foa:
-            print("Unknown annotation: %s" % ann)
-            continue
-
-        rawevents[get_time(ts)] = (annotation2foa[ann], get_time(te))
-
-    # Order the 'events' by timestamps
-    return OrderedDict(sorted(rawevents.items(), key=lambda t: t[0]))
-
-events_observed = prepare(tier_observed)
-events_expected = prepare(tier_expected)
-
-############################################################################
-#     SVG rendering
-############################################################################
-
-plotting_order = {"Observer": 0,
-                  "Facilitator": 1,
-                  "Secondary tablet": 2,
-                  "Robot": 3,
-                  "Tablet": 4,
-                  "Other": 5}
-
-
-def plot(name, events):
-    svg = Scene(name)
-
-    for ts, evt in events.items():
-
-        targets, te = evt
-        for t in targets:
-            print("%s: %d to %d" % (t, ts, te))
-
-            # a different offset for each 'target'
[…]

### 2015-09-24 data/compute_agreements.R

@@ -1,0 +1,8 @@
+
+args <- commandArgs(trailingOnly = TRUE)
+
+obs.gt <- read.csv(args[1])
+library(irr)
+agree(obs.gt[,c("rating1","rating2")])
+kappa2(obs.gt[,c("rating1","rating2")])
+

### 2015-09-24 data/process.py

@@ -1,0 +1,265 @@
+from collections import OrderedDict
+import pympi
+
+from simplesvg import Scene, Line
+
+
+
+############################################################################
+#     Processing of the EAF file
+############################################################################
+
+tier_groundtruth = "Attention focus"
+tier_expected = "Robot state"
+
+
+LOST = "Lost track"
+
+annotation2foa = {
+        ##### Robot state #####
+        "Waiting for feedback": ("Tablet", "Secondary tablet"),
+        "Writing word": ("Tablet", "Robot"),
+        "Waiting for word to write": ("Secondary tablet",),
+        "Story telling": ("Robot",),
+        "Presentation": ("Robot",),
+        "Bye": ("Robot",),
+        ##### Attention focus #####
+        "robot": ("Robot",),
+        "tablet": ("Tablet",),
+        "facilitator": ("Facilitator",),
+        "selection": ("Secondary tablet",),
+        "observer": ("Observer",),
+        "paper": ("Secondary tablet",), 
+        "other": ("Other",),
+        "lost_track": (LOST,)}
+
+
+def get_time(ts):
+    """ Returns the timestamp of an EAF 'timeslot', in seconds
+    """
+    return eaf.timeslots[ts] / 1000.
+
+
+# First, convert all EAF timeslots into timestamp in seconds and 'normalize'
+# the focus of attention using the mappings in annotation2foa
+def prepare(eaf, tier):
+    rawevents = {}
+    for _, entry in eaf.tiers[tier][0].items():
+        ts, te, ann, _ = entry
+
+        if ann not in annotation2foa:
+            print("Unknown annotation: %s" % ann)
+            continue
+
+        rawevents[get_time(ts)] = (annotation2foa[ann], get_time(te))
+
+    # Order the 'events' by timestamps
+    return OrderedDict(sorted(rawevents.items(), key=lambda t: t[0]))
+
+############################################################################
+#     CSV files
+############################################################################
+
+
+def parse_robot_observations(robot_observations_file):
+
+    events_observed = OrderedDict()
+
+    with open(robot_observations_file,'r') as observations:
+        t_start = 0
+        current_target = None
+        last_t = 0
+        for l in observations.readlines():
+            a, target = l.split(":")
+            t_data = int(a)
+            target = target.strip()
+
+            if target not in annotation2foa:
+                print("Unknown annotation: %s" % target)
+                continue
+
+
[…]

### 2015-09-24 paper.tex
Updated all the results
Agreement is now correctly reported with a Cohen Kappa, and uses the 'raw' data computed by the robot

@@
 
-Successful detection of the head, and, when detected, the attentional target of
+Successful detections of the head, and, when detected, the attentional targets of
 the children as estimated by the robot, were logged during the experiment (in
-total, $6\times19.6=117.6$ min of interaction).\fixme{What it the case that 2
-focuses of attention were sometimes detected? what did we do in such a
-case?}
+total, $6\times19.6=117.6$ min of interaction). The only post-processing
+consisted in filtering out gaze shifts (short episodes -- below 500ms -- between
+two attentional targets).
 
@@
         \midrule
-        {\bf Agreement} (\%) & 61.1 & 85.5 & 80.8 & 62.2 & 63.4 & 83.9 & {\bf 72.8} & {\it 11.7}\\
-        {\bf Cronbach's $\alpha$} & 0.63 & 0.81 & 0.82 & 0.59 & 0.64 & 0.78 & {\bf0.71} & {\it0.10}\\
+        {\bf Agreement} (\%) & 58.9 & 67.1 & 79.2 & 48.3 & 65 & 77.1 & {\bf 65.9} & {\it 11.5}\\
+        {\bf Cohen's $\kappa$} & 0.48 & 0.56 & 0.68 & 0.26 & 0.47 & 0.68 & {\bf 0.52} & {\it 0.16}\\
         %{\bf Cosine affinity} & 0.73 & 0.82 & 0.87 & 0.73 & 0.77 & 0.89 & {\bf 0.8} & {\it 0.07} \\
@@
     ground-truth (\ie \emph{ground-truth with-me-ness}), or based on the focus of
-    attention measured by the robot.}
+    attention measured by the robot. $\mathcal{W}_{robot}$ is strongly correlated to
+    $\mathcal{W}_{groundtruth}$: Pearson's $r(4)=0.50$.}
 
@@
         \midrule
-        $\mathcal{W}_{g.truth}$ & 92.6 & 88.8  & 95.8 & 93.1 & 96.5 & 89.6 & {\bf 92.7} & {\it 3.1} \\ 
-        $\mathcal{W}_{robot}$ & 89.6 & 84.4 & 86.6 & 72.8 & 81.9 & 79.7 & {\bf 82.5} & {\it 5.4} \\
+        $\mathcal{W}_{g.truth}$ & 79.3 & 82.6  & 90.5 & 87.9 & 90.7 & 83.6 & {\bf 85.8} & {\it 4.6} \\ 
+        $\mathcal{W}_{robot}$ & 52.5 & 55.5 & 74.1 & 52.6 & 60.0 & 66.4 & {\bf 60.2} & {\it 8.6} \\
         \bottomrule

### 2015-09-28 paper.tex
Updated result discussion following use of Cohen kappa

@@
         in table~\ref{tab:results}, 3.5min-long excerpt).
-        In blue (top line), the focus of attention as computed by the robot;
-        in orange (bottom line), the focus of attention as manually annotated
-        (ground-truth). The bottom graph shows agreement between both (whenever
+        In blue (top lines), the focus of attention as computed by the robot;
+        in orange (bottom lines), the focus of attention as manually annotated
+        (ground-truth). Periods where the head is lost are marked with dashed
+        vertical lines. The bottom graph shows agreement between both (whenever
         the head is detected).}
@@
 
-In terms of attention tracking, two groups of children stand out of
-Table~\ref{tab:results}: the robot's estimate of the focus of attention for
-subjects 1, 4 and 5 has questionable agreement with the ground-truth (Cronbach's
-$\alpha$ close to 0.6), while for subjects 2, 3 and 6, we find a good agreement
-($\alpha$ close to 0.8). After further analysis of the videos, it
-reflects that the first group of children rely more on their eyes to direct
-their gaze than the second group. Note that, notwithstanding the small sample
-size, it is rather intriguing that the two groups appear so well distinct.
-This calls for further experimental investigation.
+In terms of attention tracking, Cohen's $\kappa$ values are between 0.47 and
+0.68 with one subject resulting in significantly worst tracking, at 0.26. While
+the interpretation of Cohen's $\kappa$ is subject to discussions (the number of
+the coded values -- in our case 6 -- and the distribution probability of values
+-- in our case, values are \emph{not} equiprobable -- are factors impacting
+$\kappa$ independently of the level of agreement), the levels of agreement are
+\emph{moderate} to \emph{substantial}, with one subject only
+showing \emph{fair} agreement~\cite{landis1977measurement}. Further analysis of
+the videos shows that the child with the lowest level of agreement is especially
+calm and would rely more on the eyes to direct his gaze than the other children,
+thus leading to a less accurate estimation of its focus of attention.
 

### 2015-09-28 paper.tex
Updated with-me-ness figure. Should be mostly final

@@
     \caption{\small \textbf{With-me-ness}. Evolution of the level of
-        \emph{with-me-ness} over the whole $\approx$20min long interaction of
-        subject 4. \emph{Interaction Episodes} describe the distinct phases of
-        the interaction, and are discussed in the text. The central chart is
-        similar to Figure~\ref{fig:realExpected} with the \emph{expected}
-        attentional targets added in green. The bottom diagram represents the
-        levels of with-me-ness over successive windows of 30 seconds. The blue
-        line is with-me-ness as estimated by the robot, the orange line is
-        with-me-ness computed from manually annotated attentional targets. The
-        ``holes'' in the with-me-ness computed by the robot correspond to period of time
-        where the child's face is not detected.}
+        \emph{with-me-ness} over the whole $\approx$17min long interaction of
+        subject 2. The top chart is similar to Figure~\ref{fig:realExpected}
+        with the \emph{expected} attentional targets added in green. The bottom
+        diagram represents the instantaneous level of with-me-ness over a
+        sliding window of 30 seconds. The blue line is with-me-ness as estimated
+        by the robot, the orange line is with-me-ness computed from manually
+        annotated attentional targets.  Pearson's correlation between both: $r(973)=0.58, p =
+        2.2e^{-16}.$}
 
@@
     ground-truth (\ie \emph{ground-truth with-me-ness}), or based on the focus of
-    attention measured by the robot. $\mathcal{W}_{robot}$ is strongly correlated to
-    $\mathcal{W}_{groundtruth}$: Pearson's $r(4)=0.50$.}
+    attention measured by the robot. Pearson's correlation: $r(4)=0.46$.}
 
@@
         \midrule
-        $\mathcal{W}_{g.truth}$ & 79.3 & 82.6  & 90.5 & 87.9 & 90.7 & 83.6 & {\bf 85.8} & {\it 4.6} \\ 
-        $\mathcal{W}_{robot}$ & 52.5 & 55.5 & 74.1 & 52.6 & 60.0 & 66.4 & {\bf 60.2} & {\it 8.6} \\
+        $\mathcal{W}_{g.truth}$ & 79.4 & 81.6  & 90.5 & 87.9 & 90.7 & 80.9 & {\bf 85.2} & {\it 5.1} \\ 
+        $\mathcal{W}_{robot}$ & 52.6 & 55.3 & 74.3 & 52.9 & 59.5 & 63.9 & {\bf 59.8} & {\it 8.3} \\
         \bottomrule

### 2015-09-28 data/process.py
Lastest version of the scripts to compute/plot withmeness

@@ -1,7 +1,7 @@
 from collections import OrderedDict
 import pympi
 
-from simplesvg import Scene, Line
+from simplesvg import Scene, Line, Polyline
 
 
 
@@ -93,7 +93,7 @@ def parse_robot_observations(robot_observations_file):
 
     return events_observed
 
-def filter_observations(obs, min_duration = 0.3):
+def filter_observations(obs, min_duration):
 
     filtered = {}
 
@@ -187,9 +187,15 @@ def find_interval(t, intervals):
             return ts, te, target
     return None, None, None
 
-def withmeness(observations, expectations, name = None):
+def withmeness(observations, expectations, name = None, t_start = None, t_end = None):
 
-    t_end = observations[observations.keys()[-1]][1]
+    if t_end is None:
+        t_end = observations[observations.keys()[-1]][1]
+    else:
+        t_end = min(t_end, observations[observations.keys()[-1]][1])
+
+    if t_start is None:
+        t_start = 0
 
     total_time = 0.
     total_time_matching = 0.
@@ -202,11 +208,12 @@ def withmeness(observations, expectations, name = None):
         csv = open(name + ".csv", "w")
         csv.write("t,rating1,rating2\n")
 
-    for t in arange(0, t_end, dt):
+    for t in arange(t_start, t_end, dt):
+
             expected_start, _, e_target = find_interval(t, expectations_intervals)
             observations_start, _, o_target = find_interval(t, observations_intervals)
 
-            if expected_start and observations_start:
+            if e_target and o_target:
                 target = observations[observations_start][0][0]
 
                 if target != LOST:
@@ -223,10 +230,38 @@ def withmeness(observations, expectations, name = None):
     if csv:
         csv.close()
 
+    if total_time == 0:
+        import pdb;pdb.set_trace()
+        return None
+
     withmeness = total_time_matching/total_time
     print("Total time: %s sec, total time matching: %s sec, With-me-ness: %s" % (total_time, total_time_matching, withmeness))
 
     return withmeness
+
+def plot_withmeness(name, observations, expectations, sliding_window = 30):
+    svg = Scene(name)
+
+    t_end = observations[observations.keys()[-1]][1]
+
+    points = []
+
+    delta = 1
+    #delta = sliding_window
+
+    for t_start in arange(0, t_end-sliding_window+1, delta):
+
+            w = withmeness(observations, expectations, name = None, t_start=t_start, t_end=(t_start + sliding_window))
[…]

### 2015-09-28 data/simplesvg.py

@@ -47,8 +47,20 @@ class Line:
                 (self.start[0],self.start[1],self.end[0],self.end[1])]
 
 
+class Polyline:
+    def __init__(self,points, color = None):
+        self.points = points # [(x,y), (x,y)...]
+        self.color = color
+        return
+
+    def strarray(self):
+        return ["  <path d=\"M %s\"\n" %\
+                " ".join(["%d,%d" % (x,y) for x,y in self.points]),
+                "    style=\"fill:%s;\"  />\n" % colorstr(self.color)]
+
+
 class Circle:
-    def __init__(self,center,radius,color):
+    def __init__(self,center,radius,color = None):
         self.center = center #xy tuple
         self.radius = radius #xy tuple
         self.color = color   #rgb tuple in range(0,256)
@@ -60,7 +72,7 @@ class Circle:
                 "    style=\"fill:%s;\"  />\n" % colorstr(self.color)]
 
 class Rectangle:
-    def __init__(self,origin,height,width,color):
+    def __init__(self,origin,height,width,color = None):
         self.origin = origin
         self.height = height
         self.width = width
@@ -87,7 +99,10 @@ class Text:
                 "  </text>\n"]
         
     
-def colorstr(rgb): return "#%x%x%x" % (rgb[0]/16,rgb[1]/16,rgb[2]/16)
+def colorstr(rgb):
+    if rgb is None:
+        return "none"
+    return "#%x%x%x" % (rgb[0]/16,rgb[1]/16,rgb[2]/16)
 
 def test():
     scene = Scene('test')

### 2015-09-28 paper.tex
Put Severin as first author, to reflect the actual writing of the article

@@
 \title{From Real-time Attention Assessment to ``With-me-ness'' in Human-Robot Interaction}
-\author{Fernando Garcia \qquad Séverin Lemaignan \qquad Alexis Jacq \qquad Pierre Dillenbourg\\Computer-Human Interaction in Learning and Instruction Laboratory (CHILI)\\École Polytechnique Fédérale de Lausanne (EPFL)}
+\author{Séverin Lemaignan \qquad Fernando Garcia \qquad Alexis Jacq \qquad Pierre Dillenbourg\\Computer-Human Interaction in Learning and Instruction Laboratory (CHILI)\\École Polytechnique Fédérale de Lausanne (EPFL)}
 

### 2015-09-28 paper.tex
Rewritten discussion of with-me-ness results

@@
 interaction\footnote{Considering a probabilistic model of
-attention expectations would be an interesting extension of this metric.}. Also notable, we exclude from the computation of
+attention expectations (an attention distribution) would be an interesting
+extension of this metric.}. Also notable, we exclude from the computation of
 $\mathcal{W}$ all of the periods of time where the user's focus of attention can
-not be estimated (typically because the user's
-face is not visible at those times).
+not be estimated (typically because the user's face is not visible at those
+times).
 
-\subsection{Experimental Interpretation}
+\subsection{Experimental Measure \& Interpretation}
+
+\begin{table}[t]
+    \centering
+    \caption{Mapping between the interaction phases and the expected
+        attentional targets.}
+
+    \begin{tabular}{ll}
+        \toprule
+        {\bf Phase} & {\bf Expected targets} \\
+        \midrule
+        Presentation & {\sf robot} \\ 
+        Waiting for word to write & {\sf secondary tablet} \\ 
+        Writing word & {\sf tablet}, {\sf robot} \\ 
+        Waiting for feedback & {\sf tablet}, {\sf secondary tablet} \\ 
+        Story telling & {\sf robot} \\ 
+        Bye & {\sf robot} \\ 
+        \bottomrule
+    \end{tabular}
+    \label{tab:targets}
+\end{table}
 
 Over the course of the experiment presented at section~\ref{sec:expe}, the robot
-controller would associate a set of expected attentional targets to the task being
-performed. For instance, while the robot was waiting for the child's handwriting
-demonstration, the expected attentional target was the tablet (since the child
-was supposed to write there), and while the robot was writing itself, the
-attentional targets were expected to be either on the tablet or on the robot.
-These expected targets, forming altogether the robot's attentional {\it a
-priori}, are depicted on Figure~\ref{fig:with-me-ness} as the green lines. They
-are used to compute the with-me-ness. With-me-ness can be calculated over the
-whole interaction (Table~\ref{tab:results-with-me-ness}) or over shorter time
-windows (bottom graph on Figure~\ref{fig:with-me-ness}). Shorter time windows
-are interesting for two purposes: to analyse the level of with-me-ness in
-relation to specific interaction episodes; to allow a measurement of
-with-me-ness by the robot \emph{over the course} of the interaction
-(\emph{in-the-moment} measurement) -- in this later case, one may typically want
-to consider a sliding time window.
+controller would associate a set of expected attentional targets to the phase of
+the interaction (Table~\ref{tab:targets}). For instance, while the robot was
+waiting for the child's handwriting demonstration (``\textit{Waiting for
+feedback}''), the expected attentional target of the child was the tablet (since
+the child was supposed to write there) or the secondary tablet (that displayed a
+template of the word, used as reference by the child). These expected targets
+form altogether the robot's attentional {\it a priori} and are used to compute
+the with-me-ness.  They are represented over time by the green lines on
+Figure~\ref{fig:with-me-ness}. With-me-ness can be calculated over the whole
+interaction (see corresponding results in Table~\ref{tab:results-with-me-ness})
+or over shorter time windows. Shorter time windows are interesting for two
+purposes: to analyse the level of with-me-ness in relation to specific
+interaction episodes; to allow a measurement of with-me-ness by the robot
+\emph{over the course} of the interaction (\emph{in-the-moment} measurement) --
+in this later case, one may typically want to consider a sliding time window.
 
@@
 
-Figure~\ref{fig:with-me-ness} can be further analyzed:\fixme{This paragraph
-needs to be adapted}
-The matching curve (cyan color) shows the overlapping between the
-expected and the real focuses of attention. As can be noticed there is a
-fluctuation at the beginning of the interaction as well as just before the
-activity switching. These chaotic gaze manifestation is what we call an
-automatic discovery of events, when the user experiences an uncertainty in the
-present situation.
+The with-me-ness plotted at the bottom of Figure~\ref{fig:with-me-ness} is in
+fact computed on a sliding window of 30 seconds, and thus gives a picture of
+``how well the child is following the robot's expectations'' at that time. As
+seen, the with-me-ness computed at run-time by the robot (blue line) is
+generally lower than the ground-truth (orange line, based on video-annotations),
+and sometimes quite off, such as during episode marked ``A'': during that phase,
+one can notice that the attention is mostly directed to undefined target {\sf
+Other}, likely a consequence of inaccurate head detection.  This kind of error
+(inaccurate head pose estimation) is the main source of discrepancy between the
+ground-truth and the attention distribution measured by the robot: ignoring all
+the episodes where the child's gaze is measured to be directed to {\sf Other},
+we indeed obtain levels of with-me-ness close to the ground-truth (over the six
+subjects, $M=87.5, SD=4.6$).
 
-The first one of the previous two situations presented above is due to the
-novelty effect the subjects experience for the first time in front of the system
-and according to the rest of participants suggests that the adaptation time is
-around the first 2 minutes of interaction. In the second case, an unexpected
-robot behavior produces such reaction due to an activity switching, the story
-telling.
+A chart like Figure~\ref{fig:with-me-ness} remains an useful tool to analyse the
+interaction, and several observations can be made out of it: as said, the green
+lines represent how the robot imagine, at a given time, the attention
+distribution of the child. They also provide an accurate picture of the overall
+turn-taking as viewed by the robot: for instance, the episode ``B'' on
+Figure~\ref{fig:with-me-ness} corresponds to one of the ``\emph{Robot writing}''
+episodes, surrounded by ``\emph{Waiting for feedback}'' phases like ``C''; episode
+``D'' corresponds to the story telling; etc. In terms of interaction, the large
+variance of the duration of these phases reflects the fact that this child would
+sometimes take a lot of time to send feedback to the robot, and sometimes, on
+the contrary, be very quick.
 
-It is important to consider that in some occasions during the interaction, none
-of the focuses of attention were within the field of view due to the time
-transition from one focus to another resulting in an ambiguity. In order to deal
-with these few cases it is assumed that the minimum distance between the height
-of the field of view (cone) and the targets corresponds to the most likely focus
-of attention. However, we also want to distinguish such cases when the subject
-is directly not looking where the defined targets are located, but towards a
-region far from the interaction field. These cases are not common but they can
-be easily identified by setting a horizontal threshold to the field of view.
+Looking at the ground-truth focus of attention (orange lines), the first
+striking observation is that this child did generally \emph{closely} follow what
+the robot was expecting: in that regard, it seems that the child was very much
+engaged into the interaction (we discuss in the next section the exact
+relationships between with-me-ness and engagement). The only major exception is
+the story-telling phase (episode ``D''): the child was seemingly not captivated by the
+first half of the story, and his attention was not directed towards the robot.
 
-A relevant trend that can be observed is the continuous `look for approval'
-situation where the subjects look at the experimenter in order to provide
-feedback about the robot's response before providing their own one (see yellow
-boxes in figure \ref{fig:realExpected}). Such transition tablet-experimenter
-becomes more frequent in several cases, for instance when the experimenter makes
-a suggestion (orange box in figure \ref{fig:realExpected}), in this case, switch
-to a more complicated word.
+Another interesting observation pertains to the facilitator: as one can see,
+this child only rarely turned to the facilitator, indicating that the
+interaction and the task were meaningful and easy for him to follow alone.
 
-Once the turn-taking is established, a pattern becomes visible during the
-subject's feedback, when an attempt is performed: Such pattern shows a high
-intermittent gaze frequency from the tablet to the selection. It also reveals
-that the subject is using the demonstration model in the selection tablet to
-provide a better answer to the robot. In fact, subjects decrease the response
-time after each shift as well as the gaze frequency towards the model suggesting
-an assimilation over time. Moreover, when a new demonstration is selected for
-playing, the same behavior is replicated (see yellow boxes in figure
-\ref{fig:realExpected}).
+More subtle patterns and events can also be observed: for instance, during the
+feedback phases like episode ``C'', we
+can notice numerous gaze shifts between the tablet (where the child write) and the
+secondary tablet (that showed a template of the word). The episode ``B'' (robot
+writing) is also interesting: the child did not look at the robot, and instead
+remained focused on the secondary tablet. This situation is typically useful for
+the robot to detect as it may want to adapt its behavior to recover the child's
+attention.
 
-If we consider the case shown in section \ref{sec:fromFaceTo}, evaluating the
-subject's progression together with the time on task may be an indicator of
-engagement: If the subject answers so quickly with respect to the previous
-attempts with a poorer handwriting may be the case she is not engaged anymore.
-In such hypothetical case an activity switch can be easily implemented in a
-robot architecture. It confirms the findings reported in \cite{anzalone} where
-time response to event is proposed to account quality of interaction. 
+%A relevant trend that can be observed is the continuous `look for approval'
+%situation where the subjects look at the experimenter in order to provide
+%feedback about the robot's response before providing their own one (see yellow
+%boxes in figure \ref{fig:realExpected}). Such transition tablet-experimenter
+%becomes more frequent in several cases, for instance when the experimenter makes
+%a suggestion (orange box in figure \ref{fig:realExpected}), in this case, switch
+%to a more complicated word.
+%
+%Once the turn-taking is established, a pattern becomes visible during the
+%subject's feedback, when an attempt is performed: Such pattern shows a high
+%intermittent gaze frequency from the tablet to the selection. It also reveals
+%that the subject is using the demonstration model in the selection tablet to
+%provide a better answer to the robot. In fact, subjects decrease the response
+%time after each shift as well as the gaze frequency towards the model suggesting
+%an assimilation over time. Moreover, when a new demonstration is selected for
+%playing, the same behavior is replicated (see yellow boxes in figure
+%\ref{fig:realExpected}).
+
+%If we consider the case shown in section \ref{sec:fromFaceTo}, evaluating the
+%subject's progression together with the time on task may be an indicator of
+%engagement: If the subject answers so quickly with respect to the previous
+%attempts with a poorer handwriting may be the case she is not engaged anymore.
+%In such hypothetical case an activity switch can be easily implemented in a
+%robot architecture. It confirms the findings reported in \cite{anzalone} where
+%time response to event is proposed to account quality of interaction. 
 %
@@
 
-It is also important to notice that the initiative of the robot into the
-learning process in a turn-taking activity produces an effect in the engagement
-of the human partner. However, experience suggests that a robot that learns
-evoke a positive action in the partner. But the opposite, a robot that is not
-able to progress over time, produces the contrary effect.
+%It is also important to notice that the initiative of the robot into the
+%learning process in a turn-taking activity produces an effect in the engagement
+%of the human partner. However, experience suggests that a robot that learns
+%evoke a positive action in the partner. But the opposite, a robot that is not
+%able to progress over time, produces the contrary effect.
 
@@
 
-On the other hand, with-me-ness is only a precursor of engagement: it
-does not say much about the \emph{cognitive} commitment of a user to an interaction. A
+On the other hand, with-me-ness is only a precursor of engagement: it does not
+say much about the \emph{cognitive} commitment of a user to an interaction. A
 user may closely adhere to the injunctions of the robot (or, actually, of the
-experimenters), with thus high levels of with-me-ness,  without being \emph{engaged} into the interaction. This is
-typically seen in child-robot interaction: children will attempt to closely
-follow what they are asked to do -- which may \emph{look like} they are engaged
-into the interaction -- while they merely \emph{obey orders}.
+experimenters), with thus high levels of with-me-ness,  without being
+\emph{engaged} into the interaction. This is typically seen in child-robot
+interaction: children will attempt to closely follow what they are asked to do
+-- which may \emph{look like} they are engaged into the interaction -- while
+they merely \emph{obey orders}.
 
 Compared to engagement, one of the strength of with-me-ness is its specificity:
-it is well-defined, we can formalize it, and
-as such, it is valuable to assess and compare how users are willing or able to
-interact with a robot. Besides, we have hopefully demonstrated in this article
-that with-me-ness is an operational \emph{in-the-moment} metric that can be also
-used as feed-back to the robot controller to build richer, more adaptive
-interactive behaviors for our robots.
+it is well-defined (the only important variable -- that needs to be reported --
+is the mapping {\it phase/expected attentional target}, \ie our
+Table~\ref{tab:targets}), we can formalize it, and as such, it is valuable to
+assess and compare how users are willing or able to interact with a robot.
+Besides, we have hopefully demonstrated in this article that with-me-ness is an
+operational \emph{in-the-moment} metric that can be also used as feedback to the
+robot controller to build richer, more adaptive interactive behaviors for our
+robots.
 

### 2015-09-29 paper.tex
Proof-reading by James -- Part 1

@@
 
-Building capable social agents requires to equip them with a range
-of perceptual capabilities: while face and object tracking and
+Building capable social agents requires to endow them with a range
+of perceptual capabilities: however, while face and object tracking and
 recognition, path planning, speech recognition or task learning are some of the current major research
@@
 their own metric space, without considering the interaction quality at a global
-level: Anzalone~\etal~\cite{anzalone} have recently argued in favor of
+level. Anzalone~\etal~\cite{anzalone} have recently argued in favor of
 such a global evaluation, and they propose to assess these algorithms in term of
 their capability to \emph{obtain the desired effect} in a human-robot interaction
-context. They correspondingly propose metrics build around the measurement
+context. They correspondingly propose metrics built around the measurement
 of \emph{engagement} as indicator of the quality of the experience.
@@
 For instance, \cite{Castellano:2009} proposes to predict children's level of
-engagement by integrating in a Bayesian model non-verbal cues (gaze and smiles)
-with the current state of the interaction. While they report high level of
+engagement by integrating non-verbal cues (gaze and smiles)
+with the current state of the interaction in a Bayesian model. While they report high level of
 accuracy, their approach requires post-hoc video annotations, and is not
@@
 relying on gaze, \cite{Sanghvi:2011} focus only on the back or trunk posture as
-a determining factor for the assessment. Dinally, a recent study with
+a determining factor for the assessment. Finally, a recent study with
 social robots in face-to-face scenarios~\cite{anzalone} explores a set of metrics based on
@@
 Therefore, instead of introducing ``yet another metric of engagement'', we
-propose in this paper to introduce the more specific concept of
-``\emph{with-me-ness}'': to what extend the human is ``with me'', the
+introduce the more specific concept of
+\emph{with-me-ness}~\cite{sharma2014me}: to what extent the human is ``with me'', the
 robot, during the interaction.
 
-We assess this level of ``with-me-ness'' by combining real-time estimation of the
-attentional focus of the human with the expected, {\it a priori} focus
-requested by the task, as estimated by the robot.
+We measure \emph{with-me-ness} by comparing the
+attentional focus of the human (as estimated in real-time by the robot) with the
+expected, {\it a priori} targets of attention elicited by the task at hand.
 
@@
 method for on-line estimation of the focus of attention based on fast 6D head
-pose estimation. We then propose to validate this approach with 
+pose estimation. We then validate this approach with 
 a real-world field experiment with children. We finally show how to compute
@@
 interaction, and the related gaze patterns, has been part of classic textbooks
-like~\cite{argyle1969social} since decades. As such, there is little doubt that
+like~\cite{argyle1969social} for decades. As such, there is little doubt that
 measuring the direction of gaze is a useful proxy to estimate the (visual) focus
-of attention of a social agent, and indeed this is one of the basic tool used in
+of attention of a social agent, and indeed this is one of the basic tools used in
 social psychology.
@@
 Estimating attention using gaze is not new to robotics either. A
-recent survey by Ruhland~\etal~\cite{ruhland2015review} presents in a synthetic
-way eye gaze research in HCI and social robotics. It remains however an active field of research, as illustrated by
-several recent publications~\cite{baxter2014tracking,anzalone,kennedy2015head}:
-performing such a measure on a robot, in real-time, and in ecologically
+recent survey by Ruhland~\etal~\cite{ruhland2015review} gives in a broad
+overview of eye gaze research in HCI and social robotics. It remains however an active field of research, as illustrated by
+several recent publications~\cite{baxter2014tracking,anzalone,kennedy2015head}.
+Performing such a measure on a robot, in real-time, and in ecologically
 valid environments (which rules out bulky or invasive apparatus like
@@
 (no eye gaze tracking) and are generally based on depth sensors (RGB-D) .
-Fanelli~\etal provides a overview of these approaches in~\cite{fanelli2012real},
+Fanelli~\etal provides an overview of these approaches in~\cite{fanelli2012real},
 and recent examples include~\cite{baxter2014tracking, anzalone}.
@@
 occlusions or lightning conditions, and over-reliance on tracking to maintain
-real-time performances.  Our work relies on recent advances on template-based
+real-time performances.  Our work relies on recent advances in template-based
 face alignment~\cite{kazemi2014one} that allows fast (in the order of a few
-milliseconds) facial features extraction on 2D images, combined with 3D model
+milliseconds) facial feature extraction on 2D images, combined with 3D model
 fitting, to obtain a fast, robust and stable 6D head pose estimate, that we
@@
     \includegraphics[width=0.9\columnwidth]{field_of_attention}
-    \caption{\small The visual field of attention is approximated to a
+    \caption{\small Screenshot of the real-time attention estimation system. The visual field of attention is approximated to a
         40$\degree$ cone, spanning from the head's sellion. The objects whose 3D
@@
 We then approximate the visual \emph{focus} of attention (VFoA) of the human to the
-objects which lay inside this field of attention (Figure~\ref{fig:vfoa}). At a
+objects which lie inside this field of attention (Figure~\ref{fig:vfoa}). At a
 given time, more than one object can therefore be \emph{in focus}.
@@
 Our implementation has two limitations: objects are approximated to points
-(their origin), and we do not
+(they are considered in focus if their \emph{origin} lies in the field of
+attention), and we do not
 check actual visibility: one object could be hidden by another, it would still
@@
 experimental setup (involving relatively small objects with no occlusions) did
-not mandate it. Techniques for more accurate assessment of the visual
+not necessitate it. Techniques for more accurate assessment of the visual
 perspective of the human peer can be found in~\cite{sisbot2011situation} for instance.
@@
 modules the list of poses of existing objects (represented as {\it frames}), and
-dedicated perception modules are in charged of publishing up-to-date
+dedicated perception modules are in charge of publishing up-to-date
 informations regarding the location of the objects of interest (the so-called
@@
 actual gaze direction, and we further approximate from here the participant's field of
-attention. The assumption that such an double approximation of the field of attention
+attention. The assumption that such an approximation of the field of attention
 allows to derive the actual focus of attention needs to validated experimentally.
@@
 robot's handwriting). A tactile tablet is used as writing support.
+
+Figures~\ref{fig:setup} and~\ref{fig:realSetup} illustrate the experimental setup: a face-to-face
+child-robot interaction with an (autonomous) Aldebran's {\sc nao} robot, in the
+presence of a facilitator (one of the researchers).
 
@@
 
-Figures~\ref{fig:setup} and~\ref{fig:realSetup} illustrate the experimental setup: a face-to-face
-child-robot interaction with an (autonomous) Aldebran's {\sc nao} robot, in
-presence of a facilitator (one of the researchers).
-
 \begin{figure}[h!]
@@
 one.  The facilitator was located about 60 cm to the left of the subject.
-Finally, two observers were located further away from the interaction field.
+Finally, two observers (visible by the child) were located further away from the interaction field.
 Figure~\ref{fig:setup} indicates accordingly the location of main areas of
@@
 The dependent variable is the measurement of the participants' VFoA, assessed in
-terms of what are the attentional targets of the child over time. The face of the child
+terms of what the attentional targets of the child are over time. The face of the child
 is acquired through a fixed webcam (Logitech {\sf c920}), placed on the table
@@
 
-Six children (ages 5 to 6, 3 boys, 3 girls) were enrolled for this study.
-The study took place at school, in a separate room (the computer
+Six children (ages 5 to 6, 3 boys, 3 girls, none wearing glasses) were enrolled for this study.
+The study took place at school, in an isolated room (the computer
 lab). The participants were chosen by the teacher, and would come one after the
@@
 The interaction is organized in rounds of writing: during a typical round, the
-child requests the robot to write something (a single letter, a number or a full
+child requests the robot to write something (a single letter, a number, or a full
 word), and presents a tactile tablet (equipped with a custom writing
-application) to the robot; the robot writes on the tablet by mimicking the
+application) to the robot. The robot writes on the tablet by mimicking the
 writing gesture (but without actually physically touching the tablet); the child
-then pull back the tablet, corrects the robot's attempt by writing him/herself
-on top or next to the robot's writing, and ``send'' his/her demonstration to the
+then pulls back the tablet, corrects the robot's attempt by writing
+on top of or next to the robot's writing, and ``sends'' his/her demonstration to the
 robot by pressing a small button on the tablet. The robot learns from this
-demonstration and tries again. The child continues the turn-taking until he/she
-decides to train the robot on another word. In total, the children performed in
+demonstration and tries again. The child continues the turn-taking until they
+decide to train the robot on another word. In total, the children performed on
 average 12.16 ($SD=2.61$) rounds of writing (complete details on the rationale
@@
 
-Besides, once per interaction, the robot interrupts the handwriting task to
-propose to tell a story (taking about 2 min), and the turn-based
+Once per interaction, the robot interrupts the handwriting task to
+tell a story (taking about 2 min), and the turn-based
 hand-writing task continues afterwards. The intended purpose of the
-story-telling episode is to locally trigger a different set of attention behaviors.
+story-telling episode is to locally trigger a different set of attention
+behaviors for the child.
 
@@
 
-    \caption{\small ROS nodes involved in the VFoA estimation (yellow nodes were developed for this work).}
+    \caption{\small ROS nodes involved in the VFoA estimation (orange nodes were developed for this work).}
     \label{fig:system}
@@
 
-The experiment was carried with an Aldebaran's {\sc nao} robot, using ROS as a
+The experiment was carried out with an Aldebaran {\sc nao} robot, using ROS as a
 middleware to build the attention estimation pipeline (Figure~\ref{fig:system}).
@@
 dlib} and OpenCV libraries; the pose transformations are handled by the ROS {\sc
-tf} library. The same {\sc tf} library is used to represent as individual frames
-the possible point of interests: an object is considered to be in focus when its
-frame lays within the field of attention of the participant
+tf} library. The same {\sc tf} library is used to represent the possible point
+of interests as individual frames: an object is considered to be in focus when its
+frame lies within the field of attention of the participant
 (Figure~\ref{fig:vfoa}).  The implementation is open-source and available at
@@
 
-Besides, we video-recorded the interactions, and performed a {\it post-hoc}
-manual coding of the focus of attention by two annotators (Cronbach's
+We video-recorded the interactions, and performed a {\it post-hoc}
+manual coding of the focus of attention with two annotators (Cronbach's
 $\alpha=.89$)\fixme{did the 2nd annotator annotated the 6 whole interactions?}.
@@
 overlap between the ground-truth and the robot's estimate and the
-inter-judge agreement (Cronbach's $\alpha$).
+inter-rater agreement (Cohen's $\kappa$).
 %and the cosine similarity

### 2015-09-30 paper.tex
Languages fix from James -- Part 2

@@
 To assess the accuracy of the attention estimation by the robot, we computed
-over time (excluding the periods when the head is not detected) the
-overlap between the ground-truth and the robot's estimate and the
-inter-rater agreement (Cohen's $\kappa$).
+over time the overlap between the ground-truth and the robot's estimate and the
+inter-rater agreement (Cohen's $\kappa$). The periods where the head was not
+detected were \emph{excluded} from the agreement computation: at such times, the
+robot explicitly knows that it can not estimate the focus of attention, and as
+such, we do not consider that it \emph{wrongly} estimates the focus.
+
 %and the cosine similarity
@@
         in orange (bottom lines), the focus of attention as manually annotated
-        (ground-truth). Periods where the head is lost are marked with dashed
-        vertical lines. The bottom graph shows agreement between both (whenever
+        (ground-truth). The bottom section shows agreement between both (whenever
         the head is detected).}
@@
 
-During the whole interaction, the head pose of the children was satisfactorily
-and consistently tracked (86\% of the time in average, $SD=3.0$). While this
-high score is expected for a face-to-face interaction with a static head-tracking camera
+During the whole interaction, the head pose of the children was consistently
+tracked, 86\% of the time in average, $SD=3.0$. While this high score is
+expected for a face-to-face interaction with a static head-tracking camera
 (meaning that the child head would remain in the field of view of the camera
@@
 Figure~\ref{head_pose_real_world}), close proximity with the tablet while
-writing, and gaze directed to the facilitator (who was sitting directly on the left of
-the child, Figure~\ref{fig:realSetup}).
+writing, and gaze directed to the facilitator (who was sitting directly on the
+left of the child, Figure~\ref{fig:realSetup}).
 
@@
 0.68 with one subject resulting in significantly worst tracking, at 0.26. While
-the interpretation of Cohen's $\kappa$ is subject to discussions (the number of
+the interpretation of Cohen's $\kappa$ is subject to discussion (the number of
 the coded values -- in our case 6 -- and the distribution probability of values
@@
 $\kappa$ independently of the level of agreement), the levels of agreement are
-\emph{moderate} to \emph{substantial}, with one subject only
-showing \emph{fair} agreement~\cite{landis1977measurement}. Further analysis of
-the videos shows that the child with the lowest level of agreement is especially
-calm and would rely more on the eyes to direct his gaze than the other children,
-thus leading to a less accurate estimation of its focus of attention.
+\emph{moderate} to \emph{substantial}, with one subject only showing \emph{fair}
+agreement~\cite{landis1977measurement}. Further analysis of the videos shows
+that the child with the lowest level of agreement was particularly quiet and
+would indeed rely more on the eyes to direct his gaze than the other children,
+thus leading to a less accurate estimation of his focus of attention.
 
@@
 
-Over the course of the experiment presented at section~\ref{sec:expe}, the robot
+Over the course of the experiment presented in section~\ref{sec:expe}, the robot
 controller would associate a set of expected attentional targets to the phase of
@@
 the child was supposed to write there) or the secondary tablet (that displayed a
-template of the word, used as reference by the child). These expected targets
-form altogether the robot's attentional {\it a priori} and are used to compute
-the with-me-ness.  They are represented over time by the green lines on
-Figure~\ref{fig:with-me-ness}. With-me-ness can be calculated over the whole
-interaction (see corresponding results in Table~\ref{tab:results-with-me-ness})
-or over shorter time windows. Shorter time windows are interesting for two
+template of the word, used as a reference by the child). These expected targets (green lines on Figure~\ref{fig:with-me-ness})
+form the robot's attentional {\it a priori} knowledge and are used to compute
+the with-me-ness. With-me-ness can be calculated over the whole
+interaction
+or over shorter time windows. With-me-ness over the whole interaction for the
+six subjects is
+reported in Table~\ref{tab:results-with-me-ness}. The Pearson's correlation with the
+ground-truth is $r(4)=0.46$ (significance not computed due to small
+sample size). Shorter time windows are interesting for two
 purposes: to analyse the level of with-me-ness in relation to specific
@@
 \emph{over the course} of the interaction (\emph{in-the-moment} measurement) --
-in this later case, one may typically want to consider a sliding time window.
+in the latter case, one may typically want to consider a sliding time window.
 
@@
         diagram represents the instantaneous level of with-me-ness over a
-        sliding window of 30 seconds. The blue line is with-me-ness as estimated
-        by the robot, the orange line is with-me-ness computed from manually
-        annotated attentional targets.  Pearson's correlation between both: $r(973)=0.58, p =
-        2.2e^{-16}.$}
+        sliding window of 30 seconds. The blue line is the with-me-ness as estimated
+        by the robot, the orange line is the with-me-ness computed from manually
+        annotated attentional targets.  Pearson's correlation between both
+        series for this subject: $r(973)=0.58, p < 0.001$.}
 
@@
     ground-truth (\ie \emph{ground-truth with-me-ness}), or based on the focus of
-    attention measured by the robot. Pearson's correlation: $r(4)=0.46$.}
+    attention measured by the robot.}
 
@@
 
-A chart like Figure~\ref{fig:with-me-ness} remains an useful tool to analyse the
-interaction, and several observations can be made out of it: as said, the green
+A chart like Figure~\ref{fig:with-me-ness} remains a useful tool to analyse the
+interaction, and several observations can be made from it: the green
 lines represent how the robot imagine, at a given time, the attention
@@
 the robot was expecting: in that regard, it seems that the child was very much
-engaged into the interaction (we discuss in the next section the exact
+engaged in the interaction (we discuss in the next section the exact
 relationships between with-me-ness and engagement). The only major exception is
 the story-telling phase (episode ``D''): the child was seemingly not captivated by the
-first half of the story, and his attention was not directed towards the robot.
+first half of the story, and their attention was not directed towards the robot.
 
 Another interesting observation pertains to the facilitator: as one can see,
-this child only rarely turned to the facilitator, indicating that the
-interaction and the task were meaningful and easy for him to follow alone.
+this child only rarely turned to the facilitator, possibly indicating that the
+interaction and the task were meaningful and easy enough for him to follow alone.
 
@@
 feedback phases like episode ``C'', we
-can notice numerous gaze shifts between the tablet (where the child write) and the
+can notice numerous gaze shifts between the tablet (where the child writes) and the
 secondary tablet (that showed a template of the word). The episode ``B'' (robot
@@
 
-As hinted by the title of this article, we have presented here how a robot can
-effectively assess in real-time and with a regular camera the focus of attention
-of its interactant, and how we can, in turn, combine it with the robot's {\it a priori}
+We have presented how a robot can
+effectively assess in real-time with a regular camera, the focus of attention
+of its interactant, and how we can combine it with the robot's {\it a priori}
+knowledge
 about the interaction to build a metric of
@@
 
-We already stated the main limitations of our approach to estimation of the
+We already stated the main limitations of our approach to estimating the
 focus of attention: eye gaze information is neglected and we do not perform
-actual visibility check of the in focus objects (we simply approximate them to
+visibility check of the in-focus objects (we simply approximate them to
 their origins, ignoring possible occlusions).
@@
 capture techniques for real-time gaze estimation found in robotics,
-our results are positive: we shows that relying
+our results are positive: we show that relying
 purely on head pose estimation to estimate gaze direction leads to real-world
@@
 annotations, but they are definitely a valuable \emph{in-the-moment} input for
-the robot, reaching for certain children levels of accuracy traditionally
-considered as very good.
+the robot. For certain children, we reach levels of accuracy traditionally
+considered as good.
 
@@
 
-Our approach relies on a simple, non-intrusive sensor (a RGB camera on
+Our approach relies on a simple, non-intrusive sensor (a RGB camera by
 the robot) and an open-source, fast pose estimation algorithm\fixme{test on nao!
-how fast is it actually?}: we hope that this may contribute to a widespread
-adoption of such a technique on a range of robots, including the common {\sc
+how fast is it actually?}: we hope that this may contribute to the widespread
+adoption of such a technique on a range of robots, including the relatively common {\sc
 nao} platform.
@@
 article
-\emph{with-me-ness} as a measure of ``how much the user in \emph{with} the robot during
-a task''. This can be acquired over the course the interaction,
-thus providing the robot with a real-time metric for relatively high-level social construct,
+\emph{with-me-ness} as a measure of ``how much the user is \emph{with} the robot during
+a task''. This can be acquired over the course of the interaction,
+thus providing the robot with a real-time metric for a relatively high-level social construct,
 undoubtedly related to engagement.
 
-One may reasonably wonder how different is \emph{with-me-ness} from \emph{joint
+One may reasonably wonder how different \emph{with-me-ness} is from \emph{joint
 attention} on one hand, and from \emph{engagement} on the other hand.
@@
 builds on a \emph{perceptual} alignment of two agents). \emph{Conceptual
-with-me-ness} as proposed by Sharma~\etal is on the contrary \emph{referential}: ``you are
+with-me-ness} as proposed by Sharma~\etal in~\cite{sharma2014me} is on the contrary \emph{referential}: ``you are
 \emph{with} me if you focus on what I refer to, either explicitly or
-implicitly''. We propose to understand it here in an even slightly broader sense, that
-better reflect the \emph{interaction}: ``you are \emph{with} me if you focus on
+implicitly''. We understand it here in a slightly broader sense that
+reflects the \emph{interaction}: ``you are \emph{with} me if you focus on
 what is important for the interactive task at hand.''
@@
 experimenters), with thus high levels of with-me-ness,  without being
-\emph{engaged} into the interaction. This is typically seen in child-robot
+\emph{engaged} in the interaction. This is typically seen in child-robot
 interaction: children will attempt to closely follow what they are asked to do
--- which may \emph{look like} they are engaged into the interaction -- while
+-- which may \emph{look like} they are engaged in the interaction -- while
 they merely \emph{obey orders}.
 
-Compared to engagement, one of the strength of with-me-ness is its specificity:
+Compared to engagement, one of the strengths of with-me-ness is its specificity:
 it is well-defined (the only important variable -- that needs to be reported --
@@
 assess and compare how users are willing or able to interact with a robot.
-Besides, we have hopefully demonstrated in this article that with-me-ness is an
+We have hopefully demonstrated in this article that with-me-ness is an
 operational \emph{in-the-moment} metric that can be also used as feedback to the

### 2015-09-30 paper.tex
Rewritten abstract to improve coherence of the paper
Based on James' suggestions

@@
 
-Successfully sustaining face-to-face interaction between humans and robots
-mandates fine perception of each other attentional focus. This article presents
-a novel open-source technique to assess the visual focus of attention (VFoA)
-based on accurate, real-time head pose estimation with a single monocular
-camera.  Using manual video annotations as ground truth, we show that our system
-reaches satisfactory levels of accuracy in a real-world face-to-face interaction
-scenario.
-
-We secondly propose to introduce the concept of ``with-me-ness'', borrowed from the
-field of {\it Computer-Supported Collaborative Learning}, as an
-\emph{in-the-moment} measurement of the quality of interaction. We argue that
-this new metric, more specific than the broad concept of engagement, may be well
-suited for operational, real-time assessment of an on-going interaction.
+Measuring ``how much the human is in the interaction'' -- the \emph{level of
+engagement} -- is instrumental in building effective interactive robots.
+
+Engagement, however, is a complex, multi-faceted cognitive mechanism that is
+only indirectly observable. This article formalizes \emph{with-me-ness} as one
+of such indirect measures. \emph{With-me-ness}, a concept borrowed from the
+field of {\it Computer-Supported Collaborative Learning}, measures in a
+well-defined way to what extend the human is \emph{with} the robot over the
+course of an interactive task. As such, it is an meaningful precursor of
+engagement.
+
+We expose the full methodology, from real-time estimation of the human's focus
+of attention (relying on a novel, open-source, vision-based head pose
+estimator), to on-line computation of with-me-ness.
+
+We report as well on the experimental validation of this approach, using a
+naturalistic setup involving children during a complex robot-teaching task.
 

### 2015-09-30 paper.tex
Re-wrapped at 80 cols. No other changes

@@
 
-Building capable social agents requires to endow them with a range
-of perceptual capabilities: however, while face and object tracking and
-recognition, path planning, speech recognition or task learning are some of the current major research
-directions, existing literature tends to evaluate each of these algorithms in
-their own metric space, without considering the interaction quality at a global
-level. Anzalone~\etal~\cite{anzalone} have recently argued in favor of
-such a global evaluation, and they propose to assess these algorithms in term of
-their capability to \emph{obtain the desired effect} in a human-robot interaction
-context. They correspondingly propose metrics built around the measurement
-of \emph{engagement} as indicator of the quality of the experience.
+Building capable social agents requires to endow them with a range of perceptual
+capabilities: however, while face and object tracking and recognition, path
+planning, speech recognition or task learning are some of the current major
+research directions, existing literature tends to evaluate each of these
+algorithms in their own metric space, without considering the interaction
+quality at a global level. Anzalone~\etal~\cite{anzalone} have recently argued
+in favor of such a global evaluation, and they propose to assess these
+algorithms in term of their capability to \emph{obtain the desired effect} in a
+human-robot interaction context. They correspondingly propose metrics built
+around the measurement of \emph{engagement} as indicator of the quality of the
+experience.
 
@@
 there"~\cite{OBrien:2010}, has actively been studied in a diverse set of
-domains. Specifically in robotics, several variables and social signals have been proposed in the
-literature to quantify it. A recent review of these is presented in~\cite{ivaldi2015towards}.
+domains. Specifically in robotics, several variables and social signals have
+been proposed in the literature to quantify it. A recent review of these is
+presented in~\cite{ivaldi2015towards}.
 
@@
 For instance, \cite{Castellano:2009} proposes to predict children's level of
-engagement by integrating non-verbal cues (gaze and smiles)
-with the current state of the interaction in a Bayesian model. While they report high level of
+engagement by integrating non-verbal cues (gaze and smiles) with the current
+state of the interaction in a Bayesian model. While they report high level of
 accuracy, their approach requires post-hoc video annotations, and is not
@@
 In~\cite{Rich:2010}, a computational model based on the recognition of
-\emph{connection} events such as directed gaze, mutual facial gaze is proposed. Not
-relying on gaze, \cite{Sanghvi:2011} focus only on the back or trunk posture as
-a determining factor for the assessment. Finally, a recent study with
-social robots in face-to-face scenarios~\cite{anzalone} explores a set of metrics based on
-non-verbal cues but they also underline possible limitations in long-term
+\emph{connection} events such as directed gaze, mutual facial gaze is proposed.
+Not relying on gaze, \cite{Sanghvi:2011} focus only on the back or trunk posture
+as a determining factor for the assessment. Finally, a recent study with social
+robots in face-to-face scenarios~\cite{anzalone} explores a set of metrics based
+on non-verbal cues but they also underline possible limitations in long-term
 scenarios.
@@
 Therefore, instead of introducing ``yet another metric of engagement'', we
-introduce the more specific concept of
-\emph{with-me-ness}~\cite{sharma2014me}: to what extent the human is ``with me'', the
-robot, during the interaction.
-
-We measure \emph{with-me-ness} by comparing the
-attentional focus of the human (as estimated in real-time by the robot) with the
-expected, {\it a priori} targets of attention elicited by the task at hand.
-
-This article is organized in three sections: we first present a novel
-method for on-line estimation of the focus of attention based on fast 6D head
-pose estimation. We then validate this approach with 
-a real-world field experiment with children. We finally show how to compute
-``with-me-ness'' over the course of the interaction to provide a new \emph{in-the-moment} measurement of
-the quality of interaction, and compare levels computed by the robot with a
-ground truth based on manual post-hoc annotations.
+introduce the more specific concept of \emph{with-me-ness}~\cite{sharma2014me}:
+to what extent the human is ``with me'', the robot, during the interaction.
+
+We measure \emph{with-me-ness} by comparing the attentional focus of the human
+(as estimated in real-time by the robot) with the expected, {\it a priori}
+targets of attention elicited by the task at hand.
+
+This article is organized in three sections: we first present a novel method for
+on-line estimation of the focus of attention based on fast 6D head pose
+estimation. We then validate this approach with a real-world field experiment
+with children. We finally show how to compute ``with-me-ness'' over the course
+of the interaction to provide a new \emph{in-the-moment} measurement of the
+quality of interaction, and compare levels computed by the robot with a ground
+truth based on manual post-hoc annotations.
 
@@
 measuring the direction of gaze is a useful proxy to estimate the (visual) focus
-of attention of a social agent, and indeed this is one of the basic tools used in
-social psychology.
-
-Estimating attention using gaze is not new to robotics either. A
-recent survey by Ruhland~\etal~\cite{ruhland2015review} gives in a broad
-overview of eye gaze research in HCI and social robotics. It remains however an active field of research, as illustrated by
-several recent publications~\cite{baxter2014tracking,anzalone,kennedy2015head}.
-Performing such a measure on a robot, in real-time, and in ecologically
-valid environments (which rules out bulky or invasive apparatus like
-eye-trackers, or techniques requiring fine calibration and/or static
-interactions) remains a challenge in HRI.
+of attention of a social agent, and indeed this is one of the basic tools used
+in social psychology.
+
+Estimating attention using gaze is not new to robotics either. A recent survey
+by Ruhland~\etal~\cite{ruhland2015review} gives in a broad overview of eye gaze
+research in HCI and social robotics. It remains however an active field of
+research, as illustrated by several recent
+publications~\cite{baxter2014tracking,anzalone,kennedy2015head}.  Performing
+such a measure on a robot, in real-time, and in ecologically valid environments
+(which rules out bulky or invasive apparatus like eye-trackers, or techniques
+requiring fine calibration and/or static interactions) remains a challenge in
+HRI.
 
@@
 (no eye gaze tracking) and are generally based on depth sensors (RGB-D) .
-Fanelli~\etal provides an overview of these approaches in~\cite{fanelli2012real},
-and recent examples include~\cite{baxter2014tracking, anzalone}.
+Fanelli~\etal provides an overview of these approaches
+in~\cite{fanelli2012real}, and recent examples include~\cite{baxter2014tracking,
+anzalone}.
 
@@
 We derive the field of attention from the head pose: this is supported by
-previous work, like~\cite{stiefelhagen2002tracking} that shows that the head orientation's contribution in
-overall gaze direction is 68.9\%, which further translates into a 88.7\%
-accuracy in estimating the focus of attention from head pose only in a
-particular meeting scenario (using eye and head tracking).
-
-While previous preliminary research in HRI seemed on the contrary to indicate that deriving
-attentional focus from head pose alone would not be accurate
+previous work, like~\cite{stiefelhagen2002tracking} that shows that the head
+orientation's contribution in overall gaze direction is 68.9\%, which further
+translates into a 88.7\% accuracy in estimating the focus of attention from head
+pose only in a particular meeting scenario (using eye and head tracking).
+
+While previous preliminary research in HRI seemed on the contrary to indicate
+that deriving attentional focus from head pose alone would not be accurate
 enough~\cite{kennedy2015head}, we found in our case acceptable levels of
@@
 As explained, we derive the visual field of attention from the head pose. Our
-technique only involves a single monocular RGB camera used for facial
-feature extraction, and a static simplified 3D mesh of a human head.
-68 facial features are extracted using a fast
-template-based face alignment algorithm by Kazemi and
+technique only involves a single monocular RGB camera used for facial feature
+extraction, and a static simplified 3D mesh of a human head.  68 facial features
+are extracted using a fast template-based face alignment algorithm by Kazemi and
 Sullivan~\cite{kazemi2014one}, as implemented in the open-source {\tt dlib}
-library~\cite{dlib09}.  Eight of these features are
-then matched to their 3D counterparts (Figure~\ref{head_pose}) and we rely on an iterative $PnP$ algorithm
-(OpenCV's implementation) to compute the translation and rotation of the head
-with respect to the camera frame. With this approach, knowing the intrinsic
-parameters of the camera (calibrated camera) is required for an accurate
-estimation of the absolute 3D localization of the head.
+library~\cite{dlib09}.  Eight of these features are then matched to their 3D
+counterparts (Figure~\ref{head_pose}) and we rely on an iterative $PnP$
+algorithm (OpenCV's implementation) to compute the translation and rotation of
+the head with respect to the camera frame. With this approach, knowing the
+intrinsic parameters of the camera (calibrated camera) is required for an
+accurate estimation of the absolute 3D localization of the head.
 
@@
 (occlusions, large head rotations) as encountered in field experiments.
-Figure~\ref{head_pose_real_world} shows a selection of such difficult scenes with
-one child.
+Figure~\ref{head_pose_real_world} shows a selection of such difficult scenes
+with one child.
 
@@
     \includegraphics[width=0.9\columnwidth]{field_of_attention}
-    \caption{\small Screenshot of the real-time attention estimation system. The visual field of attention is approximated to a
-        40$\degree$ cone, spanning from the head's sellion. The objects whose 3D
-    pose intersect with this cone are considered \emph{in focus}.}
+    \caption{\small Screenshot of the real-time attention estimation system. The
+    visual field of attention is approximated to a 40$\degree$ cone, spanning
+    from the head's sellion. The objects whose 3D pose intersect with this cone are
+    considered \emph{in focus}.}
+
     \label{fig:vfoa}
@@
 
-We then approximate the visual \emph{focus} of attention (VFoA) of the human to the
-objects which lie inside this field of attention (Figure~\ref{fig:vfoa}). At a
-given time, more than one object can therefore be \emph{in focus}.
-
-Our implementation has two limitations: objects are approximated to points
-(they are considered in focus if their \emph{origin} lies in the field of
-attention), and we do not
-check actual visibility: one object could be hidden by another, it would still
-be considered as in focus. We did not address these limitations since our
-experimental setup (involving relatively small objects with no occlusions) did
-not necessitate it. Techniques for more accurate assessment of the visual
-perspective of the human peer can be found in~\cite{sisbot2011situation} for instance.
+We then approximate the visual \emph{focus} of attention (VFoA) of the human to
+the objects which lie inside this field of attention (Figure~\ref{fig:vfoa}). At
+a given time, more than one object can therefore be \emph{in focus}.
+
+Our implementation has two limitations: objects are approximated to points (they
+are considered in focus if their \emph{origin} lies in the field of attention),
+and we do not check actual visibility: one object could be hidden by another, it
+would still be considered as in focus. We did not address these limitations
+since our experimental setup (involving relatively small objects with no
+occlusions) did not necessitate it. Techniques for more accurate assessment of
+the visual perspective of the human peer can be found
+in~\cite{sisbot2011situation} for instance.
 
@@
 modules the list of poses of existing objects (represented as {\it frames}), and
-dedicated perception modules are in charge of publishing up-to-date
-informations regarding the location of the objects of interest (the so-called
-\emph{situation assessment}). Due to the nature of the experiment, most of the
-points of interest considered for the experimental validation presented
-hereafter are static with respect to the robot, thus simplifying the scene
-perception.
+dedicated perception modules are in charge of publishing up-to-date informations
+regarding the location of the objects of interest (the so-called \emph{situation
+assessment}). Due to the nature of the experiment, most of the points of
+interest considered for the experimental validation presented hereafter are
+static with respect to the robot, thus simplifying the scene perception.
 
@@
 
-As presented above, we use the 6D head pose as an approximation of the
-actual gaze direction, and we further approximate from here the participant's field of
+As presented above, we use the 6D head pose as an approximation of the actual
+gaze direction, and we further approximate from here the participant's field of
 attention. The assumption that such an approximation of the field of attention
-allows to derive the actual focus of attention needs to validated experimentally.
-Our proposed experiment involves child-robot
-interactions in the context of handwriting remediation.
-This section details the experimental procedure and presents our results.
+allows to derive the actual focus of attention needs to validated
+experimentally.  Our proposed experiment involves child-robot interactions in
+the context of handwriting remediation.  This section details the experimental
+procedure and presents our results.
 
@@
 The experiment, part of the CoWriter project~\cite{Hood:2015}, involves a robot
-which tries to engage a child in handwriting tasks using a \emph{learning by teaching}
-paradigm (\ie the child is the teacher, and he/she attempts to improve the
-robot's handwriting). A tactile tablet is used as writing support.
-
-Figures~\ref{fig:setup} and~\ref{fig:realSetup} illustrate the experimental setup: a face-to-face
-child-robot interaction with an (autonomous) Aldebran's {\sc nao} robot, in the
-presence of a facilitator (one of the researchers).
+which tries to engage a child in handwriting tasks using a \emph{learning by
+teaching} paradigm (\ie the child is the teacher, and he/she attempts to improve
+the robot's handwriting). A tactile tablet is used as writing support.
+
+Figures~\ref{fig:setup} and~\ref{fig:realSetup} illustrate the experimental
+setup: a face-to-face child-robot interaction with an (autonomous) Aldebran's
+{\sc nao} robot, in the presence of a facilitator (one of the researchers).
 
@@
 one.  The facilitator was located about 60 cm to the left of the subject.
-Finally, two observers (visible by the child) were located further away from the interaction field.
-Figure~\ref{fig:setup} indicates accordingly the location of main areas of
-interest (the two tablets, the robot, the facilitator and the observers).
+Finally, two observers (visible by the child) were located further away from the
+interaction field.  Figure~\ref{fig:setup} indicates accordingly the location of
+main areas of interest (the two tablets, the robot, the facilitator and the
+observers).
 
 The dependent variable is the measurement of the participants' VFoA, assessed in
-terms of what the attentional targets of the child are over time. The face of the child
-is acquired through a fixed webcam (Logitech {\sf c920}), placed on the table
-(see Figure~\ref{fig:setup}), and the attentional targets are then computed as
-presented in section~\ref{sec:vfoa}.
+terms of what the attentional targets of the child are over time. The face of
+the child is acquired through a fixed webcam (Logitech {\sf c920}), placed on
+the table (see Figure~\ref{fig:setup}), and the attentional targets are then
+computed as presented in section~\ref{sec:vfoa}.
 
@@
 
-Six children (ages 5 to 6, 3 boys, 3 girls, none wearing glasses) were enrolled for this study.
-The study took place at school, in an isolated room (the computer
-lab). The participants were chosen by the teacher, and would come one after the
-other to interact with the robot (duration: $M=19.6$ min, $SD=1.58$).
+Six children (ages 5 to 6, 3 boys, 3 girls, none wearing glasses) were enrolled
+for this study.  The study took place at school, in an isolated room (the
+computer lab). The participants were chosen by the teacher, and would come one
+after the other to interact with the robot (duration: $M=19.6$ min, $SD=1.58$).
 
 The interaction is organized in rounds of writing: during a typical round, the
-child requests the robot to write something (a single letter, a number, or a full
-word), and presents a tactile tablet (equipped with a custom writing
+child requests the robot to write something (a single letter, a number, or a
+full word), and presents a tactile tablet (equipped with a custom writing
 application) to the robot. The robot writes on the tablet by mimicking the
 writing gesture (but without actually physically touching the tablet); the child
-then pulls back the tablet, corrects the robot's attempt by writing
-on top of or next to the robot's writing, and ``sends'' his/her demonstration to the
-robot by pressing a small button on the tablet. The robot learns from this
-demonstration and tries again. The child continues the turn-taking until they
-decide to train the robot on another word. In total, the children performed on
-average 12.16 ($SD=2.61$) rounds of writing (complete details on the rationale
-and implementation of this experiment can be found in~\cite{Hood:2015}).
-
-Once per interaction, the robot interrupts the handwriting task to
-tell a story (taking about 2 min), and the turn-based
-hand-writing task continues afterwards. The intended purpose of the
-story-telling episode is to locally trigger a different set of attention
-behaviors for the child.
+then pulls back the tablet, corrects the robot's attempt by writing on top of or
+next to the robot's writing, and ``sends'' his/her demonstration to the robot by
+pressing a small button on the tablet. The robot learns from this demonstration
+and tries again. The child continues the turn-taking until they decide to train
+the robot on another word. In total, the children performed on average 12.16
+($SD=2.61$) rounds of writing (complete details on the rationale and
+implementation of this experiment can be found in~\cite{Hood:2015}).
+
+Once per interaction, the robot interrupts the handwriting task to tell a story
+(taking about 2 min), and the turn-based hand-writing task continues afterwards.
+The intended purpose of the story-telling episode is to locally trigger a
+different set of attention behaviors for the child.
 
@@
 tf} library. The same {\sc tf} library is used to represent the possible point
-of interests as individual frames: an object is considered to be in focus when its
-frame lies within the field of attention of the participant
+of interests as individual frames: an object is considered to be in focus when
+its frame lies within the field of attention of the participant
 (Figure~\ref{fig:vfoa}).  The implementation is open-source and available at
@@
 
-Successful detections of the head, and, when detected, the attentional targets of
-the children as estimated by the robot, were logged during the experiment (in
+Successful detections of the head, and, when detected, the attentional targets
+of the children as estimated by the robot, were logged during the experiment (in
 total, $6\times19.6=117.6$ min of interaction). The only post-processing
@@
 
-We video-recorded the interactions, and performed a {\it post-hoc}
-manual coding of the focus of attention with two annotators (Cronbach's
+We video-recorded the interactions, and performed a {\it post-hoc} manual coding
+of the focus of attention with two annotators (Cronbach's
 $\alpha=.89$)\fixme{did the 2nd annotator annotated the 6 whole interactions?}.
@@
 percentage of time during which the student's gaze overlapped the areas of
-teaching slides currently referred to by the teacher.
-In order to apply it to human-robot interactions, we propose to extend this
-concept, and to define \emph{conceptual with-me-ness} as the normalized
-ratio of time that the human interactant focuses its attention on the
-attentional target expected by the robot for the current task (or sub-task).
+teaching slides currently referred to by the teacher.  In order to apply it to
+human-robot interactions, we propose to extend this concept, and to define
+\emph{conceptual with-me-ness} as the normalized ratio of time that the human
+interactant focuses its attention on the attentional target expected by the
+robot for the current task (or sub-task).
 
@@
 considered to be \emph{with} the robot in terms of
-interaction\footnote{Considering a probabilistic model of
-attention expectations (an attention distribution) would be an interesting
-extension of this metric.}. Also notable, we exclude from the computation of
-$\mathcal{W}$ all of the periods of time where the user's focus of attention can
-not be estimated (typically because the user's face is not visible at those
-times).
+interaction\footnote{Considering a probabilistic model of attention expectations
+(an attention distribution) would be an interesting extension of this metric.}.
+Also notable, we exclude from the computation of $\mathcal{W}$ all of the
+periods of time where the user's focus of attention can not be estimated
+(typically because the user's face is not visible at those times).
 
@@
 the child was supposed to write there) or the secondary tablet (that displayed a
-template of the word, used as a reference by the child). These expected targets (green lines on Figure~\ref{fig:with-me-ness})
-form the robot's attentional {\it a priori} knowledge and are used to compute
-the with-me-ness. With-me-ness can be calculated over the whole
-interaction
-or over shorter time windows. With-me-ness over the whole interaction for the
-six subjects is
-reported in Table~\ref{tab:results-with-me-ness}. The Pearson's correlation with the
-ground-truth is $r(4)=0.46$ (significance not computed due to small
-sample size). Shorter time windows are interesting for two
-purposes: to analyse the level of with-me-ness in relation to specific
-interaction episodes; to allow a measurement of with-me-ness by the robot
-\emph{over the course} of the interaction (\emph{in-the-moment} measurement) --
-in the latter case, one may typically want to consider a sliding time window.
+template of the word, used as a reference by the child). These expected targets
+(green lines on Figure~\ref{fig:with-me-ness}) form the robot's attentional {\it
+a priori} knowledge and are used to compute the with-me-ness. With-me-ness can
+be calculated over the whole interaction or over shorter time windows.
+With-me-ness over the whole interaction for the six subjects is reported in
+Table~\ref{tab:results-with-me-ness}. The Pearson's correlation with the
+ground-truth is $r(4)=0.46$ (significance not computed due to small sample
+size). Shorter time windows are interesting for two purposes: to analyse the
+level of with-me-ness in relation to specific interaction episodes; to allow a
+measurement of with-me-ness by the robot \emph{over the course} of the
+interaction (\emph{in-the-moment} measurement) -- in the latter case, one may
+typically want to consider a sliding time window.
 
@@
 A chart like Figure~\ref{fig:with-me-ness} remains a useful tool to analyse the
-interaction, and several observations can be made from it: the green
-lines represent how the robot imagine, at a given time, the attention
-distribution of the child. They also provide an accurate picture of the overall
-turn-taking as viewed by the robot: for instance, the episode ``B'' on
+interaction, and several observations can be made from it: the green lines
+represent how the robot imagine, at a given time, the attention distribution of
+the child. They also provide an accurate picture of the overall turn-taking as
+viewed by the robot: for instance, the episode ``B'' on
 Figure~\ref{fig:with-me-ness} corresponds to one of the ``\emph{Robot writing}''
-episodes, surrounded by ``\emph{Waiting for feedback}'' phases like ``C''; episode
-``D'' corresponds to the story telling; etc. In terms of interaction, the large
-variance of the duration of these phases reflects the fact that this child would
-sometimes take a lot of time to send feedback to the robot, and sometimes, on
-the contrary, be very quick.
+episodes, surrounded by ``\emph{Waiting for feedback}'' phases like ``C'';
+episode ``D'' corresponds to the story telling; etc. In terms of interaction,
+the large variance of the duration of these phases reflects the fact that this
+child would sometimes take a lot of time to send feedback to the robot, and
+sometimes, on the contrary, be very quick.
 
@@
 relationships between with-me-ness and engagement). The only major exception is
-the story-telling phase (episode ``D''): the child was seemingly not captivated by the
-first half of the story, and their attention was not directed towards the robot.
+the story-telling phase (episode ``D''): the child was seemingly not captivated
+by the first half of the story, and their attention was not directed towards the
+robot.
 
@@
 this child only rarely turned to the facilitator, possibly indicating that the
-interaction and the task were meaningful and easy enough for him to follow alone.
+interaction and the task were meaningful and easy enough for him to follow
+alone.
 
 More subtle patterns and events can also be observed: for instance, during the
-feedback phases like episode ``C'', we
-can notice numerous gaze shifts between the tablet (where the child writes) and the
-secondary tablet (that showed a template of the word). The episode ``B'' (robot
-writing) is also interesting: the child did not look at the robot, and instead
-remained focused on the secondary tablet. This situation is typically useful for
-the robot to detect as it may want to adapt its behavior to recover the child's
-attention.
+feedback phases like episode ``C'', we can notice numerous gaze shifts between
+the tablet (where the child writes) and the secondary tablet (that showed a
+template of the word). The episode ``B'' (robot writing) is also interesting:
+the child did not look at the robot, and instead remained focused on the
+secondary tablet. This situation is typically useful for the robot to detect as
+it may want to adapt its behavior to recover the child's attention.
 
@@
 
-We have presented how a robot can
-effectively assess in real-time with a regular camera, the focus of attention
-of its interactant, and how we can combine it with the robot's {\it a priori}
-knowledge
-about the interaction to build a metric of
-\emph{with-me-ness} over the course of the interaction.
+We have presented how a robot can effectively assess in real-time with a regular
+camera, the focus of attention of its interactant, and how we can combine it
+with the robot's {\it a priori} knowledge about the interaction to build a
+metric of \emph{with-me-ness} over the course of the interaction.
 
@@
 
-We already stated the main limitations of our approach to estimating the
-focus of attention: eye gaze information is neglected and we do not perform
-visibility check of the in-focus objects (we simply approximate them to
-their origins, ignoring possible occlusions).
-
-While the first issue is shared with most of the other vision (2D or 3D) or motion
-capture techniques for real-time gaze estimation found in robotics,
-our results are positive: we show that relying
-purely on head pose estimation to estimate gaze direction leads to real-world
-measures that are worth being considered and used. They may not match manual
-annotations, but they are definitely a valuable \emph{in-the-moment} input for
-the robot. For certain children, we reach levels of accuracy traditionally
-considered as good.
+We already stated the main limitations of our approach to estimating the focus
+of attention: eye gaze information is neglected and we do not perform visibility
+check of the in-focus objects (we simply approximate them to their origins,
+ignoring possible occlusions).
+
+While the first issue is shared with most of the other vision (2D or 3D) or
+motion capture techniques for real-time gaze estimation found in robotics, our
+results are positive: we show that relying purely on head pose estimation to
+estimate gaze direction leads to real-world measures that are worth being
+considered and used. They may not match manual annotations, but they are
+definitely a valuable \emph{in-the-moment} input for the robot. For certain
+children, we reach levels of accuracy traditionally considered as good.
 
 Interestingly, previous research conducted on ape-human interaction by
-Tomasello~\etal~\cite{tomasello2007reliance} shows that apes rely primarily
-on the head pose to estimate the focus of attention of a human partner, in to contrast
-to human children who primarily rely on eye direction. From this perspective,
-our robot mimics attention tracking mechanism comparable to the ones found in apes.
-
-Our approach relies on a simple, non-intrusive sensor (a RGB camera by
-the robot) and an open-source, fast pose estimation algorithm\fixme{test on nao!
+Tomasello~\etal~\cite{tomasello2007reliance} shows that apes rely primarily on
+the head pose to estimate the focus of attention of a human partner, in to
+contrast to human children who primarily rely on eye direction. From this
+perspective, our robot mimics attention tracking mechanism comparable to the
+ones found in apes.
+
+Our approach relies on a simple, non-intrusive sensor (a RGB camera by the
+robot) and an open-source, fast pose estimation algorithm\fixme{test on nao!
 how fast is it actually?}: we hope that this may contribute to the widespread
-adoption of such a technique on a range of robots, including the relatively common {\sc
-nao} platform.
+adoption of such a technique on a range of robots, including the relatively
+common {\sc nao} platform.
 
@@
 Borrowing the neologism from the field of CSCL, we have also introduced in this
-article
-\emph{with-me-ness} as a measure of ``how much the user is \emph{with} the robot during
-a task''. This can be acquired over the course of the interaction,
-thus providing the robot with a real-time metric for a relatively high-level social construct,
-undoubtedly related to engagement.
+article \emph{with-me-ness} as a measure of ``how much the user is \emph{with}
+the robot during a task''. This can be acquired over the course of the
+interaction, thus providing the robot with a real-time metric for a relatively
+high-level social construct, undoubtedly related to engagement.
 
@@
 attention} on one hand, and from \emph{engagement} on the other hand.
-With-me-ness is related to both, with however noteworthy nuances: (Triadic) \emph{joint
-attention} is understood as the cognitive realization of a shared attention to
-an object, itself building on a shared perception of that object (\ie joint attention
-builds on a \emph{perceptual} alignment of two agents). \emph{Conceptual
-with-me-ness} as proposed by Sharma~\etal in~\cite{sharma2014me} is on the contrary \emph{referential}: ``you are
+With-me-ness is related to both, with however noteworthy nuances: (Triadic)
+\emph{joint attention} is understood as the cognitive realization of a shared
+attention to an object, itself building on a shared perception of that object
+(\ie joint attention builds on a \emph{perceptual} alignment of two agents).
+\emph{Conceptual with-me-ness} as proposed by Sharma~\etal
+in~\cite{sharma2014me} is on the contrary \emph{referential}: ``you are
 \emph{with} me if you focus on what I refer to, either explicitly or
-implicitly''. We understand it here in a slightly broader sense that
-reflects the \emph{interaction}: ``you are \emph{with} me if you focus on
-what is important for the interactive task at hand.''
+implicitly''. We understand it here in a slightly broader sense that reflects
+the \emph{interaction}: ``you are \emph{with} me if you focus on what is
+important for the interactive task at hand.''
 
@@
 interaction: children will attempt to closely follow what they are asked to do
--- which may \emph{look like} they are engaged in the interaction -- while
-they merely \emph{obey orders}.
+-- which may \emph{look like} they are engaged in the interaction -- while they
+merely \emph{obey orders}.
 
@@
 Table~\ref{tab:targets}), we can formalize it, and as such, it is valuable to
-assess and compare how users are willing or able to interact with a robot.
-We have hopefully demonstrated in this article that with-me-ness is an
-operational \emph{in-the-moment} metric that can be also used as feedback to the
-robot controller to build richer, more adaptive interactive behaviors for our
-robots.
+assess and compare how users are willing or able to interact with a robot.  We
+have hopefully demonstrated in this article that with-me-ness is an operational
+\emph{in-the-moment} metric that can be also used as feedback to the robot
+controller to build richer, more adaptive interactive behaviors for our robots.
 

### 2015-09-30 paper.tex
Improved the overall coherence of the paper by better relating attention tracking to withmeness

@@
 
-This article is organized in three sections: we first present a novel method for
-on-line estimation of the focus of attention based on fast 6D head pose
-estimation. We then validate this approach with a real-world field experiment
-with children. We finally show how to compute ``with-me-ness'' over the course
-of the interaction to provide a new \emph{in-the-moment} measurement of the
-quality of interaction, and compare levels computed by the robot with a ground
-truth based on manual post-hoc annotations.
+The following sections expose the full methodology, starting with the estimation
+of attention: we present in the next section a novel method for on-line estimation
+of the focus of attention based on fast 6D head pose estimation. We validate
+this technique in section~\ref{sec:expe} with a real-world field study involving
+children. Section~\ref{sec:withmeness} formally introduce the concept of
+\emph{with-me-ness}. We present how to compute it over the course of the
+interaction to eventually build a new \emph{in-the-moment} measurement of the
+quality of interaction. We finally validate and discuss this metric by comparing
+it to manual post-hoc annotations of the video-recordings of the interaction.
 
@@
 Figure~\ref{fig:realExpected} further gives a concrete picture of the
-ground-truth \vs computed attentional targets for subject 4 (the subject with
+ground-truth \vs computed attentional targets for subject~4 (the subject with
 the \emph{least} successful tracking).
@@
 
+The next section builds upon this technique for real-time estimation of
+the focus of attention: by comparing the focus of attention with the set of
+attentional targets {\it a priori} expected by the robot, we can estimate to
+what extent the user is ``with'' the robot.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \section{With-me-ness}
+\label{sec:withmeness}
 
@@
         \emph{with-me-ness} over the whole $\approx$17min long interaction of
-        subject 2. The top chart is similar to Figure~\ref{fig:realExpected}
+        subject~2. The top chart is similar to Figure~\ref{fig:realExpected}
         with the \emph{expected} attentional targets added in green. The bottom
@@
 
-\section{Discussion and Conclusion}
-
-We have presented how a robot can effectively assess in real-time with a regular
-camera, the focus of attention of its interactant, and how we can combine it
-with the robot's {\it a priori} knowledge about the interaction to build a
-metric of \emph{with-me-ness} over the course of the interaction.
+\section{Discussion}
 
@@
 
+\section{Conclusion}
+
+We have presented how a robot can effectively assess in real-time with a regular
+camera, the focus of attention of its interactants, and how we can combine it
+with the robot's {\it a priori} knowledge about the interaction to build a
+metric of \emph{with-me-ness} over the course of the interaction.
+
+The experimental validation has been conducted with six children in face-to-face
+interaction with an autonomous robot, over a total duration of about 2 hours. It
+shows that {\it 1)} most of the time, we are able to estimate the head pose of
+these children; {\it 2)} based on these head poses only, the instantaneous focus
+of attention as computed by the robot does reach a good level of accuracy, with
+however one inaccurate outlier out of the six participant; {\it 3)} the robot is
+able to compute in real-time a level of \emph{with-me-ness} that correlates
+strongly with the ground-truth.
+
+The accuracy of the attention estimation could be improved, first by adding
+heuristics to detect and ignore erroneous/inaccurate face detections, second by
+implementing pupil tracking on top of head pose estimation. However, our results
+show that the accuracy levels that we reach already support the reliable
+computation of a metric measuring a high-level social construct, the
+\emph{with-me-ness}, which we argued should be used as a well-defined, reliable
+precursor of engagement in building adaptive robot behaviors.
+
 \section*{Acknowledgments}

### 2015-09-30 paper.tex
[minor] typos

@@
 Figures~\ref{fig:setup} and~\ref{fig:realSetup} illustrate the experimental
-setup: a face-to-face child-robot interaction with an (autonomous) Aldebran's
+setup: a face-to-face child-robot interaction with an (autonomous) Aldebaran
 {\sc nao} robot, in the presence of a facilitator (one of the researchers).

### 2015-10-01 paper.tex
Recomputed Cohen's kappa for inter-rater reliability

@@
 We video-recorded the interactions, and performed a {\it post-hoc} manual coding
-of the focus of attention with two annotators (Cronbach's
-$\alpha=.89$)\fixme{did the 2nd annotator annotated the 6 whole interactions?}.
-The manual coding forms our attentional \emph{ground-truth}.
+of the focus of attention (24\% double-coded, Cohen's $\kappa=0.91$, high
+reliability).  The manual coding forms our attentional \emph{ground-truth}.
 

### 2015-10-01 paper.tex
Commented out last fixme

@@
 Our approach relies on a simple, non-intrusive sensor (a RGB camera by the
-robot) and an open-source, fast pose estimation algorithm\fixme{test on nao!
-how fast is it actually?}: we hope that this may contribute to the widespread
+robot) and an open-source, fast pose estimation algorithm
+: we hope that this may contribute to the widespread
 adoption of such a technique on a range of robots, including the relatively
 common {\sc nao} platform.
+%\fixme{test on nao! how fast is it actually? in 320x240, 240ms on nao, 20ms on core i7}
 

### 2015-10-01 paper.tex
Redacted for double-blind review

@@
 \newcommand{\vs}{\textit{vs.}\xspace}
+\newcommand{\redacted}{\textit{\sc [REDACTED FOR REVIEW]}\xspace}
 
@@
 \title{From Real-time Attention Assessment to ``With-me-ness'' in Human-Robot Interaction}
-\author{Séverin Lemaignan \qquad Fernando Garcia \qquad Alexis Jacq \qquad Pierre Dillenbourg\\Computer-Human Interaction in Learning and Instruction Laboratory (CHILI)\\École Polytechnique Fédérale de Lausanne (EPFL)}
+
+%\author{Séverin Lemaignan \qquad Fernando Garcia \qquad Alexis Jacq \qquad Pierre Dillenbourg\\Computer-Human Interaction in Learning and Instruction Laboratory (CHILI)\\École Polytechnique Fédérale de Lausanne (EPFL)}
+\author{Redacted for double-blind review}
 
@@
 field of {\it Computer-Supported Collaborative Learning}, measures in a
-well-defined way to what extend the human is \emph{with} the robot over the
-course of an interactive task. As such, it is an meaningful precursor of
+well-defined way to what extent the human is \emph{with} the robot over the
+course of an interactive task. As such, it is a meaningful precursor of
 engagement.
@@
 in favor of such a global evaluation, and they propose to assess these
-algorithms in term of their capability to \emph{obtain the desired effect} in a
+algorithms in terms of their capability to \emph{obtain the desired effect} in a
 human-robot interaction context. They correspondingly propose metrics built
@@
 engagement by integrating non-verbal cues (gaze and smiles) with the current
-state of the interaction in a Bayesian model. While they report high level of
+state of the interaction in a Bayesian model. While they report a high level of
 accuracy, their approach requires post-hoc video annotations, and is not
@@
 this technique in section~\ref{sec:expe} with a real-world field study involving
-children. Section~\ref{sec:withmeness} formally introduce the concept of
+children. Section~\ref{sec:withmeness} formally introduces the concept of
 \emph{with-me-ness}. We present how to compute it over the course of the
@@
     \caption{The 6D head pose is estimated by fitting a 3D model of an
-        adult head (left) onto the detected 2D features of the face (right). We
+        adult head (left) {\it\sc[FACE REDACTED FOR REVIEW]} onto the detected 2D features of the face (right). We
         rely on an iterative $PnP$ algorithm, using 8 correspondence pairs
@@
 attention. The assumption that such an approximation of the field of attention
-allows to derive the actual focus of attention needs to validated
+allows to derive the actual focus of attention needs to be validated
 experimentally.  Our proposed experiment involves child-robot interactions in
@@
 
-The experiment, part of the CoWriter project~\cite{Hood:2015}, involves a robot
+The experiment, part of the \redacted project, involves a robot
 which tries to engage a child in handwriting tasks using a \emph{learning by
@@
     \includegraphics[width=1\columnwidth]{realSetup}
-    \caption{\small Picture of the interaction with one of the children.}
+    \caption{\small Picture of the interaction with one of the children
+    {\it\sc[FACE REDACTED FOR REVIEW]}.}
     \label{fig:realSetup}
@@
 ($SD=2.61$) rounds of writing (complete details on the rationale and
-implementation of this experiment can be found in~\cite{Hood:2015}).
+implementation of this experiment can be found in \redacted).
 
@@
 (Figure~\ref{fig:vfoa}).  The implementation is open-source and available at
-\url{https://github.com/chili-epfl/attention-tracker}.
+\redacted.
 
 The implementation of the hand-writing activity itself has been presented by
-Hood~\etal in~\cite{Hood:2015}.
+\redacted.
 
@@
         annotated attentional targets.  Pearson's correlation between both
-        series for this subject: $r(973)=0.58, p < 0.001$.}
+        series for this subject: $r(973)=0.58, p < .001$.}
 
@@
 of attention as computed by the robot does reach a good level of accuracy, with
-however one inaccurate outlier out of the six participant; {\it 3)} the robot is
+however one inaccurate outlier out of the six participants; {\it 3)} the robot is
 able to compute in real-time a level of \emph{with-me-ness} that correlates
@@
 
-This research was partially supported by the Funda\c{c}\~{a}o para a Ci\^{e}ncia
-e a Tecnologia (FCT) with reference UID/CEC/ 50021/2013, and by the Swiss
-National Science Foundation through the National Centre of Competence in
-Research Robotics.
+\redacted
 

### 2015-10-01 data/process.py

@@ -14,6 +14,7 @@ tier_expected = "Robot state"
 
 
 LOST = "Lost track"
+OTHER = "Other"
 
 annotation2foa = {
         ##### Robot state #####
@@ -34,7 +35,7 @@ annotation2foa = {
         "lost_track": (LOST,)}
 
 
-def get_time(ts):
+def get_time(eaf, ts):
     """ Returns the timestamp of an EAF 'timeslot', in seconds
     """
     return eaf.timeslots[ts] / 1000.
@@ -51,7 +52,7 @@ def prepare(eaf, tier):
             print("Unknown annotation: %s" % ann)
             continue
 
-        rawevents[get_time(ts)] = (annotation2foa[ann], get_time(te))
+        rawevents[get_time(eaf, ts)] = (annotation2foa[ann], get_time(eaf, te))
 
     # Order the 'events' by timestamps
     return OrderedDict(sorted(rawevents.items(), key=lambda t: t[0]))
@@ -106,42 +107,7 @@ def filter_observations(obs, min_duration):
 
     filtered = OrderedDict(sorted(filtered.items(), key=lambda t: t[0]))
     return filtered
-    #merged = OrderedDict()
 
-    #match = False
-    #idx = 0
-    #while idx < len(filtered.keys()):
-    #    k = filtered.keys()[idx]
-
-    #    j = 1
-    #    while (idx + j) < len(filtered.keys())-1 and \
-    #          filtered[k][0] == filtered[filtered.keys()[idx + j]][0]:
-    #        j += 1
-    #        match = True
-
-    #    if match:
-    #        merged[k] = (filtered[k][0], filtered[filtered.keys()[idx + j]][1])
-    #        idx += j 
-    #    else:
-    #        merged[k] = (filtered[k][0], filtered[k][1])
-    #        idx += 1
-
-    #return merged
-
-#test = OrderedDict([
-#        (1,(('b',), 3)),
-#        (3.1,(('Other',), 3.2)),
-#        (3.3,(('a',), 3.9)),
-#        (4,(('Other',), 5)),
-#        (5.1,(('a',), 5.2)),
-#        (5.3,(('Other',), 5.4)),
-#        (5.5,(('a',), 7)),
-#        (7.1,(('Other',), 7.2)),
-#        (7.3,(('a',), 8))])
-#
-#print(filter_observations(test))
-
-#print(events_observed)
 ############################################################################
 #     SVG rendering
 ############################################################################
[…]

### 2015-10-01 paper.tex
Typos

@@
 \newcommand{\vs}{\textit{vs.}\xspace}
-\newcommand{\redacted}{\textit{\sc [REDACTED FOR REVIEW]}\xspace}
 
@@
 \title{From Real-time Attention Assessment to ``With-me-ness'' in Human-Robot Interaction}
-
-%\author{Séverin Lemaignan \qquad Fernando Garcia \qquad Alexis Jacq \qquad Pierre Dillenbourg\\Computer-Human Interaction in Learning and Instruction Laboratory (CHILI)\\École Polytechnique Fédérale de Lausanne (EPFL)}
-\author{Redacted for double-blind review}
+\author{Séverin Lemaignan \qquad Fernando Garcia \qquad Alexis Jacq \qquad Pierre Dillenbourg\\Computer-Human Interaction in Learning and Instruction Laboratory (CHILI)\\École Polytechnique Fédérale de Lausanne (EPFL)}
 
@@
     \caption{The 6D head pose is estimated by fitting a 3D model of an
-        adult head (left) {\it\sc[FACE REDACTED FOR REVIEW]} onto the detected 2D features of the face (right). We
+        adult head (left) onto the detected 2D features of the face (right). We
         rely on an iterative $PnP$ algorithm, using 8 correspondence pairs
@@
 
-The experiment, part of the \redacted project, involves a robot
+The experiment, part of the CoWriter project~\cite{Hood:2015}, involves a robot
 which tries to engage a child in handwriting tasks using a \emph{learning by
@@
     \includegraphics[width=1\columnwidth]{realSetup}
-    \caption{\small Picture of the interaction with one of the children
-    {\it\sc[FACE REDACTED FOR REVIEW]}.}
+    \caption{\small Picture of the interaction with one of the children.}
     \label{fig:realSetup}
@@
 ($SD=2.61$) rounds of writing (complete details on the rationale and
-implementation of this experiment can be found in \redacted).
+implementation of this experiment can be found in~\cite{Hood:2015}).
 
@@
 (Figure~\ref{fig:vfoa}).  The implementation is open-source and available at
-\redacted.
+\url{https://github.com/chili-epfl/attention-tracker}.
 
 The implementation of the hand-writing activity itself has been presented by
-\redacted.
+Hood~\etal in~\cite{Hood:2015}.
 
@@
 
-\redacted
+This research was partially supported by the Funda\c{c}\~{a}o para a Ci\^{e}ncia
+e a Tecnologia (FCT) with reference UID/CEC/ 50021/2013, and by the Swiss
+National Science Foundation through the National Centre of Competence in
+Research Robotics.
 

### 2015-11-11 paper.tex
[minor] formatting

--- 2015-10-01 paper.tex
@@
 experiments, one finds that most approaches rely on head pose estimation alone
-(no eye gaze tracking) and are generally based on depth sensors (RGB-D) .
+(no eye gaze tracking) and are generally based on depth sensors (RGB-D).
 Fanelli~\etal provides an overview of these approaches
@@
         tragion and the menton). The 3D origin of the head is set at the sellion.}
-
 \label{head_pose}

### 2015-12-09 data/process.py
Revert "Redacted for double-blind review"
This reverts commit e7466ed0e19901a42151ab056df0589aae9cc542.

@@ -14,7 +14,6 @@ tier_expected = "Robot state"
 
 
 LOST = "Lost track"
-OTHER = "Other"
 
 annotation2foa = {
         ##### Robot state #####
@@ -35,7 +34,7 @@ annotation2foa = {
         "lost_track": (LOST,)}
 
 
-def get_time(eaf, ts):
+def get_time(ts):
     """ Returns the timestamp of an EAF 'timeslot', in seconds
     """
     return eaf.timeslots[ts] / 1000.
@@ -52,7 +51,7 @@ def prepare(eaf, tier):
             print("Unknown annotation: %s" % ann)
             continue
 
-        rawevents[get_time(eaf, ts)] = (annotation2foa[ann], get_time(eaf, te))
+        rawevents[get_time(ts)] = (annotation2foa[ann], get_time(te))
 
     # Order the 'events' by timestamps
     return OrderedDict(sorted(rawevents.items(), key=lambda t: t[0]))
@@ -107,7 +106,42 @@ def filter_observations(obs, min_duration):
 
     filtered = OrderedDict(sorted(filtered.items(), key=lambda t: t[0]))
     return filtered
+    #merged = OrderedDict()
 
+    #match = False
+    #idx = 0
+    #while idx < len(filtered.keys()):
+    #    k = filtered.keys()[idx]
+
+    #    j = 1
+    #    while (idx + j) < len(filtered.keys())-1 and \
+    #          filtered[k][0] == filtered[filtered.keys()[idx + j]][0]:
+    #        j += 1
+    #        match = True
+
+    #    if match:
+    #        merged[k] = (filtered[k][0], filtered[filtered.keys()[idx + j]][1])
+    #        idx += j 
+    #    else:
+    #        merged[k] = (filtered[k][0], filtered[k][1])
+    #        idx += 1
+
+    #return merged
+
+#test = OrderedDict([
+#        (1,(('b',), 3)),
+#        (3.1,(('Other',), 3.2)),
+#        (3.3,(('a',), 3.9)),
+#        (4,(('Other',), 5)),
+#        (5.1,(('a',), 5.2)),
+#        (5.3,(('Other',), 5.4)),
+#        (5.5,(('a',), 7)),
+#        (7.1,(('Other',), 7.2)),
+#        (7.3,(('a',), 8))])
+#
+#print(filter_observations(test))
+
+#print(events_observed)
 ############################################################################
 #     SVG rendering
 ############################################################################
[…]

### 2015-12-09 paper.tex
Updated template to IEEEtran

@@
-\documentclass{sig-alternate}
+\documentclass[conference]{IEEEtran}
+
+\usepackage[utf8]{inputenc}
+\usepackage{ gensymb } % for \degree
+\usepackage{ amssymb } % for \varnothing
 
 \usepackage{hyperref}
+\usepackage{cite}
 
@@
 \usepackage{booktabs}
+
+\usepackage{amsmath}
 
@@
 \begin{document}
-%
-% --- Author Metadata here ---
-\conferenceinfo{HRI}{'16 Chrischurch, New Zealand.}
-%\CopyrightYear{2007} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
-%\crdata{0-12345-67-8/90/01}  % Allows default copyright data (0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
-% --- End of Author Metadata ---
 
 \title{From Real-time Attention Assessment to ``With-me-ness'' in Human-Robot Interaction}
-\author{Séverin Lemaignan \qquad Fernando Garcia \qquad Alexis Jacq \qquad Pierre Dillenbourg\\Computer-Human Interaction in Learning and Instruction Laboratory (CHILI)\\École Polytechnique Fédérale de Lausanne (EPFL)}
+
+\author{\IEEEauthorblockN{Séverin Lemaignan, Fernando Garcia, Alexis Jacq, Pierre Dillenbourg}
+\IEEEauthorblockA{
+Computer-Human Interaction in Learning and Instruction Laboratory (CHILI) \\
+École Polytechnique Fédérale de Lausanne (EPFL) \\
+Lausanne, Switzerland \\
+Email: \tt firstname.lastname@epfl.ch}
+}
 
@@
 \end{abstract}
-
-%% A category with the (minimum) three required fields
-%\category{H.4}{Information Systems Applications}{Miscellaneous}
-%%A category including the fourth, optional field follows...
-%\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]
-
-\terms{Experimentation, Human Factors.}
-
-\keywords{Human-Robot Interaction; Social Robotics; Attention Assessment; Engagement; Computer Vision.}
 
@@
 
-\bibliographystyle{abbrv}
-\bibliography{sigproc}
+\bibliographystyle{IEEEtran}
+\bibliography{biblio}
 \end{document}

### 2015-12-09 paper.tex
Mention the face rotation ranges that dlib can handle

@@
 Besides being fast, the face alignment algorithm has been found to perform well
-in terms of robustness, including in a range of difficult situations
-(occlusions, large head rotations) as encountered in field experiments.
+in terms of robustness, including in a range of difficult situations encountered
+in field experiments, like partial occlusions or large head rotations (we have
+measured the default {\tt dlib} model to be able to track a face
+with rotations up to $\pm 40\degree$
+horizontally and $\pm 30\degree$ vertically).
 Figure~\ref{head_pose_real_world} shows a selection of such difficult scenes

### 2015-12-10 paper.tex
Implemented the remaining reviewers' comments + fit into 8 pages

@@
 engagement} -- is instrumental in building effective interactive robots.
-
 Engagement, however, is a complex, multi-faceted cognitive mechanism that is
@@
 course of an interactive task. As such, it is a meaningful precursor of
-engagement.
-
-We expose the full methodology, from real-time estimation of the human's focus
-of attention (relying on a novel, open-source, vision-based head pose
-estimator), to on-line computation of with-me-ness.
-
-We report as well on the experimental validation of this approach, using a
+engagement.  We expose in this paper the full methodology, from real-time
+estimation of the human's focus of attention (relying on a novel, open-source,
+vision-based head pose estimator), to on-line computation of with-me-ness.  We
+report as well on the experimental validation of this approach, using a
 naturalistic setup involving children during a complex robot-teaching task.
@@
 \end{abstract}
+
+\begin{IEEEkeywords}
+Human-Robot Interaction; Visual Focus of Attention; With-me-ness; Real-time Head Pose Estimation.
+\end{IEEEkeywords}
 
@@
 Sullivan~\cite{kazemi2014one}, as implemented in the open-source {\tt dlib}
-library~\cite{dlib09}.  Eight of these features are then matched to their 3D
+library~\cite{dlib09}.  Eight of these features (chosen to be far apart and
+relatively stable across age and gender) are then matched to their 3D
 counterparts (Figure~\ref{head_pose}) and we rely on an iterative $PnP$
@@
 (sellion) of the human face. Different dimensions for the human field of view
-can be found in the litterature: Holmqvist~\cite{holmqvist2011eye} models it
+can be found in the literature: Holmqvist~\cite{holmqvist2011eye} models it
 with an horizontal aperture of $ \pm 40\degree $ and a vertical aperture of $
-\pm 25\degree $, while Spector~\cite{walker1980clinical} for instance suggests
+\pm 25\degree $, while Walker~\cite{walker1980clinical} for instance suggests
 $60\degree$ up, $75\degree$ down, $60\degree$ inwards (towards the nose) and
@@
 $30\degree$. We retained in this work a slightly wider aperture of $40\degree$.
-
 We then approximate the visual \emph{focus} of attention (VFoA) of the human to
@@
 of attention of the human requires first to transform the coordinates
-$A(X_A,Y_A,Z_A)$ into the frame of the face, and then to verify the following
-simple inequality (with $fov$ the aperture, and assuming that the main axis
-$\vec{x}$ of the field of attention points forward):
-
-\begin{align}
-    \sqrt{Y_A^2 + Z_A^2} < tan\left(\frac{fov}{2}\right) \cdot X_A
-\label{eq:fov}
-\end{align}
+$A(X_A,Y_A,Z_A)$ into the frame of the face, and then to verify the
+simple inequality $\sqrt{Y_A^2 + Z_A^2} < tan\left(\frac{fov}{2}\right) \cdot
+X_A$ (with $fov$ the aperture, and assuming that the main axis
+$\vec{x}$ of the field of attention points forward).
 
@@
 the robot's handwriting). A tactile tablet is used as writing support.
-
 Figures~\ref{fig:setup} and~\ref{fig:realSetup} illustrate the experimental
@@
 
-\begin{figure}[h!]
+\begin{figure}[ht!]
     \centering
@@
 full word), and presents a tactile tablet (equipped with a custom writing
-application) to the robot. The robot writes on the tablet by mimicking the
-writing gesture (but without actually physically touching the tablet); the child
+application) to the robot. The robot ``writes'' on the tablet by drawing in the
+air the letters that are displayed on the screen by the tablet application; the child
 then pulls back the tablet, corrects the robot's attempt by writing on top of or
@@
 (taking about 2 min), and the turn-based hand-writing task continues afterwards.
-The intended purpose of the story-telling episode is to locally trigger a
-different set of attention behaviors for the child.
+The intended purpose of the story-telling episode is to break the routine of the
+writing turns by creating a surprise, and thus, to elicit a different set of
+attention behaviors from the child.
 
@@
     \node [cmpt, fill=YellowOrange!50, left=of tf] (attention) {attention tracking};
-    \node [above=0.6cm of attention, text width=2.7cm,align=center] (input) {\tt /image \\ /camera\_info \\ dlib {\sf faces model} \\ {\sf 3D face template}};
+    \node [above=0.6cm of attention, text width=3cm,align=center] (input) {{\tt
+    /image}, {\tt /camera\_info}, \\ {\tt dlib} {\sf faces model}, \\ {\sf 3D face template}};
     \node [cmpt, right=of tf] (static) {static pose publishers};
@@
 }
-
-    \caption{\small ROS nodes involved in the VFoA estimation (orange nodes were developed for this work).}
+    \caption{\small ROS nodes involved in the VFoA estimation (orange nodes were
+    specifically developed for this work).}
     \label{fig:system}
@@
 (Figure~\ref{fig:vfoa}).  The implementation is open-source and available at
-\url{https://github.com/chili-epfl/attention-tracker}.
+\texttt{\url{https://github.com/chili-epfl/attention-tracker}}.
 
@@
     level is reported over the whole interaction, either based on the
-    ground-truth (\ie \emph{ground-truth with-me-ness}), or based on the focus of
+    annotated focus of attention (\ie \emph{ground-truth with-me-ness}), or based on the focus of
     attention measured by the robot.}
@@
 by the first half of the story, and their attention was not directed towards the
-robot.
+robot (this actually matches the observed behavior of the children who mostly
+found the story boring).
 
@@
 
-Interestingly, previous research conducted on ape-human interaction by
-Tomasello~\etal~\cite{tomasello2007reliance} shows that apes rely primarily on
-the head pose to estimate the focus of attention of a human partner, in to
-contrast to human children who primarily rely on eye direction. From this
-perspective, our robot mimics attention tracking mechanism comparable to the
-ones found in apes.
+%Interestingly, previous research conducted on ape-human interaction by
+%Tomasello~\etal~\cite{tomasello2007reliance} shows that apes rely primarily on
+%the head pose to estimate the focus of attention of a human partner, in to
+%contrast to human children who primarily rely on eye direction. From this
+%perspective, our robot mimics attention tracking mechanism comparable to the
+%ones found in apes.
 
@@
 Compared to engagement, one of the strengths of with-me-ness is its specificity:
-it is well-defined (the only important variable -- that needs to be reported --
-is the mapping {\it phase/expected attentional target}, \ie our
-Table~\ref{tab:targets}), we can formalize it, and as such, it is valuable to
+it is well-defined, we can formalize it, and as such, it is valuable to
 assess and compare how users are willing or able to interact with a robot.  We
 have hopefully demonstrated in this article that with-me-ness is an operational
-\emph{in-the-moment} metric that can be also used as feedback to the robot
-controller to build richer, more adaptive interactive behaviors for our robots.
+\emph{in-the-moment} metric that can also be used as a real-time feedback to the
+robot controller to build richer, more adaptive interactive behaviors for our
+robots.
+
+Note however that, besides the actual focus of attention, the mapping {\it
+phase/expected attentional target} (\ie our Table~\ref{tab:targets}) is a
+critical piece of information to interpret with-me-ness. The mapping is
+typically built by a domain expert, and is often subject to debate (for
+instance in our experiment, one could argue that during the ``Waiting for
+feedback'' phase, the child could have gazed toward the robot to make sure the
+robot was paying attention, and consequently, {\sf robot} should be added to the
+expected target).  For this reason, the chosen mapping should always be reported
+along with the computed with-me-ness levels, and with-me-ness should not be
+reported as an absolute metric, but rather as a mean of comparing
+different interactions within the same study.
 
